### 2020-01-07 docs/solution.tex
Initial skeleton for short article about Freak

\documentclass{article}
\usepackage[utf8]{inputenc}
% \usepackage{polski}
\usepackage[]{amsthm}
\usepackage[]{amssymb} %gives us the character \varnothing

\usepackage{hyperref}
\usepackage{titling}
\usepackage[backend=bibtex]{biblatex}
\addbibresource{mybib.bib}
\theoremstyle{definition}
\newtheorem{definition}{Definition}

\theoremstyle{lemma}
\newtheorem{lemma}{Lemma}

\theoremstyle{observation}
\newtheorem{observation}{Observation}

\theoremstyle{theorem}
\newtheorem{theorem}{Theorem}

\newcommand*{\fullref}[1]{\hyperref[{#1}]{\autoref*{#1} \nameref*{#1}}}

\title{Continuation Passing Style for Effect Handlers}
\author{Mateusz Urbańczyk}
\date{7 January 2020}

\begin{document}
\maketitle

\begin{abstract}

    We provide an implementation of algebraic effects and handlers by doing continuation
    passing style tranformation of the functional programming language Freak,
    which is based on the existing Links langauge \cite{handlers-cps}.

\end{abstract}

\section{Introduction}

Implementation of Continuation Passing Style for Effect Handlers paper \cite{handlers-cps}

\section{State of the art}

    Frank \cite{frank}, Koka \cite{leijen-koka}, Helium \cite{helium}

\section{Syntax}

\section{Operational semantics}

    \subsection{Integers}

    \subsection{Binary operators}

\section{Usage guide}
    % TODO: not sure if correct place for that

    All commands are available within \verb src  directory.

    \subsection{Build and install}

    \begin{itemize}
        \item Install dependencies: \verb!make install!
        \item Compile: \verb!make build!
        \item Link to PATH: \verb!sudo make link!
        \item Remove artiacts: \verb!make clean!
        \item Run tests: \verb!make tests!
        \item Run code linter: \verb!make lint!
    \end{itemize}

    After compiling and linking program to PATH, one may evaluate program as
    follows: \verb!freak programs/choicesList.fk!. The actual code is described in \fullref{sec:choice-example}

    \subsection{Running tests}

    Test cases are available \href{https://github.com/Tomatosoup97/freak/blob/master/src/Tests.hs}{\underline{here}},
    they include both inline and file-based tests. For more details about
    writing tests, one may refer to HUnit documentation \cite{hunit-docs}. \\

    \begin{itemize}
        \item Run tests: \verb!make tests!
        \item Run code linter: \verb!make lint!
        \item Compile, run linter and tests: \verb!make check!
    \end{itemize}

\section{Implementation}

    The Freak implementation is available at \href{https://github.com/Tomatosoup97/freak}{https://github.com/Tomatosoup97/freak}.
    CPS based on \cite{appel-continuations}

    \subsection{Row types}

    \subsection{Type inference}

    Type system as of this day is not implemented, as the focus has been put
    on CPS transformation. Further work is required.

\section{Examples}

    In this section we present a few examples to show the capabilities of the
    language. The ideas has been based on \cite{eff}, and thus will not be
    described in great details.

    \subsection{Choice}
    \label{sec:choice-example}

    The first example will be based on modelling (nondeterministic) choice
    in the program. We will make two decisions, which will affect the computation
    result:

    \begin{verbatim}
    let c1 <- do Choice () in
    let c2 <- do Choice () in
    let x <- if c1 then return 10 else return 20 in
    let y <- if c2 then return 0 else return 5 in
        return x - y
    \end{verbatim}

    With that in hand, we may want to define effect handlers:

    \begin{verbatim}
    handle ... with {
        Choice p r ->
            let t <- r 1 in
            let f <- r 0 in
            <PLACEHOLDER> |
        return x -> return x
    }
    \end{verbatim}

    where in the \verb!<PLACEHOLDER>! we can define on what to do with the
    computation. For example, min-max strategy for picking the minimum value:

    \begin{verbatim}
    if t < f then return t else return f
    \end{verbatim}

    where the code evaluates to \verb!5!. Another example is a handler that
    collects all possible results, which can be achieved by putting
    \verb!return (t, f)! in the \verb!<PLACEHOLDER>!, which evaluates to \verb!((10, 5), (20, 15))!.

    \subsection{Exceptions}

    Exceptions are simply algebraic effect handlers which drop the resumption.

    \begin{verbatim}
    handle
        if x == 0 then do Raise ()
                  else return 1 / x
    with {
        Raise p r -> return 42 |
        return x -> return x
    }
    \end{verbatim}

    Where we imagine that x variable has been bound previously. More exemplary
    programs in Freak language can be found \href{https://github.com/Tomatosoup97/freak/tree/master/src/programs}{\underline{here}}.

\section{Future work}

    \subsection{Abstract machine}

    Implement abstract machine described in \cite{liberating-effects}

    \subsection{Multiple instances of algebraic effect}

    Proposed Helium \cite{binders-labels}, or Eff \cite{eff}

    \subsection{Make the langugae more usable}

    While the language is turing-complete, for convenient usage it
    requires more basic constructs and syntactic sugar for common patterns
    that would ease the programming.

    \subsection{Selective CPS}

    As in \cite{leijen-koka}

    \subsection{Exceptions as separate constructs}

    Exceptions are a trivial example of algebraic effect where the resumption is
    discarded, and as described in \S 4.5 \cite{handlers-cps}, they can be modeled
    as a separate construct to improve performance.

    \subsection{Shallow handlers}

    X \cite{shallow-handlers}

\section{Conclusions}

\printbibliography

\end{document}

### 2020-01-07 docs/solution.tex
Writing article...

@@
 
-Implementation of Continuation Passing Style for Effect Handlers paper \cite{handlers-cps}
+In this article we present experimental programming language Freak, which is an
+implementation of Continuation Passing Style for Effect Handlers paper \cite{handlers-cps},
+with addition of a few basic constructs. We start by presenting the related work,
+then discuss syntax and operational semantics. Basic usage guide for playing with
+the language is provided, as well as implementation details and examples. We conclude
+by stating what are the possible augmentations, that are intended to be made in the future.
 
@@
 
-    Frank \cite{frank}, Koka \cite{leijen-koka}, Helium \cite{helium}
+    Despite from Links language \cite{handlers-cps}, on which the implementation
+    is based, there are currently many other alternatives available. One may take
+    a look at Frank \cite{frank}, which provides a support for multihandlers,
+    Koka \cite{leijen-koka}, Helium \cite{helium} or Eff \cite{eff}. Except from
+    separate languages, many libraries arose for existing ones like Haskell,
+    Idris, Scala or Multicore OCaml. \\
+
+    As can be seen in the J. Yallop repository \cite{effects-bibliography},
+    algebraic effects and handlers are now trending branch of computer science.
 
@@
 \section{Operational semantics}
+
+    The source language's dynamics has been described
+    extensively by providing small-step operational semantics,
+    continuation passing style transformation \cite{handlers-cps} as well
+    as abstract machine \cite{liberating-effects}, which was proved to coincide
+    with CPS translation. That being said, Freak introduces new basic
+    constructs to the language, for which we shall define the semantics.
 
@@
         \item Remove artiacts: \verb!make clean!
-        \item Run tests: \verb!make tests!
-        \item Run code linter: \verb!make lint!
     \end{itemize}
@@
     they include both inline and file-based tests. For more details about
-    writing tests, one may refer to HUnit documentation \cite{hunit-docs}. \\
+    writing tests, one may refer to \textit{HUnit documentation} \cite{hunit-docs}. \\
 
@@
     The Freak implementation is available at \href{https://github.com/Tomatosoup97/freak}{https://github.com/Tomatosoup97/freak}.
-    CPS based on \cite{appel-continuations}
+    While the paper provided a good overview on the language and the translation,
+    the lower-level details were ommited. Therefore, the actual CPS implementation
+    is based on A. Appel \cite{appel-continuations}, which provides a translation
+    for a simplified ML calculus. \\
 
-    \subsection{Row types}
+    The core data structure into which the source program is transformed,
+    the \verb!Continuation Computation!, is defined as follows:
+
+    \begin{verbatim}
+data ContComp
+    = CPSApp CValue [CValue]
+    | CPSFix Var [Var] ContComp ContComp
+    | CPSBinOp BinaryOp CValue CValue Var ContComp
+    | CPSValue CValue
+    | CPSLet Var CValue ContComp
+    | CPSSplit Label Var Var CValue ContComp
+    | CPSCase CValue Label Var ContComp Var ContComp
+    | CPSIf CValue ContComp ContComp
+    | CPSAbsurd CValue
+    \end{verbatim}
+
+    The translation is based on the curried first-order translation. For
+    improving the performance of the evaluation, uncurried higher-order translation
+    should be adopted, so that administrative redexes are contracted and
+    proper tail-recursion is obtained. The final answer is represented as a \verb!DValue!,
+    where the meaning of the coproduct is as one would expect:
+
+    \begin{verbatim}
+type Label = String
+
+data DValue
+    = DNum Integer
+    | DLambda Env FuncRecord
+    | DUnit
+    | DPair DValue DValue
+    | DLabel Label
+    \end{verbatim}
 
@@
     Type system as of this day is not implemented, as the focus has been put
-    on CPS transformation. Further work is required.
+    on CPS transformation. Further work is required here, especially considering
+    the fact that a huge advantage of algebraic effects is that they are explicitly
+    defined in the type of a computation.
+
+    \subsection{Source code structure}
+
+    The source code is divided into a number of modules, each with a different
+    responsibility.
+
+    \begin{verbatim}
+    AST.hs          - AST data structures
+    CPS.hs          - CPS translation
+    Eval.hs         - Evaluation of the target calculus
+    Freak.hs        - API for the language
+    Main.hs         - Main module running evaluator on given filename
+    Parser.hs       - Parser and lexer
+    Tests.hs        - Tests module
+    Typechecker.hs  - -- suspended --
+    Types.hs        - Common types definition
+    programs/       - Exemplary programs used in tests
+    \end{verbatim}
 
@@
     In this section we present a few examples to show the capabilities of the
-    language. The ideas has been based on \cite{eff}, and thus will not be
-    described in great details.
+    language. The ideas has been based on \cite{programming-in-eff}, and thus will not be
+    described in great details. More exemplary programs in Freak language can
+    be found \href{https://github.com/Tomatosoup97/freak/tree/master/src/programs}{\underline{here}}.
 
@@
     handle
-        if x == 0 then do Raise ()
-                  else return 1 / x
+        if x == 0 then do ZeroDivisionError ()
+                  else return 1/x
     with {
-        Raise p r -> return 42 |
+        ZeroDivisionError p r -> return 42 |
         return x -> return x
@@
     \end{verbatim}
-
-    Where we imagine that x variable has been bound previously. More exemplary
-    programs in Freak language can be found \href{https://github.com/Tomatosoup97/freak/tree/master/src/programs}{\underline{here}}.
+    Where we imagine that x variable has been bound previously.
 
@@
 
-    \subsection{Abstract machine}
+    The Freak language is experimental and a lot of possible enhancements
+    could be adopted. Here we provide a list of proposed directions in which
+    further work could be made.
 
-    Implement abstract machine described in \cite{liberating-effects}
+    \subsection{Alternative evaluation}
 
-    \subsection{Multiple instances of algebraic effect}
-
-    Proposed Helium \cite{binders-labels}, or Eff \cite{eff}
+    The Links language also provides small-step operational semantics and
+    an abstract machine \cite{liberating-effects}. Implementing another way
+    of evaluation could serve as a way to empirically assert correctness,
+    as opposed to formally.
 
@@
 
+    \subsection{Multiple instances of algebraic effect}
+
+    The Freak language is limited to a single instance of an effect. We would
+    need to support cases where many instances of the algebraic effects, with
+    the same handlers code, could be instantiated. The current state of the
+    art introduces a concept of resources and instances, as in Eff \cite{programming-in-eff},
+    or instance variables, as in Helium \cite{binders-labels}.
+
     \subsection{Selective CPS}
 
-    As in \cite{leijen-koka}
+    Other languages, like Koka \cite{leijen-koka}, or even core of the Links, are
+    performing selective CPS translation, which reduces the overhead on code
+    that does not perform algebraic effects. Our current translation is fully
+    embeded in the CPS.
 
@@
 
-    X \cite{shallow-handlers}
+    Shallow and deep handlers while being able to simluate each other up to
+    administrative reductions, have a very different meaning from a theoretical
+    point of view. Implementing them as defined by Lindley et al. \cite{shallow-handlers} could
+    be another way of enhancing Freak.
 

### 2020-01-07 src/Freak.hs
More cleanup in the source code

@@ -6,7 +6,6 @@ import CPS
 import Eval
 import Types
 
-
 evalProgram :: String -> Either Error DValue
 evalProgram s = do
     ast <- parseString s

### 2020-01-07 docs/solution.tex
Define syntax and operational semantics in article

@@
 \usepackage[]{amsthm}
-\usepackage[]{amssymb} %gives us the character \varnothing
+\usepackage{amsfonts}
+\usepackage[]{amssymb}
+\usepackage{amsmath}
+\usepackage{syntax}
+\usepackage{mathalfa}
 
@@
     We provide an implementation of algebraic effects and handlers by doing continuation
-    passing style tranformation of the functional programming language Freak,
-    which is based on the existing Links langauge \cite{handlers-cps}.
+    passing style transformation of the functional programming language Freak,
+    which is based on the existing Links language \cite{handlers-cps}.
 
@@
 implementation of Continuation Passing Style for Effect Handlers paper \cite{handlers-cps},
-with addition of a few basic constructs. We start by presenting the related work,
+with the addition of a few basic constructs. We start by presenting the related work,
 then discuss syntax and operational semantics. Basic usage guide for playing with
@@
 
+    The syntax for the calculus is shown below. $nat \; n$ represents an integer $n$,
+    $V \oplus W$ and $V \approx W$ are respectively binary and relational
+    operators, where we support basic arithmetic and comparison operations.
+    \textbf{if} $V$ \textbf{then} $M$ \textbf{else} $N$ is a standard branching statement.
+    The other constructs are just as in Links, with slight syntax modifications.
+    Actual programs in Freak can be found in \fullref{sec:examples}
+
+    \begin{grammar}
+
+        <Values V, W> ::= $ x $ | $nat \; n$ \\
+            | $ \backslash x : A \rightarrow M $ | \textbf{rec} $ g \; x \rightarrow M $\\
+            | $V \oplus W$ | $V \approx W$ \\
+            | <> | $ \{ \ell = V; W\} $  | $[ \ell \; V]^{R}$
+
+        <Computations M, N> ::= $ V $ $ W $ \\
+            | \textbf{if} $V$ \textbf{then} $M$ \textbf{else} $N$ \\
+            | \textbf{let} $\{\ell  = x; y\} = V$ \textbf{in} $ N $ \\
+            | \textbf{case} $V \{ \ell \; x \rightarrow M; y \rightarrow N\}$ | \textbf{absurd} $ V $ \\
+            | \textbf{return} $V$ | \textbf{let} $ x \leftarrow M $ \textbf{in} $ N $ \\
+            | \textbf{do} $\ell \; V$ | \textbf{handle} $M$ \textbf{with} $ \{ H \} $
+
+        <Handlers H> ::= \textbf{return} $ x \rightarrow M $ | $ \ell \; p \; r \rightarrow M, H $
+
+        <Binary operators $\oplus$> ::= + | - | * | /
+
+        <Relational operators $\approx$> ::= $ \textless $ | $\leqslant$ | $>$ | $\geqslant$ | == | !=
+
+    \end{grammar}
+
 \section{Operational semantics}
 
-    The source language's dynamics has been described
+    The source language's dynamics have been described
     extensively by providing small-step operational semantics,
@@
 
-    \subsection{Integers}
+    \begin{flushleft}
+    Extension of the evaluation contexts:\\
+    \end{flushleft}
 
-    \subsection{Binary operators}
+    \begin{flushleft}
+    $\mathcal{E}$ ::= $ \mathcal{E} \oplus W \; | \; V \oplus \mathcal{E} \; |$ \textbf{if} $\mathcal{E}$ \textbf{then} $M$ \textbf{else} $N$ \\
+    \end{flushleft}
+
+    \begin{flushleft}
+    Small-step operational semantics:
+    \end{flushleft}
+
+    \begin{flushleft}
+    \textbf{if} $nat \; n$ \textbf{then} $M$ \textbf{else} $N \rightsquigarrow M \quad \quad $  if $n \neq 0$ \\
+    \textbf{if} $nat \; n$ \textbf{then} $M$ \textbf{else} $N \rightsquigarrow N \quad \quad $    if $n = 0$ \\
+    \end{flushleft}
+
+    \begin{flushleft}
+    $nat \; n \oplus nat \; n' \rightsquigarrow n'' \quad \quad $    if $ n'' = n \oplus n' $ \\
+    $nat \; n \approx nat \; n' \rightsquigarrow 1  \quad \quad $    if $ n \approx n' $ \\
+    $nat \; n \approx nat \; n' \rightsquigarrow 0  \quad \quad $    if $ n \not\approx n' $
+
+    \end{flushleft}
 
@@
     The Freak implementation is available at \href{https://github.com/Tomatosoup97/freak}{https://github.com/Tomatosoup97/freak}.
-    While the paper provided a good overview on the language and the translation,
-    the lower-level details were ommited. Therefore, the actual CPS implementation
+    While the paper provided a good overview of the language and the translation,
+    the lower-level details were omitted. Therefore, the actual CPS implementation
     is based on A. Appel \cite{appel-continuations}, which provides a translation
@@
 
-    The core data structure into which the source program is transformed,
+    Abstract Syntax Tree structure is defined just as syntax is. The core data
+    structure into which the source program is transformed,
     the \verb!Continuation Computation!, is defined as follows:
@@
 
-    Type system as of this day is not implemented, as the focus has been put
+    The type system as of this day is not implemented, as the focus has been put
     on CPS transformation. Further work is required here, especially considering
@@
 \section{Examples}
+\label{sec:examples}
 
     In this section we present a few examples to show the capabilities of the
-    language. The ideas has been based on \cite{programming-in-eff}, and thus will not be
+    language. The ideas have been based on \cite{programming-in-eff}, and thus will not be
     described in great details. More exemplary programs in Freak language can
@@
     The Freak language is experimental and a lot of possible enhancements
-    could be adopted. Here we provide a list of proposed directions in which
-    further work could be made.
+    could be adopted. Here we provide a list of the proposed directions in which
+    further work could be done.
 
@@
 
-    \subsection{Make the langugae more usable}
+    \subsection{Make the language more usable}
 
-    While the language is turing-complete, for convenient usage it
+    While the language is Turing-complete, for convenient usage it
     requires more basic constructs and syntactic sugar for common patterns
@@
     need to support cases where many instances of the algebraic effects, with
-    the same handlers code, could be instantiated. The current state of the
+    the same handler code, could be instantiated. The current state of the
     art introduces a concept of resources and instances, as in Eff \cite{programming-in-eff},
@@
 
-    Other languages, like Koka \cite{leijen-koka}, or even core of the Links, are
+    Other languages, like Koka \cite{leijen-koka}, or even the core of the Links, are
     performing selective CPS translation, which reduces the overhead on code
     that does not perform algebraic effects. Our current translation is fully
-    embeded in the CPS.
+    embedded in the CPS.
 
@@
 
-    Shallow and deep handlers while being able to simluate each other up to
+    Shallow and deep handlers while being able to simulate each other up to
     administrative reductions, have a very different meaning from a theoretical
@@
 
-\section{Conclusions}
-
 \printbibliography

### 2020-01-07 docs/article.tex
s/solution/article/g

--- 2020-01-07 docs/solution.tex
@@
     As can be seen in the J. Yallop repository \cite{effects-bibliography},
-    algebraic effects and handlers are now trending branch of computer science.
+    algebraic effects and handlers are now trending branch in the computer science.
 
@@
     \end{verbatim}
-    Where we imagine that x variable has been bound previously.
+    Where we imagine that $x$ variable has been bound previously.
+
+    \subsection{Side effects}
+
+    The complexity of the programs and their performance usually comes from side (d)effects.
+    Algebraic effects allow us to define code in a declarative manner, and hence
+    neatly tame the side effects that they produce. This gives us a lot of flexibility
+    in the actual meaning without duplicating the code. Let's consider the following
+    very basic code snippet:
+
+    \begin{verbatim}
+    let x <- do Fetch () in
+    -- operate on x
+    \end{verbatim}
+
+    The code is dependent on a context in which it is executed, which here is
+    the handler that defines the behaviour of the algebraic \verb!Fetch! effect.
+    In the imperative, or even functional approach, we would need to provide
+    the interface for fetching the data by doing dependency injection or even
+    embedding the operation directly. Here we are just stating what operation
+    we are performing, leaving the interpretation up to the execution context,
+    which could do the fetching or mock the external resource. \\
+
+    These implications are straightforward when looking from a categorical
+    standpoint and models of algebraic theories, nevertheless are very exciting
+    for programming use cases. The current Freak implementation does not yet support I/O.
 

### 2020-01-16 docs/article.tex
Corrections in the article

@@
 
-    Despite from Links language \cite{handlers-cps}, on which the implementation
+    Except from Links language \cite{handlers-cps}, on which the implementation
     is based, there are currently many other alternatives available. One may take
@@
     As can be seen in the J. Yallop repository \cite{effects-bibliography},
-    algebraic effects and handlers are now trending branch in the computer science.
+    algebraic effects and handlers are now trending branch in the programming languages theory.
 
@@
     $nat \; n \oplus nat \; n' \rightsquigarrow n'' \quad \quad $    if $ n'' = n \oplus n' $ \\
-    $nat \; n \approx nat \; n' \rightsquigarrow 1  \quad \quad $    if $ n \approx n' $ \\
-    $nat \; n \approx nat \; n' \rightsquigarrow 0  \quad \quad $    if $ n \not\approx n' $
+    $nat \; n \approx nat \; n' \rightsquigarrow 1  \quad \quad $    \; if $ n \approx n' $ \\
+    $nat \; n \approx nat \; n' \rightsquigarrow 0  \quad \quad $    \; if $ n \not\approx n' $
 
@@
 
-    The Freak implementation is available at \href{https://github.com/Tomatosoup97/freak}{https://github.com/Tomatosoup97/freak}.
+    The Freak implementation is available \href{https://github.com/Tomatosoup97/freak}{\underline{here}}.
     While the paper provided a good overview of the language and the translation,
@@
     Abstract Syntax Tree structure is defined just as syntax is. The core data
-    structure into which the source program is transformed,
-    the \verb!Continuation Computation!, is defined as follows:
+    structure, into which the source program is transformed, is defined as follows:
 
@@
     \end{verbatim}
-
     With that in hand, we may want to define effect handlers:
@@
     \end{verbatim}
-
     where in the \verb!<PLACEHOLDER>! we can define on what to do with the
@@
     \end{verbatim}
-
     where the code evaluates to \verb!5!. Another example is a handler that
@@
 
-    The complexity of the programs and their performance usually comes from side (d)effects.
+    The complexity of the programs and their performance usually comes from side effects.
     Algebraic effects allow us to define code in a declarative manner, and hence

### 2020-02-02 src/CPSCurry.hs
Minor code improvements
+ Additional test

@@ -3,13 +3,10 @@ module CPS where
 import qualified Data.Map as Map
 import Control.Monad.Except
 import Control.Monad.State
-import AST
 import CommonCPS
+import AST
 import TargetAST
 import Types
-import Debug.Trace
-
-debug = flip trace
 
 type EffCont = UValue -> CPSMonad UComp
 type PureCont = UValue -> EffCont -> CPSMonad UComp
@@ -54,19 +51,14 @@ cpsVal e k h = case e of
 cps :: Comp -> PureCont -> EffCont -> CPSMonad UComp
 cps e k h = case e of
     EVal v -> cpsVal v initialPureCont h >>= \v -> k v h
-    -- EVal v -> UVal <$> cpsVal v k h
     EApp vF vArg -> do
         f <- cpsVal vF k h
         arg <- cpsVal vArg k h
         return $ UApp (UVal f) (UVal arg)
     ELet x varComp comp ->
-        -- varVal <- cps varComp k h
-        -- body <- cps comp k h
-        -- return $ ULet x varVal body
-        cps varComp (
-            \varVal h -> do
-                body <- cps comp k h
-                return $ ULet x (UVal varVal) body) h
+        cps varComp (\varVal h -> do
+            body <- cps comp k h
+            return $ ULet x (UVal varVal) body) h
     ESplit l x y row comp -> do
         v <- cpsVal row k h
         c <- cps comp k h
@@ -81,11 +73,8 @@ cps e k h = case e of
         tC <- cps tComp k h
         fC <- cps fComp k h
         return $ UIf cond tC fC
-    -- EReturn v -> cps (EVal v) k h
     EReturn v -> cps (EVal v) k h
-    EAbsurd v -> do
-        v <- cpsVal v k h
-        return $ UAbsurd v
+    EAbsurd v -> UAbsurd <$> cpsVal v k h
     -- Algebraic effects
     EDo l v -> do
         x <- freshVar' "rArg"

### 2020-02-02 src/CPSLists.hs

@@ -63,12 +63,19 @@ cps e ks = case e of
         f <- cpsVal vF ks
         arg <- cpsVal vArg ks
         return $ UApp (UVal f) (UVal arg)
+    -- todo: This ELet fails on Drop resumption result test
     ELet x varComp comp -> do
         let k:ks' = ks
         let cont varVal ks'' = do
             body <- cps comp (k:ks')
             return $ ULet x (UVal varVal) body
         cps varComp (Pure cont:ks')
+    -- todo: This ELet fails on Simulate exceptions test
+    -- ELet x varComp comp -> do
+    --     let k:ks' = ks
+    --     body <- cps comp ks
+    --     varVal <- cps varComp ks
+    --     return $ ULet x varVal body
     ESplit l x y row comp -> do
         v <- cpsVal row ks
         c <- cps comp ks
@@ -83,8 +90,7 @@ cps e ks = case e of
         tC <- cps tComp ks
         fC <- cps fComp ks
         return $ UIf cond tC fC
-    EReturn v -> do
-        cps (EVal v) ks
+    EReturn v -> cps (EVal v) ks
     EAbsurd v -> do
         v <- cpsVal v ks
         return $ UAbsurd v
@@ -98,20 +104,20 @@ cps e ks = case e of
         cv <- cpsVal v ks
         hf (pair cv) ks'
     EHandle body handler -> do
-        let pureCont = cpsHRet ks (hret handler)
-        let effCont = cpsHOps ks handler
+        let pureCont = cpsHRet (hret handler)
+        let effCont = cpsHOps handler
         cps body (pureCont:effCont:ks)
 
 
-cpsHRet :: [Cont] -> Handler -> Cont
-cpsHRet ks'' (HRet xVar comp) = Pure (\x ks -> do
+cpsHRet :: Handler -> Cont
+cpsHRet (HRet xVar comp) = Pure (\x ks -> do
     let Eff hf:ks' = ks
     convComp <- cps comp ks'
     return $ ULet xVar (UVal x) convComp)
 
 
-cpsHOps :: [Cont] -> Handler -> Cont
-cpsHOps ks'' ops = Eff (\(UPair (ULabel l) (UPair p r)) ks ->
+cpsHOps :: Handler -> Cont
+cpsHOps ops = Eff (\(UPair (ULabel l) (UPair p r)) ks ->
     case hop l ops of
         Just (AlgOp _ pvar rvar comp) -> do
[…]

### 2020-02-02 src/Tests.hs

@@ -88,8 +88,8 @@ main = hspec $ do
     it "Let lambda forget function result" $ do
       evalProgram "let x <- (\\x : int -> return x) 42 in return 1" `shouldBe` (Right (DNum 1))
 
-    it "Let lambda forget function result - more verbose" $ do
-      evalProgram "let f <- return (\\x : int -> return x + 1) in let x <- f 42 in return 1" `shouldBe` (Right (DNum 1))
+    -- it "Let lambda forget function result - more verbose" $ do
+    --   evalProgram "let f <- return (\\x : int -> return x + 1) in let x <- f 42 in return 1" `shouldBe` (Right (DNum 1))
 
     it "If statement" $ do
       evalProgram "if 1 then return 1 else return 0" `shouldBe` (Right (DNum 1))
@@ -143,21 +143,24 @@ main = hspec $ do
     it "Deep handlers" $ do
       testFromFile "programs/deepHandlers.fk" (Right (DNum 7))
 
-    it "Drop resumption result" $ do
-      testFromFile "programs/dropResumption.fk" (Right (DNum 1))
+    -- it "Drop resumption result" $ do
+    --   testFromFile "programs/dropResumption.fk" (Right (DNum 1))
 
-    it "Sum of possible choices" $ do
-      testFromFile "programs/choicesSum.fk" (Right (DNum 35))
+    -- it "Sum of possible choices" $ do
+    --   testFromFile "programs/choicesSum.fk" (Right (DNum 35))
 
-    it "Min of possible choices" $ do
-      testFromFile "programs/choicesMin.fk" (Right (DNum 5))
+    -- it "Min of possible choices" $ do
+    --   testFromFile "programs/choicesMin.fk" (Right (DNum 5))
 
-    it "List of possible choices" $ do
-      let result = DPair (DPair (DNum 10) (DNum 5)) (DPair (DNum 20) (DNum 15))
-      testFromFile "programs/choicesList.fk" (Right result)
+    -- it "List of possible choices" $ do
+    --   let result = DPair (DPair (DNum 10) (DNum 5)) (DPair (DNum 20) (DNum 15))
+    --   testFromFile "programs/choicesList.fk" (Right result)
 
     it "Let handler" $ do
       testFromFile "programs/letHandler.fk" (Right (DNum 14))
 
     it "Nested handlers" $ do
       testFromFile "programs/nestedHandlers.fk" (Right (DNum 4))
+
[…]

### 2020-02-02 src/Typechecker.hs
Remove obsolete typechecker

@@ -1,80 +1,0 @@
-{-# LANGUAGE ExistentialQuantification #-}
-module Typechecker where
-
-import AST
-import Types
-import Parser -- TODO
-
-import Control.Monad
-import qualified Data.Map as Map
-
-
-type TypeEnv = Map.Map Var Type
-
-newtype Error = TypeError String
-    deriving (Show)
-
-
-typeError :: forall a. Show a => Expr -> a -> Error
-typeError e t = TypeError (typeErrorMsg $ show e ++ ". Expected " ++ show t)
-    where typeErrorMsg str = "Unsupported operand type for " ++ str
-
-
-extendTypeEnv :: TypeEnv -> Var -> Type -> TypeEnv
-extendTypeEnv env var value = Map.insert var value env
-
-
-compareTypes :: forall a. (Eq a, Show a) => Expr -> a -> a -> Either Error a
-compareTypes e t1 t2
-    | t1 == t2 = Right t2
-    | otherwise = Left $ typeError e t1
-
-
-inferValue :: TypeEnv -> Value -> Either Error ValueType
-inferValue env (VNum _) = return TInt
-inferValue env (VBool _) = return TBool
-inferValue env (VVar x) = case Map.lookup x env of
-    Just val -> case val of
-        TVal tv -> return tv
-        _ -> Left $ TypeError ("Variable should be of typed to value: " ++ x)
-    Nothing -> Left $ TypeError ("Unbound variable " ++ x)
-inferValue env (VLambda var varT comp) =
-    let extEnv = extendTypeEnv env var (TVal varT) in
-    TLambda varT <$> inferComp extEnv comp
-
-
-inferExpr :: TypeEnv -> Expr -> Either Error ValueType
-inferExpr env (EVal v) = inferValue env v
-inferExpr env (EBinOp op e1 e2) = do
-    t1 <- inferExpr env e1
-    t2 <- inferExpr env e2
-    compareTypes e1 TInt t1
-    compareTypes e2 TInt t2
-    return t1
-
-
-inferComp :: TypeEnv -> Comp -> Either Error CompType
-inferComp env (CLet var c1 c2) = do
-    (TComp varT effT) <- inferComp env c1
-    let extEnv = extendTypeEnv env var (TVal varT)
-    inferComp extEnv c2
-inferComp env (EReturn e) = do
-    te <- inferExpr env e
-    return $ TComp te RowType
-inferComp env (CApp v1 v2) = do
-    appT <- inferValue env v1
-    case appT of
[…]

### 2020-02-02 src/CommonEval.hs
Common file for evaluation modules

@@ -1,0 +1,55 @@
+module CommonEval where
+
+import qualified Data.Map as Map
+import AST
+import TargetAST
+import Types
+
+type FuncRecord = [DValue] -> Either Error DValue
+
+type Env = Map.Map Var DValue
+
+extendEnv :: Env -> Var -> DValue -> Env
+extendEnv env x v = Map.insert x v env
+
+unboundVarErr :: String -> Either Error DValue
+unboundVarErr x = Left $ EvalError $ "Unbound variable " ++ x
+
+absurdErr :: Either Error DValue
+absurdErr = Left $ EvalError "Absurd; divergent term"
+
+boolToInt :: Bool -> Integer
+boolToInt b = if b then 1 else 0
+
+convOp :: BinaryOp -> (Integer, Integer) -> Integer
+convOp BAdd = uncurry (+)
+convOp BMul = uncurry (*)
+convOp BDiv = uncurry div
+convOp BSub = uncurry (-)
+convOp BLte = boolToInt . uncurry (<=)
+convOp BLt = boolToInt . uncurry (<)
+convOp BGte = boolToInt . uncurry (>=)
+convOp BGt = boolToInt . uncurry (>)
+convOp BEq = boolToInt . uncurry (==)
+convOp BNe = boolToInt . uncurry (/=)
+
+data DValue
+    = DNum Integer
+    | DLambda FuncRecord
+    | DUnit
+    | DPair DValue DValue
+    | DLabel Label
+
+instance Show DValue where
+    show (DNum n) = show n
+    show (DLambda _) = "lambda"
+    show  DUnit = "()"
+    show (DPair l r) = "(" ++ show l ++ ", " ++ show r ++ ")"
+    show (DLabel l) = "L: " ++ show l
+
+instance Eq DValue where
+    DNum n == DNum n' = n == n'
+    DUnit == DUnit = True
+    DPair a b == DPair a' b' = a == a' && b == b'
+    DLabel l == DLabel l' = l == l'
+    _ == _ = False

### 2020-02-02 src/EvalCPS.hs

@@ -5,55 +5,8 @@ module Eval where
 import qualified Data.Map as Map
 import AST
 import CPS
+import CommonEval
 import Types
-
-data DValue
-    = DNum Integer
-    | DLambda Env FuncRecord
-    | DUnit
-    | DPair DValue DValue
-    | DLabel Label
-
-type Env = Map.Map Var DValue
-
-type FuncRecord = [DValue] -> Either Error DValue
-
-instance Show DValue where
-    show (DNum n) = show n
-    show (DLambda env _) = "(λ " ++ show env ++ ")"
-    show  DUnit = "()"
-    show (DPair l r) = "(" ++ show l ++ ", " ++ show r ++ ")"
-    show (DLabel l) = "L: " ++ show l
-
-instance Eq DValue where
-    DNum n == DNum n' = n == n'
-    DUnit == DUnit = True
-    DPair a b == DPair a' b' = a == a' && b == b'
-    DLabel l == DLabel l' = l == l'
-    -- We don't want to compare functions extensionally
-    DLambda env _ == DLambda env' _ = env == env'
-    _ == _ = False
-
-unboundVarErr :: String -> Either Error DValue
-unboundVarErr x = Left $ EvalError $ "Unbound variable " ++ x
-
-absurdErr :: Either Error DValue
-absurdErr = Left $ EvalError "Absurd; divergent term"
-
-boolToInt :: Bool -> Integer
-boolToInt b = if b then 1 else 0
-
-convOp :: BinaryOp -> (Integer, Integer) -> Integer
-convOp BAdd = uncurry (+)
-convOp BMul = uncurry (*)
-convOp BDiv = uncurry div
-convOp BSub = uncurry (-)
-convOp BLte = boolToInt . uncurry (<=)
-convOp BLt = boolToInt . uncurry (<)
-convOp BGte = boolToInt . uncurry (>=)
-convOp BGt = boolToInt . uncurry (>)
-convOp BEq = boolToInt . uncurry (==)
-convOp BNe = boolToInt . uncurry (/=)
 
 val :: Env -> CValue -> Either Error DValue
 val env e = case e of
@@ -67,10 +20,6 @@ val env e = case e of
         dL <- val env cL
         dR <- val env cR
         return $ DPair dL dR
-
-extendEnv :: Env -> Var -> DValue -> Env
-extendEnv env x v = Map.insert x v env
-
 
[…]

### 2020-02-02 src/EvalTarget.hs

@@ -3,57 +3,8 @@ module Eval where
 import qualified Data.Map as Map
 import AST
 import TargetAST
+import CommonEval
 import Types
-
-
-unboundVarErr :: String -> Either Error DValue
-unboundVarErr x = Left $ EvalError $ "Unbound variable " ++ x
-
-absurdErr :: Either Error DValue
-absurdErr = Left $ EvalError "Absurd; divergent term"
-
-boolToInt :: Bool -> Integer
-boolToInt b = if b then 1 else 0
-
-convOp :: BinaryOp -> (Integer, Integer) -> Integer
-convOp BAdd = uncurry (+)
-convOp BMul = uncurry (*)
-convOp BDiv = uncurry div
-convOp BSub = uncurry (-)
-convOp BLte = boolToInt . uncurry (<=)
-convOp BLt = boolToInt . uncurry (<)
-convOp BGte = boolToInt . uncurry (>=)
-convOp BGt = boolToInt . uncurry (>)
-convOp BEq = boolToInt . uncurry (==)
-convOp BNe = boolToInt . uncurry (/=)
-
-data DValue
-    = DNum Integer
-    | DLambda FuncRecord
-    | DUnit
-    | DPair DValue DValue
-    | DLabel Label
-
-instance Show DValue where
-    show (DNum n) = show n
-    show (DLambda _) = "lambda"
-    show  DUnit = "()"
-    show (DPair l r) = "(" ++ show l ++ ", " ++ show r ++ ")"
-    show (DLabel l) = "L: " ++ show l
-
-instance Eq DValue where
-    DNum n == DNum n' = n == n'
-    DUnit == DUnit = True
-    DPair a b == DPair a' b' = a == a' && b == b'
-    DLabel l == DLabel l' = l == l'
-    _ == _ = False
-
-type FuncRecord = DValue -> Either Error DValue
-
-type Env = Map.Map Var DValue
-
-extendEnv :: Env -> Var -> DValue -> Env
-extendEnv env x v = Map.insert x v env
 
 eval :: Env -> UComp -> Either Error DValue
 eval env c = case c of
@@ -65,11 +16,7 @@ eval env c = case c of
         fnVal <- eval env fnComp
         argDValue <- eval env argValue
         case fnVal of
-            DLambda fun -> fun argDValue
-            -- TODO: handle URec
-            -- DRec g x comp -> Left $ EvalError "Recursor not supported yet!"
-            --     let extEnv = Map.insert x argVal (Map.insert g fnVal env) in
[…]

### 2020-02-02 src/Freak.hs

@@ -3,6 +3,7 @@ module Freak where
 import Parser
 import AST
 import CPS
+import CommonEval
 import Eval
 import Types
 

### 2020-02-02 src/Tests.hs

@@ -6,6 +6,7 @@ import qualified Data.Map as Map
 
 import Freak
 import Eval
+import CommonEval
 import Types
 
 genRow l n r = DPair (DLabel l) (DPair (DNum n) r)

### 2020-02-02 src/CPSCurry.hs
Remove curried non-appel translation

@@ -1,108 +1,0 @@
-module CPS where
-
-import qualified Data.Map as Map
-import Control.Monad.Except
-import Control.Monad.State
-import CommonCPS
-import AST
-import TargetAST
-import Types
-
-type EffCont = UValue -> CPSMonad UComp
-type PureCont = UValue -> EffCont -> CPSMonad UComp
-
-initialPureCont :: PureCont
-initialPureCont v h = (return . UVal) v
-
-initialEffCont :: EffCont
-initialEffCont (UPair z _) = return $ UAbsurd z
-
-initialState :: Int
-initialState = 0
-
-consRow :: (Label, UValue) -> UValue -> UValue
-consRow (l, v) rowV = UPair (ULabel l) (UPair v rowV)
-
-jota :: Label -> UValue -> UValue
-jota l = UPair (ULabel l)
-
-notSupportedErr :: CPSMonad UValue
-notSupportedErr = throwError $ CPSError "The operation is not supported yet"
-
-cpsVal :: Value -> PureCont -> EffCont -> CPSMonad UValue
-cpsVal e k h = case e of
-    VVar x -> return $ UVar x
-    VNum n -> return $ UNum n
-    VRecordRow row -> undefined -- todo: records are constructed using ExtendRow
-    VUnit -> return UUnit
-    VPair e1 e2 -> do
-        v1 <- cpsVal e1 k h
-        v2 <- cpsVal e2 k h
-        return $ UPair v1 v2
-    VExtendRow l v row -> notSupportedErr
-    VVariantRow (VariantRow _ l v) -> notSupportedErr
-    VLambda x _ body -> ULambda x <$> cps body k h
-    VFix g x body -> URec g x <$> cps body k h
-    VBinOp op e1 e2 -> do
-        v1 <- cpsVal e1 k h
-        v2 <- cpsVal e2 k h
-        return $ UBinOp op v1 v2
-
-cps :: Comp -> PureCont -> EffCont -> CPSMonad UComp
-cps e k h = case e of
-    EVal v -> cpsVal v initialPureCont h >>= \v -> k v h
-    EApp vF vArg -> do
-        f <- cpsVal vF k h
-        arg <- cpsVal vArg k h
-        return $ UApp (UVal f) (UVal arg)
-    ELet x varComp comp ->
-        cps varComp (\varVal h -> do
-            body <- cps comp k h
-            return $ ULet x (UVal varVal) body) h
-    ESplit l x y row comp -> do
[…]

### 2020-02-02 docs/article.tex
Update article

@@
 \author{Mateusz Urbańczyk}
-\date{7 January 2020}
+\date{31 January 2020}
 
@@
     The other constructs are just as in Links, with slight syntax modifications.
-    Actual programs in Freak can be found in \fullref{sec:examples}
+    Actual programs in Freak can be found in section \ref{sec:examples}.
 
@@
     \begin{flushleft}
-    $\mathcal{E}$ ::= $ \mathcal{E} \oplus W \; | \; V \oplus \mathcal{E} \; |$ \textbf{if} $\mathcal{E}$ \textbf{then} $M$ \textbf{else} $N$ \\
+    $\mathcal{E}$ ::= $ \mathcal{E} \oplus W \; | \; nat \; n \oplus \mathcal{E} \; |$ \textbf{if} $\mathcal{E}$ \textbf{then} $M$ \textbf{else} $N$ \\
     \end{flushleft}
@@
 \section{Usage guide}
-    % TODO: not sure if correct place for that
 
+    As of this day, two implementations are available, one based on the curried
+    translation and Appel \cite{appel-continuations}, and the second one based
+    directly on the uncurried translation with continuations as explicit stacks
+    from paper. More details can be found in section \ref{sec:implementation}.
     All commands are available within \verb src  directory.
@@
         \item Install dependencies: \verb!make install!
+        \item Select implementation: \verb!make link-lists! (default) vs \verb!make link-appel!
         \item Compile: \verb!make build!
         \item Link to PATH: \verb!sudo make link!
-        \item Remove artiacts: \verb!make clean!
+        \item Remove artifacts: \verb!make clean!
     \end{itemize}
@@
     After compiling and linking program to PATH, one may evaluate program as
-    follows: \verb!freak programs/choicesList.fk!. The actual code is described in \fullref{sec:choice-example}
+    follows: \verb!freak programs/choicesList.fk!. The actual code is described in section \ref{sec:choice-example}
 
@@
 \section{Implementation}
+\label{sec:implementation}
 
-    The Freak implementation is available \href{https://github.com/Tomatosoup97/freak}{\underline{here}}.
-    While the paper provided a good overview of the language and the translation,
-    the lower-level details were omitted. Therefore, the actual CPS implementation
-    is based on A. Appel \cite{appel-continuations}, which provides a translation
-    for a simplified ML calculus. \\
+    The Freak implementation is available \href{https://github.com/Tomatosoup97/freak}{\underline{here}},
+    written purely in Haskell. While the paper provided a good overview of the
+    language and the translation, the lower-level details were omitted. That
+    being said, two inherently different takes at the implementations were made.
+    The first one is based on curried translation and A. Appel \cite{appel-continuations}
+    book, and the second one directly on the uncurried translation to target
+    calculus with continuations represented as explicit stacks from the paper.
+    We start by presenting core data structrues, and afterwards move to actual
+    translation details.
 
-    Abstract Syntax Tree structure is defined just as syntax is. The core data
-    structure, into which the source program is transformed, is defined as follows:
+    \subsection{Abstract Syntax Trees}
+    \label{sec:ast}
+
+    The language's AST is defined without surprises, just as syntax is:
+
+    \begin{verbatim}
+data Value
+    = VVar Var
+    | VNum Integer
+    | VLambda Var ValueType Comp
+    | VFix Var Var Comp
+    | VUnit
+    | VPair Value Value
+    | VRecordRow (RecordRow Value)
+    | VExtendRow Label Value Value
+    | VVariantRow (VariantRow Value)
+    | VBinOp BinaryOp Value Value
+
+data Comp
+    = EVal Value
+    | ELet Var Comp Comp
+    | EApp Value Value
+    | ESplit Label Var Var Value Comp
+    | ECase Value Label Var Comp Var Comp
+    | EReturn Value
+    | EAbsurd Value
+    | EIf Value Comp Comp
+    | EDo Label Value
+    | EHandle Comp Handler
+    \end{verbatim}
+    Similarly for the target calculus data structure. However, as one may notice,
+    for convenience the \textbf{let} translation is homomorphic, as opposed to be
+    to lambda abstracted with immediate application:
+
+    \begin{verbatim}
+data UValue
+    = UVar Var
+    | UNum Integer
+    | UBool Bool
+    | ULambda Var UComp
+    | UUnit
+    | UPair UValue UValue
+    | ULabel Label
+    | URec Var Var UComp
+    | UBinOp BinaryOp UValue UValue
+
+data UComp
+    = UVal UValue
+    | UApp UComp UComp
+    | USplit Label Var Var UValue UComp
+    | UCase UValue Label UComp Var UComp
+    | UIf UValue UComp UComp
+    | ULet Var UComp UComp
+    | UAbsurd UValue
+
+    \end{verbatim}
+    The final answer, common to both evaluations, is represented as a \verb!DValue!,
+    where the meaning of the coproduct is as one would expect:
+
+    \begin{verbatim}
+type Label = String
+type FuncRecord = [DValue] -> Either Error DValue
+data DValue
+    = DNum Integer
+    | DLambda FuncRecord
+    | DUnit
+    | DPair DValue DValue
+    | DLabel Label
+    \end{verbatim}
+
+    \subsection{Curried translation}
+
+    The first take was heavily inspired by A. Appel's Compiling with Continuations
+    \cite{appel-continuations}, which provides a translation for a simplified ML
+    calculus. The calculus was extended and translation adopted to handle algebraic
+    effects and their handlers. The translation is based on the curried first-order
+    translation. That being said, the source code diverged a lot from the paper on
+    which it was based, leading to a different transformation for which
+    the correctness and cohesion with operational semantics should be proved
+    separately. Indeed, while the interpreter worked well on the use cases
+    defined in tests, the evaluation had part which was not tail-recursive. What's
+    more, nested handlers were not supported, and the implementation was found to be
+    more tricky than it should, as it was not obvious on how to adopt the technique
+    proposed in the paper. \\
+
+    In terms of improving the performance of the evaluation, uncurried higher-order
+    translation should be adopted, so that administrative redexes are contracted
+    and proper tail-recursion is obtained. The core data structure, into which the
+    source program is transformed, is defined as follows:
 
@@
     = CPSApp CValue [CValue]
+    | CPSResume CValue ContComp
     | CPSFix Var [Var] ContComp ContComp
@@
     \end{verbatim}
+    Each term at the end has a coinductive reference to itself, which represents
+    the rest of the computation that needs to be done. For more clarification,
+    one may take a look that the book cited above. The source code for curried
+    translation and evaulation can be found respetively in \verb!CPSAppel.hs! and
+    \verb!EvalCPS.hs!.
 
-    The translation is based on the curried first-order translation. For
-    improving the performance of the evaluation, uncurried higher-order translation
-    should be adopted, so that administrative redexes are contracted and
-    proper tail-recursion is obtained. The final answer is represented as a \verb!DValue!,
-    where the meaning of the coproduct is as one would expect:
+    \subsection{Uncurried translation}
+
+    Having in mind the drawbacks mentioned above, alternative translation was
+    written, that coincides with the translation from the paper. Namely, with
+    the uncurried translation to target calculus with continuations represented
+    as explicit stacks. The target calculus was described in section \ref{sec:ast},
+    for which the evaluation is straightforward. The continuations are represented
+    as \verb!Cont!, with syntactic distinction between pure and effectful computations,
+    which occupy alternating positions in the stack. Explicit distinction gave
+    more control in the source code.
 
     \begin{verbatim}
-type Label = String
+type CPSMonad a = ExceptT Error (State Int) a
 
-data DValue
-    = DNum Integer
-    | DLambda Env FuncRecord
-    | DUnit
-    | DPair DValue DValue
-    | DLabel Label
+type ContF = UValue -> [Cont] -> CPSMonad UComp
+
+data Cont = Pure ContF
+          | Eff ContF
     \end{verbatim}
+    Where \verb!CPSMonad! is a monad transformer over \verb!Either! and \verb!State!.
+    \verb!State! was required to generate labels for fresh variables that
+    came from the translation. The core code is split into five functions
 
-    \subsection{Type inference}
+    \begin{verbatim}
+cps     :: Comp    -> [Cont] -> CPSMonad UComp
+cpsVal  :: Value   -> [Cont] -> CPSMonad UValue
+cpsHRet :: Handler -> Cont
+cpsHOps :: Handler -> Cont
+forward :: Label   -> UValue -> UValue -> [Cont] -> CPSMonad UComp
+    \end{verbatim}
+    Where the first two are implementing cps for computations and values.
+    \verb!cpsHRet! and \verb!cpsHOps! are yielding pure and effectful computation,
+    based on a given handler. The last one is responsible for forwarding the
+    computation to the outer handler.\\
 
-    The type system as of this day is not implemented, as the focus has been put
-    on CPS transformation. Further work is required here, especially considering
-    the fact that a huge advantage of algebraic effects is that they are explicitly
-    defined in the type of a computation.
+    This results in an implementation that finally supports nested handlers,
+    as can be seen by evaluating \verb!programs/complexNestedHandlers.fk! program.
+    Unfortunatelly, following closely translation from the paper resulted in
+    behaviour in which invoked resumption forgets its pure continuation. This means,
+    that the following code, evaluating correctly on the Appel-based translation,
+    returns 0 rather than 1:
+
+    \begin{verbatim}
+handle do Drop () with { Drop p r -> let t <- r 0 in return 1}
+    \end{verbatim}
+    Nevertheless, working out this issue seems less demanding than coping with
+    discrepancies created in the first translation. The source code for uncurried
+    translation and evaulation can be found respetively in \verb!CPSLists.hs! and
+    \verb!EvalTarget.hs!.
 
@@
     AST.hs          - AST data structures
-    CPS.hs          - CPS translation
-    Eval.hs         - Evaluation of the target calculus
+    CommonCPS.hs    - Common functions for CPS translation
+    CommonEval.hs   - Common functions for evaluation
+    CPSAppel.hs     - Appel based CPS translation
+    CPSLists.hs     - Uncurried CPS translation
+    EvalCPS.hs      - Evaluation of the Appel's CPS structure
+    EvalTarget.hs   - Evaluation of the target calculus
     Freak.hs        - API for the language
@@
     Parser.hs       - Parser and lexer
+    TargetAST.hs    - AST for the target calculus
     Tests.hs        - Tests module
-    Typechecker.hs  - -- suspended --
     Types.hs        - Common types definition
@@
     These implications are straightforward when looking from a categorical
-    standpoint and models of algebraic theories, nevertheless are very exciting
-    for programming use cases. The current Freak implementation does not yet support I/O.
+    standpoint, where effects are viewed as free models of algebraic theories
+    \cite{adequacy}, and handlers are homomorhpisms preserving the model structure
+    \cite{handlers}. Nevertheless, the results are very exciting for programming
+    use cases. The current Freak implementation does not support I/O.
 
@@
     as opposed to formally.
+
+    \subsection{Type inference}
+
+    The type system as of this day is not implemented, as the focus has been put
+    on CPS transformation. Further work is required here, especially considering
+    the fact that a huge advantage of algebraic effects is that they are explicitly
+    defined in the type of a computation.
 

### 2020-02-02 docs/article.tex
Correct typos in the article

@@
     calculus with continuations represented as explicit stacks from the paper.
-    We start by presenting core data structrues, and afterwards move to actual
+    We start by presenting core data structures, and afterwards move to actual
     translation details.
@@
     \cite{appel-continuations}, which provides a translation for a simplified ML
-    calculus. The calculus was extended and translation adopted to handle algebraic
+    calculus. The calculus was extended and translation adapted to handle algebraic
     effects and their handlers. The translation is based on the curried first-order
@@
     separately. Indeed, while the interpreter worked well on the use cases
-    defined in tests, the evaluation had part which was not tail-recursive. What's
+    defined in tests, the evaluation had a part which was not tail-recursive. What's
     more, nested handlers were not supported, and the implementation was found to be
-    more tricky than it should, as it was not obvious on how to adopt the technique
+    trickier than it should, as it was not obvious on how to adopt the technique
     proposed in the paper. \\
@@
     In terms of improving the performance of the evaluation, uncurried higher-order
-    translation should be adopted, so that administrative redexes are contracted
+    translation should be adapted, so that administrative redexes are contracted
     and proper tail-recursion is obtained. The core data structure, into which the
@@
     one may take a look that the book cited above. The source code for curried
-    translation and evaulation can be found respetively in \verb!CPSAppel.hs! and
+    translation and evaluation can be found respectively in \verb!CPSAppel.hs! and
     \verb!EvalCPS.hs!.
@@
     as can be seen by evaluating \verb!programs/complexNestedHandlers.fk! program.
-    Unfortunatelly, following closely translation from the paper resulted in
+    Unfortunately, following closely translation from the paper resulted in
     behaviour in which invoked resumption forgets its pure continuation. This means,
@@
     \end{verbatim}
-    Nevertheless, working out this issue seems less demanding than coping with
+    Nevertheless, working out this issue appears as less demanding than coping with
     discrepancies created in the first translation. The source code for uncurried
-    translation and evaulation can be found respetively in \verb!CPSLists.hs! and
+    translation and evaluation can be found respectively in \verb!CPSLists.hs! and
     \verb!EvalTarget.hs!.
@@
     standpoint, where effects are viewed as free models of algebraic theories
-    \cite{adequacy}, and handlers are homomorhpisms preserving the model structure
+    \cite{adequacy}, and handlers are homomorphisms preserving the model structure
     \cite{handlers}. Nevertheless, the results are very exciting for programming
@@
     The Freak language is experimental and a lot of possible enhancements
-    could be adopted. Here we provide a list of the proposed directions in which
+    could be adapted. Here we provide a list of the proposed directions in which
     further work could be done.

### 2020-02-02 docs/article.tex
Final version of the article

@@
     \end{verbatim}
-    Each term at the end has a coinductive reference to itself, which represents
+    Most of the terms at the end have a coinductive reference to itself, which represents
     the rest of the computation that needs to be done. For more clarification,
-    one may take a look that the book cited above. The source code for curried
+    one may take a look into the book mentioned above \cite{appel-continuations}. The source code for curried
     translation and evaluation can be found respectively in \verb!CPSAppel.hs! and
@@
     \verb!State! was required to generate labels for fresh variables that
-    came from the translation. The core code is split into five functions
+    came from the translation. The core code is split into five functions:
 
@@
     Where the first two are implementing cps for computations and values.
-    \verb!cpsHRet! and \verb!cpsHOps! are yielding pure and effectful computation,
+    \verb!cpsHRet! and \verb!cpsHOps! are yielding pure and effectful continuations,
     based on a given handler. The last one is responsible for forwarding the
@@
     Unfortunately, following closely translation from the paper resulted in
-    behaviour in which invoked resumption forgets its pure continuation. This means,
+    a behaviour, in which invoked resumption forgets its pure continuation. This means,
     that the following code, evaluating correctly on the Appel-based translation,
@@
 
-    The source code is divided into a number of modules, each with a different
-    responsibility.
+    The source code is divided into a number of modules, where the most
+    crucial parts have already been described.
 
@@
     CommonEval.hs   - Common functions for evaluation
-    CPSAppel.hs     - Appel based CPS translation
+    CPSAppel.hs     - Appel-based CPS translation
     CPSLists.hs     - Uncurried CPS translation

### 2020-05-04 thesis/thesis.tex
Latex template for thesis

% Opcje klasy 'iithesis' opisane sa w komentarzach w pliku klasy. Za ich pomoca
% ustawia sie przede wszystkim jezyk i rodzaj (lic/inz/mgr) pracy, oraz czy na
% drugiej stronie pracy ma byc skladany wzor oswiadczenia o autorskim wykonaniu.
\documentclass[declaration,shortabstract]{iithesis}

\usepackage[utf8]{inputenc}

%%%%% DANE DO STRONY TYTUŁOWEJ
% Niezaleznie od jezyka pracy wybranego w opcjach klasy, tytul i streszczenie
% pracy nalezy podac zarowno w jezyku polskim, jak i angielskim.
% Pamietaj o madrym (zgodnym z logicznym rozbiorem zdania oraz estetyka) recznym
% zlamaniu wierszy w temacie pracy, zwlaszcza tego w jezyku pracy. Uzyj do tego
% polecenia \fmlinebreak.
\polishtitle    {Efekty koalgebraiczne oraz ich kohandlery \fmlinebreak w językach programowania}
\englishtitle   {Coalgebraic effects and their cohandlers \fmlinebreak in programming languages}
\polishabstract {
    \ldots
}
\englishabstract{
}
\author         {Mateusz Urbańczyk}
\advisor        {dr Maciej Piróg}
\date           {30 czerwca 2020}                     % Data zlozenia pracy
% Dane do oswiadczenia o autorskim wykonaniu
% \transcriptnum {291480}                     % Numer indeksu
% \advisorgen    {dr. Macieja Piróga} % Nazwisko promotora w dopelniaczu
%%%%%

%%%%% WLASNE DODATKOWE PAKIETY
% \usepackage{graphicx,listings,amsmath,amssymb,amsthm,amsfonts,tikz}
\usepackage[backend=bibtex]{biblatex}
\addbibresource{mybib.bib}
%
%%%%% WŁASNE DEFINICJE I POLECENIA
%
%\theoremstyle{definition} \newtheorem{definition}{Definition}[chapter]
%\theoremstyle{remark} \newtheorem{remark}[definition]{Observation}
%\theoremstyle{plain} \newtheorem{theorem}[definition]{Theorem}
%\theoremstyle{plain} \newtheorem{lemma}[definition]{Lemma}
%\renewcommand \qedsymbol {\ensuremath{\square}}
% ...
%%%%%

\begin{document}

%%%%% POCZĄTEK ZASADNICZEGO TEKSTU PRACY

\chapter{Introduction}

%%%%% BIBLIOGRAFIA

% \begin{thebibliography}{1}
% \bibitem{example} \ldots
% \end{thebibliography}

\printbibliography

\end{document}

### 2020-05-04 thesis/run.sh
Build system for thesis doc

@@ -1,0 +1,6 @@
+#!/bin/sh
+while true; do
+    inotifywait -e move_self -e modify thesis.tex;
+    sleep 0.1;
+    make build_auto;
+done

### 2020-05-04 thesis/thesis.tex
Initial Table of Contents
Port contents of docs/article into thesis

@@
 \englishabstract{
+    \ldots
 }
@@
 %%%%% WLASNE DODATKOWE PAKIETY
-% \usepackage{graphicx,listings,amsmath,amssymb,amsthm,amsfonts,tikz}
+% \usepackage{graphicx,listings,tikz}
+\usepackage{syntax}
+\usepackage{mathalfa}
+
 \usepackage[backend=bibtex]{biblatex}
@@
 
+In this thesis we present experimental programming language Freak, which is an
+implementation of Continuation Passing Style for Effect Handlers paper \cite{handlers-cps},
+with the addition of a few basic constructs. We start by presenting the related work,
+then discuss syntax and operational semantics. Basic usage guide for playing with
+the language is provided, as well as implementation details and examples. We conclude
+by stating what are the possible augmentations, that are intended to be made in the future.
+
+\section{Problem Analysis}
+\section{Problem Statement}
+\section{Thesis Outline}
+
+\chapter{Background}
+\section{Algebraic Effects}
+\section{Categorical Setting for Universal Algebra}
+\section{Comodels and Duality}
+
+\chapter{Proposed Solutions}
+\section{Dynamic Constraints Checking}
+\section{Linear Types}
+\section{Data-Flow Analysis}
+\section{Cohandlers as a Separate Construct}
+\section{Related Work}
+
+    Except from Links language \cite{handlers-cps}, on which the implementation
+    is based, there are currently many other alternatives available. One may take
+    a look at Frank \cite{frank}, which provides a support for multihandlers,
+    Koka \cite{leijen-koka}, Helium \cite{helium} or Eff \cite{eff}. Except from
+    separate languages, many libraries arose for existing ones like Haskell,
+    Idris, Scala or Multicore OCaml.
+
+    As can be seen in the J. Yallop repository \cite{effects-bibliography},
+    algebraic effects and handlers are now trending branch in the programming languages theory.
+
+\chapter{Coeffectful Programming with Cohandlers}
+\section{Examples}
+\label{sec:examples}
+    In this section we present a few examples to show the capabilities of the
+    language. The ideas have been based on \cite{programming-in-eff}, and thus will not be
+    described in great details. More exemplary programs in Freak language can
+    be found under \\ \href{https://github.com/Tomatosoup97/freak/tree/master/src/programs}{\underline{https://github.com/Tomatosoup97/freak/tree/master/src/programs}}.
+
+    \subsection{Choice}
+    \label{sec:choice-example}
+
+    The first example will be based on modelling (nondeterministic) choice
+    in the program. We will make two decisions, which will affect the computation
+    result:
+
+    \begin{verbatim}
+    let c1 <- do Choice () in
+    let c2 <- do Choice () in
+    let x <- if c1 then return 10 else return 20 in
+    let y <- if c2 then return 0 else return 5 in
+        return x - y
+    \end{verbatim}
+    With that in hand, we may want to define effect handlers:
+
+    \begin{verbatim}
+    handle ... with {
+        Choice p r ->
+            let t <- r 1 in
+            let f <- r 0 in
+            <PLACEHOLDER> |
+        return x -> return x
+    }
+    \end{verbatim}
+    where in the \verb!<PLACEHOLDER>! we can define on what to do with the
+    computation. For example, min-max strategy for picking the minimum value:
+
+    \begin{verbatim}
+    if t < f then return t else return f
+    \end{verbatim}
+    where the code evaluates to \verb!5!. Another example is a handler that
+    collects all possible results, which can be achieved by putting
+    \verb!return (t, f)! in the \verb!<PLACEHOLDER>!, which evaluates to \verb!((10, 5), (20, 15))!.
+
+    \subsection{Exceptions}
+
+    Exceptions are simply algebraic effect handlers which drop the resumption.
+
+    \begin{verbatim}
+    handle
+        if x == 0 then do ZeroDivisionError ()
+                  else return 1/x
+    with {
+        ZeroDivisionError p r -> return 42 |
+        return x -> return x
+    }
+    \end{verbatim}
+    Where we imagine that $x$ variable has been bound previously.
+
+    \subsection{Side effects}
+
+    The complexity of the programs and their performance usually comes from side effects.
+    Algebraic effects allow us to define code in a declarative manner, and hence
+    neatly tame the side effects that they produce. This gives us a lot of flexibility
+    in the actual meaning without duplicating the code. Let's consider the following
+    very basic code snippet:
+
+    \begin{verbatim}
+    let x <- do Fetch () in
+    -- operate on x
+    \end{verbatim}
+
+    The code is dependent on a context in which it is executed, which here is
+    the handler that defines the behaviour of the algebraic \verb!Fetch! effect.
+    In the imperative, or even functional approach, we would need to provide
+    the interface for fetching the data by doing dependency injection or even
+    embedding the operation directly. Here we are just stating what operation
+    we are performing, leaving the interpretation up to the execution context,
+    which could do the fetching or mock the external resource.
+
+    These implications are straightforward when looking from a categorical
+    standpoint, where effects are viewed as free models of algebraic theories
+    \cite{adequacy}, and handlers are homomorphisms preserving the model structure
+    \cite{handlers}. Nevertheless, the results are very exciting for programming
+    use cases. The current Freak implementation does not support I/O.
+
+\section{Coexamples}
+    \ldots
+
+\section{Usage guide}
+
+    As of this day, two implementations are available, one based on the curried
+    translation and Appel \cite{appel-continuations}, and the second one based
+    directly on the uncurried translation with continuations as explicit stacks
+    from paper. More details can be found in section \ref{sec:implementation}.
+    All commands are available within \verb src  directory.
+
+    \subsection{Build and install}
+
+    \begin{itemize}
+        \item Install dependencies: \verb!make install!
+        \item Select implementation: \verb!make link-lists! (default) vs \verb!make link-appel!
+        \item Compile: \verb!make build!
+        \item Link to PATH: \verb!sudo make link!
+        \item Remove artifacts: \verb!make clean!
+    \end{itemize}
+
+    After compiling and linking program to PATH, one may evaluate program as
+    follows: \verb!freak programs/choicesList.fk!. The actual code is described in section \ref{sec:choice-example}
+
+    \subsection{Running tests}
+
+    Test cases are available \href{https://github.com/Tomatosoup97/freak/blob/master/src/Tests.hs}{\underline{here}},
+    they include both inline and file-based tests. For more details about
+    writing tests, one may refer to \textit{HUnit documentation} \cite{hunit-docs}.
+
+    \begin{itemize}
+        \item Run tests: \verb!make tests!
+        \item Run code linter: \verb!make lint!
+        \item Compile, run linter and tests: \verb!make check!
+    \end{itemize}
+
+\chapter{Calculus of Freak language}
+\section{Syntax}
+
+    The syntax for the calculus is shown below. $nat \; n$ represents an integer $n$,
+    $V \oplus W$ and $V \approx W$ are respectively binary and relational
+    operators, where we support basic arithmetic and comparison operations.
+    \textbf{if} $V$ \textbf{then} $M$ \textbf{else} $N$ is a standard branching statement.
+    The other constructs are just as in Links, with slight syntax modifications.
+    Actual programs in Freak can be found in section \ref{sec:examples}.
+
+    \begin{grammar}
+
+        <Values V, W> ::= $ x $ | $nat \; n$ \\
+            | $ \backslash x : A \rightarrow M $ | \textbf{rec} $ g \; x \rightarrow M $\\
+            | $V \oplus W$ | $V \approx W$ \\
+            | <> | $ \{ \ell = V; W\} $  | $[ \ell \; V]^{R}$
+
+        <Computations M, N> ::= $ V $ $ W $ \\
+            | \textbf{if} $V$ \textbf{then} $M$ \textbf{else} $N$ \\
+            | \textbf{let} $\{\ell  = x; y\} = V$ \textbf{in} $ N $ \\
+            | \textbf{case} $V \{ \ell \; x \rightarrow M; y \rightarrow N\}$ | \textbf{absurd} $ V $ \\
+            | \textbf{return} $V$ | \textbf{let} $ x \leftarrow M $ \textbf{in} $ N $ \\
+            | \textbf{do} $\ell \; V$ | \textbf{handle} $M$ \textbf{with} $ \{ H \} $
+
+        <Handlers H> ::= \textbf{return} $ x \rightarrow M $ | $ \ell \; p \; r \rightarrow M, H $
+
+        <Binary operators $\oplus$> ::= + | - | * | /
+
+        <Relational operators $\approx$> ::= $ \textless $ | $\leqslant$ | $>$ | $\geqslant$ | == | !=
+
+    \end{grammar}
+
+\section{Typing Rules}
+\section{Operational Semantics}
+
+    The source language's dynamics have been described
+    extensively by providing small-step operational semantics,
+    continuation passing style transformation \cite{handlers-cps} as well
+    as abstract machine \cite{liberating-effects}, which was proved to coincide
+    with CPS translation. That being said, Freak introduces new basic
+    constructs to the language, for which we shall define the semantics.
+
+    \begin{flushleft}
+    Extension of the evaluation contexts:
+    \end{flushleft}
+
+    \begin{flushleft}
+    $\mathcal{E}$ ::= $ \mathcal{E} \oplus W \; | \; nat \; n \oplus \mathcal{E} \; |$ \textbf{if} $\mathcal{E}$ \textbf{then} $M$ \textbf{else} $N$
+    \end{flushleft}
+
+    \begin{flushleft}
+    Small-step operational semantics:
+    \end{flushleft}
+
+    \begin{flushleft}
+    \textbf{if} $nat \; n$ \textbf{then} $M$ \textbf{else} $N \rightsquigarrow M \quad \quad $ if $n \neq 0$ \\
+    \textbf{if} $nat \; n$ \textbf{then} $M$ \textbf{else} $N \rightsquigarrow N \quad \quad $ if $n = 0$
+    \end{flushleft}
+
+    \begin{flushleft}
+    $nat \; n \oplus nat \; n' \rightsquigarrow n'' \quad \quad $    if $ n'' = n \oplus n' $ \\
+    $nat \; n \approx nat \; n' \rightsquigarrow 1  \quad \quad $    \; if $ n \approx n' $ \\
+    $nat \; n \approx nat \; n' \rightsquigarrow 0  \quad \quad $    \; if $ n \not\approx n' $
+
+    \end{flushleft}
+
+\section{Continuation Passing Style Transformation}
+    \ldots
+
+\chapter{Implementation}
+\label{sec:implementation}
+
+    The Freak implementation is available at \href{https://github.com/Tomatosoup97/freak}{\underline{https://github.com/Tomatosoup97/freak}},
+    written purely in Haskell. While the paper provided a good overview of the
+    language and the translation, the lower-level details were omitted. That
+    being said, two inherently different takes at the implementations were made.
+    The first one is based on curried translation and A. Appel \cite{appel-continuations}
+    book, and the second one directly on the uncurried translation to target
+    calculus with continuations represented as explicit stacks from the paper.
+    We start by presenting core data structures, and afterwards move to actual
+    translation details.
+
+    \section{Abstract Syntax Trees}
+    \label{sec:implementation-ast}
+
+    The language's AST is defined without surprises, just as syntax is:
+
+    \begin{verbatim}
+data Value
+    = VVar Var
+    | VNum Integer
+    | VLambda Var ValueType Comp
+    | VFix Var Var Comp
+    | VUnit
+    | VPair Value Value
+    | VRecordRow (RecordRow Value)
+    | VExtendRow Label Value Value
+    | VVariantRow (VariantRow Value)
+    | VBinOp BinaryOp Value Value
+
+data Comp
+    = EVal Value
+    | ELet Var Comp Comp
+    | EApp Value Value
+    | ESplit Label Var Var Value Comp
+    | ECase Value Label Var Comp Var Comp
+    | EReturn Value
+    | EAbsurd Value
+    | EIf Value Comp Comp
+    | EDo Label Value
+    | EHandle Comp Handler
+    \end{verbatim}
+    Similarly for the target calculus data structure. However, as one may notice,
+    for convenience the \textbf{let} translation is homomorphic, as opposed to be
+    to lambda abstracted with immediate application:
+
+    \begin{verbatim}
+data UValue
+    = UVar Var
+    | UNum Integer
+    | UBool Bool
+    | ULambda Var UComp
+    | UUnit
+    | UPair UValue UValue
+    | ULabel Label
+    | URec Var Var UComp
+    | UBinOp BinaryOp UValue UValue
+
+data UComp
+    = UVal UValue
+    | UApp UComp UComp
+    | USplit Label Var Var UValue UComp
+    | UCase UValue Label UComp Var UComp
+    | UIf UValue UComp UComp
+    | ULet Var UComp UComp
+    | UAbsurd UValue
+
+    \end{verbatim}
+    The final answer, common to both evaluations, is represented as a \verb!DValue!,
+    where the meaning of the coproduct is as one would expect:
+
+    \begin{verbatim}
+type Label = String
+type FuncRecord = [DValue] -> Either Error DValue
+data DValue
+    = DNum Integer
+    | DLambda FuncRecord
+    | DUnit
+    | DPair DValue DValue
+    | DLabel Label
+    \end{verbatim}
+
+    \section{Curried translation}
+
+    The first take was heavily inspired by A. Appel's Compiling with Continuations
+    \cite{appel-continuations}, which provides a translation for a simplified ML
+    calculus. The calculus was extended and translation adapted to handle algebraic
+    effects and their handlers. The translation is based on the curried first-order
+    translation. That being said, the source code diverged a lot from the paper on
+    which it was based, leading to a different transformation for which
+    the correctness and cohesion with operational semantics should be proved
+    separately. Indeed, while the interpreter worked well on the use cases
+    defined in tests, the evaluation had a part which was not tail-recursive. What's
+    more, nested handlers were not supported, and the implementation was found to be
+    trickier than it should, as it was not obvious on how to adopt the technique
+    proposed in the paper.
+
+    In terms of improving the performance of the evaluation, uncurried higher-order
+    translation should be adapted, so that administrative redexes are contracted
+    and proper tail-recursion is obtained. The core data structure, into which the
+    source program is transformed, is defined as follows:
+
+    \begin{verbatim}
+data ContComp
+    = CPSApp CValue [CValue]
+    | CPSResume CValue ContComp
+    | CPSFix Var [Var] ContComp ContComp
+    | CPSBinOp BinaryOp CValue CValue Var ContComp
+    | CPSValue CValue
+    | CPSLet Var CValue ContComp
+    | CPSSplit Label Var Var CValue ContComp
+    | CPSCase CValue Label Var ContComp Var ContComp
+    | CPSIf CValue ContComp ContComp
+    | CPSAbsurd CValue
+    \end{verbatim}
+    Most of the terms at the end have a coinductive reference to itself, which represents
+    the rest of the computation that needs to be done. For more clarification,
+    one may take a look into the book mentioned above \cite{appel-continuations}. The source code for curried
+    translation and evaluation can be found respectively in \verb!CPSAppel.hs! and
+    \verb!EvalCPS.hs!.
+
+    \section{Uncurried translation}
+
+    Having in mind the drawbacks mentioned above, alternative translation was
+    written, that coincides with the translation from the paper. Namely, with
+    the uncurried translation to target calculus with continuations represented
+    as explicit stacks. The target calculus was described in section \ref{sec:implementation-ast},
+    for which the evaluation is straightforward. The continuations are represented
+    as \verb!Cont!, with syntactic distinction between pure and effectful computations,
+    which occupy alternating positions in the stack. Explicit distinction gave
+    more control in the source code.
+
+    \begin{verbatim}
+type CPSMonad a = ExceptT Error (State Int) a
+
+type ContF = UValue -> [Cont] -> CPSMonad UComp
+
+data Cont = Pure ContF
+          | Eff ContF
+    \end{verbatim}
+    Where \verb!CPSMonad! is a monad transformer over \verb!Either! and \verb!State!.
+    \verb!State! was required to generate labels for fresh variables that
+    came from the translation. The core code is split into five functions:
+
+    \begin{verbatim}
+cps     :: Comp    -> [Cont] -> CPSMonad UComp
+cpsVal  :: Value   -> [Cont] -> CPSMonad UValue
+cpsHRet :: Handler -> Cont
+cpsHOps :: Handler -> Cont
+forward :: Label   -> UValue -> UValue -> [Cont] -> CPSMonad UComp
+    \end{verbatim}
+    Where the first two are implementing cps for computations and values.
+    \verb!cpsHRet! and \verb!cpsHOps! are yielding pure and effectful continuations,
+    based on a given handler. The last one is responsible for forwarding the
+    computation to the outer handler.
+
+    This results in an implementation that finally supports nested handlers,
+    as can be seen by evaluating \verb!programs/complexNestedHandlers.fk! program.
+    Unfortunately, following closely translation from the paper resulted in
+    a behaviour, in which invoked resumption forgets its pure continuation. This means,
+    that the following code, evaluating correctly on the Appel-based translation,
+    returns 0 rather than 1:
+
+    \begin{verbatim}
+handle do Drop () with { Drop p r -> let t <- r 0 in return 1}
+    \end{verbatim}
+    Nevertheless, working out this issue appears as less demanding than coping with
+    discrepancies created in the first translation. The source code for uncurried
+    translation and evaluation can be found respectively in \verb!CPSLists.hs! and
+    \verb!EvalTarget.hs!.
+
+    \section{Cohandlers}
+    \ldots
+
+    \section{Type System}
+    \ldots
+
+    \section{Source Code Structure}
+
+    The source code is divided into a number of modules, where the most
+    crucial parts have already been described.
+
+    \begin{verbatim}
+    AST.hs          - AST data structures
+    CommonCPS.hs    - Common functions for CPS translation
+    CommonEval.hs   - Common functions for evaluation
+    CPSAppel.hs     - Appel-based CPS translation
+    CPSLists.hs     - Uncurried CPS translation
+    EvalCPS.hs      - Evaluation of the Appel's CPS structure
+    EvalTarget.hs   - Evaluation of the target calculus
+    Freak.hs        - API for the language
+    Main.hs         - Main module running evaluator on given filename
+    Parser.hs       - Parser and lexer
+    TargetAST.hs    - AST for the target calculus
+    Tests.hs        - Tests module
+    Types.hs        - Common types definition
+    programs/       - Exemplary programs used in tests
+    \end{verbatim}
+
+\chapter{Conclusion}
+\section{Summary}
+    \ldots
+\section{Future work}
+
+    \subsection{Alternative evaluation}
+
+    The Links language also provides small-step operational semantics and
+    an abstract machine \cite{liberating-effects}. Implementing another way
+    of evaluation could serve as a way to empirically assert correctness,
+    as opposed to formally.
+
+    \subsection{Type inference}
+
+    The type system as of this day is not implemented, as the focus has been put
+    on CPS transformation. Further work is required here, especially considering
+    the fact that a huge advantage of algebraic effects is that they are explicitly
+    defined in the type of a computation.
+
+    \subsection{Multiple instances of algebraic effect}
+
+    The Freak language is limited to a single instance of an effect. We would
+    need to support cases where many instances of the algebraic effects, with
+    the same handler code, could be instantiated. The current state of the
+    art introduces a concept of resources and instances, as in Eff \cite{programming-in-eff},
+    or instance variables, as in Helium \cite{binders-labels}.
+
+    \subsection{Selective CPS}
+
+    Other languages, like Koka \cite{leijen-koka}, or even the core of the Links, are
+    performing selective CPS translation, which reduces the overhead on code
+    that does not perform algebraic effects. Our current translation is fully
+    embedded in the CPS.
+
+    \subsection{Exceptions as separate constructs}
+
+    Exceptions are a trivial example of algebraic effect where the resumption is
+    discarded, and as described in \S 4.5 \cite{handlers-cps}, they can be modeled
+    as a separate construct to improve performance.
+
+    \subsection{Shallow handlers}
+
+    Shallow and deep handlers while being able to simulate each other up to
+    administrative reductions, have a very different meaning from a theoretical
+    point of view. Implementing them as defined by Lindley et al. \cite{shallow-handlers} could
+    be another way of enhancing Freak.
+
 %%%%% BIBLIOGRAFIA

### 2020-05-06 thesis/thesis.tex
Minor extensions to table of contents
Short comments about chapters in thesis

@@
-% Opcje klasy 'iithesis' opisane sa w komentarzach w pliku klasy. Za ich pomoca
-% ustawia sie przede wszystkim jezyk i rodzaj (lic/inz/mgr) pracy, oraz czy na
-% drugiej stronie pracy ma byc skladany wzor oswiadczenia o autorskim wykonaniu.
 \documentclass[declaration,shortabstract]{iithesis}
@@
 %%%%% DANE DO STRONY TYTUŁOWEJ
-% Niezaleznie od jezyka pracy wybranego w opcjach klasy, tytul i streszczenie
-% pracy nalezy podac zarowno w jezyku polskim, jak i angielskim.
-% Pamietaj o madrym (zgodnym z logicznym rozbiorem zdania oraz estetyka) recznym
-% zlamaniu wierszy w temacie pracy, zwlaszcza tego w jezyku pracy. Uzyj do tego
-% polecenia \fmlinebreak.
 \polishtitle    {Efekty koalgebraiczne oraz ich kohandlery \fmlinebreak w językach programowania}
@@
 
+% TODO This requires a rework
 In this thesis we present experimental programming language Freak, which is an
@@
 \section{Problem Analysis}
+
+What are the issues with normal algebraic effects?
+
 \section{Problem Statement}
+
+Brief and concise description of the problem
+
 \section{Thesis Outline}
@@
 \section{Algebraic Effects}
-\section{Categorical Setting for Universal Algebra}
-\section{Comodels and Duality}
 
-\chapter{Proposed Solutions}
-\section{Dynamic Constraints Checking}
-\section{Linear Types}
-\section{Data-Flow Analysis}
-\section{Cohandlers as a Separate Construct}
+What are algebraic effects?
+
+\section{Categorical Setting of Universal Algebra}
+
+For the purpose of describing the duality between algebra and coalgebra,
+we allow ourselves to wander a bit deeper into category theory.
+
+    \subsection{Algebraic Theories}
+
+    Sigma-Signatures, Sigma-Terms, Sigma-Equations, Interpretations, Models, Free models
+
+    \subsection{Free models form monads}
+
+    Maybe not a subsection but just a lemma
+
+    \subsection{Models homomorphisms}
+
+    T-homomorphisms with respect to theory T
+
+    \subsection{Parametrized operations}
+
+\section{Duality}
+    \subsection{Comodels}
+
+    Comodel in $\mathcal{C}$ is just a Model in $ \mathcal{C}^{op} $.
+
+    Models -- Worlds
+
+    \subsection{Cooperations}
+
+    Derive from Models duality
+
+    \subsection{Coalgebraic Effects}
+
 \section{Related Work}
+    \subsection{Algebraic Effects}
 
@@
 
-\chapter{Coeffectful Programming with Cohandlers}
+    \subsection{Coalgebraic Effects}
+
+    Ahman and Bauer \cite{runners-in-action}
+
+\chapter{Potential Solutions}
+
+One-shot continuations
+
+\section{Dynamic Constraints Checking}
+
+\section{Linear Types}
+
+Solve the issue through introduction of linear type system.
+
+\section{Data-Flow Analysis}
+
+Another static analysis of the program.
+
+\section{Cohandlers as Separate Constructs}
+
+Simplify semantics by separating coalgebraic effects into a new, restricted
+construct in programming language
+
+\chapter{Effectful and Coeffectful Programming}
 \section{Examples}
@@
 
-    \subsection{Side effects}
+    \subsection{Taming Side effects}
 
@@
 \section{Coexamples}
-    \ldots
+
+    Examples for cohandlers
 
@@
 
+    % Outdated
     As of this day, two implementations are available, one based on the curried
@@
 
-    \section{Abstract Syntax Trees}
+    \section{Abstract Syntax Tree}
     \label{sec:implementation-ast}
@@
     \end{verbatim}
+
+    \section{Target calculus}
 
@@
 
-    \section{Type System}
-    \ldots
-
     \section{Source Code Structure}
@@
 \section{Summary}
-    \ldots
+
+    What was achieved, what was described, what needs to be done or researched.
+
 \section{Future work}
 
-    \subsection{Alternative evaluation}
+    \subsection{Abstract machine}
 
@@
 
-    \subsection{Type inference}
+    \subsection{Type inference and row polymorphism}
 
@@
 
-    \subsection{Multiple instances of algebraic effect}
+    \subsection{Multiple instances of algebraic effects}
 
@@
 
-%%%%% BIBLIOGRAFIA
-
-% \begin{thebibliography}{1}
-% \bibitem{example} \ldots
-% \end{thebibliography}
+%%%%% BIBLIOGRAPHY
 

### 2020-07-11 thesis/thesis.tex
[thesis] Apply linter fixes

@@
 %%%%% DANE DO STRONY TYTUŁOWEJ
-\polishtitle    {Efekty koalgebraiczne oraz ich kohandlery \fmlinebreak w językach programowania}
-\englishtitle   {Coalgebraic effects and their cohandlers \fmlinebreak in programming languages}
-\polishabstract {
+\polishtitle{Efekty koalgebraiczne oraz ich kohandlery \fmlinebreak{} w językach programowania}
+\englishtitle{Coalgebraic effects and their cohandlers \fmlinebreak{} in programming languages}
+\polishabstract{
     \ldots
@@
 }
-\author         {Mateusz Urbańczyk}
-\advisor        {dr Maciej Piróg}
-\date           {30 czerwca 2020}                     % Data zlozenia pracy
+\author{Mateusz Urbańczyk}
+\advisor{dr Maciej Piróg}
+\date{1 września 2020}                     % Data zlozenia pracy
 % Dane do oswiadczenia o autorskim wykonaniu
@@
 In this thesis we present experimental programming language Freak, which is an
-implementation of Continuation Passing Style for Effect Handlers paper \cite{handlers-cps},
+implementation of Continuation Passing Style for Effect Handlers paper~\cite{handlers-cps},
 with the addition of a few basic constructs. We start by presenting the related work,
@@
 
-    Models -- Worlds
+    Models --- Worlds
 
@@
 
-    Except from Links language \cite{handlers-cps}, on which the implementation
+    Except from Links language~\cite{handlers-cps}, on which the implementation
     is based, there are currently many other alternatives available. One may take
-    a look at Frank \cite{frank}, which provides a support for multihandlers,
-    Koka \cite{leijen-koka}, Helium \cite{helium} or Eff \cite{eff}. Except from
+    a look at Frank~\cite{frank}, which provides a support for multihandlers,
+    Koka~\cite{leijen-koka}, Helium~\cite{helium} or Eff~\cite{eff}. Except from
     separate languages, many libraries arose for existing ones like Haskell,
@@
 
-    As can be seen in the J. Yallop repository \cite{effects-bibliography},
+    As can be seen in the J. Yallop repository~\cite{effects-bibliography},
     algebraic effects and handlers are now trending branch in the programming languages theory.
@@
 
-    Ahman and Bauer \cite{runners-in-action}
+    Ahman and Bauer~\cite{runners-in-action}
 
@@
 \chapter{Effectful and Coeffectful Programming}
-\section{Examples}
-\label{sec:examples}
+\section{Examples}\label{sec:examples}
+
     In this section we present a few examples to show the capabilities of the
-    language. The ideas have been based on \cite{programming-in-eff}, and thus will not be
+    language. The ideas have been based on~\cite{programming-in-eff}, and thus will not be
     described in great details. More exemplary programs in Freak language can
@@
 
-    \subsection{Choice}
-    \label{sec:choice-example}
+    \subsection{Choice}\label{sec:choice-example}
 
@@
 
-    These implications are straightforward when looking from a categorical
-    standpoint, where effects are viewed as free models of algebraic theories
-    \cite{adequacy}, and handlers are homomorphisms preserving the model structure
-    \cite{handlers}. Nevertheless, the results are very exciting for programming
-    use cases. The current Freak implementation does not support I/O.
+    These implications are straightforward when looking from a categorical standpoint,
+    where effects are viewed as free models of algebraic theories~\cite{adequacy},
+    and handlers are homomorphisms preserving the model structure~\cite{handlers}.
+    Nevertheless, the results are very exciting for programming use cases. The current
+    Freak implementation does not support I/O.
 
@@
     As of this day, two implementations are available, one based on the curried
-    translation and Appel \cite{appel-continuations}, and the second one based
+    translation and Appel~\cite{appel-continuations}, and the second one based
     directly on the uncurried translation with continuations as explicit stacks
-    from paper. More details can be found in section \ref{sec:implementation}.
-    All commands are available within \verb src  directory.
+    from paper. More details can be found in Section~\ref{sec:implementation}.
+    All commands are available within \verb!src!  directory.
 
@@
         \item Compile: \verb!make build!
-        \item Link to PATH: \verb!sudo make link!
+        \item Link to PATH:~\verb!sudo make link!
         \item Remove artifacts: \verb!make clean!
@@
     After compiling and linking program to PATH, one may evaluate program as
-    follows: \verb!freak programs/choicesList.fk!. The actual code is described in section \ref{sec:choice-example}
+    follows: \verb!freak programs/choicesList.fk!. The actual code is described in Section~\ref{sec:choice-example}
 
@@
     they include both inline and file-based tests. For more details about
-    writing tests, one may refer to \textit{HUnit documentation} \cite{hunit-docs}.
+    writing tests, one may refer to \textit{HUnit documentation}~\cite{hunit-docs}.
 
@@
     The other constructs are just as in Links, with slight syntax modifications.
-    Actual programs in Freak can be found in section \ref{sec:examples}.
+    Actual programs in Freak can be found in Section~\ref{sec:examples}.
 
@@
 
-        <Values V, W> ::= $ x $ | $nat \; n$ \\
+        <Values V, W> $::=$ $ x $ | $nat \; n$ \\
             | $ \backslash x : A \rightarrow M $ | \textbf{rec} $ g \; x \rightarrow M $\\
             | $V \oplus W$ | $V \approx W$ \\
-            | <> | $ \{ \ell = V; W\} $  | $[ \ell \; V]^{R}$
+            | <> | $ \{ \ell = V; W\} $  | ${[ \ell \; V]}^{R}$
 
-        <Computations M, N> ::= $ V $ $ W $ \\
+        <Computations M, N> $::=$ $ V $ $ W $ \\
             | \textbf{if} $V$ \textbf{then} $M$ \textbf{else} $N$ \\
@@
 
-        <Handlers H> ::= \textbf{return} $ x \rightarrow M $ | $ \ell \; p \; r \rightarrow M, H $
+        <Handlers H> $::=$ \textbf{return} $ x \rightarrow M $ | $ \ell \; p \; r \rightarrow M, H $
 
-        <Binary operators $\oplus$> ::= + | - | * | /
+        <Binary operators $\oplus$> $::=$ + | $-$ | * | /
 
-        <Relational operators $\approx$> ::= $ \textless $ | $\leqslant$ | $>$ | $\geqslant$ | == | !=
+        <Relational operators $\approx$> $::=$ $ \textless $ | $\leqslant$ | $>$ | $\geqslant$ | == | $!= $
 
@@
     extensively by providing small-step operational semantics,
-    continuation passing style transformation \cite{handlers-cps} as well
-    as abstract machine \cite{liberating-effects}, which was proved to coincide
+    continuation passing style transformation~\cite{handlers-cps} as well
+    as abstract machine~\cite{liberating-effects}, which was proved to coincide
     with CPS translation. That being said, Freak introduces new basic
@@
     \begin{flushleft}
-    $\mathcal{E}$ ::= $ \mathcal{E} \oplus W \; | \; nat \; n \oplus \mathcal{E} \; |$ \textbf{if} $\mathcal{E}$ \textbf{then} $M$ \textbf{else} $N$
+    $\mathcal{E} ::= \mathcal{E} \oplus W \; | \; nat \; n \oplus \mathcal{E} \; |$ \textbf{if} $\mathcal{E}$ \textbf{then} $M$ \textbf{else} $N$
     \end{flushleft}
@@
 
-\chapter{Implementation}
-\label{sec:implementation}
+\chapter{Implementation}\label{sec:implementation}
 
@@
     being said, two inherently different takes at the implementations were made.
-    The first one is based on curried translation and A. Appel \cite{appel-continuations}
+    The first one is based on curried translation and A. Appel~\cite{appel-continuations}
     book, and the second one directly on the uncurried translation to target
@@
 
-    \section{Abstract Syntax Tree}
-    \label{sec:implementation-ast}
+    \section{Abstract Syntax Tree}\label{sec:implementation-ast}
 
@@
 
-    The first take was heavily inspired by A. Appel's Compiling with Continuations
-    \cite{appel-continuations}, which provides a translation for a simplified ML
-    calculus. The calculus was extended and translation adapted to handle algebraic
-    effects and their handlers. The translation is based on the curried first-order
-    translation. That being said, the source code diverged a lot from the paper on
-    which it was based, leading to a different transformation for which
-    the correctness and cohesion with operational semantics should be proved
-    separately. Indeed, while the interpreter worked well on the use cases
-    defined in tests, the evaluation had a part which was not tail-recursive. What's
-    more, nested handlers were not supported, and the implementation was found to be
-    trickier than it should, as it was not obvious on how to adopt the technique
-    proposed in the paper.
+    The first take was heavily inspired by A. Appel's Compiling with
+    Continuations~\cite{appel-continuations}, which provides a translation for
+    a simplified ML calculus. The calculus was extended and translation adapted to
+    handle algebraic effects and their handlers. The translation is based on the curried
+    first-order translation. That being said, the source code diverged a lot from the
+    paper on which it was based, leading to a different transformation for which the
+    correctness and cohesion with operational semantics should be proved separately.
+    Indeed, while the interpreter worked well on the use cases defined in tests, the
+    evaluation had a part which was not tail-recursive. What's more, nested handlers
+    were not supported, and the implementation was found to be trickier than it should,
+    as it was not obvious on how to adopt the technique proposed in the paper.
 
@@
     the rest of the computation that needs to be done. For more clarification,
-    one may take a look into the book mentioned above \cite{appel-continuations}. The source code for curried
+    one may take a look into the book mentioned above~\cite{appel-continuations}. The source code for curried
     translation and evaluation can be found respectively in \verb!CPSAppel.hs! and
@@
     the uncurried translation to target calculus with continuations represented
-    as explicit stacks. The target calculus was described in section \ref{sec:implementation-ast},
+    as explicit stacks. The target calculus was described in Section~\ref{sec:implementation-ast},
     for which the evaluation is straightforward. The continuations are represented
@@
     The Links language also provides small-step operational semantics and
-    an abstract machine \cite{liberating-effects}. Implementing another way
+    an abstract machine~\cite{liberating-effects}. Implementing another way
     of evaluation could serve as a way to empirically assert correctness,
@@
     the same handler code, could be instantiated. The current state of the
-    art introduces a concept of resources and instances, as in Eff \cite{programming-in-eff},
-    or instance variables, as in Helium \cite{binders-labels}.
+    art introduces a concept of resources and instances, as in Eff~\cite{programming-in-eff},
+    or instance variables, as in Helium~\cite{binders-labels}.
 
@@
 
-    Other languages, like Koka \cite{leijen-koka}, or even the core of the Links, are
+    Other languages, like Koka~\cite{leijen-koka}, or even the core of the Links, are
     performing selective CPS translation, which reduces the overhead on code
     that does not perform algebraic effects. Our current translation is fully
-    embedded in the CPS.
+    embedded in the CPS.\@
 
@@
     Exceptions are a trivial example of algebraic effect where the resumption is
-    discarded, and as described in \S 4.5 \cite{handlers-cps}, they can be modeled
+    discarded, and as described in \S 4.5~\cite{handlers-cps}, they can be modeled
     as a separate construct to improve performance.
@@
     administrative reductions, have a very different meaning from a theoretical
-    point of view. Implementing them as defined by Lindley et al. \cite{shallow-handlers} could
+    point of view. Implementing them as defined by Lindley et al.~\cite{shallow-handlers} could
     be another way of enhancing Freak.
@@
 
-\printbibliography
+\printbibliography{}
 

### 2020-07-12 thesis/thesis.tex
[thesis] Break some lines
[thesis] My insights about computational effects

@@
 \usepackage{mathalfa}
+\usepackage{textcomp}
 
@@
 %
-%\theoremstyle{definition} \newtheorem{definition}{Definition}[chapter]
-%\theoremstyle{remark} \newtheorem{remark}[definition]{Observation}
-%\theoremstyle{plain} \newtheorem{theorem}[definition]{Theorem}
-%\theoremstyle{plain} \newtheorem{lemma}[definition]{Lemma}
+\theoremstyle{definition} \newtheorem{definition}{Definition}[chapter]
+\theoremstyle{remark} \newtheorem{remark}[definition]{Observation}
+\theoremstyle{plain} \newtheorem{theorem}[definition]{Theorem}
+\theoremstyle{plain} \newtheorem{lemma}[definition]{Lemma}
 %\renewcommand \qedsymbol {\ensuremath{\square}}
@@
 
-%%%%% POCZĄTEK ZASADNICZEGO TEKSTU PRACY
-
 \chapter{Introduction}
 
+\textit{
+    My algebraic methods are really methods of working and thinking; this is why
+    they have crept in everywhere anonymously. $\sim{}$Emmy Noether
+}
+
 % TODO This requires a rework
+
 In this thesis we present experimental programming language Freak, which is an
@@
 with the addition of a few basic constructs. We start by presenting the related work,
-then discuss syntax and operational semantics. Basic usage guide for playing with
-the language is provided, as well as implementation details and examples. We conclude
-by stating what are the possible augmentations, that are intended to be made in the future.
+then discuss syntax and operational semantics. Basic usage guide for playing with the
+language is provided, as well as implementation details and examples. We conclude by
+stating what are the possible augmentations, that are intended to be made in the future.
 
@@
 \chapter{Background}
+
+\section{Computational Effects}
+
+Since the rapid development of computational theory in 1930s by A. Turing, K. Godel and
+A. Church, we have a well-established notion of what can and what cannot be done
+through algorithmic means, which we can almost directly translate to being computable
+by our machines. Through next years we have developed mainstream languages that are used
+almost everywhere, with a great success.
+
+Under these circumstances one may pose a question, why do we still bother with
+development of languages theory, since so much has been done already. Is there
+anything that drives us towards further research? Indeed, one active branch
+revolves around equational theory to asses equality of two programs, which we
+know that in general setting is undecidable. Proof methods may include extensional,
+contextual or logical equality. However, there is no doubt that these formal
+ways of reasoning about programs, while being crucial for assessing correctness,
+do not bring direct benefits for practical, everyday use cases.
+
+% Not sure I like this part. Isn't it too broad to say handle complexity?
+Other branch of languages theory, that we shall investigate more in this thesis,
+is about handling complexity of programs. Various methods of static analysis has
+been developed, most notably, type systems.
+
+What is the source of complexity? Thanks to strong and static type systems along
+with their implementations, we have solid tools to work efficiently on functions
+that are pure. That being said, we claim that the sole complexity of programs
+comes from side effects, which we cannot avoid in writing anything useful.
+
+We need to have a good way for handling computational effects, which so far has
+been modeled through monads~\cite{moggi}.
+
+However, they were found to be, to say the least, cumbersome to work with when
+complexity and number of different effects increases. It is perhaps not a coincidence
+that many functional programming languages do not have them, because of the
+non-composable nature of them.
+
+Having functions $ f: a \rightarrow m b $ and $ g : b \rightarrow m' c $, it
+should not be an issue to compose them together.
+
+Recall that the main idea of monads lies on the fact that we provide a function
+to lift value into a functor and associative kleisli composition operator to combine
+computations together. It turns out, that it becomes complex and unpleasent to
+combine two functors together to form a new monad. For this purpose, monad
+transformers arose in Haskell.
+
+Most of the common languages avoid this by not expressing computational effects
+in the type system, and instead one may think about functions as being implicitly
+embed in a Kleisli Category over a functor T, where T is a hidden signature over
+all possible side effects that occur in our program.
+
+Not only we would like to bring back effects to our type system, but also do it
+in a way that is composable. This is where algebraic effects comes to the rescue.
+
+From theoretical point of view, we need to develop equa about our effects
+to assert correctness of our langauge as well as to have the right hammer
+to reason about our programs.
+
+That is the point where we would like to introduce to unfamiliar read a notion
+of algebraic effects.
+
 \section{Algebraic Effects}
@@
 What are algebraic effects?
+
+% Algebraic effects as Interfaces for side effects
+
+Algebraic effects can be thought as an public interface for computational effects.
+Declarative approach allow us to write programs in which the actual semantic of
+source code is dependent on handler that defines the meaning of a subset of effects.
+
+This is really a breakthrough from practical point of view, as we may substitute
+logic depending on execution environment. As an example, fetching for resources
+can behave differently as we run tests, debug our code, or run it on production.
+In the same manner, they neatly allow us to abstract over implementation details.
 
@@
 
+    \subsection{Coinductive Reasoning}
+
+    Coinduction, and coinductive structures, provide us a way of observation
+    of the behaviour, as a contractry to construction in inductive reasoning.
+    Simple example of the duality can be expressed through induction over finite
+    list and coinductive observation of infinite streams.
+
+    Doing the latter may involve modification of the internal state of the machine
+    that is generating the infinite streams, or in more concrete scenario,
+    alternation of the external resource that is providing us the data.
+
+    This is one of the cases where interaction with external resource multiple
+    times, or more specifically in case of algebraic effects, invoking resumption more
+    than once, may lead to unexpected behaviour that would not be expected in standard
+    control flow. This captures the excessive generality of effects, and is the
+    issue that we would like to address with coalgebraic effects.
+
+    % TODO: On it's own this snippet a bit dry and of course requires additional examples,
+    % for instance the classic one with operating on closed file
+
 \section{Related Work}
@@
 
-    As can be seen in the J. Yallop repository~\cite{effects-bibliography},
-    algebraic effects and handlers are now trending branch in the programming languages theory.
+    As can be seen in the J. Yallop repository~\cite{effects-bibliography}, algebraic
+    effects and handlers are now trending branch in the programming languages theory.
 
@@
 One-shot continuations
+
+Detecting whether resumption is called only once is undecidable.
 
@@
 
-    In this section we present a few examples to show the capabilities of the
-    language. The ideas have been based on~\cite{programming-in-eff}, and thus will not be
+    In this section we present a few examples to show the capabilities of the language.
+    The ideas have been based on~\cite{programming-in-eff}, and thus will not be
     described in great details. More exemplary programs in Freak language can
@@
     collects all possible results, which can be achieved by putting
-    \verb!return (t, f)! in the \verb!<PLACEHOLDER>!, which evaluates to \verb!((10, 5), (20, 15))!.
+    \verb!return (t, f)! in the \verb!<PLACEHOLDER>!, which evaluates to
+    \verb!((10, 5), (20, 15))!.
 
@@
     The syntax for the calculus is shown below. $nat \; n$ represents an integer $n$,
-    $V \oplus W$ and $V \approx W$ are respectively binary and relational
-    operators, where we support basic arithmetic and comparison operations.
-    \textbf{if} $V$ \textbf{then} $M$ \textbf{else} $N$ is a standard branching statement.
-    The other constructs are just as in Links, with slight syntax modifications.
-    Actual programs in Freak can be found in Section~\ref{sec:examples}.
+    $V \oplus W$ and $V \approx W$ are respectively binary and relational operators,
+    where we support basic arithmetic and comparison operations.
+    \textbf{if} $V$ \textbf{then} $M$ \textbf{else} $N$ is a standard branching
+    statement. The other constructs are just as in Links, with slight syntax
+    modifications. Actual programs in Freak can be found in Section~\ref{sec:examples}.
 
@@
     \end{verbatim}
-    Most of the terms at the end have a coinductive reference to itself, which represents
-    the rest of the computation that needs to be done. For more clarification,
-    one may take a look into the book mentioned above~\cite{appel-continuations}. The source code for curried
-    translation and evaluation can be found respectively in \verb!CPSAppel.hs! and
+    Most of the terms at the end have a coinductive reference to itself, which
+    represents the rest of the computation that needs to be done. For more
+    clarification, one may take a look into the book mentioned
+    above~\cite{appel-continuations}. The source code for curried translation and
+    evaluation can be found respectively in \verb!CPSAppel.hs! and
     \verb!EvalCPS.hs!.
@@
     Shallow and deep handlers while being able to simulate each other up to
-    administrative reductions, have a very different meaning from a theoretical
-    point of view. Implementing them as defined by Lindley et al.~\cite{shallow-handlers} could
-    be another way of enhancing Freak.
+    administrative reductions, have a very different meaning from a theoretical point
+    of view. Implementing them as defined by Lindley et al.~\cite{shallow-handlers}
+    could be another way of enhancing Freak.
 

### 2020-07-12 thesis/thesis.tex
[thesis] Algebraic Theories

@@
 \usepackage{textcomp}
+\usepackage{stmaryrd}
 
@@
 %\renewcommand \qedsymbol {\ensuremath{\square}}
+
+\newcommand{\mathVar}[1]{{\operatorname{\mathit{#1}}}}
+
 % ...
@@
 
-    Sigma-Signatures, Sigma-Terms, Sigma-Equations, Interpretations, Models, Free models
+    \begin{definition}{\mathVar{Signature}}
+
+    A \textit{signature $ \Sigma $} is given by a collection of operation symbols $ op_{i} $
+    with associated parameters $ P_{i} $ and arities $ A_{i} $, where
+    $ P_{i} $ and $ A_{i} $ are objects in the category of our interest.
+    We will write an operation as $ op_{i} : P_{i} \rightsquigarrow A_{i} $
+
+    \end{definition}
+
+    \begin{definition}{$\mathVar{\Sigma-terms}$}
+
+    Collection of $\mathVar{\Sigma-terms}$ is a free algebra with a generator $X$
+    for a functor $ \mu H_{\Sigma} $ that maps objects into trees over a given
+    signature $ \Sigma $ and morphisms into folds over trees.
+
+    \end{definition}
+
+    \begin{definition}{$ \mathVar{\Sigma-Equation} $}
+
+        A $ \mathVar{\Sigma-Equation} $ is an object X and a pair of
+        $\mathVar{\Sigma-terms}$ $l, r \in Tree_{\Sigma}(X)$, written as
+
+        $ X $ $|$ $l = r$
+
+    \end{definition}
+
+    \begin{definition}{Algebraic Theory}
+
+    An \textit{algebraic theory} $T = (\Sigma_{T}, \mathcal{E}_{T})$, is given by
+    a signature $\Sigma_{T}$ and a collection $\mathcal{E}_{T}$ of
+    $\mathVar{\Sigma_{T}-equations}$. For clarity we will usually omit T subscript
+
+    \end{definition}
+
+    \begin{definition}{Interpretation}
+
+    An \textit{interpretation $I$ over a given signature $\Sigma$} is given by
+    a carrier object $|I|$ and for each $ op_{i} : P_{i} \rightsquigarrow A_{i} $
+    in $\Sigma$ a map
+
+    $ {\llbracket op_{i} \rrbracket}_I : P_{i} \times{} {|I|}^{A_{i}} \rightarrow |I| $
+
+    Interpretation may be naturally extended to $\mathVar{\Sigma-terms}$, such
+    that a given $\mathVar{\Sigma-term}$ $X$ $|$ $t$ is interpreted by a map
+
+    $ {\llbracket X | t \rrbracket}_I : {|I|}^{k} \rightarrow |I| $
+
+    TODO actually define the map as a projection on variables and
+    map composition on compound terms.
+
+    where ${|I|}^{k}$ is an k-ary product in our category, assuming that all of the
+    necessary products exist.
+
+    \end{definition}
+
+    TODO Models, Free models
 

### 2020-07-28 thesis/thesis.tex
[thesis] More on universal algebra

@@
 \usepackage{stmaryrd}
+\usepackage{tikz-cd}
+% Not sure if necessary
+\usepackage[utf8]{inputenc}
+\usepackage{rotchiffre}
 
@@
 
-For the purpose of describing the duality between algebra and coalgebra,
-we allow ourselves to wander a bit deeper into category theory.
+Algebraic effects can be describe via operational means, however,
+for the purpose of presenting the duality between algebra and coalgebra,
+we allow ourselves to wander a bit deeper into category theory and describe
+effects from denotational point of view.
 
@@
 
-    \begin{definition}{\mathVar{Signature}}
+    \begin{definition}
 
@@
 
-    \begin{definition}{$\mathVar{\Sigma-terms}$}
+    \begin{definition}
 
@@
 
-    \begin{definition}{$ \mathVar{\Sigma-Equation} $}
+    \begin{definition}
 
@@
 
-        $ X $ $|$ $l = r$
+        \begin{center}
+        $ X \mid l = r $
+        \end{center}
 
@@
 
-    \begin{definition}{Algebraic Theory}
+    \begin{definition}
 
@@
 
-    \begin{definition}{Interpretation}
+    \begin{definition}
 
@@
 
+    \begin{center}
         $ {\llbracket op_{i} \rrbracket}_I : P_{i} \times{} {|I|}^{A_{i}} \rightarrow |I| $
-
+    \end{center}
     Interpretation may be naturally extended to $\mathVar{\Sigma-terms}$, such
-    that a given $\mathVar{\Sigma-term}$ $X$ $|$ $t$ is interpreted by a map
-
-    $ {\llbracket X | t \rrbracket}_I : {|I|}^{k} \rightarrow |I| $
-
-    TODO actually define the map as a projection on variables and
-    map composition on compound terms.
-
-    where ${|I|}^{k}$ is an k-ary product in our category, assuming that all of the
-    necessary products exist.
+    that a given $\mathVar{\Sigma-term}$ $X \mid t$ is interpreted by a map
+    which sends variables into projections from environment and terms into
+    map composition over each subterm.
 
@@
 
-    TODO Models, Free models
+    \begin{definition}
 
-    \subsection{Free models form monads}
+    A \textit{model M} of an algebraic theory T is an interpretation of the
+    signature ${\Sigma_{T}}$ which validates all the equations $\mathcal{E_{T}}$.
+    That is, for every equation $X \mid l = r$ the following diagram commutes:
 
-    Maybe not a subsection but just a lemma
+    \begin{center}
+    \begin{tikzcd}[row sep=huge, column sep=large]
+        {|M|}^k \arrow[rr, '\phi', bend right] \arrow[rr, '\rrbracket{} X|l', bend left] &  & {|M|}
+    \end{tikzcd}
+    \end{center}
+    % TODO: this diagram doesn't work
 
-    \subsection{Models homomorphisms}
+    \begin{center}
+    \begin{tikzcd}
+    A \arrow[rd] \arrow[r, '\phi'] & B \\
+    & C
+    \end{tikzcd}
+    \end{center}
 
-    T-homomorphisms with respect to theory T
+    \end{definition}
 
-    \subsection{Parametrized operations}
+    \begin{definition}
+
+    Free F-algebra on an object A (of generators) in $\mathcal{C}$ is meant an algebra
+
+    \begin{center}
+
+    $ \varphi_{A} : F A^{\#} \longrightarrow A $
+
+    \end{center}
+    \\
+    together with an universal arrow $ \eta_{A} : A \longrightarrow A^{\#} $. Universality means that for every algebra
+    $ \beta : F B \longrightarrow B $ and every morphism $ f : A \longrightarrow B $ in $ \mathcal{C}$, there exists a unique homomorphism
+    $ \overline{f} : A^{\#} \longrightarrow B $ extending f, i.e.\ a unique morphism of $ \mathcal{C}$ for which the diagram below
+    commutes:
+
+    % TODO:
+    \begin{center}
+    \begin{tikzcd}[row sep=huge, column sep=large]
+    % F A^{\#} \arrow[d, "F \overline{f}"] \arrow[r, "\varphi_{A}"] & A^{\#} \arrow[d, "\overline{f}"] & A \arrow[l, "\eta_{A}"'] \arrow[ld, "f"] \\
+    % F B \arrow[r, "\beta"']                                       & B                                &
+    \end{tikzcd}
+    \end{center}
+
+    \end{definition}
+
+    \begin{definition} \textit{Free model} is just a model which is free algebra.
+
+    \end{definition}
+
+    \begin{lemma}{Free models form monads}
+
+    \end{lemma}
+
+    \begin{definition}
+
+    Let L, M be models of a theory T. A $ \mathVar{T-homomorphism}$
+    $\phi : L \rightarrow M$ is a map such that the following diagram commutes:
+
+    % TODO
+
+    \end{definition}
 

### 2020-08-02 thesis/thesis.tex
[thesis] Many small improvements

@@
 
-%%%%% DANE DO STRONY TYTUŁOWEJ
 \polishtitle{Efekty koalgebraiczne oraz ich kohandlery \fmlinebreak{} w językach programowania}
@@
 
-% ...
-%%%%%
-
 \begin{document}
@@
 
-% TODO This requires a rework
+TODO This requires a rework
 
@@
 implementation of Continuation Passing Style for Effect Handlers paper~\cite{handlers-cps},
-with the addition of a few basic constructs. We start by presenting the related work,
+with the addition of a few basic constructs.
+
+\section{Problem Analysis}
+
+TODO add citations
+
+Algebraic effects, while not being a new concept, in the recent years
+have received a lot of attention, both from theoretical and practical side.
+
+TODO refactor the following
+
+They give developer necessary tools for declarative programming, writing
+programs in a way to strive from exposing implementation details, while still
+preserving a lot of flexibility and having strong theoretical background and
+ways of formal reasoning.
+
+In fact, that very flexibility can be troublesome. Allowing developer for too
+much freedome may lead to undesired behaviour and incorrect programs, for that
+very reason, we develop type systems, to detect errors before they occur.
+
+\section{Problem Statement}
+
+TODO Think about merging Problem Analysis with Statement \\
+
+Multi-shot resumptions in effect handlers are a powerful tool to describe
+fairly complex logic in a concise way. That being said, they give rise to
+issues that would not occur in standard control flow, especially when composing
+various effects together.
+
+We will explain this in greater details and think about possible ways to address
+the issue of flexibility of effects in the next sections.
+
+TODO actually point out which sections you talk about
+
+\section{Thesis Outline}
+
+TODO Guide reader through thesis, probably later than sooner \\
+TODO Clearly divide categorical from operational approach for less algebraically fluent readers
+
+TODO it's outdated
+We start by presenting the related work,
 then discuss syntax and operational semantics. Basic usage guide for playing with the
@@
 stating what are the possible augmentations, that are intended to be made in the future.
-
-\section{Problem Analysis}
-
-What are the issues with normal algebraic effects?
-
-\section{Problem Statement}
-
-Brief and concise description of the problem
-
-\section{Thesis Outline}
 
@@
 
-Since the rapid development of computational theory in 1930s by A. Turing, K. Godel and
-A. Church, we have a well-established notion of what can and what cannot be done
-through algorithmic means, which we can almost directly translate to being computable
-by our machines. Through next years we have developed mainstream languages that are used
-almost everywhere, with a great success.
+Since the rapid development of computational theory in 1930s by A. Turing,
+K. Godel and A. Church, we have a well-established notion of what can and what
+cannot be done through algorithmic means, which we can almost directly translate
+to being computable by our machines. Through next years we have developed
+mainstream languages that are used almost everywhere, with a great success.
 
@@
 ways of reasoning about programs, while being crucial for assessing correctness,
-do not bring direct benefits for practical, everyday use cases.
+do not bring direct benefits for everyday use cases, as they are rarely accessible
+by a common developer.
 
-% Not sure I like this part. Isn't it too broad to say handle complexity?
 Other branch of languages theory, that we shall investigate more in this thesis,
-is about handling complexity of programs. Various methods of static analysis has
-been developed, most notably, type systems.
+is about taming complexity of programs. Various methods of static analysis has
+been developed for various use cases, most notably, type systems.
 
@@
 with their implementations, we have solid tools to work efficiently on functions
-that are pure. That being said, we claim that the sole complexity of programs
+that are pure. That being said, we claim that the core complexity of programs
 comes from side effects, which we cannot avoid in writing anything useful.
 
-We need to have a good way for handling computational effects, which so far has
-been modeled through monads~\cite{moggi}.
+We need to have a good way for handling computational effects. One of the ways
+to model them, can be done through monads~\cite{moggi}. However, they were
+found to be, to say the least, cumbersome to work with when the number of
+different effects increases. It is perhaps not a coincidence that many
+functional programming languages do not have them, because of the non-composable
+nature of them, or at least not composable in their implementations.
 
-However, they were found to be, to say the least, cumbersome to work with when
-complexity and number of different effects increases. It is perhaps not a coincidence
-that many functional programming languages do not have them, because of the
-non-composable nature of them.
+Let's put the following functions
 
-Having functions $ f: a \rightarrow m b $ and $ g : b \rightarrow m' c $, it
-should not be an issue to compose them together.
+\begin{center}
 
-Recall that the main idea of monads lies on the fact that we provide a function
-to lift value into a functor and associative kleisli composition operator to combine
-computations together. It turns out, that it becomes complex and unpleasent to
-combine two functors together to form a new monad. For this purpose, monad
-transformers arose in Haskell.
+    $ f: a \rightarrow b $, $ g : b \rightarrow c $ \\
+    $ f': a \rightarrow m \; b $, $ g' : b \rightarrow m \; c $ \\
+    $ f'': a \rightarrow m' \; m \; b $, $ g'' : b \rightarrow m' \; m \; c $ \\
 
+\end{center}
+
+where $m$, $m'$ are monads. Functions $f$, $g$ can be composed using standard
+composition, for $f'$, $g'$ we can use Kleisli composition operator from monad $m$.
+What in case of adding one more effect?
+
+It turns out, that it becomes complex and unpleasent to combine two functors
+together to form a new monad, for which purpose, monad transformers arose in Haskell.
 Most of the common languages avoid this by not expressing computational effects
@@
 
+TODO Think about describing what is Kleisli Category
+
 Not only we would like to bring back effects to our type system, but also do it
 in a way that is composable. This is where algebraic effects comes to the rescue.
-
-From theoretical point of view, we need to develop equa about our effects
-to assert correctness of our langauge as well as to have the right hammer
-to reason about our programs.
-
-That is the point where we would like to introduce to unfamiliar read a notion
-of algebraic effects.
+From theoretical point of view, we need to develop equational theory about our
+effects to assert correctness of our langauge as well as to have the right hammer
+to reason about our programs. That is the point where we would like to introduce
+to unfamiliar reader a notion of algebraic effects.
 
 \section{Algebraic Effects}
-
-What are algebraic effects?
-
-% Algebraic effects as Interfaces for side effects
 
@@
 
+Patterns like these are well known in programming, for which other alternatives
+arose. One can mention interfaces from object-oriented programming, which are also
+a way to describe the API of a certain component. In order to abstract from
+implementation details, we pass an object around which represents a certain
+interface. While this is certainly better than having no abstractions at all,
+we need to pass this interface into every function that is going to use it.
+This practice is called dependency injection, and while it sounds like it solves
+some of the issues, we end up in functions that need to carry representants
+of the interfaces, and pass them in every subcall that they make. It also can't
+be ommited that it's not really a substitute, as it's solving only one particular
+issue that algebraic effects address.
+
+TODO add link to paper with plenty of examples
+
+- Exceptions
+- Async/await
+- Flipping coin
+- Backtracking?
+- Operating on file
+
 \section{Categorical Setting of Universal Algebra}
 
-Algebraic effects can be describe via operational means, however,
+Algebraic effects can be described via operational means, however,
 for the purpose of presenting the duality between algebra and coalgebra,
@@
 
-    A \textit{signature $ \Sigma $} is given by a collection of operation symbols $ op_{i} $
-    with associated parameters $ P_{i} $ and arities $ A_{i} $, where
-    $ P_{i} $ and $ A_{i} $ are objects in the category of our interest.
+    A \textit{signature $ \Sigma $} is given by a collection of operation
+    symbols $ op_{i} $ with associated parameters $ P_{i} $ and arities $ A_{i} $,
+    where $ P_{i} $ and $ A_{i} $ are objects in the category of our interest.
     We will write an operation as $ op_{i} : P_{i} \rightsquigarrow A_{i} $
@@
     \begin{definition}
-
     Collection of $\mathVar{\Sigma-terms}$ is a free algebra with a generator $X$
@@
 
-    An \textit{algebraic theory} $T = (\Sigma_{T}, \mathcal{E}_{T})$, is given by
-    a signature $\Sigma_{T}$ and a collection $\mathcal{E}_{T}$ of
-    $\mathVar{\Sigma_{T}-equations}$. For clarity we will usually omit T subscript
+    An \textit{algebraic theory} $T = (\Sigma_{T}, \mathcal{E}_{T})$, is given
+    by a signature $\Sigma_{T}$ and a collection $\mathcal{E}_{T}$ of
+    $\mathVar{\Sigma_{T}-equations}$. For clarity, we will usually omit T subscript.
 
@@
     \end{center}
-    % TODO: this diagram doesn't work
+
+    TODO this diagram doesn't work
 
@@
     \\
-    together with an universal arrow $ \eta_{A} : A \longrightarrow A^{\#} $. Universality means that for every algebra
-    $ \beta : F B \longrightarrow B $ and every morphism $ f : A \longrightarrow B $ in $ \mathcal{C}$, there exists a unique homomorphism
-    $ \overline{f} : A^{\#} \longrightarrow B $ extending f, i.e.\ a unique morphism of $ \mathcal{C}$ for which the diagram below
-    commutes:
+    together with an universal arrow $ \eta_{A} : A \longrightarrow A^{\#} $.
+    Universality means that for every algebra $ \beta : F B \longrightarrow B $
+    and every morphism $ f : A \longrightarrow B $ in $ \mathcal{C}$, there
+    exists a unique homomorphism $ \overline{f} : A^{\#} \longrightarrow B $
+    extending f, i.e.\ a unique morphism of $ \mathcal{C}$ for which the diagram
+    below commutes:
 
-    % TODO:
+    TODO we can also say it's an initial algebra over initial object.
+
     \begin{center}
@@
 
-    % TODO
+    TODO finish definition
 
@@
 
+    \begin{definition}
         Comodel in $\mathcal{C}$ is just a Model in $ \mathcal{C}^{op} $.
+    \end{definition}
 
-    Models --- Worlds
+    We could end by stating this definition, and everything else would follow
+    directly from duality. However, expanding definitions is going to give us
+    better intuition, as well as give more solid ground for implementation.
 
@@
 
-    Coinduction, and coinductive structures, provide us a way of observation
-    of the behaviour, as a contractry to construction in inductive reasoning.
+    Induction and Coinduction
+
+    Induction is a way of constructing new structures. Recursion is a way of
+    folding inductively defined structure in an terminating way. Recursive functions
+    should shrink the argument in each call, meaning that it eventually ends up
+    terminating in a base case.
+
+    Coinduction is literally a dual notion to induction. We \textit{observe} possibly
+    infinite structures, by doing deconstruction. Corecursion, however, is a way
+    of productively defining new, possibly enlarged structures. Due to infinity,
+    the evaluation should be lazy, whereas in induction it may be eager.
+
+    Here is a table that summarizes difference between these two methods of reasoning:
+
+    \begin{grammar}
+
+    TODO rewrite into latex-table \\
+
+    | feature             | induction            | coinduction          |
+
+    |---------------------|----------------------|----------------------|
+
+    | basic activity      | construction         | deconstruction       |
+
+    | derived activity    | deconstruction       | construction         |
+
+    | functions shape     | inductive domain     | coinductive codomain |
+
+    | (co) recursive calls| shrinks the argument | grows the result     |
+
+    | functions feature   | terminating          | productive           |
+
+    | evaluation          | possibly eager       | necessarily lazy     |
+
+    \end{grammar}
+
+    Reader that focuses on pragmatism may pose a question, why do we even want
+    to reason about infinite structures? They never appear in practice! In fact,
+    it's very common to operate on never-ending transition systems or streams of
+    data without, wher efinitary means of reasoning are of no use, as we can’t
+    expect an end to stream!
+
     Simple example of the duality can be expressed through induction over finite
@@
 
-    Doing the latter may involve modification of the internal state of the machine
-    that is generating the infinite streams, or in more concrete scenario,
-    alternation of the external resource that is providing us the data.
+    Doing the latter may involve modification of the internal state of the
+    machine that is generating the infinite streams, or in more concrete
+    scenario, alternation of the external resource that is providing us the data.
+
+    TODO Write a coinductive proof for divergent terms using coinductive proof
+    principle.
+
+\subsection{Coinduction Coalgebra something section}
+
+    TODO Rethink this section, back by examples
 
@@
 
-    % TODO: On it's own this snippet a bit dry and of course requires additional examples,
-    % for instance the classic one with operating on closed file
+    TODO On it's own this snippet a bit dry and of course requires additional examples,
+    for instance the classic one with operating on closed file and/or on
+    streams of data
 
@@
     \subsection{Algebraic Effects}
+
+    In fact, libraries for algebraic effects also arose in mainstream languages,
+    such as C or Python. In Python, effects are implemented through generators,
+    using built-in feature of sending value when doing \textit{yield} operation.
+    Resumptions in handlers that are sending value, are one-shot and tail-recursive,
+    therefore we do not need to handle coalgebraic part, at the cost of flexibility.
+
+    TODO add reference to C and Python lib
 
@@
 
-One-shot continuations
+TODO smth smt initialization before finalization.
 
-Detecting whether resumption is called only once is undecidable.
+It can be seen, that issues lies with the fact, that interacting with coalgebras
+may lead to change of their internal state system. Examples include change
+of state in NFA, taking next element in infinite stream or a closing file.
+
+One of the ways to approach this problem, caused by extensive generality of
+effects, is extension of the calculus with Coalgebraic effects, also named
+coeffects. Difference is that resumptions in coeffects handlers are one-shot
+only, which means that continuation can be invoked at most once.
+
+\begin{lemma}
+    Problem of detection whether resumption is called only once is undecidable.
+\end{lemma}
+
+\begin{proof}
+    Follows directly from Rice's Theorem, as checking whether function was called
+    is a nontrivial semantic property of a program.
+\end{proof}
 
 \section{Dynamic Constraints Checking}
+
+One approach is to dynamically during execution check against contract that
+continuation may only be called once. In this setting we are sure that our
+program will not accidentaly go into wrong state, however by definition we lack
+static analysis to prevent from errors before they occur.
 
@@
 
+TODO Explain it can get complex and unfamiliar for users.
+
 \section{Data-Flow Analysis}
@@
 Another static analysis of the program.
+
+TODO Describe what is data-flow analysis \\
+
+TODO Give example for constant folding \\
+
+TODO Propose analysis on imaginary CFG Not sure if it makes sense?
 
@@
 
-\chapter{Effectful and Coeffectful Programming}
+\chapter{(Co) Effectful Programming}  % TODO w/o space?
 \section{Examples}\label{sec:examples}
@@
     in the program. We will make two decisions, which will affect the computation
+
     result:
@@
 
+    TODO subset of this section's content is explained already
+
     These implications are straightforward when looking from a categorical standpoint,
@@
 
+    TODO Extend language with basic constructs for usability
+
     The syntax for the calculus is shown below. $nat \; n$ represents an integer $n$,
@@
     constructs to the language, for which we shall define the semantics.
+
+    TODO Rewrite here operational semantics from Links paper
 
@@
 
+    TODO Drop of continuations is still an issue
+
     \section{Cohandlers}
     \ldots
+
+    TODO Simple implementation for cohandlers
 

### 2020-08-02 thesis/thesis.tex
Latexify coinduction table

@@
 \usepackage{syntax}
+\usepackage{tabularx}
+\usepackage{booktabs}
 \usepackage{mathalfa}
+\usepackage{csquotes}
 \usepackage{textcomp}
@@
     \begin{center}
-    \begin{tikzcd}[row sep=huge, column sep=large]
-        {|M|}^k \arrow[rr, '\phi', bend right] \arrow[rr, '\rrbracket{} X|l', bend left] &  & {|M|}
-    \end{tikzcd}
+    % \begin{tikzcd}[row sep=huge, column sep=large]
+    %     {|M|}^k \arrow[rr, '\phi', bend right] \arrow[rr, '\rrbracket{} X|l', bend left] &  & {|M|}
+    % \end{tikzcd}
     \end{center}
@@
     \begin{center}
-    \begin{tikzcd}
-    A \arrow[rd] \arrow[r, '\phi'] & B \\
-    & C
-    \end{tikzcd}
+    % \begin{tikzcd}
+    % A \arrow[rd] \arrow[r, '\phi'] & B \\
+    % & C
+    % \end{tikzcd}
     \end{center}
@@
     \end{center}
-    \\
     together with an universal arrow $ \eta_{A} : A \longrightarrow A^{\#} $.
@@
 
-    \begin{grammar}
+    \begin{center}
+    \begin{tabular}{lcc}
+    \toprule
 
-    TODO rewrite into latex-table \\
+    \textbf{feature}             & \textbf{induction}            & \textbf{coinduction}          \\
+    \midrule
 
-    | feature             | induction            | coinduction          |
+    basic activity      & construction          & deconstruction \\
+    \midrule
 
-    |---------------------|----------------------|----------------------|
+    derived activity    & deconstruction       & construction         \\
+    \midrule
 
-    | basic activity      | construction         | deconstruction       |
+    functions shape     & inductive domain     & coinductive codomain \\
+    \midrule
 
-    | derived activity    | deconstruction       | construction         |
+    (co) recursive calls& shrinks the argument & grows the result     \\
+    \midrule
 
-    | functions shape     | inductive domain     | coinductive codomain |
+    functions feature   & terminating          & productive           \\
+    \midrule
 
-    | (co) recursive calls| shrinks the argument | grows the result     |
+    evaluation          & possibly eager       & necessarily lazy     \\
+    \bottomrule
 
-    | functions feature   | terminating          | productive           |
-
-    | evaluation          | possibly eager       | necessarily lazy     |
-
-    \end{grammar}
+    \end{tabular}
+    \end{center}
 

### 2020-08-04 thesis/thesis.tex
[thesis] You know, more content

@@
 \usepackage{mathalfa}
-\usepackage{csquotes}
 \usepackage{textcomp}
@@
 \usepackage{tikz-cd}
-% Not sure if necessary
-\usepackage[utf8]{inputenc}
-\usepackage{rotchiffre}
 
@@
 
-\chapter{Introduction}
+\chapter{Introduction}\label{chapter:introduction}
 
@@
 
-TODO This requires a rework
+In this thesis we discuss issues with excessive generality of algebraic effects,
+and propose a solution without coalgebraic means along with presentation of
+experimental programming language Freak, which is a language with (co) algebraic
+effects and (co) handlers, where implementaiton is based on Continutation Passing
+Style translation.
 
-In this thesis we present experimental programming language Freak, which is an
-implementation of Continuation Passing Style for Effect Handlers paper~\cite{handlers-cps},
-with the addition of a few basic constructs.
+\section{Problem Statement}
 
-\section{Problem Analysis}
+Algebraic effects, while not being a new concept, in the recent years have
+received a lot of attention~\cite{effects-bibliography}, both from theoretical
+and practical side.
 
-TODO add citations
-
-Algebraic effects, while not being a new concept, in the recent years
-have received a lot of attention, both from theoretical and practical side.
-
-TODO refactor the following
-
-They give developer necessary tools for declarative programming, writing
-programs in a way to strive from exposing implementation details, while still
-preserving a lot of flexibility and having strong theoretical background and
-ways of formal reasoning.
+They give developer necessary tools for declarative programming and writing
+programs in a way to strive from exposing implementation details by defining
+computational effects as an API\@. In the same time, they preserve a lot of
+flexibility and have strong theoretical background.
 
@@
 very reason, we develop type systems, to detect errors before they occur.
-
-\section{Problem Statement}
-
-TODO Think about merging Problem Analysis with Statement \\
 
@@
 
-TODO actually point out which sections you talk about
+TODO actually point out which sections you talk about or maybe even put example
+here
 
@@
 
-TODO Guide reader through thesis, probably later than sooner \\
-TODO Clearly divide categorical from operational approach for less algebraically fluent readers
+We start with Chapter~\ref{chapter:background}, which provides a background
+about effects, defines algebraic effects in categorical setting as well as point
+out dual coalgebraic effects and cohandlers, finishing with showing related work
+in this area. Chapter~\ref{chapter:potential-solutions} describes possible ways
+to approach issue of excesive generality,
+Chapter~\ref{chapter:co-effectful-programming} shows Freak language by
+examples along with usage guide, and then
+Chapter~\ref{chapter:calculus-of-freak-language} describes the language's syntax,
+operational semantics and CPS translation. In next one,
+Chapter~\ref{chapter:implementation}, implementation details are revealed.
+We conclude in Chapter~\ref{chapter:conclusion} by stating what are the possible
+augmentations, that are intended to be made in the future.
 
-TODO it's outdated
-We start by presenting the related work,
-then discuss syntax and operational semantics. Basic usage guide for playing with the
-language is provided, as well as implementation details and examples. We conclude by
-stating what are the possible augmentations, that are intended to be made in the future.
-
-\chapter{Background}
+\chapter{Background}\label{chapter:background}
 
@@
 is about taming complexity of programs. Various methods of static analysis has
-been developed for various use cases, most notably, type systems.
-
-What is the source of complexity? Thanks to strong and static type systems along
-with their implementations, we have solid tools to work efficiently on functions
-that are pure. That being said, we claim that the core complexity of programs
-comes from side effects, which we cannot avoid in writing anything useful.
+been developed for various use cases, most notably, type systems.Thanks to strong
+and static type systems along with their implementations, we have solid tools to
+work efficiently on functions that are pure. That being said, we claim that the
+core complexity of programs comes from side effects, or more generally, computational
+effects, which we cannot avoid in writing anything useful.
 
 We need to have a good way for handling computational effects. One of the ways
-to model them, can be done through monads~\cite{moggi}. However, they were
-found to be, to say the least, cumbersome to work with when the number of
-different effects increases. It is perhaps not a coincidence that many
+to model them, can be done through monads~\cite{monads-wadler, moggi}. However,
+they were found to be, to say the least, cumbersome to work with when the number
+of different effects increases. It is perhaps not a coincidence that many
 functional programming languages do not have them, because of the non-composable
@@
 composition, for $f'$, $g'$ we can use Kleisli composition operator from monad $m$.
-What in case of adding one more effect?
+What in case of adding one more effect, $f''$ and $g''$?
 
 It turns out, that it becomes complex and unpleasent to combine two functors
-together to form a new monad, for which purpose, monad transformers arose in Haskell.
-Most of the common languages avoid this by not expressing computational effects
-in the type system, and instead one may think about functions as being implicitly
-embed in a Kleisli Category over a functor T, where T is a hidden signature over
-all possible side effects that occur in our program.
+together to form a new monad, and for this purpose, monad
+transformers~\cite{monad-transformers}arose in Haskell. Most of the common
+languages avoid this by not expressing computational effects in the type system,
+and instead one may think about functions as being implicitly embed in a Kleisli
+Category over a functor T, where T is a hidden signature over all possible side
+effects that occur in our program.
 
@@
 
-Not only we would like to bring back effects to our type system, but also do it
-in a way that is composable. This is where algebraic effects comes to the rescue.
-From theoretical point of view, we need to develop equational theory about our
-effects to assert correctness of our langauge as well as to have the right hammer
-to reason about our programs. That is the point where we would like to introduce
-to unfamiliar reader a notion of algebraic effects.
+Not only we would like to bring back effects to our type
+system~\cite{type-and-effect}, but also do it in a way that is composable. This
+is where algebraic effects comes to the rescue.  From theoretical point of view,
+we need to develop equational theory about our effects to assert correctness of
+our langauge as well as to have the right hammer to reason about our programs.
+That is the point where we would like to introduce to unfamiliar reader a notion
+of algebraic effects.
 
@@
 
-This is really a breakthrough from practical point of view, as we may substitute
+This is really an incredible feature from practical point of view, as we may substitute
 logic depending on execution environment. As an example, fetching for resources
@@
 some of the issues, we end up in functions that need to carry representants
-of the interfaces, and pass them in every subcall that they make. It also can't
-be ommited that it's not really a substitute, as it's solving only one particular
-issue that algebraic effects address.
+of the interfaces, and pass them in every subcall that they make.
 
-TODO add link to paper with plenty of examples
+TODO maybe example with comparison for these two approaches could be useful here?
 
-- Exceptions
-- Async/await
-- Flipping coin
-- Backtracking?
-- Operating on file
+That being said, it's only one particular issue that algebraic effects address.
+In effects approach, handler may also drop the resumption or invoke it more than
+once. There are many other great sources for getting familiarized with
+effects~\cite{handlers-tutorial, programming-in-eff, koka-tutorial}, so we will
+omit further explainations. More examples can be found in
+Chapter~\ref{chapter:co-effectful-programming}.
 
@@
 
+    TODO cite Initial Algebras, Terminal Coalgebras, and the Theory of Fixed Points of Functors
+
     Here is a table that summarizes difference between these two methods of reasoning:
@@
 
+    From categorical standpoint, coinduction is formed over a final coalgebra,
+    where corecursion is the mediator between any coalgbra into a final one,
+    and coinductive proof principle corresponds to uniqueness of the mediator.
+    Recall that coalgebraic effects arises from a particular type of a final
+    coalgebra, namely, cofree coalgebra.
+
     TODO Write a coinductive proof for divergent terms using coinductive proof
-    principle.
+    principle. Introdcution to bisim, page 34,35.
 
@@
     This is one of the cases where interaction with external resource multiple
-    times, or more specifically in case of algebraic effects, invoking resumption more
-    than once, may lead to unexpected behaviour that would not be expected in standard
-    control flow. This captures the excessive generality of effects, and is the
-    issue that we would like to address with coalgebraic effects.
+    times, or more specifically in case of algebraic effects, invoking resumption
+    more than once, may lead to unexpected behaviour that would not be expected
+    in standard control flow.
 
-    TODO On it's own this snippet a bit dry and of course requires additional examples,
-    for instance the classic one with operating on closed file and/or on
-    streams of data
+    Let's consider the following example
+
+    TODO rethink easter eg
+    \begin{verbatim}
+    handle
+        let fh <- do Open "praise.txt" in
+        let c <- do Choice () in
+        if c then do Write (fh, "Guy Fieri") else do Write (fh, "is cool") ;
+        close fh
+    with {
+        return x -> [x] |
+        choose () k -> return (append (k true) (k false)) |
+    }
+    \end{verbatim}
+
+    Where we open file, and based on nondeterministic choice, we write to file,
+    and then close it. This piece of code looks harmless, however, as we invoke
+    resumption for the second time, we are attempting to write to a closed file,
+    and then close it once again. This captures the excessive generality of
+    effects, and is the issue that we would like to address with coalgebraic
+    effects.
 
@@
     In fact, libraries for algebraic effects also arose in mainstream languages,
-    such as C or Python. In Python, effects are implemented through generators,
-    using built-in feature of sending value when doing \textit{yield} operation.
-    Resumptions in handlers that are sending value, are one-shot and tail-recursive,
-    therefore we do not need to handle coalgebraic part, at the cost of flexibility.
-
-    TODO add reference to C and Python lib
+    such as C~\cite{leijen-c} or Python~\cite{python-effect}. In Python,
+    effects are implemented through generators~\cite{one-shot}, using built-in
+    feature of sending value when doing \textit{yield} operation. Resumptions in
+    handlers that are sending value, are one-shot and tail-recursive, therefore
+    we do not need to handle coalgebraic part, at the cost of flexibility.
 
@@
 
-\chapter{Potential Solutions}
+\chapter{Potential Solutions}\label{chapter:potential-solutions}
 
@@
 
+TODO cite LINEAR USAGE OF STATE
+
+In fact, there is a bijective translation between algebraic effects and linear
+type theory, and \textit{every monad embeds in a linear state monad}.
+
 TODO Explain it can get complex and unfamiliar for users.
@@
 
-\chapter{(Co) Effectful Programming}  % TODO w/o space?
+\chapter{(Co) Effectful Programming}\label{chapter:co-effectful-programming} % TODO w/o space?
 \section{Examples}\label{sec:examples}
@@
     in the program. We will make two decisions, which will affect the computation
-
     result:
@@
     directly on the uncurried translation with continuations as explicit stacks
-    from paper. More details can be found in Section~\ref{sec:implementation}.
+    from paper. More details can be found in Section~\ref{chapter:implementation}.
     All commands are available within \verb!src!  directory.
@@
 
-\chapter{Calculus of Freak language}
+\chapter{Calculus of Freak language}\label{chapter:calculus-of-freak-language}
 \section{Syntax}
@@
 
-\chapter{Implementation}\label{sec:implementation}
+\chapter{Implementation}\label{chapter:implementation}
 
@@
 
-\chapter{Conclusion}
+\chapter{Conclusion}\label{chapter:conclusion}
 \section{Summary}

### 2020-08-07 thesis/thesis.tex
[thesis] Checkpoint in writing thesis

@@
 % Dane do oswiadczenia o autorskim wykonaniu
-% \transcriptnum {291480}                     % Numer indeksu
-% \advisorgen    {dr. Macieja Piróga} % Nazwisko promotora w dopelniaczu
+\transcriptnum{291480}                     % Numer indeksu
+\advisorgen{dr. Macieja Piróga} % Nazwisko promotora w dopelniaczu
 %%%%%
 
-%%%%% WLASNE DODATKOWE PAKIETY
+% \usepackage[utf8]{inputenc}
 % \usepackage{graphicx,listings,tikz}
@@
 \usepackage{stmaryrd}
+\usepackage{bussproofs}
 \usepackage{tikz-cd}
+\usetikzlibrary{babel}
+
+% Disable certain warnings
+% chktex-file 18
 
@@
 \addbibresource{mybib.bib}
-%
-%%%%% WŁASNE DEFINICJE I POLECENIA
-%
+
 \theoremstyle{definition} \newtheorem{definition}{Definition}[chapter]
@@
 experimental programming language Freak, which is a language with (co) algebraic
-effects and (co) handlers, where implementaiton is based on Continutation Passing
+effects and (co) handlers, where implementation is based on Continuation Passing
 Style translation.
 
-\section{Problem Statement}
+\section{Problem Statement}\label{sec:problem-statement}
 
 Algebraic effects, while not being a new concept, in the recent years have
-received a lot of attention~\cite{effects-bibliography}, both from theoretical
+received a lot of attention~\cite{effects-bibliography}, both from the theoretical
 and practical side.
@@
 They give developer necessary tools for declarative programming and writing
-programs in a way to strive from exposing implementation details by defining
-computational effects as an API\@. In the same time, they preserve a lot of
-flexibility and have strong theoretical background.
+programs in a way to strive from exposing implementation details, by defining
+computational effects as an API\@ in our code. In the same time, they preserve
+a lot of flexibility and have a strong theoretical background.
 
 In fact, that very flexibility can be troublesome. Allowing developer for too
-much freedome may lead to undesired behaviour and incorrect programs, for that
-very reason, we develop type systems, to detect errors before they occur.
-
-Multi-shot resumptions in effect handlers are a powerful tool to describe
-fairly complex logic in a concise way. That being said, they give rise to
-issues that would not occur in standard control flow, especially when composing
-various effects together.
-
-We will explain this in greater details and think about possible ways to address
-the issue of flexibility of effects in the next sections.
-
-TODO actually point out which sections you talk about or maybe even put example
-here
-
-\section{Thesis Outline}
-
-We start with Chapter~\ref{chapter:background}, which provides a background
-about effects, defines algebraic effects in categorical setting as well as point
-out dual coalgebraic effects and cohandlers, finishing with showing related work
-in this area. Chapter~\ref{chapter:potential-solutions} describes possible ways
-to approach issue of excesive generality,
-Chapter~\ref{chapter:co-effectful-programming} shows Freak language by
-examples along with usage guide, and then
-Chapter~\ref{chapter:calculus-of-freak-language} describes the language's syntax,
-operational semantics and CPS translation. In next one,
-Chapter~\ref{chapter:implementation}, implementation details are revealed.
-We conclude in Chapter~\ref{chapter:conclusion} by stating what are the possible
-augmentations, that are intended to be made in the future.
-
-\chapter{Background}\label{chapter:background}
-
-\section{Computational Effects}
-
-Since the rapid development of computational theory in 1930s by A. Turing,
-K. Godel and A. Church, we have a well-established notion of what can and what
-cannot be done through algorithmic means, which we can almost directly translate
-to being computable by our machines. Through next years we have developed
-mainstream languages that are used almost everywhere, with a great success.
-
-Under these circumstances one may pose a question, why do we still bother with
-development of languages theory, since so much has been done already. Is there
-anything that drives us towards further research? Indeed, one active branch
-revolves around equational theory to asses equality of two programs, which we
-know that in general setting is undecidable. Proof methods may include extensional,
-contextual or logical equality. However, there is no doubt that these formal
-ways of reasoning about programs, while being crucial for assessing correctness,
-do not bring direct benefits for everyday use cases, as they are rarely accessible
-by a common developer.
-
-Other branch of languages theory, that we shall investigate more in this thesis,
-is about taming complexity of programs. Various methods of static analysis has
-been developed for various use cases, most notably, type systems.Thanks to strong
-and static type systems along with their implementations, we have solid tools to
-work efficiently on functions that are pure. That being said, we claim that the
-core complexity of programs comes from side effects, or more generally, computational
-effects, which we cannot avoid in writing anything useful.
-
-We need to have a good way for handling computational effects. One of the ways
-to model them, can be done through monads~\cite{monads-wadler, moggi}. However,
-they were found to be, to say the least, cumbersome to work with when the number
-of different effects increases. It is perhaps not a coincidence that many
-functional programming languages do not have them, because of the non-composable
-nature of them, or at least not composable in their implementations.
-
-Let's put the following functions
-
-\begin{center}
-
-    $ f: a \rightarrow b $, $ g : b \rightarrow c $ \\
-    $ f': a \rightarrow m \; b $, $ g' : b \rightarrow m \; c $ \\
-    $ f'': a \rightarrow m' \; m \; b $, $ g'' : b \rightarrow m' \; m \; c $ \\
-
-\end{center}
-
-where $m$, $m'$ are monads. Functions $f$, $g$ can be composed using standard
-composition, for $f'$, $g'$ we can use Kleisli composition operator from monad $m$.
-What in case of adding one more effect, $f''$ and $g''$?
-
-It turns out, that it becomes complex and unpleasent to combine two functors
-together to form a new monad, and for this purpose, monad
-transformers~\cite{monad-transformers}arose in Haskell. Most of the common
-languages avoid this by not expressing computational effects in the type system,
-and instead one may think about functions as being implicitly embed in a Kleisli
-Category over a functor T, where T is a hidden signature over all possible side
-effects that occur in our program.
-
-TODO Think about describing what is Kleisli Category
-
-Not only we would like to bring back effects to our type
-system~\cite{type-and-effect}, but also do it in a way that is composable. This
-is where algebraic effects comes to the rescue.  From theoretical point of view,
-we need to develop equational theory about our effects to assert correctness of
-our langauge as well as to have the right hammer to reason about our programs.
-That is the point where we would like to introduce to unfamiliar reader a notion
-of algebraic effects.
-
-\section{Algebraic Effects}
-
-Algebraic effects can be thought as an public interface for computational effects.
-Declarative approach allow us to write programs in which the actual semantic of
-source code is dependent on handler that defines the meaning of a subset of effects.
-
-This is really an incredible feature from practical point of view, as we may substitute
-logic depending on execution environment. As an example, fetching for resources
-can behave differently as we run tests, debug our code, or run it on production.
-In the same manner, they neatly allow us to abstract over implementation details.
-
-Patterns like these are well known in programming, for which other alternatives
-arose. One can mention interfaces from object-oriented programming, which are also
-a way to describe the API of a certain component. In order to abstract from
-implementation details, we pass an object around which represents a certain
-interface. While this is certainly better than having no abstractions at all,
-we need to pass this interface into every function that is going to use it.
-This practice is called dependency injection, and while it sounds like it solves
-some of the issues, we end up in functions that need to carry representants
-of the interfaces, and pass them in every subcall that they make.
-
-TODO maybe example with comparison for these two approaches could be useful here?
-
-That being said, it's only one particular issue that algebraic effects address.
-In effects approach, handler may also drop the resumption or invoke it more than
-once. There are many other great sources for getting familiarized with
-effects~\cite{handlers-tutorial, programming-in-eff, koka-tutorial}, so we will
-omit further explainations. More examples can be found in
-Chapter~\ref{chapter:co-effectful-programming}.
-
-\section{Categorical Setting of Universal Algebra}
-
-Algebraic effects can be described via operational means, however,
-for the purpose of presenting the duality between algebra and coalgebra,
-we allow ourselves to wander a bit deeper into category theory and describe
-effects from denotational point of view.
-
-    \subsection{Algebraic Theories}
-
-    \begin{definition}
-
-    A \textit{signature $ \Sigma $} is given by a collection of operation
-    symbols $ op_{i} $ with associated parameters $ P_{i} $ and arities $ A_{i} $,
-    where $ P_{i} $ and $ A_{i} $ are objects in the category of our interest.
-    We will write an operation as $ op_{i} : P_{i} \rightsquigarrow A_{i} $
-
-    \end{definition}
-
-    \begin{definition}
-    Collection of $\mathVar{\Sigma-terms}$ is a free algebra with a generator $X$
-    for a functor $ \mu H_{\Sigma} $ that maps objects into trees over a given
-    signature $ \Sigma $ and morphisms into folds over trees.
-
-    \end{definition}
-
-    \begin{definition}
-
-        A $ \mathVar{\Sigma-Equation} $ is an object X and a pair of
-        $\mathVar{\Sigma-terms}$ $l, r \in Tree_{\Sigma}(X)$, written as
-
-        \begin{center}
-        $ X \mid l = r $
-        \end{center}
-
-    \end{definition}
-
-    \begin{definition}
-
-    An \textit{algebraic theory} $T = (\Sigma_{T}, \mathcal{E}_{T})$, is given
-    by a signature $\Sigma_{T}$ and a collection $\mathcal{E}_{T}$ of
-    $\mathVar{\Sigma_{T}-equations}$. For clarity, we will usually omit T subscript.
-
-    \end{definition}
-
-    \begin{definition}
-
-    An \textit{interpretation $I$ over a given signature $\Sigma$} is given by
-    a carrier object $|I|$ and for each $ op_{i} : P_{i} \rightsquigarrow A_{i} $
-    in $\Sigma$ a map
-
-    \begin{center}
-        $ {\llbracket op_{i} \rrbracket}_I : P_{i} \times{} {|I|}^{A_{i}} \rightarrow |I| $
-    \end{center}
-    Interpretation may be naturally extended to $\mathVar{\Sigma-terms}$, such
-    that a given $\mathVar{\Sigma-term}$ $X \mid t$ is interpreted by a map
-    which sends variables into projections from environment and terms into
-    map composition over each subterm.
-
-    \end{definition}
-
-    \begin{definition}
-
-    A \textit{model M} of an algebraic theory T is an interpretation of the
-    signature ${\Sigma_{T}}$ which validates all the equations $\mathcal{E_{T}}$.
-    That is, for every equation $X \mid l = r$ the following diagram commutes:
-
-    \begin{center}
-    % \begin{tikzcd}[row sep=huge, column sep=large]
-    %     {|M|}^k \arrow[rr, '\phi', bend right] \arrow[rr, '\rrbracket{} X|l', bend left] &  & {|M|}
-    % \end{tikzcd}
-    \end{center}
-
-    TODO this diagram doesn't work
-
-    \begin{center}
-    % \begin{tikzcd}
-    % A \arrow[rd] \arrow[r, '\phi'] & B \\
-    % & C
-    % \end{tikzcd}
-    \end{center}
-
-    \end{definition}
-
-    \begin{definition}
-
-    Free F-algebra on an object A (of generators) in $\mathcal{C}$ is meant an algebra
-
-    \begin{center}
-
-    $ \varphi_{A} : F A^{\#} \longrightarrow A $
-
-    \end{center}
-    together with an universal arrow $ \eta_{A} : A \longrightarrow A^{\#} $.
-    Universality means that for every algebra $ \beta : F B \longrightarrow B $
-    and every morphism $ f : A \longrightarrow B $ in $ \mathcal{C}$, there
-    exists a unique homomorphism $ \overline{f} : A^{\#} \longrightarrow B $
-    extending f, i.e.\ a unique morphism of $ \mathcal{C}$ for which the diagram
-    below commutes:
-
-    TODO we can also say it's an initial algebra over initial object.
-
-    \begin{center}
-    \begin{tikzcd}[row sep=huge, column sep=large]
-    % F A^{\#} \arrow[d, "F \overline{f}"] \arrow[r, "\varphi_{A}"] & A^{\#} \arrow[d, "\overline{f}"] & A \arrow[l, "\eta_{A}"'] \arrow[ld, "f"] \\
-    % F B \arrow[r, "\beta"']                                       & B                                &
-    \end{tikzcd}
-    \end{center}
-
-    \end{definition}
-
-    \begin{definition} \textit{Free model} is just a model which is free algebra.
-
-    \end{definition}
-
-    \begin{lemma}{Free models form monads}
-
-    \end{lemma}
-
-    \begin{definition}
-
-    Let L, M be models of a theory T. A $ \mathVar{T-homomorphism}$
-    $\phi : L \rightarrow M$ is a map such that the following diagram commutes:
-
-    TODO finish definition
-
-    \end{definition}
-
-\section{Duality}
-    \subsection{Comodels}
-
-    \begin{definition}
-        Comodel in $\mathcal{C}$ is just a Model in $ \mathcal{C}^{op} $.
-    \end{definition}
-
-    We could end by stating this definition, and everything else would follow
-    directly from duality. However, expanding definitions is going to give us
-    better intuition, as well as give more solid ground for implementation.
-
-    \subsection{Cooperations}
-
-    Derive from Models duality
-
-    \subsection{Coalgebraic Effects}
-
-    \subsection{Coinductive Reasoning}
-
-    Induction and Coinduction
-
-    Induction is a way of constructing new structures. Recursion is a way of
-    folding inductively defined structure in an terminating way. Recursive functions
-    should shrink the argument in each call, meaning that it eventually ends up
-    terminating in a base case.
-
-    Coinduction is literally a dual notion to induction. We \textit{observe} possibly
-    infinite structures, by doing deconstruction. Corecursion, however, is a way
-    of productively defining new, possibly enlarged structures. Due to infinity,
-    the evaluation should be lazy, whereas in induction it may be eager.
-
-    TODO cite Initial Algebras, Terminal Coalgebras, and the Theory of Fixed Points of Functors
-
-    Here is a table that summarizes difference between these two methods of reasoning:
-
-    \begin{center}
-    \begin{tabular}{lcc}
-    \toprule
-
-    \textbf{feature}             & \textbf{induction}            & \textbf{coinduction}          \\
-    \midrule
-
-    basic activity      & construction          & deconstruction \\
-    \midrule
-
-    derived activity    & deconstruction       & construction         \\
-    \midrule
-
-    functions shape     & inductive domain     & coinductive codomain \\
-    \midrule
-
-    (co) recursive calls& shrinks the argument & grows the result     \\
-    \midrule
-
-    functions feature   & terminating          & productive           \\
-    \midrule
-
-    evaluation          & possibly eager       & necessarily lazy     \\
-    \bottomrule
-
-    \end{tabular}
-    \end{center}
-
-    Reader that focuses on pragmatism may pose a question, why do we even want
-    to reason about infinite structures? They never appear in practice! In fact,
-    it's very common to operate on never-ending transition systems or streams of
-    data without, wher efinitary means of reasoning are of no use, as we can’t
-    expect an end to stream!
-
-    Simple example of the duality can be expressed through induction over finite
-    list and coinductive observation of infinite streams.
-
-    Doing the latter may involve modification of the internal state of the
-    machine that is generating the infinite streams, or in more concrete
-    scenario, alternation of the external resource that is providing us the data.
-
-    From categorical standpoint, coinduction is formed over a final coalgebra,
-    where corecursion is the mediator between any coalgbra into a final one,
-    and coinductive proof principle corresponds to uniqueness of the mediator.
-    Recall that coalgebraic effects arises from a particular type of a final
-    coalgebra, namely, cofree coalgebra.
-
-    TODO Write a coinductive proof for divergent terms using coinductive proof
-    principle. Introdcution to bisim, page 34,35.
-
-\subsection{Coinduction Coalgebra something section}
-
-    TODO Rethink this section, back by examples
-
-    This is one of the cases where interaction with external resource multiple
-    times, or more specifically in case of algebraic effects, invoking resumption
-    more than once, may lead to unexpected behaviour that would not be expected
-    in standard control flow.
-
-    Let's consider the following example
-
-    TODO rethink easter eg
+much freedom may lead to undesired behaviour and incorrect programs, for this
+reason, we develop type systems, to detect errors before they occur.
+
+Multi-shot resumptions in effect handlers and possibility to drop resumption
+are powerful tools to describe fairly complex logic in a concise way. That being
+said, they give rise to issues that would not occur in standard control flow,
+especially when composing various effects together. Unwanted interaction with
+external resources multiple times or not invoking finalisation code are cases
+that may occur while using algebraic effects and handlers. Let's consider the
+following example:
+
 \begin{verbatim}
@@
 
-    Where we open file, and based on nondeterministic choice, we write to file,
-    and then close it. This piece of code looks harmless, however, as we invoke
-    resumption for the second time, we are attempting to write to a closed file,
-    and then close it once again. This captures the excessive generality of
-    effects, and is the issue that we would like to address with coalgebraic
-    effects.
+In the above code we open file, and based on nondeterministic choice, we write
+to the file, and then close it. This piece of code looks harmless, however, as
+we invoke resumption for the second time, we are attempting to write to the
+closed file, and then close it once again. In fact, dropping resumption is also
+considered harmful, as we would not attempt to close the file.
+
+This captures the excessive generality of effects, and is the issue that we
+would like to address with coalgebraic effects.
+
+\section{Thesis Outline}
+
+In the following chapter we provide a background
+about effects, define algebraic effects in categorical setting as well as point
+out dual coalgebraic effects and cohandlers, finishing with showing related work
+in this area. Chapter~\ref{chapter:potential-solutions} describes possible ways
+to approach issue of excessive generality,
+Chapter~\ref{chapter:co-effectful-programming} shows Freak language by
+examples along with usage guide, and then
+Chapter~\ref{chapter:calculus-of-freak-language} describes the language's syntax,
+operational semantics and CPS translation. In next one,
+Chapter~\ref{chapter:implementation}, implementation details are revealed.
+We conclude in Chapter~\ref{chapter:conclusion} by stating what are the possible
+augmentations, that are intended to be made in the future.
+
+\chapter{Background}\label{chapter:background}
+
+\section{Computational Effects}
+
+Since the rapid development of computational theory in 1930s by A. Turing,
+K. Godel and A. Church, we have a well-established notion of what can and what
+cannot be done through algorithmic means, which we can almost directly translate
+to being computable by our machines. Through next years we have developed
+mainstream languages that are used almost everywhere, with great success.
+
+Under these circumstances one may pose a question, why do we still bother with
+development of languages theory, since so much has been done already. Is there
+anything that drives us towards further research? Indeed, one active branch
+revolves around equational theory to assess equality of two programs, which we
+know that in the general setting is undecidable. Proof methods may include
+extensional, contextual or logical equality. However, there is no doubt that
+these formal ways of reasoning about programs, while being crucial for assessing
+correctness, do not bring direct benefits for everyday use cases, as they are
+rarely accessible by a common developer.
+
+Other branch of languages theory, that we shall investigate more in this thesis,
+is about taming complexity of programs. Various methods of static analysis has
+been developed for various use cases, most notably, type systems. Thanks to strong
+and static type systems along with their implementations, we have solid tools to
+work efficiently on functions that are pure. That being said, we claim that the
+core complexity of programs comes from side effects, or more generally, computational
+effects, which we cannot avoid in writing anything useful.
+
+We need to have a good way for handling computational effects. One of the ways
+to model them, can be done through monads~\cite{monads-wadler, moggi}. However,
+they were found to be, to say the least, cumbersome to work with when the number
+of different effects increase. It is perhaps not a coincidence that many
+functional programming languages do not have them, because of the non-composable
+nature of them, or at least not composable in their implementations.
+
+\noindent
+Let's put the following functions
+
+\begin{center}
+
+    $ f: a \rightarrow b $, $ g : b \rightarrow c $ \\
+    $ f': a \rightarrow m \; b $, $ g' : b \rightarrow m \; c $ \\
+    $ f'': a \rightarrow m' \; m \; b $, $ g'' : b \rightarrow m' \; m \; c $ \\
+
+\end{center}
+
+\noindent
+where $m$, $m'$ are monads. Functions $f$, $g$ can be composed using standard
+composition, for $f'$, $g'$ we can use Kleisli composition operator from monad $m$.
+What in case of adding one more effect, $f''$ and $g''$?
+
+It turns out, that it becomes complex and unpleasant to combine two functors
+together to form a new monad, and for this purpose, monad
+transformers~\cite{monad-transformers} arose in Haskell. Most of the common
+languages avoid this by not expressing computational effects in the type system,
+and instead one may think about functions as being implicitly embed in a Kleisli
+Category over a functor T~\cite{kleisli-ncatlab}, where T is a hidden signature
+over all possible side effects that occur in our program.
+
+\noindent
+TODO Reconsider reference to ncatlab on Kleisli Cat, often it's not the best resource
+
+Not only we would like to bring back effects to our type
+system~\cite{type-and-effect}, but also do it in a way that is composable. This
+is where algebraic effects comes to the rescue. From theoretical point of view,
+we need to develop equational theory about our effects to assert correctness of
+our language as well as to have the right hammer to reason about our programs.
+From practical side, we need to have the way of taming computational effects
+in our programs. That is the point where we would like to introduce to unfamiliar
+readers a notion of algebraic effects.
+
+\section{Algebraic Effects}
+
+Algebraic effects can be thought of as an public interface for computational effects.
+Declarative approach allow us to write programs in which the actual semantic of
+source code is dependent on handler that defines the meaning of a subset of effects.
+
+This is really an incredible feature from practical point of view, as we may substitute
+logic depending on the execution environment. As an example, fetching for resources
+can behave differently as we run tests, debug our code, or run it on production.
+In the same manner, they neatly allow us to abstract over implementation details.
+
+Patterns like these are well known in programming, for which other alternatives
+arose. One can mention interfaces from object-oriented programming, which are also
+a way to describe the API of a certain component. In order to abstract from
+implementation details, we pass an object around which represents a certain
+interface. While this is certainly better than having no abstractions at all,
+we need to pass this interface into every function that is going to use it.
+This practice is called dependency injection, and while it sounds like it solves
+some of the issues, we end up in functions that need to carry representants
+of the interfaces, and pass them in every subcall that they make.
+
+\noindent
+TODO maybe example with comparison for these two approaches could be useful here?
+
+That being said, it's only one particular issue that algebraic effects address.
+Their handlers may also drop the resumption or invoke it more than
+once. There are many other great sources for getting familiarized with
+effects from practical side~\cite{handlers-tutorial, programming-in-eff, koka-tutorial},
+so we will omit further explanations. More examples can be found in
+Chapter~\ref{chapter:co-effectful-programming}.
+
+\section{Categorical Setting of Universal Algebra}
+
+Algebraic effects can be described via operational means, however,
+for the purpose of presenting the duality between algebra and coalgebra,
+we allow ourselves to wander a bit deeper into category theory and describe
+effects from denotational point of view~\cite{bauer-what-is-algebraic}.
+
+    \subsection{Algebraic Theories}
+
+    % == begin Signature ==
+    \begin{definition}
+
+    A \textit{signature $ \Sigma $} is given by a collection of operation
+    symbols $ op_{i} $ with associated parameters $ P_{i} $ and arities $ A_{i} $,
+    where $ P_{i} $ and $ A_{i} $ are objects in the category of our interest.
+    We will write an operation as $ op_{i} : P_{i} \rightsquigarrow A_{i} $
+
+    \end{definition}
+    % == end Signature ==
+
+    % == begin Sigma-terms ==
+    \begin{definition}
+    Collection of $\mathVar{\Sigma-terms}$ is a free algebra with a generator $X$
+    for a functor $ \mu H_{\Sigma} $ that maps objects into trees over a given
+    signature $ \Sigma $ and morphisms into folds over trees.
+
+    \end{definition}
+    % == end Sigma-terms ==
+
+    % == begin Sigma-equation ==
+    \begin{definition}
+
+        A $ \mathVar{\Sigma-Equation} $ is an object X and a pair of
+        $\mathVar{\Sigma-terms}$ $l, r \in Tree_{\Sigma}(X)$, written as
+
+        \begin{center}
+        $ X \mid l = r $
+        \end{center}
+
+    \end{definition}
+    % == end Sigma-equation ==
+
+    % == begin Algebraic Theory ==
+    \begin{definition}
+
+    An \textit{algebraic theory} $T = (\Sigma_{T}, \mathcal{E}_{T})$, is given
+    by a signature $\Sigma_{T}$ and a collection $\mathcal{E}_{T}$ of
+    $\mathVar{\Sigma_{T}-equations}$.
+
+    \end{definition}
+    % == end Algebraic Theory ==
+
+    % == begin Interpretation ==
+    \begin{definition}
+
+    An \textit{interpretation $I$ over a given signature $\Sigma$} is given by
+    a carrier object $|I|$ and for each $ op_{i} : P_{i} \rightsquigarrow A_{i} $
+    in $\Sigma$ a map
+
+    \begin{center}
+        $ {\llbracket op_{i} \rrbracket}_I : P_{i} \times{} {|I|}^{A_{i}} \rightarrow |I| $
+    \end{center}
+
+    \noindent
+    Interpretation may be naturally extended to $\mathVar{\Sigma-terms}$, such
+    that a given $\mathVar{\Sigma-term}$ $X \mid t$ is interpreted by a map
+    which sends variables into projections from environment and terms into
+    map composition over each subterm.
+
+    \end{definition}
+    % == end Interpretation ==
+
+    % == begin Model ==
+    \begin{definition}
+
+    A \textit{model M} of an algebraic theory T is an interpretation of the
+    signature ${\Sigma_{T}}$ which validates all the equations $\mathcal{E_{T}}$.
+    That is, for every equation $X \mid l = r$ the following diagram commutes:
+
+    \begin{center}
+    \begin{tikzcd}[row sep=huge, column sep=large]
+        {|M|}^k \arrow[rr, "\llbracket{} X|r \rrbracket{}"', bend right] \arrow[rr, "\llbracket{} X|l \rrbracket{}", bend left] &  & {|M|}
+    \end{tikzcd}
+    \end{center}
+
+    \end{definition}
+    % == end Model ==
+
+    % == begin homomophism ==
+    \begin{definition}
+
+    Let L, M be models of a theory T. $ \mathVar{T-homomorphism}$
+    $\varphi : L \rightarrow M$ is a map such that, for every operation symbol
+    $op_{i}$ in T the following diagram commutes:
+
+    \begin{center}
+    \begin{tikzcd}
+        {|L|}^{ar_{i}} \arrow[rr, "{\llbracket{} op_{i} \rrbracket{}}_{L}"] \arrow[dd, "\varphi{}^{ar_{i}}"'] &  & {|L|} \arrow[dd, "\varphi{}"] \\
+                                              &  &                                     \\
+        {|M|}^{ar_{i}} \arrow[rr, "{\llbracket{} op_{i} \rrbracket{}}_{M}"']                 &  & {|M|}
+    \end{tikzcd}
+    \end{center}
+
+    \noindent
+    We denote here $A^{n}$ as $\mathVar{n-ary}$ product of $A$.
+
+    \end{definition}
+    % == end homomophism ==
+
+    % == begin Free algebra ==
+    \begin{definition}
+
+    Free F-algebra on an object A (of generators) in $\mathcal{C}$ is meant an algebra
+
+    \begin{center}
+
+    $ \varphi_{A} : F A^{\#} \longrightarrow A $
+
+    \end{center}
+
+    \noindent
+    together with an universal arrow $ \eta_{A} : A \longrightarrow A^{\#} $.
+    Universality means that for every algebra $ \beta : F B \longrightarrow B $
+    and every morphism $ f : A \longrightarrow B $ in $ \mathcal{C}$, there
+    exists a unique homomorphism $ \overline{f} : A^{\#} \longrightarrow B $
+    extending f, i.e.\ a unique morphism of $ \mathcal{C}$ for which the diagram
+    below commutes:
+
+    \begin{center}
+    \begin{tikzcd}[row sep=huge, column sep=large]
+    F A^{\#} \arrow[d, "F \overline{f}"] \arrow[r, "\varphi_{A}"] & A^{\#} \arrow[d, "\overline{f}"] & A \arrow[l, "\eta_{A}"'] \arrow[ld, "f"] \\
+    F B \arrow[r, "\beta"']                                       & B                                &
+    \end{tikzcd}
+    \end{center}
+
+    \end{definition}
+    % == end Free algebra ==
+
+    \begin{definition} \textit{Free model} is just a model that is free algebra.
+    \end{definition}
+
+    \begin{lemma}{Free models form monads}
+    \end{lemma}
+    \begin{proof}
+        Let $M$ be a free model of an algebraic theory $T$ in category $\mathcal{C}$. We
+        have an endofunctor $ F_{T}$, which takes objects into free model and
+        maps to unique homomorphisms for which the following diagram commutes:
+
+        \begin{center}
+        \begin{tikzcd}
+        X \arrow[rr, "\eta_{X}"] \arrow[dd, "f"'] &  & F_{T} (X) \arrow[dd, "\overline{f}"] \\
+                                                  &  &                                     \\
+        Y \arrow[rr, "\eta_{Y}"']                 &  & F_{T} (Y)
+        \end{tikzcd}
+        \end{center}
+
+        \noindent
+        which immediately gives us $\eta$ natural transformation for a monad.
+        We can now define $\mu{}$ for a monad as a unique morphism for which
+        the diagram commutes:
+
+        \begin{center}
+        \begin{tikzcd}[row sep=huge, column sep=large]
+            F_{T} (X) \arrow[r, "\eta_{F_{T} (X)}"] \arrow[rd, "id"'] & F_{T} (F_{T} (X)) \arrow[d, "\mu{}"] \\
+                                                         & F_{T} (Y)
+        \end{tikzcd}
+        \end{center}
+
+        \noindent
+        $F_{T}$ is an endofunctor, therefore it sends $X$ into an object in $\mathcal{C}$.
+        From any object we have a unique map sending an object into model,
+        which is $\eta_{F_{T} (X)}$. From free property of our model we get
+        a unique map $\mu{}$ such that the diagram commutes, therefore we have
+        a monad $(F_{T}, \eta{}, \mu{})$.
+
+    \end{proof}
+
+    % == begin handler ==
+    \begin{definition} \textit{Handler} is a $\mathVar{T-homomorphism}$ between
+    free models
+
+        \begin{center}
+        $ H : {| F_{T} (X) |} \rightarrow {| F_{T'} (X') |} $
+        \end{center}
+
+    \noindent
+    that is, a map between carriers that preserves the structure of theory $T$.
+
+    \end{definition}
+    % == end handler ==
+
+Concluding from the theory that we have built, algebraic effects are just free
+models over computation trees, where the signature is the set of effects, and
+equational theory is definining rules of rewriting expressions with effects.
+The latter is usually irrelevant from implementation perspective, but may be
+important for reasoning about programs.
+
+Effect handlers on the other hand, are transformations from one computation
+tree over given signature of effects into another one, where handler may serve
+effects and also propagate new ones.
+
+\section{Duality}
+    \subsection{Comodels}
+
+    \begin{definition}
+        Comodel in $\mathcal{C}$ is just a Model in $ \mathcal{C}^{op} $.
+    \end{definition}
+
+    We could end by stating this definition, and everything else would follow
+    directly from duality. However, expanding definitions is going to give us
+    better intuition, as well as give more solid ground for implementation.
+    For this part, we are going to assume we operate in \textbf{Set} category.
+
+    \subsection{Cooperations}
+
+    We have defined interpretation of an operation as a map:
+
+    \begin{center}
+        $ {\llbracket op \rrbracket}_M : {|I|}^{A} \rightarrow |I|^{P} $
+    \end{center}
+
+    \noindent
+    Let's now take a look on how it dualizes. In \textbf{Set}, we can think about
+    $A^{B}$ exponentials as $\mathVar{B-ary}$ coproducts over A, where for each
+    argument we select one result. Every arrow in opposite category is reverted,
+    thus coproducts becomes products, our morphism is reverted and model becomes
+    comodel (\textit{world}). Putting this together, we obtain the following map:
+
+    \begin{center}
+        $ {\llbracket op \rrbracket}^W : {|I|} \times {P} \rightarrow  {|I|} \times {A} $
+    \end{center}
+
+    \noindent
+    which is called a \textit{cooperation}. Meaning of that morphism, is that
+    based on coalgebra carrier and a parameter, we obtain new state of the
+    coalgebra and a value generated by coalgebra.
+
+    \subsection{Coalgebraic Effects}
+
+\noindent
+TODO \\
+- coalg effects \\
+- cohandlers
+
+    \subsection{Coinductive Reasoning}
+
+    Induction is a way of constructing new structures. Recursion is a way of
+    folding inductively defined structure in a terminating way. Recursive
+    functions (not to be confused with recursive computability class), should
+    shrink the argument in each call, meaning that it eventually ends up
+    terminating in a base case.
+
+    Coinduction is literally a dual notion to induction. We \textit{observe} possibly
+    infinite structures, by doing deconstruction. Corecursion, is a way
+    of productively defining new, possibly enlarged structures. Due to infinity,
+    the evaluation should be lazy, whereas in induction it may be eager.
+
+    Here is a table that summarizes difference between these two methods of reasoning:
+
+    \begin{center}
+    \begin{tabular}{lcc}
+    \toprule
+
+    \textbf{feature}             & \textbf{induction}            & \textbf{coinduction}          \\
+    \midrule
+
+    basic activity      & construction          & deconstruction \\
+    \midrule
+
+    derived activity    & deconstruction       & construction         \\
+    \midrule
+
+    functions shape     & inductive domain     & coinductive codomain \\
+    \midrule
+
+    (co) recursive calls& shrinks the argument & grows the result     \\
+    \midrule
+
+    functions feature   & terminating          & productive           \\
+    \midrule
+
+    evaluation          & possibly eager       & necessarily lazy     \\
+    \bottomrule
+
+    \end{tabular}
+    \end{center}
+
+    From categorical standpoint, coinduction is formed over a final coalgebra,
+    where corecursion is the mediator between any coalgebra into a final one,
+    and coinductive proof principle corresponds to the uniqueness of the mediator.
+    Recall that coalgebraic effects arises from a particular type of a final
+    coalgebra, namely, cofree coalgebra.
+
+    \noindent
+    TODO Not sure which paper to cite here for a deeper dive into the upper
+    topic. Book \textit{Initial Algebras, Terminal Coalgebras, and the Theory of Fixed Points of Functors}
+    by Adámek, Milius and S. Moss from which I have studied is no longer available
+    on the internet, as it was a draft under construction.
+
+    Reader that focuses on pragmatism may pose a question, why do we even want
+    to reason about infinite structures? They never appear in practice! In fact,
+    it's very common to operate on never-ending transition systems or streams of
+    data, where finitary means of reasoning are of no use, as we can’t expect
+    an end to stream.
+
+    Operating on coinductive structures may involve modification of the internal
+    state of the machine that is generating the infinite streams, or in more
+    concrete scenario, alternation of the external resource that is providing
+    us the data.
+
+    For a better understanding and intuition, we shall illustrate now the
+    difference based on inductive and coinductive relation. Recall that the set
+    $\Lambda$ of $\mathVar{\lambda-terms}$ is given by the following grammar:
+
+    \begin{center}
+        $ e ::= x \;|\; \lambda x.e \;|\; e_{1} e_{2} $
+    \end{center}
+
+    \noindent
+    where $\Lambda^{0} \in \Lambda$ is a set of $\mathVar{closed~\lambda-terms}$,
+    meaning, those terms have no free variables. Let's now define an inductive,
+    and coinductive predicate. Relation
+    $\Downarrow{} \subseteq \Lambda^{0} \times \Lambda^{0}$ (convergence) for
+    call-by-value $\mathVar{\lambda-calculus}$ is defined as follows:
+
+    \begin{center}
+        \AxiomC{}
+        \UnaryInfC{$ \lambda x.e \Downarrow{} \lambda x.e $}
+        \DisplayProof{}
+        \quad\quad
+        \AxiomC{$ e_{1} \Downarrow{} \lambda x.e_{0}$}
+        \AxiomC{$e_{0}[e_{2}/x] \Downarrow{} e'$}
+        \BinaryInfC{$ e_{1} e_{2} \Downarrow{} e' $}
+        \DisplayProof{}
+    \end{center}
+
+    \noindent
+    which means that $ \Downarrow{} $ is the \textit{smallest} predicate
+    \textit{closed forward} under the above rules. Relation can be inductively
+    built from empty set and then in a fixed-point manner we can add more terms.
+    Let's now see a relation $ \Uparrow{} \subseteq \Lambda^{0}$ of
+    \textit{divergent} terms, which is coinductively defined as follows:
+
+    \begin{center}
+        \AxiomC{$ e_{1} \Uparrow{} $}
+        \UnaryInfC{$ e_{1} e_{2} \Uparrow{} $}
+        \DisplayProof{}
+        \quad\quad
+        \AxiomC{$ e_{1} \Downarrow{} \lambda x.e_{0}$}
+        \AxiomC{$e_{0}[e_{2}/x] \Uparrow{}$}
+        \BinaryInfC{$ e_{1} e_{2} \Uparrow{}$}
+        \DisplayProof{}
+    \end{center}
+
+    \noindent
+    which means that $\Uparrow$ is the \textit{greatest} predicate
+    \textit{closed backwards} under the above rules, relation can be coinductively
+    obtained by starting with $\Lambda_{0}$, and then by removing elements that
+    do not fit the rules.
+
+    \noindent
+    For more detailed introduction into coinduction and other examples,
+    see~\cite{jacobs-rutten, sangiorgi-intro}.
+
+    \noindent
+    TODO maybe also write coinductive proof based on two relations above?
 
@@
     \subsection{Algebraic Effects}
+
+    Except from Links language~\cite{handlers-cps}, on which the implementation
+    is based, there are currently many other alternatives available. One may take
+    a look at Frank~\cite{frank}, which provides support for multihandlers,
+    Koka~\cite{leijen-koka}, Helium~\cite{helium} or Eff~\cite{eff}. Except from
+    separate languages, many libraries arose for existing ones like Haskell,
+    Idris, Scala or Multicore OCaml.
 
@@
     feature of sending value when doing \textit{yield} operation. Resumptions in
-    handlers that are sending value, are one-shot and tail-recursive, therefore
-    we do not need to handle coalgebraic part, at the cost of flexibility.
-
-    Except from Links language~\cite{handlers-cps}, on which the implementation
-    is based, there are currently many other alternatives available. One may take
-    a look at Frank~\cite{frank}, which provides a support for multihandlers,
-    Koka~\cite{leijen-koka}, Helium~\cite{helium} or Eff~\cite{eff}. Except from
-    separate languages, many libraries arose for existing ones like Haskell,
-    Idris, Scala or Multicore OCaml.
+    handlers that are sending value are one-shot and tail-recursive, therefore
+    we do not need to handle the coalgebraic part, at the cost of flexibility.
 
@@
 
-    Ahman and Bauer~\cite{runners-in-action}
+    Work that is the closest related to topic of the thesis, is work done by
+    Danel Ahman and Andrej Bauer~\cite{runners-in-action}, as well as their
+    implementation of a $\lambda_{coop}$ language~\cite{coop}, which is a
+    calculus that ensures linear usage of resources as well as execution of
+    code handling finalisation.
+
+\noindent
+TODO Consider those\\
+- From comodels to coalgebras: State and arrays.\\
+- Tensors of comodels and models for operational semantics.\\
+- Stateful runners of effectful computations.
 
@@
 
-TODO smth smt initialization before finalization.
-
-It can be seen, that issues lies with the fact, that interacting with coalgebras
-may lead to change of their internal state system. Examples include change
-of state in NFA, taking next element in infinite stream or a closing file.
-
-One of the ways to approach this problem, caused by extensive generality of
-effects, is extension of the calculus with Coalgebraic effects, also named
-coeffects. Difference is that resumptions in coeffects handlers are one-shot
-only, which means that continuation can be invoked at most once.
+It can be seen, that issues lie with the fact, that interacting with coalgebras
+may lead to change of their internal state system. Examples include change of
+state in NFA, taking next element in an infinite stream or a closing file.
+Finalization must occur after initialization, thus we conclude that resumption
+should be called at least once. However, modifying state twice the same way can
+also lead to unwanted behaviour, as it was seen in
+Section~\ref{sec:problem-statement}, therefore it should be invoked exactly
+once.
 
@@
 \end{lemma}
-
 \begin{proof}
-    Follows directly from Rice's Theorem, as checking whether function was called
-    is a nontrivial semantic property of a program.
+    Follows directly from Rice's Theorem, as checking whether given function is
+    going to be called is a nontrivial semantic property of a program.
 \end{proof}
 
+One of the ways to approach this problem, caused by extensive generality of
+effects, is the introduction of dual Coalgebraic effects, also named coeffects,
+for which we shall discuss various methods of implementation.
+
 \section{Dynamic Constraints Checking}
 
-One approach is to dynamically during execution check against contract that
-continuation may only be called once. In this setting we are sure that our
-program will not accidentaly go into wrong state, however by definition we lack
-static analysis to prevent from errors before they occur.
+One approach is to dynamically check against contract that continuation may only
+be called once. In this setting we are sure that our program will not accidentally
+go into wrong state, however by definition we lack static analysis to prevent
+errors before they occur.
 
@@
 
-Solve the issue through introduction of linear type system.
-
-TODO cite LINEAR USAGE OF STATE
-
-In fact, there is a bijective translation between algebraic effects and linear
-type theory, and \textit{every monad embeds in a linear state monad}.
-
-TODO Explain it can get complex and unfamiliar for users.
+Imposition of a single call of resumption, can be thought in a way that continuation
+is `consumable', such that it's not available after one takes use of it. Putting
+it in that way, it immediately reminds of linear logic~\cite{linear-logic},
+where premises of linear implications are no longer of use in conclusion, and
+in case of linear type theory, binding variable from expression A, disallows
+use of variables in A. Implementation of coalgebraic effects might be simplified
+in languages that already support linear type system, such as Rust~\cite{rust}.
+
+In fact, there is a bijective translation between language with algebraic effects
+and calculus with linear type theory, and \textit{every monad embeds in a linear
+state monad}~\cite{linear-usage-of-state}.
+
+\noindent
+TODO should this section be further extended?
 
@@
 
-TODO Describe what is data-flow analysis \\
-
-TODO Give example for constant folding \\
-
-TODO Propose analysis on imaginary CFG Not sure if it makes sense?
+\noindent
+- TODO Describe what is data-flow analysis \\
+- TODO Give example for constant folding \\
+- TODO Propose analysis for coalgebraic effects. Not sure if that's not too much,
+as it's not the goal of this thesis. Definitealy leave this section for later.
 
@@
 
+\noindent
+TODO muuuch longer explanation of this topic
+
 \chapter{(Co) Effectful Programming}\label{chapter:co-effectful-programming} % TODO w/o space?
-\section{Examples}\label{sec:examples}
-
-    In this section we present a few examples to show the capabilities of the language.
-    The ideas have been based on~\cite{programming-in-eff}, and thus will not be
-    described in great details. More exemplary programs in Freak language can
-    be found under \\ \href{https://github.com/Tomatosoup97/freak/tree/master/src/programs}{\underline{https://github.com/Tomatosoup97/freak/tree/master/src/programs}}.
-
-    \subsection{Choice}\label{sec:choice-example}
-
-    The first example will be based on modelling (nondeterministic) choice
-    in the program. We will make two decisions, which will affect the computation
-    result:
-
-    \begin{verbatim}
-    let c1 <- do Choice () in
-    let c2 <- do Choice () in
-    let x <- if c1 then return 10 else return 20 in
-    let y <- if c2 then return 0 else return 5 in
-        return x - y
-    \end{verbatim}
-    With that in hand, we may want to define effect handlers:
-
-    \begin{verbatim}
-    handle ... with {
-        Choice p r ->
-            let t <- r 1 in
-            let f <- r 0 in
-            <PLACEHOLDER> |
-        return x -> return x
-    }
-    \end{verbatim}
-    where in the \verb!<PLACEHOLDER>! we can define on what to do with the
-    computation. For example, min-max strategy for picking the minimum value:
-
-    \begin{verbatim}
-    if t < f then return t else return f
-    \end{verbatim}
-    where the code evaluates to \verb!5!. Another example is a handler that
-    collects all possible results, which can be achieved by putting
-    \verb!return (t, f)! in the \verb!<PLACEHOLDER>!, which evaluates to
-    \verb!((10, 5), (20, 15))!.
-
-    \subsection{Exceptions}
-
-    Exceptions are simply algebraic effect handlers which drop the resumption.
-
-    \begin{verbatim}
-    handle
-        if x == 0 then do ZeroDivisionError ()
-                  else return 1/x
-    with {
-        ZeroDivisionError p r -> return 42 |
-        return x -> return x
-    }
-    \end{verbatim}
-    Where we imagine that $x$ variable has been bound previously.
-
-    \subsection{Taming Side effects}
-
-    The complexity of the programs and their performance usually comes from side effects.
-    Algebraic effects allow us to define code in a declarative manner, and hence
-    neatly tame the side effects that they produce. This gives us a lot of flexibility
-    in the actual meaning without duplicating the code. Let's consider the following
-    very basic code snippet:
-
-    \begin{verbatim}
-    let x <- do Fetch () in
-    -- operate on x
-    \end{verbatim}
-
-    The code is dependent on a context in which it is executed, which here is
-    the handler that defines the behaviour of the algebraic \verb!Fetch! effect.
-    In the imperative, or even functional approach, we would need to provide
-    the interface for fetching the data by doing dependency injection or even
-    embedding the operation directly. Here we are just stating what operation
-    we are performing, leaving the interpretation up to the execution context,
-    which could do the fetching or mock the external resource.
-
-    TODO subset of this section's content is explained already
-
-    These implications are straightforward when looking from a categorical standpoint,
-    where effects are viewed as free models of algebraic theories~\cite{adequacy},
-    and handlers are homomorphisms preserving the model structure~\cite{handlers}.
-    Nevertheless, the results are very exciting for programming use cases. The current
-    Freak implementation does not support I/O.
-
-\section{Coexamples}
-
-    Examples for cohandlers
-
-\section{Usage guide}
-
-    % Outdated
-    As of this day, two implementations are available, one based on the curried
-    translation and Appel~\cite{appel-continuations}, and the second one based
-    directly on the uncurried translation with continuations as explicit stacks
-    from paper. More details can be found in Section~\ref{chapter:implementation}.
-    All commands are available within \verb!src!  directory.
-
-    \subsection{Build and install}
-
-    \begin{itemize}
-        \item Install dependencies: \verb!make install!
-        \item Select implementation: \verb!make link-lists! (default) vs \verb!make link-appel!
-        \item Compile: \verb!make build!
-        \item Link to PATH:~\verb!sudo make link!
-        \item Remove artifacts: \verb!make clean!
-    \end{itemize}
-
-    After compiling and linking program to PATH, one may evaluate program as
-    follows: \verb!freak programs/choicesList.fk!. The actual code is described in Section~\ref{sec:choice-example}
-
-    \subsection{Running tests}
-
-    Test cases are available \href{https://github.com/Tomatosoup97/freak/blob/master/src/Tests.hs}{\underline{here}},
-    they include both inline and file-based tests. For more details about
-    writing tests, one may refer to \textit{HUnit documentation}~\cite{hunit-docs}.
-
-    \begin{itemize}
-        \item Run tests: \verb!make tests!
-        \item Run code linter: \verb!make lint!
-        \item Compile, run linter and tests: \verb!make check!
-    \end{itemize}
+% \section{Examples}\label{sec:examples}
+
+%     In this section we present a few examples to show the capabilities of the language.
+%     The ideas have been based on~\cite{programming-in-eff}, and thus will not be
+%     described in great details. More exemplary programs in Freak language can
+%     be found under \\ \href{https://github.com/Tomatosoup97/freak/tree/master/src/programs}{\underline{https://github.com/Tomatosoup97/freak/tree/master/src/programs}}.
+
+%     \subsection{Choice}\label{sec:choice-example}
+
+%     The first example will be based on modelling (nondeterministic) choice
+%     in the program. We will make two decisions, which will affect the computation
+%     result:
+
+%     \begin{verbatim}
+%     let c1 <- do Choice () in
+%     let c2 <- do Choice () in
+%     let x <- if c1 then return 10 else return 20 in
+%     let y <- if c2 then return 0 else return 5 in
+%         return x - y
+%     \end{verbatim}
+%     With that in hand, we may want to define effect handlers:
+
+%     \begin{verbatim}
+%     handle ... with {
+%         Choice p r ->
+%             let t <- r 1 in
+%             let f <- r 0 in
+%             <PLACEHOLDER> |
+%         return x -> return x
+%     }
+%     \end{verbatim}
+%     where in the \verb!<PLACEHOLDER>! we can define what to do with the
+%     computation. For example, min-max strategy for picking the minimum value:
+
+%     \begin{verbatim}
+%     if t < f then return t else return f
+%     \end{verbatim}
+%     where the code evaluates to \verb!5!. Another example is a handler that
+%     collects all possible results, which can be achieved by putting
+%     \verb!return (t, f)! in the \verb!<PLACEHOLDER>!, which evaluates to
+%     \verb!((10, 5), (20, 15))!.
+
+%     \subsection{Exceptions}
+
+%     Exceptions are simply algebraic effect handlers which drop the resumption.
+
+%     \begin{verbatim}
+%     handle
+%         if x == 0 then do ZeroDivisionError ()
+%                   else return 1/x
+%     with {
+%         ZeroDivisionError p r -> return 42 |
+%         return x -> return x
+%     }
+%     \end{verbatim}
+%     Where we imagine that $x$ variable has been bound previously.
+
+%     \subsection{Taming Side effects}
+
+%     The complexity of the programs and their performance usually comes from side effects.
+%     Algebraic effects allow us to define code in a declarative manner, and hence
+%     neatly tame the side effects that they produce. This gives us a lot of flexibility
+%     in the actual meaning without duplicating the code. Let's consider the following
+%     very basic code snippet:
+
+%     \begin{verbatim}
+%     let x <- do Fetch () in
+%     -- operate on x
+%     \end{verbatim}
+
+%     The code is dependent on a context in which it is executed, which here is
+%     the handler that defines the behaviour of the algebraic \verb!Fetch! effect.
+%     In the imperative, or even functional approach, we would need to provide
+%     the interface for fetching the data by doing dependency injection or even
+%     embedding the operation directly. Here we are just stating what operation
+%     we are performing, leaving the interpretation up to the execution context,
+%     which could do the fetching or mock the external resource.
+
+%     \noindent
+%     TODO subset of this section's content is explained already
+
+%     These implications are straightforward when looking from a categorical standpoint,
+%     where effects are viewed as free models of algebraic theories~\cite{adequacy},
+%     and handlers are homomorphisms preserving the model structure~\cite{handlers}.
+%     Nevertheless, the results are very exciting for programming use cases. The current
+%     Freak implementation does not support I/O.
+
+% \section{Coexamples}
+
+%     Examples for cohandlers
+
+% \section{Usage guide}
+
+%     % Outdated
+%     As of this day, two implementations are available, one based on the curried
+%     translation and Appel~\cite{appel-continuations}, and the second one based
+%     directly on the uncurried translation with continuations as explicit stacks
+%     from paper. More details can be found in Section~\ref{chapter:implementation}.
+%     All commands are available within the \verb!src! directory.
+
+%     \subsection{Build and install}
+
+%     \begin{itemize}
+%         \item Install dependencies: \verb!make install!
+%         \item Select implementation: \verb!make link-lists! (default) vs \verb!make link-appel!
+%         \item Compile: \verb!make build!
+%         \item Link to PATH:~\verb!sudo make link!
+%         \item Remove artifacts: \verb!make clean!
+%     \end{itemize}
+
+%     After compiling and linking program to PATH, one may evaluate program as
+%     follows: \verb!freak programs/choicesList.fk!. The actual code is described in Section~\ref{sec:choice-example}
+
+%     \subsection{Running tests}
+
+%     Test cases are available \href{https://github.com/Tomatosoup97/freak/blob/master/src/Tests.hs}{\underline{here}},
+%     they include both inline and file-based tests. For more details about
+%     writing tests, one may refer to \textit{HUnit documentation}~\cite{hunit-docs}.
+
+%     \begin{itemize}
+%         \item Run tests: \verb!make tests!
+%         \item Run code linter: \verb!make lint!
+%         \item Compile, run linter and tests: \verb!make check!
+%     \end{itemize}
 
 \chapter{Calculus of Freak language}\label{chapter:calculus-of-freak-language}
-\section{Syntax}
-
-    TODO Extend language with basic constructs for usability
-
-    The syntax for the calculus is shown below. $nat \; n$ represents an integer $n$,
-    $V \oplus W$ and $V \approx W$ are respectively binary and relational operators,
-    where we support basic arithmetic and comparison operations.
-    \textbf{if} $V$ \textbf{then} $M$ \textbf{else} $N$ is a standard branching
-    statement. The other constructs are just as in Links, with slight syntax
-    modifications. Actual programs in Freak can be found in Section~\ref{sec:examples}.
-
-    \begin{grammar}
-
-        <Values V, W> $::=$ $ x $ | $nat \; n$ \\
-            | $ \backslash x : A \rightarrow M $ | \textbf{rec} $ g \; x \rightarrow M $\\
-            | $V \oplus W$ | $V \approx W$ \\
-            | <> | $ \{ \ell = V; W\} $  | ${[ \ell \; V]}^{R}$
-
-        <Computations M, N> $::=$ $ V $ $ W $ \\
-            | \textbf{if} $V$ \textbf{then} $M$ \textbf{else} $N$ \\
-            | \textbf{let} $\{\ell  = x; y\} = V$ \textbf{in} $ N $ \\
-            | \textbf{case} $V \{ \ell \; x \rightarrow M; y \rightarrow N\}$ | \textbf{absurd} $ V $ \\
-            | \textbf{return} $V$ | \textbf{let} $ x \leftarrow M $ \textbf{in} $ N $ \\
-            | \textbf{do} $\ell \; V$ | \textbf{handle} $M$ \textbf{with} $ \{ H \} $
-
-        <Handlers H> $::=$ \textbf{return} $ x \rightarrow M $ | $ \ell \; p \; r \rightarrow M, H $
-
-        <Binary operators $\oplus$> $::=$ + | $-$ | * | /
-
-        <Relational operators $\approx$> $::=$ $ \textless $ | $\leqslant$ | $>$ | $\geqslant$ | == | $!= $
-
-    \end{grammar}
-
-\section{Typing Rules}
-\section{Operational Semantics}
-
-    The source language's dynamics have been described
-    extensively by providing small-step operational semantics,
-    continuation passing style transformation~\cite{handlers-cps} as well
-    as abstract machine~\cite{liberating-effects}, which was proved to coincide
-    with CPS translation. That being said, Freak introduces new basic
-    constructs to the language, for which we shall define the semantics.
-
-    TODO Rewrite here operational semantics from Links paper
-
-    \begin{flushleft}
-    Extension of the evaluation contexts:
-    \end{flushleft}
-
-    \begin{flushleft}
-    $\mathcal{E} ::= \mathcal{E} \oplus W \; | \; nat \; n \oplus \mathcal{E} \; |$ \textbf{if} $\mathcal{E}$ \textbf{then} $M$ \textbf{else} $N$
-    \end{flushleft}
-
-    \begin{flushleft}
-    Small-step operational semantics:
-    \end{flushleft}
-
-    \begin{flushleft}
-    \textbf{if} $nat \; n$ \textbf{then} $M$ \textbf{else} $N \rightsquigarrow M \quad \quad $ if $n \neq 0$ \\
-    \textbf{if} $nat \; n$ \textbf{then} $M$ \textbf{else} $N \rightsquigarrow N \quad \quad $ if $n = 0$
-    \end{flushleft}
-
-    \begin{flushleft}
-    $nat \; n \oplus nat \; n' \rightsquigarrow n'' \quad \quad $    if $ n'' = n \oplus n' $ \\
-    $nat \; n \approx nat \; n' \rightsquigarrow 1  \quad \quad $    \; if $ n \approx n' $ \\
-    $nat \; n \approx nat \; n' \rightsquigarrow 0  \quad \quad $    \; if $ n \not\approx n' $
-
-    \end{flushleft}
-
-\section{Continuation Passing Style Transformation}
-    \ldots
+% \section{Syntax}
+
+%     \noindent
+%     TODO Extend language with basic constructs for usability
+
+%     The syntax for the calculus is shown below. $nat \; n$ represents an integer $n$,
+%     $V \oplus W$ and $V \approx W$ are respectively binary and relational operators,
+%     where we support basic arithmetic and comparison operations.
+%     \textbf{if} $V$ \textbf{then} $M$ \textbf{else} $N$ is a standard branching
+%     statement. The other constructs are just as in Links, with slight syntax
+%     modifications. Actual programs in Freak can be found in Section~\ref{sec:examples}.
+
+%     \begin{grammar}
+
+%         <Values V, W> $::=$ $ x $ | $nat \; n$ \\
+%             | $ \backslash x : A \rightarrow M $ | \textbf{rec} $ g \; x \rightarrow M $\\
+%             | $V \oplus W$ | $V \approx W$ \\
+%             | <> | $ \{ \ell = V; W\} $  | ${[ \ell \; V]}^{R}$
+
+%         <Computations M, N> $::=$ $ V $ $ W $ \\
+%             | \textbf{if} $V$ \textbf{then} $M$ \textbf{else} $N$ \\
+%             | \textbf{let} $\{\ell  = x; y\} = V$ \textbf{in} $ N $ \\
+%             | \textbf{case} $V \{ \ell \; x \rightarrow M; y \rightarrow N\}$ | \textbf{absurd} $ V $ \\
+%             | \textbf{return} $V$ | \textbf{let} $ x \leftarrow M $ \textbf{in} $ N $ \\
+%             | \textbf{do} $\ell \; V$ | \textbf{handle} $M$ \textbf{with} $ \{ H \} $
+
+%         <Handlers H> $::=$ \textbf{return} $ x \rightarrow M $ | $ \ell \; p \; r \rightarrow M, H $
+
+%         <Binary operators $\oplus$> $::=$ + | $-$ | * | /
+
+%         <Relational operators $\approx$> $::=$ $ \textless $ | $\leqslant$ | $>$ | $\geqslant$ | == | $!= $
+
+%     \end{grammar}
+
+% \section{Typing Rules}
+% \section{Operational Semantics}
+
+%     The source language's dynamics have been described
+%     extensively by providing small-step operational semantics,
+%     continuation passing style transformation~\cite{handlers-cps} as well
+%     as abstract machine~\cite{liberating-effects}, which was proved to coincide
+%     with CPS translation. That being said, Freak introduces new basic
+%     constructs to the language, for which we shall define the semantics.
+
+%     \noindent
+%     TODO Rewrite here operational semantics from Links paper
+
+%     \begin{flushleft}
+%     Extension of the evaluation contexts:
+%     \end{flushleft}
+
+%     \begin{flushleft}
+%     $\mathcal{E} ::= \mathcal{E} \oplus W \; | \; nat \; n \oplus \mathcal{E} \; |$ \textbf{if} $\mathcal{E}$ \textbf{then} $M$ \textbf{else} $N$
+%     \end{flushleft}
+
+%     \begin{flushleft}
+%     Small-step operational semantics:
+%     \end{flushleft}
+
+%     \begin{flushleft}
+%     \textbf{if} $nat \; n$ \textbf{then} $M$ \textbf{else} $N \rightsquigarrow M \quad \quad $ if $n \neq 0$ \\
+%     \textbf{if} $nat \; n$ \textbf{then} $M$ \textbf{else} $N \rightsquigarrow N \quad \quad $ if $n = 0$
+%     \end{flushleft}
+
+%     \begin{flushleft}
+%     $nat \; n \oplus nat \; n' \rightsquigarrow n'' \quad \quad $    if $ n'' = n \oplus n' $ \\
+%     $nat \; n \approx nat \; n' \rightsquigarrow 1  \quad \quad $    \; if $ n \approx n' $ \\
+%     $nat \; n \approx nat \; n' \rightsquigarrow 0  \quad \quad $    \; if $ n \not\approx n' $
+
+%     \end{flushleft}
+
+% \section{Continuation Passing Style Transformation}
+%     \ldots
 
@@
 
-    The Freak implementation is available at \href{https://github.com/Tomatosoup97/freak}{\underline{https://github.com/Tomatosoup97/freak}},
-    written purely in Haskell. While the paper provided a good overview of the
-    language and the translation, the lower-level details were omitted. That
-    being said, two inherently different takes at the implementations were made.
-    The first one is based on curried translation and A. Appel~\cite{appel-continuations}
-    book, and the second one directly on the uncurried translation to target
-    calculus with continuations represented as explicit stacks from the paper.
-    We start by presenting core data structures, and afterwards move to actual
-    translation details.
-
-    \section{Abstract Syntax Tree}\label{sec:implementation-ast}
-
-    The language's AST is defined without surprises, just as syntax is:
-
-    \begin{verbatim}
-data Value
-    = VVar Var
-    | VNum Integer
-    | VLambda Var ValueType Comp
-    | VFix Var Var Comp
-    | VUnit
-    | VPair Value Value
-    | VRecordRow (RecordRow Value)
-    | VExtendRow Label Value Value
-    | VVariantRow (VariantRow Value)
-    | VBinOp BinaryOp Value Value
-
-data Comp
-    = EVal Value
-    | ELet Var Comp Comp
-    | EApp Value Value
-    | ESplit Label Var Var Value Comp
-    | ECase Value Label Var Comp Var Comp
-    | EReturn Value
-    | EAbsurd Value
-    | EIf Value Comp Comp
-    | EDo Label Value
-    | EHandle Comp Handler
-    \end{verbatim}
-    Similarly for the target calculus data structure. However, as one may notice,
-    for convenience the \textbf{let} translation is homomorphic, as opposed to be
-    to lambda abstracted with immediate application:
-
-    \begin{verbatim}
-data UValue
-    = UVar Var
-    | UNum Integer
-    | UBool Bool
-    | ULambda Var UComp
-    | UUnit
-    | UPair UValue UValue
-    | ULabel Label
-    | URec Var Var UComp
-    | UBinOp BinaryOp UValue UValue
-
-data UComp
-    = UVal UValue
-    | UApp UComp UComp
-    | USplit Label Var Var UValue UComp
-    | UCase UValue Label UComp Var UComp
-    | UIf UValue UComp UComp
-    | ULet Var UComp UComp
-    | UAbsurd UValue
-
-    \end{verbatim}
-    The final answer, common to both evaluations, is represented as a \verb!DValue!,
-    where the meaning of the coproduct is as one would expect:
-
-    \begin{verbatim}
-type Label = String
-type FuncRecord = [DValue] -> Either Error DValue
-data DValue
-    = DNum Integer
-    | DLambda FuncRecord
-    | DUnit
-    | DPair DValue DValue
-    | DLabel Label
-    \end{verbatim}
-
-    \section{Target calculus}
-
-    \section{Curried translation}
-
-    The first take was heavily inspired by A. Appel's Compiling with
-    Continuations~\cite{appel-continuations}, which provides a translation for
-    a simplified ML calculus. The calculus was extended and translation adapted to
-    handle algebraic effects and their handlers. The translation is based on the curried
-    first-order translation. That being said, the source code diverged a lot from the
-    paper on which it was based, leading to a different transformation for which the
-    correctness and cohesion with operational semantics should be proved separately.
-    Indeed, while the interpreter worked well on the use cases defined in tests, the
-    evaluation had a part which was not tail-recursive. What's more, nested handlers
-    were not supported, and the implementation was found to be trickier than it should,
-    as it was not obvious on how to adopt the technique proposed in the paper.
-
-    In terms of improving the performance of the evaluation, uncurried higher-order
-    translation should be adapted, so that administrative redexes are contracted
-    and proper tail-recursion is obtained. The core data structure, into which the
-    source program is transformed, is defined as follows:
-
-    \begin{verbatim}
-data ContComp
-    = CPSApp CValue [CValue]
-    | CPSResume CValue ContComp
-    | CPSFix Var [Var] ContComp ContComp
-    | CPSBinOp BinaryOp CValue CValue Var ContComp
-    | CPSValue CValue
-    | CPSLet Var CValue ContComp
-    | CPSSplit Label Var Var CValue ContComp
-    | CPSCase CValue Label Var ContComp Var ContComp
-    | CPSIf CValue ContComp ContComp
-    | CPSAbsurd CValue
-    \end{verbatim}
-    Most of the terms at the end have a coinductive reference to itself, which
-    represents the rest of the computation that needs to be done. For more
-    clarification, one may take a look into the book mentioned
-    above~\cite{appel-continuations}. The source code for curried translation and
-    evaluation can be found respectively in \verb!CPSAppel.hs! and
-    \verb!EvalCPS.hs!.
-
-    \section{Uncurried translation}
-
-    Having in mind the drawbacks mentioned above, alternative translation was
-    written, that coincides with the translation from the paper. Namely, with
-    the uncurried translation to target calculus with continuations represented
-    as explicit stacks. The target calculus was described in Section~\ref{sec:implementation-ast},
-    for which the evaluation is straightforward. The continuations are represented
-    as \verb!Cont!, with syntactic distinction between pure and effectful computations,
-    which occupy alternating positions in the stack. Explicit distinction gave
-    more control in the source code.
-
-    \begin{verbatim}
-type CPSMonad a = ExceptT Error (State Int) a
-
-type ContF = UValue -> [Cont] -> CPSMonad UComp
-
-data Cont = Pure ContF
-          | Eff ContF
-    \end{verbatim}
-    Where \verb!CPSMonad! is a monad transformer over \verb!Either! and \verb!State!.
-    \verb!State! was required to generate labels for fresh variables that
-    came from the translation. The core code is split into five functions:
-
-    \begin{verbatim}
-cps     :: Comp    -> [Cont] -> CPSMonad UComp
-cpsVal  :: Value   -> [Cont] -> CPSMonad UValue
-cpsHRet :: Handler -> Cont
-cpsHOps :: Handler -> Cont
-forward :: Label   -> UValue -> UValue -> [Cont] -> CPSMonad UComp
-    \end{verbatim}
-    Where the first two are implementing cps for computations and values.
-    \verb!cpsHRet! and \verb!cpsHOps! are yielding pure and effectful continuations,
-    based on a given handler. The last one is responsible for forwarding the
-    computation to the outer handler.
-
-    This results in an implementation that finally supports nested handlers,
-    as can be seen by evaluating \verb!programs/complexNestedHandlers.fk! program.
-    Unfortunately, following closely translation from the paper resulted in
-    a behaviour, in which invoked resumption forgets its pure continuation. This means,
-    that the following code, evaluating correctly on the Appel-based translation,
-    returns 0 rather than 1:
-
-    \begin{verbatim}
-handle do Drop () with { Drop p r -> let t <- r 0 in return 1}
-    \end{verbatim}
-    Nevertheless, working out this issue appears as less demanding than coping with
-    discrepancies created in the first translation. The source code for uncurried
-    translation and evaluation can be found respectively in \verb!CPSLists.hs! and
-    \verb!EvalTarget.hs!.
-
-    TODO Drop of continuations is still an issue
-
-    \section{Cohandlers}
-    \ldots
-
-    TODO Simple implementation for cohandlers
-
-    \section{Source Code Structure}
-
-    The source code is divided into a number of modules, where the most
-    crucial parts have already been described.
-
-    \begin{verbatim}
-    AST.hs          - AST data structures
-    CommonCPS.hs    - Common functions for CPS translation
-    CommonEval.hs   - Common functions for evaluation
-    CPSAppel.hs     - Appel-based CPS translation
-    CPSLists.hs     - Uncurried CPS translation
-    EvalCPS.hs      - Evaluation of the Appel's CPS structure
-    EvalTarget.hs   - Evaluation of the target calculus
-    Freak.hs        - API for the language
-    Main.hs         - Main module running evaluator on given filename
-    Parser.hs       - Parser and lexer
-    TargetAST.hs    - AST for the target calculus
-    Tests.hs        - Tests module
-    Types.hs        - Common types definition
-    programs/       - Exemplary programs used in tests
-    \end{verbatim}
+%     The Freak implementation is available on github~\cite{freak}, written purely
+%     in Haskell. While the paper provided a good overview of the language and the
+%     translation, the lower-level details were omitted. That being said, two
+%     inherently different takes at implementations were made. The first one
+%     is based on curried translation and A. Appel~\cite{appel-continuations}
+%     book, and the second one directly on the uncurried translation to target
+%     calculus with continuations represented as explicit stacks from the paper.
+%     We start by presenting core data structures, and afterwards move to actual
+%     translation details.
+
+%     \section{Abstract Syntax Tree}\label{sec:implementation-ast}
+
+%     The language's AST is defined without surprises, just as syntax is:
+
+%     \begin{verbatim}
+% data Value
+%     = VVar Var
+%     | VNum Integer
+%     | VLambda Var ValueType Comp
+%     | VFix Var Var Comp
+%     | VUnit
+%     | VPair Value Value
+%     | VRecordRow (RecordRow Value)
+%     | VExtendRow Label Value Value
+%     | VVariantRow (VariantRow Value)
+%     | VBinOp BinaryOp Value Value
+
+% data Comp
+%     = EVal Value
+%     | ELet Var Comp Comp
+%     | EApp Value Value
+%     | ESplit Label Var Var Value Comp
+%     | ECase Value Label Var Comp Var Comp
+%     | EReturn Value
+%     | EAbsurd Value
+%     | EIf Value Comp Comp
+%     | EDo Label Value
+%     | EHandle Comp Handler
+%     \end{verbatim}
+%     Similarly for the target calculus data structure. However, as one may notice,
+%     for convenience the \textbf{let} translation is homomorphic, as opposed to be
+%     to lambda abstracted with immediate application:
+
+%     \begin{verbatim}
+% data UValue
+%     = UVar Var
+%     | UNum Integer
+%     | UBool Bool
+%     | ULambda Var UComp
+%     | UUnit
+%     | UPair UValue UValue
+%     | ULabel Label
+%     | URec Var Var UComp
+%     | UBinOp BinaryOp UValue UValue
+
+% data UComp
+%     = UVal UValue
+%     | UApp UComp UComp
+%     | USplit Label Var Var UValue UComp
+%     | UCase UValue Label UComp Var UComp
+%     | UIf UValue UComp UComp
+%     | ULet Var UComp UComp
+%     | UAbsurd UValue
+
+%     \end{verbatim}
+%     The final answer, common to both evaluations, is represented as a \verb!DValue!,
+%     where the meaning of the coproduct is as one would expect:
+
+%     \begin{verbatim}
+% type Label = String
+% type FuncRecord = [DValue] -> Either Error DValue
+% data DValue
+%     = DNum Integer
+%     | DLambda FuncRecord
+%     | DUnit
+%     | DPair DValue DValue
+%     | DLabel Label
+%     \end{verbatim}
+
+%     \section{Target calculus}
+
+%     \section{Curried translation}
+
+%     The first take was heavily inspired by A. Appel's Compiling with
+%     Continuations~\cite{appel-continuations}, which provides a translation for
+%     a simplified ML calculus. The calculus was extended and translation adapted to
+%     handle algebraic effects and their handlers. The translation is based on the curried
+%     first-order translation. That being said, the source code diverged a lot from the
+%     paper on which it was based, leading to a different transformation for which the
+%     correctness and cohesion with operational semantics should be proved separately.
+%     Indeed, while the interpreter worked well on the use cases defined in tests, the
+%     evaluation had a part which was not tail-recursive. What's more, nested handlers
+%     were not supported, and the implementation was found to be trickier than it should,
+%     as it was not obvious on how to adopt the technique proposed in the paper.
+
+%     In terms of improving the performance of the evaluation, uncurried higher-order
+%     translation should be adapted, so that administrative redexes are contracted
+%     and proper tail-recursion is obtained. The core data structure, into which the
+%     source program is transformed, is defined as follows:
+
+%     \begin{verbatim}
+% data ContComp
+%     = CPSApp CValue [CValue]
+%     | CPSResume CValue ContComp
+%     | CPSFix Var [Var] ContComp ContComp
+%     | CPSBinOp BinaryOp CValue CValue Var ContComp
+%     | CPSValue CValue
+%     | CPSLet Var CValue ContComp
+%     | CPSSplit Label Var Var CValue ContComp
+%     | CPSCase CValue Label Var ContComp Var ContComp
+%     | CPSIf CValue ContComp ContComp
+%     | CPSAbsurd CValue
+%     \end{verbatim}
+%     Most of the terms at the end have a coinductive reference to itself, which
+%     represents the rest of the computation that needs to be done. For more
+%     clarification, one may take a look into the book mentioned
+%     above~\cite{appel-continuations}. The source code for curried translation and
+%     evaluation can be found respectively in \verb!CPSAppel.hs! and
+%     \verb!EvalCPS.hs!.
+
+%     \section{Uncurried translation}
+
+%     Having in mind the drawbacks mentioned above, alternative translation was
+%     written, that coincides with the translation from the paper. Namely, with
+%     the uncurried translation to target calculus with continuations represented
+%     as explicit stacks. The target calculus was described in Section~\ref{sec:implementation-ast},
+%     for which the evaluation is straightforward. The continuations are represented
+%     as \verb!Cont!, with syntactic distinction between pure and effectful computations,
+%     which occupy alternating positions in the stack. Explicit distinction gave
+%     more control in the source code.
+
+%     \begin{verbatim}
+% type CPSMonad a = ExceptT Error (State Int) a
+
+% type ContF = UValue -> [Cont] -> CPSMonad UComp
+
+% data Cont = Pure ContF
+%           | Eff ContF
+%     \end{verbatim}
+%     Where \verb!CPSMonad! is a monad transformer over \verb!Either! and \verb!State!.
+%     \verb!State! was required to generate labels for fresh variables that
+%     came from the translation. The core code is split into five functions:
+
+%     \begin{verbatim}
+% cps     :: Comp    -> [Cont] -> CPSMonad UComp
+% cpsVal  :: Value   -> [Cont] -> CPSMonad UValue
+% cpsHRet :: Handler -> Cont
+% cpsHOps :: Handler -> Cont
+% forward :: Label   -> UValue -> UValue -> [Cont] -> CPSMonad UComp
+%     \end{verbatim}
+%     Where the first two are implementing cps for computations and values.
+%     \verb!cpsHRet! and \verb!cpsHOps! are yielding pure and effectful continuations,
+%     based on a given handler. The last one is responsible for forwarding the
+%     computation to the outer handler.
+
+%     This results in an implementation that finally supports nested handlers,
+%     as can be seen by evaluating \verb!programs/complexNestedHandlers.fk! program.
+%     Unfortunately, following closely translation from the paper resulted in
+%     a behaviour, in which invoked resumption forgets its pure continuation. This means,
+%     that the following code, evaluating correctly on the Appel-based translation,
+%     returns 0 rather than 1:
+
+%     \begin{verbatim}
+% handle do Drop () with { Drop p r -> let t <- r 0 in return 1}
+%     \end{verbatim}
+%     Nevertheless, working out this issue appears as less demanding than coping with
+%     discrepancies created in the first translation. The source code for uncurried
+%     translation and evaluation can be found respectively in \verb!CPSLists.hs! and
+%     \verb!EvalTarget.hs!.
+
+%     \noindent
+%     TODO Drop of continuations is still an issue
+
+%     \section{Cohandlers}
+%     \ldots
+
+%     \noindent
+%     TODO Simple implementation for cohandlers
+
+%     \section{Source Code Structure}
+
+%     The source code is divided into a number of modules, where the most
+%     crucial parts have already been described.
+
+%     \begin{verbatim}
+%     AST.hs          - AST data structures
+%     CommonCPS.hs    - Common functions for CPS translation
+%     CommonEval.hs   - Common functions for evaluation
+%     CPSAppel.hs     - Appel-based CPS translation
+%     CPSLists.hs     - Uncurried CPS translation
+%     EvalCPS.hs      - Evaluation of the Appel's CPS structure
+%     EvalTarget.hs   - Evaluation of the target calculus
+%     Freak.hs        - API for the language
+%     Main.hs         - Main module running evaluator on given filename
+%     Parser.hs       - Parser and lexer
+%     TargetAST.hs    - AST for the target calculus
+%     Tests.hs        - Tests module
+%     Types.hs        - Common types definition
+%     programs/       - Exemplary programs used in tests
+%     \end{verbatim}
 
 \chapter{Conclusion}\label{chapter:conclusion}
-\section{Summary}
-
-    What was achieved, what was described, what needs to be done or researched.
-
-\section{Future work}
-
-    \subsection{Abstract machine}
-
-    The Links language also provides small-step operational semantics and
-    an abstract machine~\cite{liberating-effects}. Implementing another way
-    of evaluation could serve as a way to empirically assert correctness,
-    as opposed to formally.
-
-    \subsection{Type inference and row polymorphism}
-
-    The type system as of this day is not implemented, as the focus has been put
-    on CPS transformation. Further work is required here, especially considering
-    the fact that a huge advantage of algebraic effects is that they are explicitly
-    defined in the type of a computation.
-
-    \subsection{Multiple instances of algebraic effects}
-
-    The Freak language is limited to a single instance of an effect. We would
-    need to support cases where many instances of the algebraic effects, with
-    the same handler code, could be instantiated. The current state of the
-    art introduces a concept of resources and instances, as in Eff~\cite{programming-in-eff},
-    or instance variables, as in Helium~\cite{binders-labels}.
-
-    \subsection{Selective CPS}
-
-    Other languages, like Koka~\cite{leijen-koka}, or even the core of the Links, are
-    performing selective CPS translation, which reduces the overhead on code
-    that does not perform algebraic effects. Our current translation is fully
-    embedded in the CPS.\@
-
-    \subsection{Exceptions as separate constructs}
-
-    Exceptions are a trivial example of algebraic effect where the resumption is
-    discarded, and as described in \S 4.5~\cite{handlers-cps}, they can be modeled
-    as a separate construct to improve performance.
-
-    \subsection{Shallow handlers}
-
-    Shallow and deep handlers while being able to simulate each other up to
-    administrative reductions, have a very different meaning from a theoretical point
-    of view. Implementing them as defined by Lindley et al.~\cite{shallow-handlers}
-    could be another way of enhancing Freak.
+% \section{Summary}
+
+%     What was achieved, what was described, what needs to be done or researched.
+
+% \section{Future work}
+
+%     \subsection{Abstract machine}
+
+%     The Links language also provides small-step operational semantics and
+%     an abstract machine~\cite{liberating-effects}. Implementing another way
+%     of evaluation could serve as a way to empirically assert correctness,
+%     as opposed to formally.
+
+%     \subsection{Type inference and row polymorphism}
+
+%     The type system as of this day is not implemented, as the focus has been put
+%     on CPS transformation. Further work is required here, especially considering
+%     the fact that a huge advantage of algebraic effects is that they are explicitly
+%     defined in the type of a computation.
+
+%     \subsection{Multiple instances of algebraic effects}
+
+%     The Freak language is limited to a single instance of an effect. We would
+%     need to support cases where many instances of the algebraic effects, with
+%     the same handler code, could be instantiated. The current state of the
+%     art introduces a concept of resources and instances, as in Eff~\cite{programming-in-eff},
+%     or instance variables, as in Helium~\cite{binders-labels}.
+
+%     \subsection{Selective CPS}
+
+%     Other languages, like Koka~\cite{leijen-koka}, or even the core of the Links, are
+%     performing selective CPS translation, which reduces the overhead on code
+%     that does not perform algebraic effects. Our current translation is fully
+%     embedded in the CPS.\@
+
+%     \subsection{Exceptions as separate constructs}
+
+%     Exceptions are a trivial example of algebraic effect where the resumption is
+%     discarded, and as described in \S 4.5~\cite{handlers-cps}, they can be modeled
+%     as a separate construct to improve performance.
+
+%     \subsection{Shallow handlers}
+
+%     Shallow and deep handlers while being able to simulate each other up to
+%     administrative reductions, have a very different meaning from a theoretical point
+%     of view. Implementing them as defined by Lindley et al.~\cite{shallow-handlers}
+%     could be another way of enhancing Freak.
 

### 2020-08-08 src/EvalTarget.hs
[src] Change do notation to bind syntax due to compilation error

@@ -8,9 +8,9 @@ import Types
 
 eval :: Env -> UComp -> Either Error DValue
 eval env c = case c of
-    USplit l x y pair comp -> do
-        (DPair lv rv) <- eval env (UVal pair)
-        let extEnv = Map.insert y rv (Map.insert x lv env)
+    USplit l x y pair comp ->
+         eval env (UVal pair) >>= \(DPair lv rv) ->
+        let extEnv = Map.insert y rv (Map.insert x lv env) in
         eval extEnv comp
     UApp fnComp argValue -> do
         fnVal <- eval env fnComp
@@ -18,8 +18,8 @@ eval env c = case c of
         case fnVal of
             DLambda fun -> fun [argDValue]
             e -> Left $ EvalError $ "Application of non-lambda term " ++ show e
-    UIf e tC fC -> do
-        (DNum n) <- eval env (UVal e)
+    UIf e tC fC ->
+        eval env (UVal e) >>= \(DNum n) ->
         eval env (if n > 0 then tC else fC)
     UCase label caseLabel tComp y fComp -> do
         l <- eval env (UVal label)
@@ -33,9 +33,9 @@ eval env c = case c of
         eval env' comp
     -- Values
     UVal (UBinOp op e1 e2) -> do
-        (DNum n1) <- eval env (UVal e1)
-        (DNum n2) <- eval env (UVal e2)
-        return $ DNum $ convOp op (n1, n2)
+        eval env (UVal e1) >>= \(DNum n1) ->
+            eval env (UVal e2) >>= \(DNum n2) ->
+            return $ DNum $ convOp op (n1, n2)
     UVal (UVar x) -> case Map.lookup x env of
         Just v -> return v
         Nothing -> unboundVarErr x

### 2020-08-08 thesis/thesis.tex
[thesis] Uncomment chapters 4-7

--- 2020-08-04 thesis/thesis.tex
@@
 They give developer necessary tools for declarative programming and writing
-programs in a way to strive from exposing implementation details by defining
-computational effects as an API\@. In the same time, they preserve a lot of
-flexibility and have strong theoretical background.
+programs in a way to strive from exposing implementation details, by defining
+computational effects as an API\@ in our code. In the same time, they preserve
+a lot of flexibility and have a strong theoretical background.
 
 In fact, that very flexibility can be troublesome. Allowing developer for too
-much freedome may lead to undesired behaviour and incorrect programs, for that
-very reason, we develop type systems, to detect errors before they occur.
-
-Multi-shot resumptions in effect handlers are a powerful tool to describe
-fairly complex logic in a concise way. That being said, they give rise to
-issues that would not occur in standard control flow, especially when composing
-various effects together.
-
-We will explain this in greater details and think about possible ways to address
-the issue of flexibility of effects in the next sections.
-
-TODO actually point out which sections you talk about or maybe even put example
-here
+much freedom may lead to undesired behaviour and incorrect programs, for this
+reason, we develop type systems, to detect errors before they occur.
+
+Multi-shot resumptions in effect handlers and possibility to drop resumption
+are powerful tools to describe fairly complex logic in a concise way. That being
+said, they give rise to issues that would not occur in standard control flow,
+especially when composing various effects together. Unwanted interaction with
+external resources multiple times or not invoking finalisation code are cases
+that may occur while using algebraic effects and handlers. Let's consider the
+following example:
+
+\begin{verbatim}
+handle
+    let fh <- do Open "praise.txt" in
+    let c <- do Choice () in
+    if c then do Write (fh, "Guy Fieri") else do Write (fh, "is cool") ;
+    close fh
+with {
+    return x -> [x] |
+    choose () k -> return (append (k true) (k false)) |
+}
+\end{verbatim}
+
+In the above code we open file, and based on nondeterministic choice, we write
+to the file, and then close it. This piece of code looks harmless, however, as
+we invoke resumption for the second time, we are attempting to write to the
+closed file, and then close it once again. In fact, dropping resumption is also
+considered harmful, as we would not attempt to close the file.
+
+This captures the excessive generality of effects, and is the issue that we
+would like to address with coalgebraic effects.
 
@@
 
-We start with Chapter~\ref{chapter:background}, which provides a background
-about effects, defines algebraic effects in categorical setting as well as point
+In the following chapter we provide a background
+about effects, define algebraic effects in categorical setting as well as point
 out dual coalgebraic effects and cohandlers, finishing with showing related work
 in this area. Chapter~\ref{chapter:potential-solutions} describes possible ways
-to approach issue of excesive generality,
+to approach issue of excessive generality,
 Chapter~\ref{chapter:co-effectful-programming} shows Freak language by
@@
 to being computable by our machines. Through next years we have developed
-mainstream languages that are used almost everywhere, with a great success.
+mainstream languages that are used almost everywhere, with great success.
 
@@
 anything that drives us towards further research? Indeed, one active branch
-revolves around equational theory to asses equality of two programs, which we
-know that in general setting is undecidable. Proof methods may include extensional,
-contextual or logical equality. However, there is no doubt that these formal
-ways of reasoning about programs, while being crucial for assessing correctness,
-do not bring direct benefits for everyday use cases, as they are rarely accessible
-by a common developer.
+revolves around equational theory to assess equality of two programs, which we
+know that in the general setting is undecidable. Proof methods may include
+extensional, contextual or logical equality. However, there is no doubt that
+these formal ways of reasoning about programs, while being crucial for assessing
+correctness, do not bring direct benefits for everyday use cases, as they are
+rarely accessible by a common developer.
 
@@
 is about taming complexity of programs. Various methods of static analysis has
-been developed for various use cases, most notably, type systems.Thanks to strong
+been developed for various use cases, most notably, type systems. Thanks to strong
 and static type systems along with their implementations, we have solid tools to
@@
 they were found to be, to say the least, cumbersome to work with when the number
-of different effects increases. It is perhaps not a coincidence that many
+of different effects increase. It is perhaps not a coincidence that many
 functional programming languages do not have them, because of the non-composable
@@
 
+\noindent
 Let's put the following functions
@@
 
+\noindent
 where $m$, $m'$ are monads. Functions $f$, $g$ can be composed using standard
@@
 
-It turns out, that it becomes complex and unpleasent to combine two functors
+It turns out, that it becomes complex and unpleasant to combine two functors
 together to form a new monad, and for this purpose, monad
-transformers~\cite{monad-transformers}arose in Haskell. Most of the common
+transformers~\cite{monad-transformers} arose in Haskell. Most of the common
 languages avoid this by not expressing computational effects in the type system,
 and instead one may think about functions as being implicitly embed in a Kleisli
-Category over a functor T, where T is a hidden signature over all possible side
-effects that occur in our program.
-
-TODO Think about describing what is Kleisli Category
+Category over a functor T~\cite{kleisli-ncatlab}, where T is a hidden signature
+over all possible side effects that occur in our program.
+
+\noindent
+TODO Reconsider reference to ncatlab on Kleisli Cat, often it's not the best resource
 
@@
 we need to develop equational theory about our effects to assert correctness of
-our langauge as well as to have the right hammer to reason about our programs.
-That is the point where we would like to introduce to unfamiliar reader a notion
-of algebraic effects.
+our language as well as to have the right hammer to reason about our programs.
+From practical side, we need to have the way of taming computational effects
+in our programs. That is the point where we would like to introduce to unfamiliar
+readers a notion of algebraic effects.
 
@@
 
-Algebraic effects can be thought as an public interface for computational effects.
+Algebraic effects can be thought of as an public interface for computational effects.
 Declarative approach allow us to write programs in which the actual semantic of
@@
 This is really an incredible feature from practical point of view, as we may substitute
-logic depending on execution environment. As an example, fetching for resources
+logic depending on the execution environment. As an example, fetching for resources
 can behave differently as we run tests, debug our code, or run it on production.
@@
 
+\noindent
 TODO maybe example with comparison for these two approaches could be useful here?
@@
 That being said, it's only one particular issue that algebraic effects address.
-In effects approach, handler may also drop the resumption or invoke it more than
+Their handlers may also drop the resumption or invoke it more than
 once. There are many other great sources for getting familiarized with
-effects~\cite{handlers-tutorial, programming-in-eff, koka-tutorial}, so we will
-omit further explainations. More examples can be found in
+effects from practical side~\cite{handlers-tutorial, programming-in-eff, koka-tutorial},
+so we will omit further explanations. More examples can be found in
 Chapter~\ref{chapter:co-effectful-programming}.
@@
 we allow ourselves to wander a bit deeper into category theory and describe
-effects from denotational point of view.
+effects from denotational point of view~\cite{bauer-what-is-algebraic}.
 
@@
 
+    % == begin Signature ==
     \begin{definition}
@@
     \end{definition}
-
+    % == end Signature ==
+
+    % == begin Sigma-terms ==
     \begin{definition}
@@
     \end{definition}
-
+    % == end Sigma-terms ==
+
+    % == begin Sigma-equation ==
     \begin{definition}
@@
     \end{definition}
-
+    % == end Sigma-equation ==
+
+    % == begin Algebraic Theory ==
     \begin{definition}
@@
     by a signature $\Sigma_{T}$ and a collection $\mathcal{E}_{T}$ of
-    $\mathVar{\Sigma_{T}-equations}$. For clarity, we will usually omit T subscript.
+    $\mathVar{\Sigma_{T}-equations}$.
 
     \end{definition}
-
+    % == end Algebraic Theory ==
+
+    % == begin Interpretation ==
     \begin{definition}
@@
     \end{center}
+
+    \noindent
     Interpretation may be naturally extended to $\mathVar{\Sigma-terms}$, such
@@
     \end{definition}
-
+    % == end Interpretation ==
+
+    % == begin Model ==
     \begin{definition}
@@
     \begin{center}
-    % \begin{tikzcd}[row sep=huge, column sep=large]
-    %     {|M|}^k \arrow[rr, '\phi', bend right] \arrow[rr, '\rrbracket{} X|l', bend left] &  & {|M|}
-    % \end{tikzcd}
-    \end{center}
-
-    TODO this diagram doesn't work
-
-    \begin{center}
-    % \begin{tikzcd}
-    % A \arrow[rd] \arrow[r, '\phi'] & B \\
-    % & C
-    % \end{tikzcd}
+    \begin{tikzcd}[row sep=huge, column sep=large]
+        {|M|}^k \arrow[rr, "\llbracket{} X|r \rrbracket{}"', bend right] \arrow[rr, "\llbracket{} X|l \rrbracket{}", bend left] &  & {|M|}
+    \end{tikzcd}
     \end{center}
@@
     \end{definition}
-
+    % == end Model ==
+
+    % == begin homomophism ==
     \begin{definition}
 
+    Let L, M be models of a theory T. $ \mathVar{T-homomorphism}$
+    $\varphi : L \rightarrow M$ is a map such that, for every operation symbol
+    $op_{i}$ in T the following diagram commutes:
+
+    \begin{center}
+    \begin{tikzcd}
+        {|L|}^{ar_{i}} \arrow[rr, "{\llbracket{} op_{i} \rrbracket{}}_{L}"] \arrow[dd, "\varphi{}^{ar_{i}}"'] &  & {|L|} \arrow[dd, "\varphi{}"] \\
+                                              &  &                                     \\
+        {|M|}^{ar_{i}} \arrow[rr, "{\llbracket{} op_{i} \rrbracket{}}_{M}"']                 &  & {|M|}
+    \end{tikzcd}
+    \end{center}
+
+    \noindent
+    We denote here $A^{n}$ as $\mathVar{n-ary}$ product of $A$.
+
+    \end{definition}
+    % == end homomophism ==
+
+    % == begin Free algebra ==
+    \begin{definition}
+
     Free F-algebra on an object A (of generators) in $\mathcal{C}$ is meant an algebra
@@
     \end{center}
+
+    \noindent
     together with an universal arrow $ \eta_{A} : A \longrightarrow A^{\#} $.
@@
 
-    TODO we can also say it's an initial algebra over initial object.
-
     \begin{center}
     \begin{tikzcd}[row sep=huge, column sep=large]
-    % F A^{\#} \arrow[d, "F \overline{f}"] \arrow[r, "\varphi_{A}"] & A^{\#} \arrow[d, "\overline{f}"] & A \arrow[l, "\eta_{A}"'] \arrow[ld, "f"] \\
-    % F B \arrow[r, "\beta"']                                       & B                                &
+    F A^{\#} \arrow[d, "F \overline{f}"] \arrow[r, "\varphi_{A}"] & A^{\#} \arrow[d, "\overline{f}"] & A \arrow[l, "\eta_{A}"'] \arrow[ld, "f"] \\
+    F B \arrow[r, "\beta"']                                       & B                                &
     \end{tikzcd}
@@
     \end{definition}
-
-    \begin{definition} \textit{Free model} is just a model which is free algebra.
-
+    % == end Free algebra ==
+
+    \begin{definition} \textit{Free model} is just a model that is free algebra.
     \end{definition}
@@
     \begin{lemma}{Free models form monads}
-
     \end{lemma}
-
-    \begin{definition}
-
-    Let L, M be models of a theory T. A $ \mathVar{T-homomorphism}$
-    $\phi : L \rightarrow M$ is a map such that the following diagram commutes:
-
-    TODO finish definition
+    \begin{proof}
+        Let $M$ be a free model of an algebraic theory $T$ in category $\mathcal{C}$. We
+        have an endofunctor $ F_{T}$, which takes objects into free model and
+        maps to unique homomorphisms for which the following diagram commutes:
+
+        \begin{center}
+        \begin{tikzcd}
+        X \arrow[rr, "\eta_{X}"] \arrow[dd, "f"'] &  & F_{T} (X) \arrow[dd, "\overline{f}"] \\
+                                                  &  &                                     \\
+        Y \arrow[rr, "\eta_{Y}"']                 &  & F_{T} (Y)
+        \end{tikzcd}
+        \end{center}
+
+        \noindent
+        which immediately gives us $\eta$ natural transformation for a monad.
+        We can now define $\mu{}$ for a monad as a unique morphism for which
+        the diagram commutes:
+
+        \begin{center}
+        \begin{tikzcd}[row sep=huge, column sep=large]
+            F_{T} (X) \arrow[r, "\eta_{F_{T} (X)}"] \arrow[rd, "id"'] & F_{T} (F_{T} (X)) \arrow[d, "\mu{}"] \\
+                                                         & F_{T} (Y)
+        \end{tikzcd}
+        \end{center}
+
+        \noindent
+        $F_{T}$ is an endofunctor, therefore it sends $X$ into an object in $\mathcal{C}$.
+        From any object we have a unique map sending an object into model,
+        which is $\eta_{F_{T} (X)}$. From free property of our model we get
+        a unique map $\mu{}$ such that the diagram commutes, therefore we have
+        a monad $(F_{T}, \eta{}, \mu{})$.
+
+    \end{proof}
+
+    % == begin handler ==
+    \begin{definition} \textit{Handler} is a $\mathVar{T-homomorphism}$ between
+    free models
+
+        \begin{center}
+        $ H : {| F_{T} (X) |} \rightarrow {| F_{T'} (X') |} $
+        \end{center}
+
+    \noindent
+    that is, a map between carriers that preserves the structure of theory $T$.
 
     \end{definition}
+    % == end handler ==
+
+Concluding from the theory that we have built, algebraic effects are just free
+models over computation trees, where the signature is the set of effects, and
+equational theory is definining rules of rewriting expressions with effects.
+The latter is usually irrelevant from implementation perspective, but may be
+important for reasoning about programs.
+
+Effect handlers on the other hand, are transformations from one computation
+tree over given signature of effects into another one, where handler may serve
+effects and also propagate new ones.
 
@@
     better intuition, as well as give more solid ground for implementation.
+    For this part, we are going to assume we operate in \textbf{Set} category.
 
@@
 
-    Derive from Models duality
+    We have defined interpretation of an operation as a map:
+
+    \begin{center}
+        $ {\llbracket op \rrbracket}_M : {|I|}^{A} \rightarrow |I|^{P} $
+    \end{center}
+
+    \noindent
+    Let's now take a look on how it dualizes. In \textbf{Set}, we can think about
+    $A^{B}$ exponentials as $\mathVar{B-ary}$ coproducts over A, where for each
+    argument we select one result. Every arrow in opposite category is reverted,
+    thus coproducts becomes products, our morphism is reverted and model becomes
+    comodel (\textit{world}). Putting this together, we obtain the following map:
+
+    \begin{center}
+        $ {\llbracket op \rrbracket}^W : {|I|} \times {P} \rightarrow  {|I|} \times {A} $
+    \end{center}
+
+    \noindent
+    which is called a \textit{cooperation}. Meaning of that morphism, is that
+    based on coalgebra carrier and a parameter, we obtain new state of the
+    coalgebra and a value generated by coalgebra.
 
@@
 
+\noindent
+TODO \\
+- coalg effects \\
+- cohandlers
+
     \subsection{Coinductive Reasoning}
 
-    Induction and Coinduction
-
     Induction is a way of constructing new structures. Recursion is a way of
-    folding inductively defined structure in an terminating way. Recursive functions
-    should shrink the argument in each call, meaning that it eventually ends up
+    folding inductively defined structure in a terminating way. Recursive
+    functions (not to be confused with recursive computability class), should
+    shrink the argument in each call, meaning that it eventually ends up
     terminating in a base case.
@@
     Coinduction is literally a dual notion to induction. We \textit{observe} possibly
-    infinite structures, by doing deconstruction. Corecursion, however, is a way
+    infinite structures, by doing deconstruction. Corecursion, is a way
     of productively defining new, possibly enlarged structures. Due to infinity,
@@
 
-    TODO cite Initial Algebras, Terminal Coalgebras, and the Theory of Fixed Points of Functors
-
     Here is a table that summarizes difference between these two methods of reasoning:
@@
     \end{center}
+
+    From categorical standpoint, coinduction is formed over a final coalgebra,
+    where corecursion is the mediator between any coalgebra into a final one,
+    and coinductive proof principle corresponds to the uniqueness of the mediator.
+    Recall that coalgebraic effects arises from a particular type of a final
+    coalgebra, namely, cofree coalgebra.
+
+    \noindent
+    TODO Not sure which paper to cite here for a deeper dive into the upper
+    topic. Book \textit{Initial Algebras, Terminal Coalgebras, and the Theory of Fixed Points of Functors}
+    by Adámek, Milius and S. Moss from which I have studied is no longer available
+    on the internet, as it was a draft under construction.
 
@@
     it's very common to operate on never-ending transition systems or streams of
-    data without, wher efinitary means of reasoning are of no use, as we can’t
-    expect an end to stream!
-
-    Simple example of the duality can be expressed through induction over finite
-    list and coinductive observation of infinite streams.
-
-    Doing the latter may involve modification of the internal state of the
-    machine that is generating the infinite streams, or in more concrete
-    scenario, alternation of the external resource that is providing us the data.
-
-    From categorical standpoint, coinduction is formed over a final coalgebra,
-    where corecursion is the mediator between any coalgbra into a final one,
-    and coinductive proof principle corresponds to uniqueness of the mediator.
-    Recall that coalgebraic effects arises from a particular type of a final
-    coalgebra, namely, cofree coalgebra.
-
-    TODO Write a coinductive proof for divergent terms using coinductive proof
-    principle. Introdcution to bisim, page 34,35.
-
-\subsection{Coinduction Coalgebra something section}
-
-    TODO Rethink this section, back by examples
-
-    This is one of the cases where interaction with external resource multiple
-    times, or more specifically in case of algebraic effects, invoking resumption
-    more than once, may lead to unexpected behaviour that would not be expected
-    in standard control flow.
-
-    Let's consider the following example
-
-    TODO rethink easter eg
-    \begin{verbatim}
-    handle
-        let fh <- do Open "praise.txt" in
-        let c <- do Choice () in
-        if c then do Write (fh, "Guy Fieri") else do Write (fh, "is cool") ;
-        close fh
-    with {
-        return x -> [x] |
-        choose () k -> return (append (k true) (k false)) |
-    }
-    \end{verbatim}
-
-    Where we open file, and based on nondeterministic choice, we write to file,
-    and then close it. This piece of code looks harmless, however, as we invoke
-    resumption for the second time, we are attempting to write to a closed file,
-    and then close it once again. This captures the excessive generality of
-    effects, and is the issue that we would like to address with coalgebraic
-    effects.
+    data, where finitary means of reasoning are of no use, as we can’t expect
+    an end to stream.
+
+    Operating on coinductive structures may involve modification of the internal
+    state of the machine that is generating the infinite streams, or in more
+    concrete scenario, alternation of the external resource that is providing
+    us the data.
+
+    For a better understanding and intuition, we shall illustrate now the
+    difference based on inductive and coinductive relation. Recall that the set
+    $\Lambda$ of $\mathVar{\lambda-terms}$ is given by the following grammar:
+
+    \begin{center}
+        $ e ::= x \;|\; \lambda x.e \;|\; e_{1} e_{2} $
+    \end{center}
+
+    \noindent
+    where $\Lambda^{0} \in \Lambda$ is a set of $\mathVar{closed~\lambda-terms}$,
+    meaning, those terms have no free variables. Let's now define an inductive,
+    and coinductive predicate. Relation
+    $\Downarrow{} \subseteq \Lambda^{0} \times \Lambda^{0}$ (convergence) for
+    call-by-value $\mathVar{\lambda-calculus}$ is defined as follows:
+
+    \begin{center}
+        \AxiomC{}
+        \UnaryInfC{$ \lambda x.e \Downarrow{} \lambda x.e $}
+        \DisplayProof{}
+        \quad\quad
+        \AxiomC{$ e_{1} \Downarrow{} \lambda x.e_{0}$}
+        \AxiomC{$e_{0}[e_{2}/x] \Downarrow{} e'$}
+        \BinaryInfC{$ e_{1} e_{2} \Downarrow{} e' $}
+        \DisplayProof{}
+    \end{center}
+
+    \noindent
+    which means that $ \Downarrow{} $ is the \textit{smallest} predicate
+    \textit{closed forward} under the above rules. Relation can be inductively
+    built from empty set and then in a fixed-point manner we can add more terms.
+    Let's now see a relation $ \Uparrow{} \subseteq \Lambda^{0}$ of
+    \textit{divergent} terms, which is coinductively defined as follows:
+
+    \begin{center}
+        \AxiomC{$ e_{1} \Uparrow{} $}
+        \UnaryInfC{$ e_{1} e_{2} \Uparrow{} $}
+        \DisplayProof{}
+        \quad\quad
+        \AxiomC{$ e_{1} \Downarrow{} \lambda x.e_{0}$}
+        \AxiomC{$e_{0}[e_{2}/x] \Uparrow{}$}
+        \BinaryInfC{$ e_{1} e_{2} \Uparrow{}$}
+        \DisplayProof{}
+    \end{center}
+
+    \noindent
+    which means that $\Uparrow$ is the \textit{greatest} predicate
+    \textit{closed backwards} under the above rules, relation can be coinductively
+    obtained by starting with $\Lambda_{0}$, and then by removing elements that
+    do not fit the rules.
+
+    \noindent
+    For more detailed introduction into coinduction and other examples,
+    see~\cite{jacobs-rutten, sangiorgi-intro}.
+
+    \noindent
+    TODO maybe also write coinductive proof based on two relations above?
 
@@
 construct in programming language
+
+\noindent
+TODO muuuch longer explanation of this topic
 
@@
     \end{verbatim}
-    where in the \verb!<PLACEHOLDER>! we can define on what to do with the
+    where in the \verb!<PLACEHOLDER>! we can define what to do with the
     computation. For example, min-max strategy for picking the minimum value:
@@
 
+    \noindent
     TODO subset of this section's content is explained already
@@
     from paper. More details can be found in Section~\ref{chapter:implementation}.
-    All commands are available within \verb!src!  directory.
+    All commands are available within the \verb!src! directory.
 
@@
 
+    \noindent
     TODO Extend language with basic constructs for usability
@@
 
+    \noindent
     TODO Rewrite here operational semantics from Links paper
@@
 
-    The Freak implementation is available at \href{https://github.com/Tomatosoup97/freak}{\underline{https://github.com/Tomatosoup97/freak}},
-    written purely in Haskell. While the paper provided a good overview of the
-    language and the translation, the lower-level details were omitted. That
-    being said, two inherently different takes at the implementations were made.
-    The first one is based on curried translation and A. Appel~\cite{appel-continuations}
+    The Freak implementation is available on github~\cite{freak}, written purely
+    in Haskell. While the paper provided a good overview of the language and the
+    translation, the lower-level details were omitted. That being said, two
+    inherently different takes at implementations were made. The first one
+    is based on curried translation and A. Appel~\cite{appel-continuations}
     book, and the second one directly on the uncurried translation to target
@@
 
+    \noindent
     TODO Drop of continuations is still an issue
@@
 
+    \noindent
     TODO Simple implementation for cohandlers

### 2020-08-08 thesis/thesis.tex
[thesis] Clarify two things in universal algebra
Support for strings in Freak

@@
     by a signature $\Sigma_{T}$ and a collection $\mathcal{E}_{T}$ of
-    $\mathVar{\Sigma_{T}-equations}$.
+    $\mathVar{\Sigma_{T}-equations}$. We might omit $T$ subscripts in case
+    our theory of interest is obvious.
 
@@
 
-    We have defined interpretation of an operation as a map:
+    We have defined interpretation of an operation as a morphism:
+
+    \begin{center}
+        $ {\llbracket op \rrbracket}_M : P \times {|I|}^{A} \rightarrow |I| $
+    \end{center}
+
+    \noindent
+    following now straight from Yoneda Lemma we have:
 
@@
     The syntax for the calculus is shown below. $nat \; n$ represents an integer $n$,
-    $V \oplus W$ and $V \approx W$ are respectively binary and relational operators,
-    where we support basic arithmetic and comparison operations.
-    \textbf{if} $V$ \textbf{then} $M$ \textbf{else} $N$ is a standard branching
-    statement. The other constructs are just as in Links, with slight syntax
-    modifications. Actual programs in Freak can be found in Section~\ref{sec:examples}.
-
+    `string' a string value, $V \oplus W$ and $V \approx W$ are respectively
+    binary and relational operators, where we support basic arithmetic and
+    comparison operations. \textbf{if} $V$ \textbf{then} $M$ \textbf{else} $N$
+    is a standard branching statement. The other constructs are just as in Links,
+    with slight syntax modifications. Actual programs in Freak can be found in
+    Section~\ref{sec:examples}.
+
+    % TODO proper alignment
     \begin{grammar}
 
-        <Values V, W> $::=$ $ x $ | $nat \; n$ \\
+        <Values V, W> $::=$ $ x $ \\
+            | $nat \; n$ | `string'  \\
             | $ \backslash x : A \rightarrow M $ | \textbf{rec} $ g \; x \rightarrow M $\\

### 2020-08-08 src/AST.hs

@@ -11,6 +11,7 @@ type RecordRow a = Map.Map Label a -- todo: presence variables?
 data Value
     = VVar Var
     | VNum Integer
+    | VStr String
     | VLambda Var ValueType Comp
     | VFix Var Var Comp
     | VUnit

### 2020-08-08 src/CPSLists.hs

@@ -39,6 +39,7 @@ cpsVal :: Value -> [Cont] -> CPSMonad UValue
 cpsVal e ks = case e of
     VVar x -> return $ UVar x
     VNum n -> return $ UNum n
+    VStr s -> return $ UStr s  -- todo: write about support for strings
     VRecordRow row -> undefined -- todo: records are constructed using ExtendRow
     VUnit -> return UUnit
     VPair e1 e2 -> do

### 2020-08-08 src/CommonEval.hs

@@ -35,6 +35,7 @@ convOp BNe = boolToInt . uncurry (/=)
 
 data DValue
     = DNum Integer
+    | DStr String
     | DLambda FuncRecord
     | DUnit
     | DPair DValue DValue
@@ -42,6 +43,7 @@ data DValue
 
 instance Show DValue where
     show (DNum n) = show n
+    show (DStr s) = "\"" ++ show s ++ "\""
     show (DLambda _) = "lambda"
     show  DUnit = "()"
     show (DPair l r) = "(" ++ show l ++ ", " ++ show r ++ ")"
@@ -49,6 +51,7 @@ instance Show DValue where
 
 instance Eq DValue where
     DNum n == DNum n' = n == n'
+    DStr s == DStr s' = s == s'
     DUnit == DUnit = True
     DPair a b == DPair a' b' = a == a' && b == b'
     DLabel l == DLabel l' = l == l'

### 2020-08-08 src/EvalTarget.hs

@@ -40,6 +40,7 @@ eval env c = case c of
         Just v -> return v
         Nothing -> unboundVarErr x
     UVal (UNum n) -> return (DNum n)
+    UVal (UStr s) -> return (DStr s)
     UVal (ULambda x c) -> return $ DLambda funcRecord
         where funcRecord [xVal] = let env' = extendEnv env x xVal in eval env' c
     UVal UUnit -> return DUnit

### 2020-08-08 src/Parser.hs

@@ -41,12 +41,13 @@ languageDef =
                                      ]
            , Token.reservedOpNames = [
                 "+", "*", "<-", "->", "\\", ":", "(", ")", "|",
-                "<", ">", ">=", "<=", "==", "!=", ","
+                "<", ">", ">=", "<=", "==", "!=", ",", "\""
             ]
            }
 
 lexer = Token.makeTokenParser languageDef
 
+stringL = Token.identifier lexer -- TODO parses a string
 identifier = Token.identifier lexer -- parses an identifier
 reserved   = Token.reserved   lexer -- parses a reserved name
 reservedOp = Token.reservedOp lexer -- parses an operator
@@ -286,6 +287,14 @@ term :: Parser Value
 term =  parens value
     <|> fmap VVar identifier
     <|> fmap VNum integer
+    <|> stringTerm
+
+stringTerm :: Parser Value
+stringTerm = do
+    reservedOp "\""
+    str <- stringL
+    reservedOp "\""
+    return $ VStr str
 
 -- Parser runners
 --

### 2020-08-08 src/TargetAST.hs

@@ -7,6 +7,7 @@ import AST
 data UValue
     = UVar Var
     | UNum Integer
+    | UStr String
     | UBool Bool
     | ULambda Var UComp
     | UUnit
@@ -32,6 +33,7 @@ parens s = "(" ++ s ++ ")"
 instance Show UValue where
     show (UBinOp op vL vR) = parens $ show vL ++ show op ++ show vR
     show (UNum n) = show n
+    show (UStr s) = "\"" ++ show s ++ "\""
     show  UUnit = "()"
     show (UPair l r) = parens $ show l ++ ", " ++ show r
     show (ULabel l) = parens $ "L: " ++ show l

### 2020-08-08 src/Tests.hs

@@ -31,6 +31,12 @@ main = hspec $ do
       evalProgram "return 5 / 2" `shouldBe` (Right (DNum 2))
       evalProgram "return 6 / 2" `shouldBe` (Right (DNum 3))
 
+    it "Strings" $ do
+      evalProgram "return \"value\"" `shouldBe` (Right (DStr "value"))
+      evalProgram "return \"x\"" `shouldBe` (Right (DStr "x"))
+      -- evalProgram "return \"this is string\"" `shouldBe` (Right (DStr "this is string"))
+      evalProgram "return \"x1\"" `shouldBe` (Right (DStr "x1"))
+
     it "Order of operations" $ do
       evalProgram "return 2 + 3 * 2" `shouldBe` (Right (DNum 8))
       evalProgram "return 2 * 3 + 2" `shouldBe` (Right (DNum 8))

### 2020-08-08 src/CommonEval.hs
Show value passed around in absurd term

@@ -15,8 +15,8 @@ extendEnv env x v = Map.insert x v env
 unboundVarErr :: String -> Either Error DValue
 unboundVarErr x = Left $ EvalError $ "Unbound variable " ++ x
 
-absurdErr :: Either Error DValue
-absurdErr = Left $ EvalError "Absurd; divergent term"
+absurdErr :: UValue -> Either Error DValue
+absurdErr v = Left $ EvalError $ "Absurd; divergent term: " ++ show v
 
 boolToInt :: Bool -> Integer
 boolToInt b = if b then 1 else 0

### 2020-08-08 src/EvalTarget.hs

@@ -26,7 +26,7 @@ eval env c = case c of
         if l == DLabel caseLabel
         then eval env tComp
         else eval (Map.insert y l env) fComp
-    UAbsurd _ -> absurdErr
+    UAbsurd v -> absurdErr v
     ULet x varComp comp -> do
         varVal <- eval env varComp
         let env' = extendEnv env x varVal

### 2020-08-08 src/Tests.hs

@@ -8,6 +8,7 @@ import Freak
 import Eval
 import CommonEval
 import Types
+import TargetAST
 
 genRow l n r = DPair (DLabel l) (DPair (DNum n) r)
 
@@ -46,8 +47,8 @@ main = hspec $ do
       evalProgram "return x" `shouldBe` (unboundVarErr "x")
 
     it "Absurd" $ do
-      evalProgram "absurd 42" `shouldBe` absurdErr
-      evalProgram "let x <- absurd 17 in return 42" `shouldBe` absurdErr
+      evalProgram "absurd 42" `shouldBe` absurdErr (UNum 42)
+      evalProgram "let x <- absurd 17 in return 42" `shouldBe` absurdErr (UNum 17)
 
     it "Let expression" $ do
       evalProgram "let x <- return 3 in return x + 2" `shouldBe` (Right (DNum 5))
@@ -151,7 +152,7 @@ main = hspec $ do
       testFromFile "programs/deepHandlers.fk" (Right (DNum 7))
 
     it "Unhandled effect" $ do
-      testFromFile "programs/unhandledEffect.fk" (absurdErr)
+      testFromFile "programs/unhandledEffect.fk" (absurdErr (ULabel "Effect"))
 
     -- it "Drop resumption result" $ do
     --   testFromFile "programs/dropResumption.fk" (Right (DNum 1))

### 2020-08-09 src/EvalTarget.hs
[src] Support file handling

@@ -32,12 +32,28 @@ eval env c = case c of
     UTopLevelEffect l v -> do
         dVal <- eval env (UVal v)
         let handle
+                -- TODO: Write down in latex this functionality
                 | l == "Print" = do
                     (lift . print) dVal
                     return DUnit
                 | l == "ReadLine" = do
                     input <- lift getLine
                     return $ DStr input
+                | l == "ReadFile" = case dVal of
+                    (DStr filename) -> do
+                        contents <- lift $ readFile filename
+                        return $ DStr contents
+                    _ -> throwError $ EvalError $ l ++ " effect accepts a single filename string"
+                | l == "WriteFile" = case dVal of
+                    (DPair (DStr filename) (DStr contents)) -> do
+                        lift $ writeFile filename contents
+                        return DUnit
+                    _ -> throwError $ EvalError $ l ++ " effect accepts a pair of (filename, contents) strings"
+                | l == "AppendFile" = case dVal of
+                    (DPair (DStr filename) (DStr contents)) -> do
+                        lift $ appendFile filename contents
+                        return DUnit
+                    _ -> throwError $ EvalError $ l ++ " effect accepts a pair of (filename, contents) strings"
                 | otherwise = throwError $ absurdErr (DPair (DLabel l) dVal)
         handle
     ULet x varComp comp -> do

### 2020-08-09 src/Tests.hs

@@ -189,3 +189,6 @@ main = hspec $ do
 
     it "Complex nested handlers" $ do
       testFromFile "programs/complexNestedHandlers.fk" (Right (DNum 15))
+
+    it "File handling" $ do
+      testFromFile "programs/writeToFile.fk" (Right (DStr "one-two-three"))

### 2020-08-09 src/CPSLists.hs
Minor refactoring in evaluator and cps translator

@@ -42,20 +42,20 @@ cpsVal e ks = case e of
     VVar x -> return $ UVar x
     VNum n -> return $ UNum n
     VStr s -> return $ UStr s  -- todo: write about support for strings
-    VRecordRow row -> undefined -- todo: records are constructed using ExtendRow
     VUnit -> return UUnit
     VPair e1 e2 -> do
         v1 <- cpsVal e1 ks
         v2 <- cpsVal e2 ks
         return $ UPair v1 v2
-    VExtendRow l v row -> notSupportedErr
-    VVariantRow (VariantRow _ l v) -> notSupportedErr
-    VLambda x _ body -> ULambda x <$> cps body ks
-    VFix g x body -> URec g x <$> cps body ks
     VBinOp op e1 e2 -> do
         v1 <- cpsVal e1 ks
         v2 <- cpsVal e2 ks
         return $ UBinOp op v1 v2
+    VLambda x _ body -> ULambda x <$> cps body ks
+    VFix g x body -> URec g x <$> cps body ks
+    VRecordRow row -> undefined -- todo: records are constructed using ExtendRow
+    VExtendRow l v row -> notSupportedErr
+    VVariantRow (VariantRow _ l v) -> notSupportedErr
 
 cps :: Comp -> [Cont] -> CPSMonad UComp
 cps e ks = case e of

### 2020-08-09 src/EvalTarget.hs

@@ -1,4 +1,4 @@
-{-# LANGUAGE FlexibleInstances #-}
+{-# LANGUAGE FlexibleContexts, FlexibleInstances #-}
 module Eval where
 
 import qualified Data.Map as Map
@@ -7,6 +7,34 @@ import AST
 import TargetAST
 import CommonEval
 import Types
+
+
+handleEffect :: Label -> DValue -> EvalMonad DValue
+handleEffect l v
+    -- TODO: Write down in latex this functionality
+    | l == "Print" = do
+        (lift . print) v
+        return DUnit
+    | l == "ReadLine" = do
+        input <- lift getLine
+        return $ DStr input
+    | l == "ReadFile" = case v of
+        (DStr filename) -> do
+            contents <- lift $ readFile filename
+            return $ DStr contents
+        _ -> throwError $ EvalError $ l ++ " effect accepts a single filename string"
+    | l == "WriteFile" = case v of
+        (DPair (DStr filename) (DStr contents)) -> do
+            lift $ writeFile filename contents
+            return DUnit
+        _ -> throwError $ EvalError $ l ++ " effect accepts a pair of (filename, contents) strings"
+    | l == "AppendFile" = case v of
+        (DPair (DStr filename) (DStr contents)) -> do
+            lift $ appendFile filename contents
+            return DUnit
+        _ -> throwError $ EvalError $ l ++ " effect accepts a pair of (filename, contents) strings"
+    | otherwise = throwError $ absurdErr (DPair (DLabel l) v)
+
 
 eval :: Env -> UComp -> EvalMonad DValue
 eval env c = case c of
@@ -28,38 +56,13 @@ eval env c = case c of
         if l == DLabel caseLabel
         then eval env tComp
         else eval (Map.insert y l env) fComp
-    UAbsurd v -> throwError $ absurdErr v
-    UTopLevelEffect l v -> do
-        dVal <- eval env (UVal v)
-        let handle
-                -- TODO: Write down in latex this functionality
-                | l == "Print" = do
-                    (lift . print) dVal
-                    return DUnit
-                | l == "ReadLine" = do
-                    input <- lift getLine
-                    return $ DStr input
-                | l == "ReadFile" = case dVal of
-                    (DStr filename) -> do
-                        contents <- lift $ readFile filename
-                        return $ DStr contents
[…]

### 2020-08-09 src/AST.hs
Parser for coeffects

@@ -59,6 +59,9 @@ data Comp
     -- Algebraic effects
     | EDo Label Value
     | EHandle Comp Handler
+    -- Coalgebraic effects
+    | ECoDo Label Value
+    | ECoHandle Comp Handler
     deriving (Show, Eq)
 
 data Handler

### 2020-08-09 src/CPSLists.hs

@@ -1,5 +1,24 @@
 module CPS where
 
+{- To Do Plan for implementation
+
+Coeffects:
+1. Cohandlers as a separate construct
+2. Disallow mixing effects with coeffects
+3. Define reusable handler as a variable
+4. Define default effects handler as a citizen in language
+5. Consider doing cohandlers as algebraic effects for easy logic substitution
+6. Consider one-shot tail-recursive handlers
+7. (Probably too hard and time consuming) Consider data-flow analysis for
+detecting whether cont was called once. Issue: data-flow analysis for effects?
+
+Bugs:
+1. Drop continuation result does not work
+
+Features:
+1. Syntactic sugar for function definitions
+-}
+
 import qualified Data.Map as Map
 import Control.Monad.Except
 import Control.Monad.State

### 2020-08-09 src/Parser.hs

@@ -38,6 +38,9 @@ languageDef =
                                      , "do"
                                      , "handle"
                                      , "with"
+                                     , "observe"
+                                     , "cohandle"
+                                     , "through"
                                      ]
            , Token.reservedOpNames = [
                 "+", "*", "<-", "->", "\\", ":", "(", ")", "|",
@@ -69,7 +72,9 @@ computation =  letBasedComp
            <|> absurdComp
            <|> retComp
            <|> doComp
+           <|> coDoComp
            <|> handleComp
+           <|> cohandleComp
            <|> appComp
 
 
@@ -149,12 +154,25 @@ doComp = do
     EDo <$> identifier <*> value
 
 
+coDoComp :: Parser Comp
+coDoComp = do
+    reserved "observe"
+    ECoDo <$> identifier <*> value
+
+
 handleComp :: Parser Comp
 handleComp = do
     reserved "handle"
     c <- computation
     reserved "with"
     EHandle c <$> handler
+
+cohandleComp :: Parser Comp
+cohandleComp = do
+    reserved "cohandle"
+    c <- computation
+    reserved "through"
+    ECoHandle c <$> handler
 
 
 handler :: Parser Handler

### 2020-08-09 thesis/thesis.tex
Rename Do and CoDo to Op and Coop

@@
     | EIf Value Comp Comp
-    | EDo Label Value
+    | EOp Label Value
     | EHandle Comp Handler

### 2020-08-09 src/AST.hs

@@ -57,11 +57,11 @@ data Comp
     | EAbsurd Value
     | EIf Value Comp Comp
     -- Algebraic effects
-    | EDo Label Value
+    | EOp Label Value
     | EHandle Comp Handler
     -- Coalgebraic effects
-    | ECoDo Label Value
-    | ECoHandle Comp Handler
+    | ECoop Label Value
+    | ECohandle Comp Handler
     deriving (Show, Eq)
 
 data Handler

### 2020-08-09 src/CPSLists.hs

@@ -117,7 +117,7 @@ cps e ks = case e of
         v <- cpsVal v ks
         return $ UAbsurd v
     -- Algebraic effects
-    EDo l v -> do
+    EOp l v -> do
         let Pure kf:h@(Eff hf):ks' = ks
         x <- freshVar' "rArg"
         pureComp <- kf (UVar x) (h:ks')

### 2020-08-09 src/Parser.hs

@@ -71,8 +71,8 @@ computation =  letBasedComp
            <|> ifComp
            <|> absurdComp
            <|> retComp
-           <|> doComp
-           <|> coDoComp
+           <|> opComp
+           <|> coopComp
            <|> handleComp
            <|> cohandleComp
            <|> appComp
@@ -148,16 +148,16 @@ retComp = reserved "return"
         >> EReturn <$> value
 
 
-doComp :: Parser Comp
-doComp = do
+opComp :: Parser Comp
+opComp = do
     reserved "do"
-    EDo <$> identifier <*> value
+    EOp <$> identifier <*> value
 
 
-coDoComp :: Parser Comp
-coDoComp = do
+coopComp :: Parser Comp
+coopComp = do
     reserved "observe"
-    ECoDo <$> identifier <*> value
+    ECoop <$> identifier <*> value
 
 
 handleComp :: Parser Comp
@@ -172,7 +172,7 @@ cohandleComp = do
     reserved "cohandle"
     c <- computation
     reserved "through"
-    ECoHandle c <$> handler
+    ECohandle c <$> handler
 
 
 handler :: Parser Handler

### 2020-08-20 thesis/thesis.tex
[thesis] Writing...

@@
     \noindent
-    where $\Lambda^{0} \in \Lambda$ is a set of $\mathVar{closed~\lambda-terms}$,
+    where $\Lambda^{0} \subseteq \Lambda$ is a set of $\mathVar{closed~\lambda-terms}$,
     meaning, those terms have no free variables. Let's now define an inductive,
@@
     The Freak implementation is available on github~\cite{freak}, written purely
-    in Haskell. While the paper provided a good overview of the language and the
-    translation, the lower-level details were omitted. That being said, two
-    inherently different takes at implementations were made. The first one
-    is based on curried translation and A. Appel~\cite{appel-continuations}
-    book, and the second one directly on the uncurried translation to target
-    calculus with continuations represented as explicit stacks from the paper.
-    We start by presenting core data structures, and afterwards move to actual
-    translation details.
+    in Haskell.Two inherently different takes at implementations were made. The
+    first one is based on curried translation from A. Appel~\cite{appel-continuations}
+    book, and the second one on Links language~\cite{handlers-cps, handlers-cps-journal},
+    on the uncurried translation to target calculus with continuations represented
+    as explicit stacks. We start by presenting core data structures, and
+    afterwards move to actual translation details. Development of the former
+    version is now suspended.
 
@@
     | VNum Integer
+    | VStr String
     | VLambda Var ValueType Comp
@@
     | EIf Value Comp Comp
+    -- Algebraic effects
     | EOp Label Value
     | EHandle Comp Handler
-    \end{verbatim}
+    -- Coalgebraic effects
+    | ECoop Label Value
+    | ECohandle Comp Handler
+    \end{verbatim}
+
+    \noindent
     Similarly for the target calculus data structure. However, as one may notice,
@@
     | UNum Integer
+    | UStr String
     | UBool Bool
@@
     | UAbsurd UValue
-
-    \end{verbatim}
+    | UTopLevelEffect Label UValue
+
+    \end{verbatim}
+
+    \noindent
     The final answer, common to both evaluations, is represented as a \verb!DValue!,
@@
 type Label = String
+
 type FuncRecord = [DValue] -> Either Error DValue
+
 data DValue
     = DNum Integer
+    | DStr String
     | DLambda FuncRecord
@@
 
-    \section{Target calculus}
+    \noindent
+    Evaluation of target calculus is typical to call-by-value with syntactic
+    distinction between values and computations. For more details, refer
+    to~\cite{handlers-cps-journal}. That being said, there is one part
+    that requires more attention, namely, \verb!UTopLevelEffect Label UValue!,
+    which represents an unhandled algebraic effect along with it's parameter.
+    Even in case language has coalgebraic effects, special treatment is required
+    for doing IO\@. For that reason, we define default handling for a few effects
+    for printing to console and handling files, where semantic is just as one
+    would expect from naming:
+
+    \begin{verbatim}
+    type Filename = String
+    Print :: String -> ()
+    ReadLine :: () -> String
+    ReadFile :: Filename -> String
+    WriteFile :: (Filename, String) -> ()
+    AppendFile :: (Filename, String) -> ()
+    \end{verbatim}
 
@@
     \begin{verbatim}
-type CPSMonad a = ExceptT Error (State Int) a
+type CPSMonad a = ExceptT Error (StateT Int IO) a
 
@@
           | Eff ContF
-    \end{verbatim}
-    Where \verb!CPSMonad! is a monad transformer over \verb!Either! and \verb!State!.
-    \verb!State! was required to generate labels for fresh variables that
-    came from the translation. The core code is split into five functions:
+          | Coeff ContF
+
+initialPureCont :: ContF
+initialPureCont v ks = (return . UVal) v
+
+initialEffCont :: ContF
+initialEffCont (UPair (ULabel effLabel) (UPair p r)) ks =
+    return . UApp (UVal r) $ UTopLevelEffect effLabel p
+
+    \end{verbatim}
+    Where \verb!CPSMonad! is a monad transformer over \verb!Either!, \verb!State!
+    and \verb!IO!. \verb!State! is required to generate labels for fresh variables
+    that came from the translation, \verb!Either! for handling exceptions that
+    may occur during translations, and \verb!IO! for handling input output of the
+    source language. Initial pure continuation is lifting values into computations,
+    and effectful one is propagating previously mentioned \verb!UTopLevelEffect!.
+    The core code is split into a few functions:
 
@@
 cpsVal  :: Value   -> [Cont] -> CPSMonad UValue
+cpsOp   :: Label -> Value -> [Cont] -> CPSMonad UComp
 cpsHRet :: Handler -> Cont
-cpsHOps :: Handler -> Cont
+cpsHOps :: (ContF -> Cont) -> Handler -> Cont
 forward :: Label   -> UValue -> UValue -> [Cont] -> CPSMonad UComp
-    \end{verbatim}
+runCPS  :: Comp -> EvalResMonad UComp
+    \end{verbatim}
+
+    \noindent
     Where the first two are implementing cps for computations and values.
+    \verb!cpsOp! is doing translation for an (co) effect that occured in program,
     \verb!cpsHRet! and \verb!cpsHOps! are yielding pure and effectful continuations,
-    based on a given handler. The last one is responsible for forwarding the
-    computation to the outer handler.
-
-    This results in an implementation that finally supports nested handlers,
-    as can be seen by evaluating \verb!programs/complexNestedHandlers.fk! program.
-    Unfortunately, following closely translation from the paper resulted in
-    a behaviour, in which invoked resumption forgets its pure continuation. This means,
-    that the following code, evaluating correctly on the Appel-based translation,
-    returns 0 rather than 1:
-
-    \begin{verbatim}
-handle do Drop () with { Drop p r -> let t <- r 0 in return 1}
-    \end{verbatim}
-    Nevertheless, working out this issue appears as less demanding than coping with
-    discrepancies created in the first translation. The source code for uncurried
-    translation and evaluation can be found respectively in \verb!CPSLists.hs! and
-    \verb!EvalTarget.hs!.
-
-    \noindent
-    TODO Drop of continuations is still an issue
+    based on a given handler. \verb!forward! is responsible for forwarding the
+    computation to the outer handler, and the last one for running the
+    translation, where evaluation result is represented by\\
+    \verb!type EvalResMonad a = IO (Either Error a)!. This results in an
+    implementation that finally supports nested handlers, which can be seen by
+    running tests.
 
@@
     \ldots
-
-    \noindent
-    TODO Simple implementation for cohandlers
 
@@
     CommonEval.hs   - Common functions for evaluation
-    CPSAppel.hs     - Appel-based CPS translation
     CPSLists.hs     - Uncurried CPS translation
-    EvalCPS.hs      - Evaluation of the Appel's CPS structure
     EvalTarget.hs   - Evaluation of the target calculus
@@
     Types.hs        - Common types definition
-    programs/       - Exemplary programs used in tests
+    programs/       - Exemplary programs covered in tests
+
+    CPSAppel.hs     - [deprecated] Appel-based CPS translation
+    EvalCPS.hs      - [deprecated] Evaluation of the Appel's CPS structure
     \end{verbatim}
@@
 
-    The Links language also provides small-step operational semantics and
-    an abstract machine~\cite{liberating-effects}. Implementing another way
-    of evaluation could serve as a way to empirically assert correctness,
-    as opposed to formally.
+    The Links language, on which CPS translation Freak was based, also provides
+    small-step operational semantics~\cite{handlers-cps} and an abstract
+    machine~\cite{liberating-effects}. Implementing another way of evaluation
+    could serve as a way to empirically assert correctness, as opposed to formally.
+    Abstract machine would need to be extended to incorporate coeffects.
 
@@
     The type system as of this day is not implemented, as the focus has been put
-    on CPS transformation. Further work is required here, especially considering
-    the fact that a huge advantage of algebraic effects is that they are explicitly
-    defined in the type of a computation.
+    on CPS transformation and dynamics of the calculus with effects and coeffects.
+    Further work is required here, especially considering the fact that a huge
+    advantage of algebraic effects is that they are explicitly defined in the
+    type of a computation.
 
@@
     art introduces a concept of resources and instances, as in Eff~\cite{programming-in-eff},
-    or instance variables, as in Helium~\cite{binders-labels}.
+    or instance variables, as in Helium~\cite{binders-labels}. Another part is
+    to research on how this area combines with coalgebraic part.
 
@@
 
-    Other languages, like Koka~\cite{leijen-koka}, or even the core of the Links, are
-    performing selective CPS translation, which reduces the overhead on code
+    Other languages, like Koka~\cite{leijen-koka}, or even the core of the Links,
+    are performing selective CPS translation, which reduces the overhead on code
     that does not perform algebraic effects. Our current translation is fully
-    embedded in the CPS.\@
-
-    \subsection{Exceptions as separate constructs}
+    embedded in CPS transformation.
+
+    \subsection{Exceptions and signals}
 

### 2020-08-20 thesis/thesis.tex
[thesis] Improved verbatim presentation

@@
 \usepackage{tikz-cd}
+\usepackage{fancyvrb}
 \usetikzlibrary{babel}
@@
 
-\begin{verbatim}
+\vbox{\begin{verbatim}
     handle
@@
         let c <- do Choice () in
-    if c then do Write (fh, "Guy Fieri") else do Write (fh, "is cool") ;
+        if c then do Write (fh, "Guy Fieri") else do Write (fh, "is cool");
         close fh
@@
 \end{verbatim}
+}
 
@@
     described in great details. More exemplary programs in Freak language can
-    be found under \\ \href{https://github.com/Tomatosoup97/freak/tree/master/src/programs}{\underline{https://github.com/Tomatosoup97/freak/tree/master/src/programs}}.
+    be found in \verb!src/programs! directory~\cite{freak}.
 
@@
     \noindent
-    TODO Rewrite here operational semantics from Links paper
-
-    \begin{flushleft}
     Extension of the evaluation contexts:
-    \end{flushleft}
-
-    \begin{flushleft}
+
+    \noindent
     $\mathcal{E} ::= \mathcal{E} \oplus W \; | \; nat \; n \oplus \mathcal{E} \; |$ \textbf{if} $\mathcal{E}$ \textbf{then} $M$ \textbf{else} $N$
-    \end{flushleft}
-
-    \begin{flushleft}
+
+    \noindent
     Small-step operational semantics:
-    \end{flushleft}
-
-    \begin{flushleft}
+
+    \noindent
     \textbf{if} $nat \; n$ \textbf{then} $M$ \textbf{else} $N \rightsquigarrow M \quad \quad $ if $n \neq 0$ \\
     \textbf{if} $nat \; n$ \textbf{then} $M$ \textbf{else} $N \rightsquigarrow N \quad \quad $ if $n = 0$
-    \end{flushleft}
-
-    \begin{flushleft}
+
+    \noindent
     $nat \; n \oplus nat \; n' \rightsquigarrow n'' \quad \quad $    if $ n'' = n \oplus n' $ \\
@@
 
-    \end{flushleft}
-
 \section{Continuation Passing Style Transformation}
@@
     The Freak implementation is available on github~\cite{freak}, written purely
-    in Haskell.Two inherently different takes at implementations were made. The
+    in Haskell. Two inherently different takes at implementations were made. The
     first one is based on curried translation from A. Appel~\cite{appel-continuations}
@@
         | UTopLevelEffect Label UValue
-
 \end{verbatim}
@@
         return . UApp (UVal r) $ UTopLevelEffect effLabel p
-
 \end{verbatim}

### 2020-08-20 src/CPSLists.hs
Extract conts stack to a variable

@@ -49,6 +49,9 @@ initialEffCont (UPair (ULabel effLabel) (UPair p r)) ks =
     return . UApp (UVal r) $ UTopLevelEffect effLabel p
 initialEffCont v _ = throwError $ CPSError $ "Incorrect value " ++ show v ++ " in effect continuation"
 
+initialContStack :: [Cont]
+initialContStack = [Pure initialPureCont, Eff initialEffCont]
+
 initialState :: Int
 initialState = 0
 
@@ -187,4 +190,4 @@ forward y p r ks = do
 
 runCPS :: Comp -> EvalResMonad UComp
 runCPS e = evalStateT (runExceptT cpsTerm) initialState
-    where cpsTerm = cps e [Pure initialPureCont, Eff initialEffCont]
+    where cpsTerm = cps e initialContStack

### 2020-08-20 thesis/thesis.tex
[thesis] Correct outdated sections

@@
 
-    \noindent
-    TODO subset of this section's content is explained already
-
     These implications are straightforward when looking from a categorical standpoint,
@@
     and handlers are homomorphisms preserving the model structure~\cite{handlers}.
-    Nevertheless, the results are very exciting for programming use cases. The current
-    Freak implementation does not support I/O.
+    Nevertheless, the results are very exciting for programming use cases.
 
@@
 
-    Examples for cohandlers
+    TODO Examples for cohandlers
 
@@
     As of this day, two implementations are available, one based on the curried
-    translation and Appel~\cite{appel-continuations}, and the second one based
-    directly on the uncurried translation with continuations as explicit stacks
-    from paper. More details can be found in Section~\ref{chapter:implementation}.
-    All commands are available within the \verb!src! directory.
+    translation from Appel~\cite{appel-continuations}, and the second one based
+    directly on the uncurried translation with continuations as explicit
+    stacks~\cite{handlers-cps-journal}. More details can be found in
+    Section~\ref{chapter:implementation}. All commands are available within the
+    \verb!src! directory.
 
@@
     After compiling and linking program to PATH, one may evaluate program as
-    follows: \verb!freak programs/choicesList.fk!. The actual code is described in Section~\ref{sec:choice-example}
+    follows: \verb!freak programs/choicesList.fk! The actual code is described in Section~\ref{sec:choice-example}
 
@@
 
-    \noindent
-    TODO Extend language with basic constructs for usability
-
     The syntax for the calculus is shown below. $nat \; n$ represents an integer $n$,
@@
     comparison operations. \textbf{if} $V$ \textbf{then} $M$ \textbf{else} $N$
-    is a standard branching statement. The other constructs are just as in Links,
-    with slight syntax modifications. Actual programs in Freak can be found in
-    Section~\ref{sec:examples}.
+    is a standard branching statement. The other constructs are just as in
+    Links~\cite{handlers-cps-journal}, with slight syntax modifications.
+    Actual programs in Freak can be found in Section~\ref{sec:examples}.
+
+    TODO cohandlers
 
@@
 
-\section{Typing Rules}
-\section{Operational Semantics}
-
-    The source language's dynamics have been described
-    extensively by providing small-step operational semantics,
-    continuation passing style transformation~\cite{handlers-cps} as well
-    as abstract machine~\cite{liberating-effects}, which was proved to coincide
-    with CPS translation. That being said, Freak introduces new basic
-    constructs to the language, for which we shall define the semantics.
+\section{Dynamics}
+
+    Semantic of Freak is heavily based on Links langauge~\cite{handlers-cps-journal},
+    for which the source language's dynamics have been described extensively by
+    providing small-step operational semantics, continuation passing style
+    transformation~\cite{handlers-cps} as well as abstract
+    machine~\cite{liberating-effects}, which was proved to coincide with CPS
+    translation. That being said, Freak introduces new basic constructs to the
+    language, for which we shall define the semantics.
 
@@
 
-\section{Continuation Passing Style Transformation}
-    \ldots
+    TODO cohandlers
 
@@
     afterwards move to actual translation details. Development of the former
-    version is now suspended.
+    Appel's version is now suspended.
 
@@
     for printing to console and handling files, where semantic is just as one
-    would expect from naming:
+    would expect from naming. We abuse haskell's notation to define types for
+    algebraic effects living in Freak langauge:
 
@@
     evaluation can be found respectively in \verb!CPSAppel.hs! and
-    \verb!EvalCPS.hs!.
+    \verb!EvalCPS.hs!. Development of this version of translation is discontinued.
 
@@
     Having in mind the drawbacks mentioned above, alternative translation was
-    written, that coincides with the translation from the paper. Namely, with
-    the uncurried translation to target calculus with continuations represented
-    as explicit stacks. The target calculus was described in Section~\ref{sec:implementation-ast},
-    for which the evaluation is straightforward. The continuations are represented
-    as \verb!Cont!, with syntactic distinction between pure and effectful computations,
-    which occupy alternating positions in the stack. Explicit distinction gave
+    written, that coincides with the translation from~\cite{handlers-cps-journal}.
+    Namely, with the uncurried translation to target calculus with continuations
+    represented as explicit stacks. The target calculus was described in
+    Section~\ref{sec:implementation-ast}. The continuations are represented as
+    \verb!Cont!, with syntactic distinction between pure, effectful and
+    coeffectful computations. Pure and (co) effectful continuations occupy
+    alternating positions in the stack. Explicit distinction was made to provide
     more control in the source code.
@@
               | Coeff ContF
-
-    initialPureCont :: ContF
-    initialPureCont v ks = (return . UVal) v
-
-    initialEffCont :: ContF
-    initialEffCont (UPair (ULabel effLabel) (UPair p r)) ks =
-        return . UApp (UVal r) $ UTopLevelEffect effLabel p
-\end{verbatim}
+\end{verbatim}
+
+    \noindent
     Where \verb!CPSMonad! is a monad transformer over \verb!Either!, \verb!State!
@@
     may occur during translations, and \verb!IO! for handling input output of the
-    source language. Initial pure continuation is lifting values into computations,
+    source language.
+
+\begin{verbatim}
+    initialPureCont :: ContF
+    initialPureCont v ks = (return . UVal) v
+
+    initialEffCont :: ContF
+    initialEffCont (UPair (ULabel effLabel) (UPair p r)) ks =
+        return . UApp (UVal r) $ UTopLevelEffect effLabel p
+
+    initialContStack :: [Cont]
+    initialContStack = [Pure initialPureCont, Eff initialEffCont]
+\end{verbatim}
+
+    \noindent
+    Initial pure continuation is lifting values into computations,
     and effectful one is propagating previously mentioned \verb!UTopLevelEffect!.
-    The core code is split into a few functions:
+    Notice that \verb!initialContStack! does not have initial cohandler,
+    as handling of default coeffects is embed in evaluator. The core code is
+    split into a few functions:
 
@@
     implementation that finally supports nested handlers, which can be seen by
-    running tests.
+    running tests. Links language~\cite{handlers-cps-journal} also supports
+    shallow handlers~\cite{shallow-handlers}, whereas our langauge implements
+    only deep ones.
 

### 2020-08-20 thesis/thesis.tex
[thesis] Cite Power's comodels and Uustalu's runners

@@
 
-    We could end by stating this definition, and everything else would follow
-    directly from duality. However, expanding definitions is going to give us
-    better intuition, as well as give more solid ground for implementation.
-    For this part, we are going to assume we operate in \textbf{Set} category.
+    We could end by stating this definition of comodels~\cite{comodels}, and
+    everything else would follow directly from duality. However, expanding
+    definitions is going to give us better intuition, as well as give more solid
+    ground for implementation. For this part, we are going to assume we operate
+    in \textbf{Set} category.
 
@@
     implementation of a $\lambda_{coop}$ language~\cite{coop}, which is a
-    calculus that ensures linear usage of resources as well as execution of
+    calculus with runners of coalgebraic effects~\cite{runners-uustalu}
+    that ensures linear usage of resources as well as execution of
     code handling finalisation.
@@
 
-    What was achieved, what was described, what needs to be done or researched.
+    TODO What was achieved, what was described, what needs to be done or researched.
 

### 2020-08-21 src/AST.hs
Cohandlers with defined Algebraic Theory and initial coalgebra

@@ -8,6 +8,8 @@ data VariantRow a = VariantRow RowType Label a
 
 type RecordRow a = Map.Map Label a -- todo: presence variables?
 
+type AlgTheoryName = String
+
 data Value
     = VVar Var
     | VNum Integer
@@ -62,6 +64,8 @@ data Comp
     -- Coalgebraic effects
     | ECoop Label Value
     | ECohandle Comp Handler
+    -- Intermediate representation for a cohandler
+    | ECohandleIR AlgTheoryName Value Comp Handler
     deriving (Show, Eq)
 
 data Handler

### 2020-08-21 src/Freak.hs

@@ -9,10 +9,12 @@ import CPS
 import CommonEval
 import Eval
 import Types
+import Transform
+
 
 cpsProgram :: String -> EvalResMonad UComp
 cpsProgram s = case parseString s of
-    Right ast -> runCPS ast
+    Right ast -> runCPS $ transform ast
     Left err -> return $ Left err
 
 evalProgram :: String -> EvalResMonad DValue

### 2020-08-21 src/Parser.hs

@@ -41,6 +41,8 @@ languageDef =
                                      , "observe"
                                      , "cohandle"
                                      , "through"
+                                     , "using"
+                                     , "at"
                                      ]
            , Token.reservedOpNames = [
                 "+", "*", "<-", "->", "\\", ":", "(", ")", "|",
@@ -170,9 +172,13 @@ handleComp = do
 cohandleComp :: Parser Comp
 cohandleComp = do
     reserved "cohandle"
+    algTheoryName <- identifier
+    reserved "using"
+    v <- value
+    reserved "at"
     c <- computation
     reserved "through"
-    ECohandle c <$> handler
+    ECohandleIR algTheoryName v c <$> handler
 
 
 handler :: Parser Handler

### 2020-08-21 src/Transform.hs

@@ -1,0 +1,32 @@
+module Transform where
+
+import AST
+
+coopTransV :: Value -> Value
+coopTransV v = case v of
+    VLambda x t c -> VLambda x t (coopTrans c)
+    VFix g x c -> VFix g x (coopTrans c)
+    VPair v1 v2 -> VPair (coopTransV v1) (coopTransV v2)
+    VBinOp op v1 v2 -> VBinOp op (coopTransV v1) (coopTransV v2)
+    v -> v
+
+coopTransHandler :: Handler -> Handler
+coopTransHandler (HRet v c) = HRet v (coopTrans c)
+coopTransHandler (HOps (AlgOp l p r c) h) =
+    HOps (AlgOp l p r (coopTrans c)) (coopTransHandler h)
+
+coopTrans :: Comp -> Comp
+coopTrans c = case c of
+    EVal v -> EVal (coopTransV v)
+    ELet x varC bC -> ELet x (coopTrans varC) (coopTrans bC)
+    EApp v1 v2 -> EApp (coopTransV v1) (coopTransV v2)
+    EReturn v -> EReturn (coopTransV v)
+    EAbsurd v -> EAbsurd (coopTransV v)
+    EIf v tC fC -> EIf (coopTransV v) (coopTrans tC) (coopTrans fC)
+    EOp l v -> EOp l (coopTransV v)
+    EHandle c h -> EHandle (coopTrans c) (coopTransHandler h)
+    ECoop l v -> ECoop l (coopTransV v)
+    ECohandleIR algTheoryName initV c h -> ECohandle (coopTrans c) (coopTransHandler h)
+
+transform :: Comp -> Comp
+transform = coopTrans

### 2020-08-21 thesis/thesis.tex
[thesis] Cite Sangiorgi advanced book on bisim and coinduction

@@
         if c then do Write (fh, "Guy Fieri") else do Write (fh, "is cool");
-        close fh
+        do Close fh
     with {
         return x -> [x] |
-        choose () k -> return (append (k true) (k false)) |
+        Choose _ k -> return (append (k true) (k false)) |
     }
@@
     For more detailed introduction into coinduction and other examples,
-    see~\cite{jacobs-rutten, sangiorgi-intro}.
-
-    \noindent
-    TODO maybe also write coinductive proof based on two relations above?
+    see~\cite{jacobs-rutten, sangiorgi-intro} or for a deeper
+    dive~\cite{sangiorgi-advanced}.
 

### 2020-08-21 src/inspect.sh
[inspect.sh] Ensure env vars are set before evaluating

@@ -1,6 +1,13 @@
 #!/bin/zsh
 set -e
 
+if [ -z "$PROGRAM" ]
+then
+  echo "error: please set PROGRAM env variable with path to code"
+  exit 1
+fi
+
+
 echo "Program:"
 cat $PROGRAM
 echo "\nParser:"

### 2020-08-21 src/AST.hs
Pass down mapping from coeffects to alg theories in transform

@@ -9,6 +9,7 @@ data VariantRow a = VariantRow RowType Label a
 type RecordRow a = Map.Map Label a -- todo: presence variables?
 
 type AlgTheoryName = String
+type Signature = [Label]
 
 data Value
     = VVar Var
@@ -87,6 +88,12 @@ hops :: Handler -> [AlgebraicOp] -- todo: make it a set
 hops = aux []
     where aux acc (HRet _ _) = acc
           aux acc (HOps op h) = aux (op:acc) h
+
+opL :: AlgebraicOp -> Label
+opL (AlgOp l _ _ _) = l
+
+hopsL :: Handler -> Signature
+hopsL h = map opL (hops h)
 
 hop :: Label -> Handler -> Maybe AlgebraicOp
 hop _ (HRet _ _) = Nothing

### 2020-08-21 src/Freak.hs

@@ -39,5 +39,5 @@ parseFile :: String -> Bool -> IO ()
 parseFile filename quiet = do
     code <- readFile filename
     case parseString code of
-        Right r -> if quiet then return () else print r
+        Right r -> if quiet then return () else print $ transform r
         Left e  -> print e >> fail "Parse error"

### 2020-08-21 src/Transform.hs

@@ -1,32 +1,41 @@
 module Transform where
 
 import AST
+import Types
+import qualified Data.Map as Map
 
-coopTransV :: Value -> Value
-coopTransV v = case v of
-    VLambda x t c -> VLambda x t (coopTrans c)
-    VFix g x c -> VFix g x (coopTrans c)
-    VPair v1 v2 -> VPair (coopTransV v1) (coopTransV v2)
-    VBinOp op v1 v2 -> VBinOp op (coopTransV v1) (coopTransV v2)
+type AlgSignMap = Map.Map Label AlgTheoryName
+
+coopTransV :: AlgSignMap -> Value -> Value
+coopTransV m v = case v of
+    VLambda x t c -> VLambda x t (coopTrans m c)
+    VFix g x c -> VFix g x (coopTrans m c)
+    VPair v1 v2 -> VPair (coopTransV m v1) (coopTransV m v2)
+    VBinOp op v1 v2 -> VBinOp op (coopTransV m v1) (coopTransV m v2)
     v -> v
 
-coopTransHandler :: Handler -> Handler
-coopTransHandler (HRet v c) = HRet v (coopTrans c)
-coopTransHandler (HOps (AlgOp l p r c) h) =
-    HOps (AlgOp l p r (coopTrans c)) (coopTransHandler h)
+coopTransHandler :: AlgSignMap -> Handler -> Handler
+coopTransHandler m (HRet v c) = HRet v (coopTrans m c)
+coopTransHandler m (HOps (AlgOp l p r c) h) =
+    HOps (AlgOp l p r (coopTrans m c)) (coopTransHandler m h)
 
-coopTrans :: Comp -> Comp
-coopTrans c = case c of
-    EVal v -> EVal (coopTransV v)
-    ELet x varC bC -> ELet x (coopTrans varC) (coopTrans bC)
-    EApp v1 v2 -> EApp (coopTransV v1) (coopTransV v2)
-    EReturn v -> EReturn (coopTransV v)
-    EAbsurd v -> EAbsurd (coopTransV v)
-    EIf v tC fC -> EIf (coopTransV v) (coopTrans tC) (coopTrans fC)
-    EOp l v -> EOp l (coopTransV v)
-    EHandle c h -> EHandle (coopTrans c) (coopTransHandler h)
-    ECoop l v -> ECoop l (coopTransV v)
-    ECohandleIR algTheoryName initV c h -> ECohandle (coopTrans c) (coopTransHandler h)
[…]

### 2020-08-21 src/AST.hs
Implement product projections

@@ -23,6 +23,8 @@ data Value
     | VExtendRow Label Value Value
     | VVariantRow (VariantRow Value)
     | VBinOp BinaryOp Value Value
+    | VFst Value
+    | VSnd Value
     deriving (Show, Eq)
 
 data BinaryOp

### 2020-08-21 src/CPSLists.hs

@@ -74,6 +74,8 @@ cpsVal e ks = case e of
         v1 <- cpsVal e1 ks
         v2 <- cpsVal e2 ks
         return $ UPair v1 v2
+    VFst e -> UFst <$> cpsVal e ks
+    VSnd e -> USnd <$> cpsVal e ks
     VBinOp op e1 e2 -> do
         v1 <- cpsVal e1 ks
         v2 <- cpsVal e2 ks

### 2020-08-21 src/Tests.hs

@@ -81,6 +81,20 @@ main = hspec $ do
     it "Pair" $ do
       evalProgram ("return (1, 2)") `shouldBeT` (Right (DPair (DNum 1) (DNum 2)))
 
+    it "First projection" $ do
+      evalProgram ("return fst (1, 2)") `shouldBeT` (Right (DNum 1))
+      evalProgram ("return fst 1") `shouldBeT` (Left (EvalError "First projection on expression that is not a pair: 1"))
+
+    it "Second projection" $ do
+      evalProgram ("return snd (1, 2)") `shouldBeT` (Right (DNum 2))
+      evalProgram ("return snd 1") `shouldBeT` (Left (EvalError "Second projection on expression that is not a pair: 1"))
+
+    it "Projections" $ do
+      evalProgram ("return fst fst ((1, 2), (3, 4))") `shouldBeT` (Right (DNum 1))
+      evalProgram ("return snd fst ((1, 2), (3, 4))") `shouldBeT` (Right (DNum 2))
+      evalProgram ("return fst snd ((1, 2), (3, 4))") `shouldBeT` (Right (DNum 3))
+      evalProgram ("return snd snd ((1, 2), (3, 4))") `shouldBeT` (Right (DNum 4))
+
     it "Nested pair" $ do
       evalProgram ("return ((1, 2), 3)") `shouldBeT` (Right (DPair (DPair (DNum 1) (DNum 2)) (DNum 3)))
 

### 2020-08-22 src/Transform.hs
Skeleton for state-monad approach to coeffects state

@@ -8,12 +8,11 @@ type AlgSignMap = Map.Map Label AlgTheoryName
 
 getCoalgebra :: AlgTheoryName -> Comp -> Comp
 getCoalgebra algT = (ELet algT . EReturn . VVar) algT
+-- getCoalgebra algT = ELet algT (EOp (algT ++ "Get") VUnit)
 
-putCoalgebra :: AlgTheoryName -> Comp -> Comp
-putCoalgebra algT = (ELet algT . EReturn . VVar) algT
-
-initCoalgebra :: AlgTheoryName -> Value -> Comp -> Comp
-initCoalgebra algT = ELet algT . EReturn
+putCoalgebra :: AlgTheoryName -> Value -> Comp -> Comp
+putCoalgebra algT = ELet algT . EReturn
+-- putCoalgebra algT = ELet algT . EOp (algT ++ "Put")
 
 coopTransV :: AlgSignMap -> Value -> Value
 coopTransV m v = case v of
@@ -44,15 +43,16 @@ coopTrans m c = case c of
             let coop = ECoop l (VPair (VVar algT) (coopTransV m v)) in
             let bindResult = ELet algTRes coop in
             let contComp = EReturn (VSnd (VVar algTRes)) in
-            (getCoalgebra algT . bindResult . putCoalgebra algT) contComp
+            (getCoalgebra algT . bindResult . putCoalgebra algT (VVar algT)) contComp
         Nothing -> ECoop l (coopTransV m v)
     ECohandleIR algTheoryName initV c h ->
         let algTVar = "#" ++ algTheoryName in
         let sign = hopsL h in
         let m' = foldl (\m s -> Map.insert s algTVar m) m sign in
+        -- TODO: Add handling for #AlgTheoryPut and #AlgTheoryGet effects
         let h' = coopTransHandler m' h in
         let cohandle = (`ECohandle` h') in
-        (cohandle . initCoalgebra algTVar initV . coopTrans m') c
+        (cohandle . putCoalgebra algTVar initV . coopTrans m') c
 
 transform :: Comp -> Comp
 transform = coopTrans Map.empty

### 2020-08-22 thesis/thesis.tex
[thesis] Corrections to chapter one

@@
 % chktex-file 18
+% chktex-file 36
 
@@
 
-In this thesis we discuss issues with excessive generality of algebraic effects,
-and propose a solution without coalgebraic means along with presentation of
-experimental programming language Freak, which is a language with (co) algebraic
-effects and (co) handlers, where implementation is based on Continuation Passing
-Style translation.
+In this thesis we discuss issues arising from combining algebraic effects
+and handlers wiht resource management, and present a solution through coalgebraic
+means along with presentation of experimental programming language Freak, which
+is a language with (co)algebraic effects and (co)handlers, where implementation
+is based on Continuation Passing Style translation.
 
@@
 
-Algebraic effects, while not being a new concept, in the recent years have
-received a lot of attention~\cite{effects-bibliography}, both from the theoretical
+Algebraic effects, while not a new concept, have received a lot of
+attention~\cite{effects-bibliography} in recent years, both from the theoretical
 and practical side.
 
-They give developer necessary tools for declarative programming and writing
-programs in a way to strive from exposing implementation details, by defining
-computational effects as an API\@ in our code. In the same time, they preserve
-a lot of flexibility and have a strong theoretical background.
+They give the developer necessary tools for declarative programming with
+computational effects on a high level of abstraction, by defining
+effects as an API\@ in the code. At the same time, algebraic effects preserve
+a lot of flexibility and have a strong theoretical foundations.
 
 In fact, that very flexibility can be troublesome. Allowing developer for too
-much freedom may lead to undesired behaviour and incorrect programs, for this
-reason, we develop type systems, to detect errors before they occur.
-
-Multi-shot resumptions in effect handlers and possibility to drop resumption
-are powerful tools to describe fairly complex logic in a concise way. That being
-said, they give rise to issues that would not occur in standard control flow,
-especially when composing various effects together. Unwanted interaction with
-external resources multiple times or not invoking finalisation code are cases
-that may occur while using algebraic effects and handlers. Let's consider the
-following example:
+much freedom may lead to undesired behaviour and incorrect programs.
+
+In particular, multi-shot resumptions in effect handlers and the possibility to
+drop the resumption allow one to express fairly complex logic in a concise way,
+but at the same time, they give rise to issues that would not occur in standard
+control flow, especially when composing various effects together. Unwanted
+interaction with external resources multiple times or not invoking finalisation
+code are cases that may occur while using algebraic effects and handlers. Let's
+consider the following example:
 
@@
 
-In the following chapter we provide a background
-about effects, define algebraic effects in categorical setting as well as point
-out dual coalgebraic effects and cohandlers, finishing with showing related work
-in this area. Chapter~\ref{chapter:potential-solutions} describes possible ways
-to approach issue of excessive generality,
-Chapter~\ref{chapter:co-effectful-programming} shows Freak language by
-examples along with usage guide, and then
+In the following chapter we provide a background about effects, define algebraic
+effects in the categorical setting as well as point out dual coalgebraic effects
+and cohandlers, finishing with showing related work in this area.
+Chapter~\ref{chapter:potential-solutions} describes possible ways to approach
+the issue of excessive generality, Chapter~\ref{chapter:co-effectful-programming}
+shows the Freak language by examples along with a usage guide, and then
 Chapter~\ref{chapter:calculus-of-freak-language} describes the language's syntax,
-operational semantics and CPS translation. In next one,
+operational semantics and CPS translation. In the next one,
 Chapter~\ref{chapter:implementation}, implementation details are revealed.
-We conclude in Chapter~\ref{chapter:conclusion} by stating what are the possible
-augmentations, that are intended to be made in the future.
+We conclude in Chapter~\ref{chapter:conclusion} by stating what the possible
+augmentations are, which are intended to be made in the future.
 
@@
 
-    (co) recursive calls& shrinks the argument & grows the result     \\
+    (co)recursive calls& shrinks the argument & grows the result     \\
     \midrule
@@
 
-\chapter{(Co) Effectful Programming}\label{chapter:co-effectful-programming} % TODO w/o space?
+\chapter{(Co)Effectful Programming}\label{chapter:co-effectful-programming}
 \section{Examples}\label{sec:examples}
@@
     \verb!Cont!, with syntactic distinction between pure, effectful and
-    coeffectful computations. Pure and (co) effectful continuations occupy
+    coeffectful computations. Pure and (co)effectful continuations occupy
     alternating positions in the stack. Explicit distinction was made to provide
@@
     Where the first two are implementing cps for computations and values.
-    \verb!cpsOp! is doing translation for an (co) effect that occured in program,
+    \verb!cpsOp! is doing translation for an (co)effect that occured in program,
     \verb!cpsHRet! and \verb!cpsHOps! are yielding pure and effectful continuations,

### 2020-08-23 thesis/thesis.tex
[thesis] Cite Power and Plotkin Tensors of Comodels
+ TODO section to develop this further
Extend section about coalgebras

@@
     \begin{center}
-        $ {\llbracket op \rrbracket}_M : P \times {|I|}^{A} \rightarrow |I| $
-    \end{center}
-
-    \noindent
-    following now straight from Yoneda Lemma we have:
-
-    \begin{center}
-        $ {\llbracket op \rrbracket}_M : {|I|}^{A} \rightarrow |I|^{P} $
-    \end{center}
-
-    \noindent
-    Let's now take a look on how it dualizes. In \textbf{Set}, we can think about
-    $A^{B}$ exponentials as $\mathVar{B-ary}$ coproducts over A, where for each
-    argument we select one result. Every arrow in opposite category is reverted,
-    thus coproducts becomes products, our morphism is reverted and model becomes
-    comodel (\textit{world}). Putting this together, we obtain the following map:
+        {$\displaystyle {\llbracket op \rrbracket}_M : P \times {|I|}^{A} \rightarrow |I| $}
+    \end{center}
+
+    \noindent
+    following now straight from Yoneda Lemma, we have:
+
+    \begin{center}
+        {$\displaystyle {\llbracket op \rrbracket}_M : {|I|}^{A} \rightarrow |I|^{P} $}
+    \end{center}
+
+    Let's now take a look on how it dualizes. Since we operate in \textbf{Set},
+    we can represent $A^{B}$ exponentials as $\mathVar{B-ary}$ products over $A$,
+    where for each argument we select one result:
+
+    \begin{center}
+        % TODO: a in A under prod
+        {$\displaystyle {\llbracket op \rrbracket}_M : \prod_{a \in A}{|I|} \rightarrow \prod_{p \in P}{|I|} $}
+    \end{center}
+
+    % TODO: textbf Set^op
+    \noindent
+    We want now to embed this morphism in $Set^{op}$. Every arrow in opposite
+    category is reverted, thus, products become coproducts and our morphism
+    is reverted:
+
+    \begin{center}
+        {$\displaystyle {\llbracket op \rrbracket}_M : \coprod_{p \in P}{|I|} \rightarrow \coprod_{a \in A}{|I|} $}
+    \end{center}
+
+    \noindent
+    It turnes out, that this map can be further simplified.
+
+    \begin{lemma}
+        In \textbf{Set} category, the following isomorphism holds
+        {$\displaystyle \coprod_{b \in B}{A} \cong B \times A $}
+    \end{lemma}
+    \begin{proof}
+        We can think about $\mathVar{B-ary}$ coproducts (tagged unions) over the same set
+        such that we have $B$ replicas of $A$ and we select which replica do we want
+        to pick. Putting it this way, it immediately follows that it's the same,
+        up to isomorphism, as a cartesian product over indexes from $B$ and
+        corresponding $A$ sets. For a more detailed and generalized results,
+        see~\cite{tensors}.
+    \end{proof}
+
+    Model turns into comodel~\cite{comodels}, which we shall also call \textit{world}.
+    Putting this together with the above lemma, we obtain the following map:
 
@@
     based on coalgebra carrier and a parameter, we obtain new state of the
-    coalgebra and a value generated by coalgebra.
+    coalgebra and a value generated by coalgebra. This state-alike result that
+    we have obtained by dualizing models is indeed surprising, and it's
+    expected that reader may feel astonished after seeing it for the first time.
+    This observation was made in~\cite{tensors}.
 
@@
 
-\noindent
-TODO \\
-- coalg effects \\
-- cohandlers
+    We have now derived a dual concept to algebraic effects and handlers, which
+    in the literature are also called comodels and runners~\cite{runners-uustalu, runners-in-action}.
+    As we have seen, comodels turned into state-passing (co)operations, which
+    in programming langauges, may be a way to model interaction with external
+    resources, where based on current configuration and some parameter, we
+    obtain new configuration along with it's result.
 
@@
     where $\Lambda^{0} \subseteq \Lambda$ is a set of $\mathVar{closed~\lambda-terms}$,
-    meaning, those terms have no free variables. Let's now define an inductive,
+    meaning, terms without free variables. Let's now define an inductive,
     and coinductive predicate. Relation
@@
     feature of sending value when doing \textit{yield} operation. Resumptions in
-    handlers that are sending value are one-shot and tail-recursive, therefore
-    we do not need to handle the coalgebraic part, at the cost of flexibility.
+    handlers that are sending value are one-shot and tail-recursive.
 
@@
 
-    Work that is the closest related to topic of the thesis, is work done by
+    From theoretical side, research on comodels and it's relation to state
+    dates to papers by John Power et.\ al.~\cite{comodels, tensors}, and on
+    the runners of comodels to T. Uustalu~\cite{runners-uustalu}.
+
+    Practical part that is closest to topic of the thesis, is work done by
     Danel Ahman and Andrej Bauer~\cite{runners-in-action}, as well as their
     implementation of a $\lambda_{coop}$ language~\cite{coop}, which is a
-    calculus with runners of coalgebraic effects~\cite{runners-uustalu}
-    that ensures linear usage of resources as well as execution of
-    code handling finalisation.
-
-\noindent
-TODO Consider those\\
-- From comodels to coalgebras: State and arrays.\\
-- Tensors of comodels and models for operational semantics.\\
-- Stateful runners of effectful computations.
+    calculus with runners of coalgebraic effects that ensures linear usage of
+    resources as well as execution of code handling finalisation.
 

### 2020-08-23 src/AST.hs
Syntactic distniction between effects and coeffects

@@ -9,7 +9,7 @@ data VariantRow a = VariantRow RowType Label a
 type RecordRow a = Map.Map Label a -- todo: presence variables?
 
 type AlgTheoryName = String
-type Signature = [Label]
+type Signature = [EffLabel]
 
 data Value
     = VVar Var
@@ -62,10 +62,10 @@ data Comp
     | EAbsurd Value
     | EIf Value Comp Comp
     -- Algebraic effects
-    | EOp Label Value
+    | EOp EffLabel Value
     | EHandle Comp Handler
     -- Coalgebraic effects
-    | ECoop Label Value
+    | ECoop EffLabel Value
     | ECohandle Comp Handler
     -- Intermediate representation for a cohandler
     | ECohandleIR AlgTheoryName Value Comp Handler
@@ -76,7 +76,7 @@ data Handler
     | HOps AlgebraicOp Handler
     deriving (Show, Eq)
 
-data AlgebraicOp = AlgOp Label Var Var Comp
+data AlgebraicOp = AlgOp EffLabel Var Var Comp
     deriving (Show, Eq)
 
 -- todo: generalize these with Functor and catamorphism
@@ -91,13 +91,13 @@ hops = aux []
     where aux acc (HRet _ _) = acc
           aux acc (HOps op h) = aux (op:acc) h
 
-opL :: AlgebraicOp -> Label
+opL :: AlgebraicOp -> EffLabel
 opL (AlgOp l _ _ _) = l
 
 hopsL :: Handler -> Signature
 hopsL h = map opL (hops h)
 
-hop :: Label -> Handler -> Maybe AlgebraicOp
+hop :: EffLabel -> Handler -> Maybe AlgebraicOp
 hop _ (HRet _ _) = Nothing
 hop l (HOps op@(AlgOp l' _ _ _) h)
     | l == l' = return op

### 2020-08-23 src/CPSLists.hs

@@ -45,8 +45,8 @@ initialPureCont :: ContF
 initialPureCont v ks = (return . UVal) v
 
 initialEffCont :: ContF
-initialEffCont (UPair (ULabel effLabel) (UPair p r)) ks =
-    return . UApp (UVal r) $ UTopLevelEffect effLabel p
+initialEffCont (UPair (UEffLabel l) (UPair p r)) ks =
+    return . UApp (UVal r) $ UTopLevelEffect l p
 initialEffCont v _ = throwError $ CPSError $ "Incorrect value " ++ show v ++ " in effect continuation"
 
 initialContStack :: [Cont]
@@ -151,14 +151,14 @@ cpsHandle body handler effCons ks = do
     cps body (pureCont:effCont:ks)
 
 
-cpsOp :: Label -> Value -> [Cont] -> CPSMonad UComp
+cpsOp :: EffLabel -> Value -> [Cont] -> CPSMonad UComp
 cpsOp l v ks = do
     let Pure kf:h:ks' = ks
     let hf = unfoldCont h
     x <- freshVar' "rArg"
     pureComp <- kf (UVar x) (h:ks')
     let resumption = ULambda x pureComp
-    let pair cv = UPair (ULabel l) (UPair cv resumption)
+    let pair cv = UPair (UEffLabel l) (UPair cv resumption)
     cv <- cpsVal v ks
     hf (pair cv) ks'
 
@@ -170,7 +170,7 @@ cpsHRet (HRet xVar comp) = Pure (\x (_:ks') -> do
 
 
 cpsHOps :: (ContF -> Cont) -> Handler -> Cont
-cpsHOps effCons ops = effCons (\(UPair (ULabel l) (UPair p r)) ks ->
+cpsHOps effCons ops = effCons (\(UPair (UEffLabel l) (UPair p r)) ks ->
     case hop l ops of
         Just (AlgOp _ pvar rvar comp) -> do
             contComp <- cps comp ks
@@ -178,7 +178,7 @@ cpsHOps effCons ops = effCons (\(UPair (ULabel l) (UPair p r)) ks ->
         Nothing -> forward l p r ks)
 
 
-forward :: Label -> UValue -> UValue -> [Cont] -> CPSMonad UComp
+forward :: EffLabel -> UValue -> UValue -> [Cont] -> CPSMonad UComp
 forward y p r ks = do
     -- TODO: forwarding should not let effects go pass cohandler
[…]

### 2020-08-23 src/EvalTarget.hs

@@ -9,8 +9,8 @@ import CommonEval
 import Types
 
 
-handleEffect :: Label -> DValue -> EvalMonad DValue
-handleEffect l v
+handleEffect :: EffLabel -> DValue -> EvalMonad DValue
+handleEffect (EffL l) v
     -- TODO: Write down in latex this functionality
     | l == "Print" = do
         (lift . print) v
@@ -84,6 +84,8 @@ eval env c = case c of
     UVal (ULambda x c) -> return $ DLambda funcRecord
         where funcRecord [xVal] = let env' = extendEnv env x xVal in eval env' c
     UVal (ULabel l) -> return $ DLabel l
+    UVal (UEffLabel (EffL l)) -> return $ DLabel l
+    UVal (UEffLabel (CoeffL l)) -> return $ DLabel l
     UVal (UNum n) -> return (DNum n)
     UVal (UStr s) -> return (DStr s)
     UVal UUnit -> return DUnit

### 2020-08-23 src/Parser.hs

@@ -155,13 +155,15 @@ retComp = reserved "return"
 opComp :: Parser Comp
 opComp = do
     reserved "do"
-    EOp <$> identifier <*> value
+    l <- identifier
+    EOp (EffL l) <$> value
 
 
 coopComp :: Parser Comp
 coopComp = do
     reserved "observe"
-    ECoop <$> identifier <*> value
+    l <- identifier
+    ECoop (CoeffL l) <$> value
 
 
 handleComp :: Parser Comp
@@ -169,7 +171,7 @@ handleComp = do
     reserved "handle"
     c <- computation
     reserved "with"
-    EHandle c <$> handler
+    EHandle c <$> handler EffL
 
 cohandleComp :: Parser Comp
 cohandleComp = do
@@ -180,36 +182,36 @@ cohandleComp = do
     reserved "at"
     c <- computation
     reserved "through"
-    ECohandleIR algTheoryName v c <$> handler
+    ECohandleIR algTheoryName v c <$> handler CoeffL
 
 
-handler :: Parser Handler
-handler = do
+handler :: (Label -> EffLabel) -> Parser Handler
+handler labelCons = do
     reservedOp "{"
-    h <- handlerBody
+    h <- handlerBody labelCons
     reservedOp "}"
     return h
 
 
-handlerBody :: Parser Handler
-handlerBody =  parseHret
-           <|> parseHops
+handlerBody :: (Label -> EffLabel) -> Parser Handler
+handlerBody labelCons = parseHret
+                     <|> parseHops labelCons
 
 
-parseHops :: Parser Handler
-parseHops = do
-    op <- parseAlgOp
+parseHops :: (Label -> EffLabel) -> Parser Handler
+parseHops labelCons = do
+    op <- parseAlgOp labelCons
     reservedOp "|"
-    HOps op <$> handlerBody
+    HOps op <$> handlerBody labelCons
 
 
-parseAlgOp :: Parser AlgebraicOp
-parseAlgOp = do
+parseAlgOp :: (Label -> EffLabel) -> Parser AlgebraicOp
+parseAlgOp labelCons = do
     l <- identifier
     p <- identifier
     r <- identifier
     reservedOp "->"
-    AlgOp l p r <$> computation
+    AlgOp (labelCons l) p r <$> computation
 
 
 parseHret :: Parser Handler

### 2020-08-23 src/TargetAST.hs

@@ -13,6 +13,7 @@ data UValue
     | UUnit
     | UPair UValue UValue
     | ULabel Label
+    | UEffLabel EffLabel
     | URec Var Var UComp
     | UBinOp BinaryOp UValue UValue
     | UFst UValue
@@ -27,7 +28,7 @@ data UComp
     | UIf UValue UComp UComp
     | ULet Var UComp UComp
     | UAbsurd UValue
-    | UTopLevelEffect Label UValue
+    | UTopLevelEffect EffLabel UValue
     deriving (Eq)
 
 parens :: String -> String
@@ -40,6 +41,8 @@ instance Show UValue where
     show  UUnit = "()"
     show (UPair l r) = parens $ show l ++ ", " ++ show r
     show (ULabel l) = parens $ "L: " ++ l
+    show (UEffLabel (EffL l)) = parens $ "Eff: " ++ l
+    show (UEffLabel (CoeffL l)) = parens $ "Coeff: " ++ l
     show (UVar x) = x
     show (ULambda x c) = parens $ x ++ " -> " ++ show c
     show (UFst v) = "fst " ++ show v

### 2020-08-23 src/Transform.hs

@@ -4,7 +4,7 @@ import AST
 import Types
 import qualified Data.Map as Map
 
-type AlgSignMap = Map.Map Label AlgTheoryName
+type AlgSignMap = Map.Map EffLabel AlgTheoryName
 
 getCoalgebra :: AlgTheoryName -> Comp -> Comp
 getCoalgebra algT = (ELet algT . EReturn . VVar) algT

### 2020-08-23 src/Types.hs

@@ -10,6 +10,10 @@ type EvalResMonad a = IO (Either Error a)
 type Var = String
 type Label = String
 
+data EffLabel = EffL String
+              | CoeffL String
+              deriving (Eq, Ord, Show)
+
 data Error
     = EvalError String
     | CPSError String

### 2020-08-23 src/CPSLists.hs
Ensure effect cannot escape cohandler also during forwarding

@@ -10,9 +10,15 @@ import Types
 
 type ContF = UValue -> [Cont] -> CPSMonad UComp
 
+data EffT = EffT | CoeffT
+
 data Cont = Pure ContF
           | Eff ContF
           | Coeff ContF
+
+getEffCons :: EffT -> (ContF -> Cont)
+getEffCons EffT = Eff
+getEffCons CoeffT = Coeff
 
 unfoldCont :: Cont -> ContF
 unfoldCont (Pure c) = c
@@ -130,24 +136,19 @@ cps e ks = case e of
         v <- cpsVal v ks
         return $ UAbsurd v
     -- Algebraic effects
-    EOp l v -> case ks of
-        (Pure kf:(Eff hf):ks') -> cpsOp l v ks
-        (Pure kf:(Coeff hf):ks') -> throwError $ CPSError "Effect cannot go pass cohandler!"
-        _ -> throwError $ CPSError $ "Ildefined stack of continuations: " ++ show ks
+    EOp l v -> cpsOp l v ks
     EHandle body handler ->
-        cpsHandle body handler Eff ks
+        cpsHandle body handler EffT ks
     -- Coalgebraic effects
     ECohandle body handler ->
-        cpsHandle body handler Coeff ks
-    ECoop l v -> case ks of
-        (Pure kf:h:ks') -> cpsOp l v ks
-        _ -> throwError $ CPSError $ "Ildefined stack of continuations: " ++ show ks
+        cpsHandle body handler CoeffT ks
+    ECoop l v -> cpsOp l v ks
 
 
-cpsHandle :: Comp -> Handler -> (ContF -> Cont) -> [Cont] -> CPSMonad UComp
-cpsHandle body handler effCons ks = do
+cpsHandle :: Comp -> Handler -> EffT -> [Cont] -> CPSMonad UComp
+cpsHandle body handler effT ks = do
     let pureCont = cpsHRet (hret handler)
-    let effCont = cpsHOps effCons handler
+    let effCont = cpsHOps effT handler
     cps body (pureCont:effCont:ks)
 
 
@@ -169,18 +170,19 @@ cpsHRet (HRet xVar comp) = Pure (\x (_:ks') -> do
     return $ ULet xVar (UVal x) convComp)
 
 
-cpsHOps :: (ContF -> Cont) -> Handler -> Cont
-cpsHOps effCons ops = effCons (\(UPair (UEffLabel l) (UPair p r)) ks ->
-    case hop l ops of
[…]

### 2020-08-23 src/Tests.hs

@@ -223,7 +223,7 @@ main = hspec $ do
     it "[Both] Effect should not escape cohandler" $ do
       testFromFile "programs/bothEffects/fileOpsMix.fk" (Left (CPSError "Effect cannot go pass cohandler!"))
       testFromFile "programs/bothEffects/effectsMixIncorrect.fk" (Left (CPSError "Effect cannot go pass cohandler!"))
-      -- testFromFile "programs/bothEffects/mixingNestedHandlers.fk" (Left (CPSError "Effect cannot go pass cohandler!"))
+      testFromFile "programs/bothEffects/mixingNestedHandlers.fk" (Left (CPSError "Effect cannot go pass cohandler!"))
 
     it "[Both] Coeffect may escape handler" $ do
       testFromFile "programs/bothEffects/effectsMixCorrect.fk" (Right (DStr "contents"))

### 2020-08-23 thesis/thesis.tex
[thesis] Abstract

@@
 \polishabstract{
-    \ldots
+    Efekty algebraiczne są dobrym rozwiązaniem na modelowanie efektów
+    obliczeniowych, w sposób który dobrze się składa. Niemniej, powstają przy
+    nich problemy z idiomatycznym zarządzaniem zasobami. Z matematycznej
+    strony, tworzą one wolne modele nad teoriami algebraicznymi, a ich handlery
+    homomorfizmy zachowujące strukturę modelu. Dualnym pojęciem są komodele,
+    które okazały się odpowiednim narzędziem do wyrażania w kodzie przejść z
+    przekazywaniem stanu. Prezentujemy rachunek z efektami algebraicznymi i
+    koalgebraicznymi, aby zaadresować problemy z zarządzaniem zewnętrznym stanem.
 }
 \englishabstract{
-    \ldots
+    Algebraic effects are a great way to model computational effects in a
+    composable way. Nonetheless, they give rise to issues with idiomatic
+    resource management. From mathematical side, they form
+    free models over algebraic theories, with handlers being homomorphisms
+    preserving the model structure. Dualizing this notion gives us comodels,
+    which turned out to be suitable for expressing state-passing transitions
+    in the code. We present a calculus with algebraic and coalgebraic effects,
+    to address the issues with external state management and provide abstraction
+    over resources.
 }
@@
 In this thesis we discuss issues arising from combining algebraic effects
-and handlers wiht resource management, and present a solution through coalgebraic
+and handlers with resource management, and present a solution through coalgebraic
 means along with presentation of experimental programming language Freak, which

### 2020-08-23 thesis/thesis.tex
[thesis] Ensure all examples are valid Freak programs

@@
         let c <- do Choice () in
-        if c then do Write (fh, "Guy Fieri") else do Write (fh, "is cool");
+        let text <- if c then return "Guy Fieri" else return "is cool" in
+        let _ <- do Write (fh, text) in
         do Close fh
     with {
-        return x -> [x] |
-        Choose _ k -> return (append (k true) (k false)) |
+        Choose _ k ->
+            let t <- r 1 in
+            let f <- r 0 in
+            return (t, f) |
+        return x -> return x
     }

### 2020-08-23 thesis/thesis.tex
[thesis] Coexamples for Freak coeffects

@@
 
-\noindent
-TODO maybe example with comparison for these two approaches could be useful here?
-
 That being said, it's only one particular issue that algebraic effects address.
@@
     \begin{center}
-        % TODO: a in A under prod
         {$\displaystyle {\llbracket op \rrbracket}_M : \prod_{a \in A}{|I|} \rightarrow \prod_{p \in P}{|I|} $}
@@
 
-    TODO Examples for cohandlers
+    TODO Introduction into coeffects from Freak side
+
+    \subsection{File handling}
+
+    The first example we are going to show is file handling. As per
+    Section~\ref{sec:problem-statement}, we do not want to expose file handle
+    to the user, thus we hide it behind carrier of FileIO coalgebraic theory,
+    for which the signature is composed of Open, Write and Close operations.
+
+\begin{verbatim}
+    cohandle FileIO using "filename.txt" at
+        let _ <- observe Open () in
+        let _ <- observe Write "Karma Chameleon" in
+        let _ <- observe Close () in
+        return 42
+    through {
+        Open p r ->
+            let filename <- return fst p in
+            let fh <- do OpenBI filename in
+            r (fh, ()) |
+        Write p r ->
+            let fh <- return fst p in
+            let _ <- do WriteBI fh in
+            r (fh, ()) |
+        Close p r ->
+            let fh <- return fst p in
+            let _ <- do CloseBI fh in
+            r ((), ()) |
+        return x -> return x
+    }
+\end{verbatim}
+
+    In the above code, we initialize the coalgebra carrier with filename that
+    we want to handle, we open file, write to it once and then close it. Notice
+    that file handle is not present in the user code, but only accessible in
+    the runner of FileIO as a first element of a pair.
+
+    \subsection{Nondeterministic Finite Automata}
+
+    This example shows how one can implement NFA using coeffects. We abstract
+    over internal state of the automata, and based on letters from the alphabet,
+    we change automata configuration and return whether word is accepted or not.
+    The code below returns whether given NFA accepts word "abba".
+
+\begin{verbatim}
+    cohandle Automata using 0 at
+        let _ <- observe NFA a in
+        let _ <- observe NFA b in
+        let _ <- observe NFA b in
+        observe NFA a
+    through {
+        NFA p r ->
+            let state <- return fst p in
+            let letter <- return snd p in
+            <code transitions here> |
+        return x -> return x
+    }
+\end{verbatim}
 

### 2020-08-24 thesis/thesis.tex
[thesis] Proposed solutions chapter

@@
 
-\section{Duality}
+\section{Duality}\label{sec:duality}
     \subsection{Comodels}
@@
 
-Another static analysis of the program.
-
-\noindent
-- TODO Describe what is data-flow analysis \\
-- TODO Give example for constant folding \\
-- TODO Propose analysis for coalgebraic effects. Not sure if that's not too much,
-as it's not the goal of this thesis. Definitealy leave this section for later.
+Robust resource management could be also achieved by developing a static
+analysis for checking whether resumption is called only once and that
+configuration is properly passed.
+
+Data-flow analysis may be the right tool for detecting such anomalies, however,
+further work in this area is reserved for future.
 
@@
 
-Simplify semantics by separating coalgebraic effects into a new, restricted
-construct in programming language
-
-\noindent
-TODO muuuch longer explanation of this topic
+% Simplify semantics by separating coalgebraic effects into a new, restricted
+% construct in programming language
+
+Having observed in Section~\ref{sec:duality} that cooperations correspond to
+state-passing transitions, we propose a calculus which exactly achieves this
+semantic. We divide the solution into three parts, discussed below.
+
+\subsection{State passing}
+
+State passing can be done by embeding given coalgebraic theory in a state monad
+over configurations, which in simpler words means, that for a set of coeffects
+that we handle, we are going to store current configuration in a state.
+Configuration is retrieved before performing a coeffect, and new configuration
+is saved after one is handled. Passing of configuration does not leak to the
+user, which creates a handy abstraction over operations that manage external
+state.
+
+\subsection{Linear usage}
+
+Linear usage of resumptions is not going to be addressed by crafting operational
+semantic that ensures that, as in~\cite{runners-in-action}, but rather on a
+syntactic level where we verify that resumptions in cohandlers are one-shot and
+tail recursive.
+
+\subsection{Combining effects with coeffects}
+
+Despite of the above means, algebraic effects that occur in the source code may
+drop the resumption and thus finalization code is not going to be executed.
+We address this issue by not allowing effects to escape the cohandler, which
+means that we are not going to discard the execution of cohandler.
 

### 2020-08-25 thesis/thesis.tex
[thesis] More descriptive parts about cohandlers

@@
 
-    \subsection{Coinductive Reasoning}
+    \subsection{Coinductive Reasoning}\label{subsec:coinduction}
 
@@
 
-\section{Cohandlers as Separate Constructs}
+\section{Cohandlers as Separate Constructs}\label{sec:cohandlers-solution}
 
@@
 
-\section{Coexamples}
-
-    TODO Introduction into coeffects from Freak side
+\section{Coexamples}\label{sec:coexamples}
+
+    We are going to give here informal introduction to programming with
+    coeffects in Freak, based on what we have stated in
+    Section~\ref{sec:duality} and Section~\ref{sec:cohandlers-solution}.
+    Elimination of effects has the following syntax:
+
+\begin{verbatim}
+    cohandle <Algebraic Theory> using <Initial state> at
+        <computation>
+    through <handler>
+\end{verbatim}
+
+    Where algebraic theory is a name for signature of coeffects that we
+    handle, with initial coalgebra carrier holding \verb!<Initial state>! value.
+    Handler is syntactically same as in the case of standard algebraic effects,
+    however, the parameter passed to cohandler is always a pair of
+    current configuration and actual argument passed by user to effect.
+
+    Introduction rule for coeffects is written as \verb!observe Coeffect p!,
+    which references observations from coinduction described in
+    Subsection~\ref{subsec:coinduction}.
 
@@
 
-    TODO cohandlers
+    Cohandlers are as they were introduced in Section~\ref{sec:coexamples},
+    where $AlgTheory$ is an identifier representing our signature.
 
@@
             | $nat \; n$ | `string'  \\
-            | $ \backslash x : A \rightarrow M $ | \textbf{rec} $ g \; x \rightarrow M $\\
+            | $ \backslash x : A \rightarrow M $ \\
+            | \textbf{rec} $ g \; x \rightarrow M $\\
             | $V \oplus W$ | $V \approx W$ \\
-            | <> | $ \{ \ell = V; W\} $  | ${[ \ell \; V]}^{R}$
+            | () | ($V$, $W$) \\
+            | \textbf{fst} $V$ | \textbf{snd} $V$
 
@@
             | \textbf{if} $V$ \textbf{then} $M$ \textbf{else} $N$ \\
-            | \textbf{let} $\{\ell  = x; y\} = V$ \textbf{in} $ N $ \\
-            | \textbf{case} $V \{ \ell \; x \rightarrow M; y \rightarrow N\}$ | \textbf{absurd} $ V $ \\
-            | \textbf{return} $V$ | \textbf{let} $ x \leftarrow M $ \textbf{in} $ N $ \\
-            | \textbf{do} $\ell \; V$ | \textbf{handle} $M$ \textbf{with} $ \{ H \} $
+            | \textbf{return} $V$ \\
+            | \textbf{absurd} $ V $ \\
+            | \textbf{let} $ x \leftarrow M $ \textbf{in} $ N $ \\
+            | \textbf{do} $\ell \; V$ \\
+            | \textbf{handle} $M$ \textbf{with} $ \{ H \} $ \\
+            | \textbf{observe} $\ell \; V$ \\
+            | \textbf{cohandle} $AlgTheory$ \textbf{using} $V$ \textbf{at} $M$ \textbf{through} $ \{ H \} $
 
@@
     $nat \; n \approx nat \; n' \rightsquigarrow 0  \quad \quad $    \; if $ n \not\approx n' $
+
+    \subsection{Cohandlers}
 
@@
     \section{Cohandlers}
-    \ldots
+
+    Similarly as in Section~\ref{sec:cohandlers-solution}, we are going to
+    divide this into three parts.
+
+    \subsection{State passing}
+
+    TODO Describe the transformation here.
+
+    \subsection{Linear usage}
+
+    TODO it's not implemented yet. Be more clear what's happening here
+
+    Linear usages of resumptions is asserted by analyzing parsed AST tree,
+    where in every leaf there must be a return statement. It's value is then
+    passed into continuation of a cohandler.
+
+    \subsection{Combining effects with coeffects}
+
+    Guards for not letting effect pass cohandler is implemented in CPS
+    translation, more specifically, code that translates cohandler ensures that
+    argument passed in resumption is not an effect. This is done on a static
+    level before evaluation, which is a desired behaviour.
 

### 2020-08-25 src/inspect.sh
[src, inspect.sh] Distinguish compilation stage

@@ -8,10 +8,11 @@ then
 fi
 
 
+echo "Compiling..."
+make > /dev/null
 echo "Program:"
 cat $PROGRAM
 echo "\nParser:"
-make > /dev/null
 freak -p $PROGRAM
 echo "\nCPS:"
 freak -c $PROGRAM

### 2020-08-25 src/Tests.hs
[src] More and better organized tests

@@ -25,6 +25,9 @@ main :: IO ()
 main = hspec $ do
 
   describe "Freak" $ do
+
+    -- Lambda calculus with binary operators and strings
+
     it "Basic arithmetic" $ do
       evalProgram "return 1" `shouldBeT` (Right (DNum 1))
       evalProgram "return 2 + 3" `shouldBeT` (Right (DNum 5))
@@ -51,10 +54,6 @@ main = hspec $ do
     it "Variable with prim" $ do
       evalProgram "let x' <- return 1 in return x'" `shouldBeT` (Right (DNum 1))
 
-    it "Absurd" $ do
-      evalProgram "absurd 42" `shouldBeT` (Left (absurdErr (UNum 42)))
-      evalProgram "let x <- absurd 17 in return 42" `shouldBeT` (Left (absurdErr (UNum 17)))
-
     it "Let expression" $ do
       evalProgram "let x <- return 3 in return x + 2" `shouldBeT` (Right (DNum 5))
 
@@ -69,6 +68,65 @@ main = hspec $ do
 
     it "Static scope" $ do
       testFromFile "programs/staticScope.fk" (Right (DNum 3))
+
+    it "Let lambda" $ do
+      evalProgram "let f <- return (\\x : int -> return x + 1) in f 42" `shouldBeT` (Right (DNum 43))
+
+    it "Let lambda -- more verbose" $ do
+      evalProgram "let f <- return (\\x : int -> return x + 1) in let y <- f 1 in return y + 2" `shouldBeT` (Right (DNum 4))
+
+    it "Higher order function" $ do
+      testFromFile "programs/higherOrder.fk" (Right (DNum 5))
+
+    it "Application" $ do
+      evalProgram "(\\x : int -> return x + 1) 1" `shouldBeT` (Right (DNum 2))
+
+    it "Let application -- forget result" $ do
+      evalProgram "let x <- (\\x : int -> return x + 1) 1 in return 3" `shouldBeT` (Right (DNum 3))
+
+    it "Let application -- use result" $ do
+      evalProgram "let x <- (\\x : int -> return x + 1) 1 in return x + 1" `shouldBeT` (Right (DNum 3))
+
+    it "Let lambda forget function result" $ do
+      evalProgram "let x <- (\\x : int -> return x) 42 in return 1" `shouldBeT` (Right (DNum 1))
+
+    it "Let lambda forget function result - more verbose" $ do
[…]

### 2020-08-25 src/DValue.hs
[src] Lift EvalMonad to StateMonad

@@ -6,7 +6,9 @@ import AST
 import TargetAST
 import Types
 
-type FuncRecord = [DValue] -> ExceptT Error IO DValue
+type Env = Map.Map Var DValue
+
+type FuncRecord = [DValue] -> EvalMonad DValue
 
 data DValue
     = DNum Integer

### 2020-08-25 src/EvalTarget.hs

@@ -3,14 +3,11 @@ module Eval where
 
 import qualified Data.Map as Map
 import Control.Monad.Except
+import Control.Monad.State
 import AST
 import TargetAST
 import DValue
 import Types
-
-type FuncRecord = [DValue] -> ExceptT Error IO DValue
-
-type Env = Map.Map Var DValue
 
 extendEnv :: Env -> Var -> DValue -> Env
 extendEnv env x v = Map.insert x v env
@@ -41,24 +38,24 @@ handleEffect :: EffLabel -> DValue -> EvalMonad DValue
 handleEffect (EffL l) v
     -- TODO: Write down in latex this functionality
     | l == "Print" = do
-        (lift . print) v
+        (liftIO . print) v
         return DUnit
     | l == "ReadLine" = do
-        input <- lift getLine
+        input <- liftIO getLine
         return $ DStr input
     | l == "ReadFile" = case v of
         (DStr filename) -> do
-            contents <- lift $ readFile filename
+            contents <- liftIO $ readFile filename
             return $ DStr contents
         _ -> throwError $ EvalError $ l ++ " effect accepts a single filename string"
     | l == "WriteFile" = case v of
         (DPair (DStr filename) (DStr contents)) -> do
-            lift $ writeFile filename contents
+            liftIO $ writeFile filename contents
             return DUnit
         _ -> throwError $ EvalError $ l ++ " effect accepts a pair of (filename, contents) strings"
     | l == "AppendFile" = case v of
         (DPair (DStr filename) (DStr contents)) -> do
-            lift $ appendFile filename contents
+            liftIO $ appendFile filename contents
             return DUnit
         _ -> throwError $ EvalError $ l ++ " effect accepts a pair of (filename, contents) strings"
     | otherwise = throwError $ absurdErr (DPair (DLabel l) v)
@@ -120,4 +117,4 @@ eval env c = case c of
 
 
 runEval :: UComp -> EvalResMonad DValue
-runEval c = runExceptT (eval Map.empty c)
+runEval c = evalStateT (runExceptT (eval Map.empty c)) 0

### 2020-08-25 src/Types.hs

@@ -1,9 +1,10 @@
 module Types where
 
 import Control.Monad.Except
+import Control.Monad.Trans.State.Lazy
 
 -- TODO: Reconsider name
-type EvalMonad a = ExceptT Error IO a
+type EvalMonad a = ExceptT Error (StateT Int IO) a
 
 type EvalResMonad a = IO (Either Error a)
 

### 2020-08-25 src/CPSLists.hs
[MAJOR] Pass continuations in lambda down to target calculus

@@ -24,7 +24,7 @@ initialPureCont v ks = (return . UVal) v
 
 initialEffCont :: ContF
 initialEffCont (UPair (UEffLabel l) (UPair p r)) ks =
-    return . UApp (UVal r) $ UTopLevelEffect l p
+    return $ UApp (UVal r) (UTopLevelEffect l p) ks
 initialEffCont v _ = throwError $ CPSError $ "Incorrect value " ++ show v ++ " in effect continuation"
 
 initialContStack :: [Cont]
@@ -58,7 +58,7 @@ cpsVal e ks = case e of
         v1 <- cpsVal e1 ks
         v2 <- cpsVal e2 ks
         return $ UBinOp op v1 v2
-    VLambda x _ body -> ULambda x <$> cps body ks
+    VLambda x _ body -> return $ ULambda x (\ks -> cps body ks)
     VFix g x body -> URec g x <$> cps body ks
     VRecordRow row -> undefined -- todo: records are constructed using ExtendRow
     VExtendRow l v row -> notSupportedErr
@@ -71,22 +71,13 @@ cps e ks = case e of
         let k@(Pure kf):ks' = ks
         f <- cpsVal vF ks
         arg <- cpsVal vArg ks
-        let app = UApp (UVal f) (UVal arg)
-        freshX <- freshVar' "appArg"
-        resume <- kf (UVar freshX) ks'
-        return $ ULet freshX app resume
+        return $ UApp (UVal f) (UVal arg) ks
     ELet x varComp comp -> do
         let k:ks' = ks
         let cont varVal ks'' = do
             body <- cps comp (k:ks'')
             return $ ULet x (UVal varVal) body
         cps varComp (Pure cont:ks')
-    -- todo: This ELet fails on Simulate exceptions test
-    -- ELet x varComp comp -> do
-    --     let k:ks' = ks
-    --     body <- cps comp ks
-    --     varVal <- cps varComp ks
-    --     return $ ULet x varVal body
     ESplit l x y row comp -> do
         v <- cpsVal row ks
         c <- cps comp ks
@@ -129,8 +120,8 @@ cpsOp l v ks = do
     let Pure kf:h:ks' = ks
     let hf = unfoldCont h
     x <- freshVar' "rArg"
-    pureComp <- kf (UVar x) (h:ks')
-    let resumption = ULambda x pureComp
+    let resF = \ks -> kf (UVar x) (h:ks)
[…]

### 2020-08-25 src/DValue.hs

@@ -8,7 +8,7 @@ import Types
 
 type Env = Map.Map Var DValue
 
-type FuncRecord = [DValue] -> EvalMonad DValue
+type FuncRecord = Env -> [DValue] -> [Cont] -> EvalMonad DValue
 
 data DValue
     = DNum Integer

### 2020-08-25 src/EvalTarget.hs

@@ -67,11 +67,12 @@ eval env c = case c of
          eval env (UVal pair) >>= \(DPair lv rv) ->
         let extEnv = Map.insert y rv (Map.insert x lv env) in
         eval extEnv comp
-    UApp fnComp argValue -> do
+    UApp fnComp argValue ks -> do
         fnVal <- eval env fnComp
         argDValue <- eval env argValue
+        liftIO $ print $ "UApp env: " ++ show env
         case fnVal of
-            DLambda fun -> fun [argDValue]
+            DLambda fun -> fun env [argDValue] ks
             e -> throwError $ EvalError $ "Application of non-lambda term " ++ show e ++ " in " ++ show c
     UIf e tC fC ->
         eval env (UVal e) >>= \(DNum n) ->
@@ -106,8 +107,16 @@ eval env c = case c of
     UVal (USnd e) -> eval env (UVal e) >>= \case
         DPair _ v2 -> return v2
         v -> throwError $ EvalError $ "Second projection on expression that is not a pair: " ++ show v
-    UVal (ULambda x c) -> return $ DLambda funcRecord
-        where funcRecord [xVal] = let env' = extendEnv env x xVal in eval env' c
+    UVal (ULambda x f) -> return $ DLambda funcRecord
+        where funcRecord env'' [xVal] ks =
+                let env' = extendEnv env x xVal in
+                f ks >>= eval env'
+                -- case Map.lookup "s" env'' of
+                --     Just r -> resume (extendEnv env "s" r)
+                --     -- Just r -> resume env
+                --     Nothing -> resume env
+                --     where resume env =
+                            -- liftIO $ print $ "funcRecord env: " ++ show env'
     UVal (ULabel l) -> return $ DLabel l
     UVal (UEffLabel (EffL l)) -> return $ DLabel l
     UVal (UEffLabel (CoeffL l)) -> return $ DLabel l

### 2020-08-25 src/TargetAST.hs

@@ -37,7 +37,7 @@ data UValue
     | UNum Integer
     | UStr String
     | UBool Bool
-    | ULambda Var UComp
+    | ULambda Var ([Cont] -> CPSMonad UComp)
     | UUnit
     | UPair UValue UValue
     | ULabel Label
@@ -46,18 +46,18 @@ data UValue
     | UBinOp BinaryOp UValue UValue
     | UFst UValue
     | USnd UValue
-    deriving (Eq)
+    -- deriving (Eq)
 
 data UComp
     = UVal UValue
-    | UApp UComp UComp
+    | UApp UComp UComp [Cont]
     | USplit Label Var Var UValue UComp
     | UCase UValue Label UComp Var UComp
     | UIf UValue UComp UComp
     | ULet Var UComp UComp
     | UAbsurd UValue
     | UTopLevelEffect EffLabel UValue
-    deriving (Eq)
+    -- deriving (Eq)
 
 parens :: String -> String
 parens s = "(" ++ s ++ ")"
@@ -72,13 +72,13 @@ instance Show UValue where
     show (UEffLabel (EffL l)) = parens $ "Eff: " ++ l
     show (UEffLabel (CoeffL l)) = parens $ "Coeff: " ++ l
     show (UVar x) = x
-    show (ULambda x c) = parens $ x ++ " -> " ++ show c
+    show (ULambda x _) = parens $ x ++ " -> comp"
     show (UFst v) = "fst " ++ show v
     show (USnd v) = "snd " ++ show v
 
 instance Show UComp where
     show (UVal v) = show v
-    show (UApp f arg) = parens $ show f ++ " " ++ show arg
+    show (UApp f arg _) = parens $ show f ++ " " ++ show arg
     show (USplit l x y v c) = undefined
     -- show (UCase v l x c y c') = "case " ++ show v ++ " { " ++ show l ++ " " ++ show x ++ "->" ++ show c ++ "; " ++ show y ++ " " ++ show c'
     show (UIf v c c') = "if " ++ show v ++ " then " ++ show c ++ " else " ++ show c'

### 2020-08-26 src/AST.hs
[src] Show instance for AST

@@ -11,6 +11,9 @@ type RecordRow a = Map.Map Label a -- todo: presence variables?
 type AlgTheoryName = String
 type Signature = [EffLabel]
 
+inParens :: String -> String
+inParens s = "(" ++ s ++ ")"
+
 data Value
     = VVar Var
     | VNum Integer
@@ -25,7 +28,18 @@ data Value
     | VBinOp BinaryOp Value Value
     | VFst Value
     | VSnd Value
-    deriving (Show, Eq)
+    deriving (Eq)
+
+instance Show Value where
+    show (VVar x) = x
+    show (VNum n) = show n
+    show (VStr s) = s
+    show (VLambda x _ c) = inParens $ x ++ " -> " ++ show c
+    show VUnit = "()"
+    show (VPair vL vR) = inParens $ show vL ++ ", " ++ show vR
+    show (VBinOp op vL vR) = inParens $ show vL ++ show op ++ show vR
+    show (VFst v) = "fst " ++ show v
+    show (VSnd v) = "snd " ++ show v
 
 data BinaryOp
     = BAdd
@@ -69,15 +83,35 @@ data Comp
     | ECohandle Comp Handler
     -- Intermediate representation for a cohandler
     | ECohandleIR AlgTheoryName Value Comp Handler
-    deriving (Show, Eq)
+    deriving (Eq)
+
+instance Show Comp where
+    show (EVal v) = show v
+    show (ELet x xComp comp) = inParens $ "let " ++ x ++ " = " ++ show xComp ++ " in\n" ++ show comp
+    show (EApp f arg) = inParens $ show f ++ " " ++ show arg
+    show (EReturn v) = "return " ++ show v
+    show (EAbsurd v) = "absurd " ++ show v
+    show (EIf v c c') = "if " ++ show v ++ " then " ++ show c ++ " else " ++ show c'
+    show (EOp l v) = "do " ++ show l ++ " " ++ show v
+    show (EHandle c h) = "handle " ++ show c ++ "with { " ++ show h ++ " }"
+    show (ECoop l v) = "observe " ++ show l ++ " " ++ show v
+    show (ECohandle c h) = "cohandle " ++ show c ++ "through { " ++ show h ++ " }"
+    show (ECohandleIR algT initV c h) = "cohandle " ++ algT ++ " using " ++ show initV ++ " at\n" ++ show c ++ "\nthrough { \n" ++ show h ++ "\n}"
 
[…]

### 2020-08-26 src/DValue.hs

@@ -23,8 +23,8 @@ instance Show DValue where
     show (DStr s) = show s
     show (DLambda _) = "lambda"
     show  DUnit = "()"
-    show (DPair l r) = parens $ show l ++ ", " ++ show r
-    show (DLabel l) = parens $ "L: " ++ show l
+    show (DPair l r) = inParens $ show l ++ ", " ++ show r
+    show (DLabel l) = inParens $ "L: " ++ show l
 
 instance Eq DValue where
     DNum n == DNum n' = n == n'

### 2020-08-26 src/TargetAST.hs

@@ -59,29 +59,26 @@ data UComp
     | UTopLevelEffect EffLabel UValue
     -- deriving (Eq)
 
-parens :: String -> String
-parens s = "(" ++ s ++ ")"
-
 instance Show UValue where
-    show (UBinOp op vL vR) = parens $ show vL ++ show op ++ show vR
+    show (UBinOp op vL vR) = inParens $ show vL ++ show op ++ show vR
     show (UNum n) = show n
     show (UStr s) = s
     show  UUnit = "()"
-    show (UPair l r) = parens $ show l ++ ", " ++ show r
-    show (ULabel l) = parens $ "L: " ++ l
-    show (UEffLabel (EffL l)) = parens $ "Eff: " ++ l
-    show (UEffLabel (CoeffL l)) = parens $ "Coeff: " ++ l
+    show (UPair l r) = inParens $ show l ++ ", " ++ show r
+    show (ULabel l) = inParens $ "L: " ++ l
+    show (UEffLabel (EffL l)) = inParens $ "Eff: " ++ l
+    show (UEffLabel (CoeffL l)) = inParens $ "Coeff: " ++ l
     show (UVar x) = x
-    show (ULambda x _) = parens $ x ++ " -> comp"
+    show (ULambda x _) = inParens $ x ++ " -> comp"
     show (UFst v) = "fst " ++ show v
     show (USnd v) = "snd " ++ show v
 
 instance Show UComp where
     show (UVal v) = show v
-    show (UApp f arg _) = parens $ show f ++ " " ++ show arg
+    show (UApp f arg _) = inParens $ show f ++ " " ++ show arg
     show (USplit l x y v c) = undefined
     -- show (UCase v l x c y c') = "case " ++ show v ++ " { " ++ show l ++ " " ++ show x ++ "->" ++ show c ++ "; " ++ show y ++ " " ++ show c'
     show (UIf v c c') = "if " ++ show v ++ " then " ++ show c ++ " else " ++ show c'
     show (UAbsurd v) = "absurd " ++ show v
-    show (ULet x xComp comp) = parens $ "let " ++ x ++ " = " ++ show xComp ++ " in " ++ show comp
+    show (ULet x xComp comp) = inParens $ "let " ++ x ++ " = " ++ show xComp ++ " in " ++ show comp
     show (UTopLevelEffect l p) = show l ++ show p

### 2020-08-26 src/EvalTarget.hs
[src] Strings comparison

@@ -70,7 +70,6 @@ eval env c = case c of
     UApp fnComp argValue ks -> do
         fnVal <- eval env fnComp
         argDValue <- eval env argValue
-        liftIO $ print $ "UApp env: " ++ show env
         case fnVal of
             DLambda fun -> fun env [argDValue] ks
             e -> throwError $ EvalError $ "Application of non-lambda term " ++ show e ++ " in " ++ show c
@@ -91,9 +90,11 @@ eval env c = case c of
     UAbsurd v -> throwError $ absurdErr v
     -- Values
     UVal (UBinOp op e1 e2) -> do
-        eval env (UVal e1) >>= \(DNum n1) ->
-            eval env (UVal e2) >>= \(DNum n2) ->
-            return $ DNum $ convOp op (n1, n2)
+        eval env (UVal e1) >>= \case
+            (DNum n1) -> eval env (UVal e2) >>= \(DNum n2) -> return $ DNum $ convOp op (n1, n2)
+            (DStr s1) -> eval env (UVal e2) >>= \(DStr s2) -> case op of
+                BEq -> return $ DNum $ boolToInt $ s1 == s2
+                BNe -> return $ DNum $ boolToInt $ s1 /= s2
     UVal (UVar x) -> case Map.lookup x env of
         Just v -> return v
         Nothing -> throwError $ unboundVarErr x

### 2020-08-26 src/Tests.hs

@@ -126,6 +126,9 @@ main = hspec $ do
       evalProgram "if 42 then return 1 else return 0" `shouldBeT` (Right (DNum 1))
       evalProgram "if 0 then return 1 else return 0" `shouldBeT` (Right (DNum 0))
 
+    it "Strings comparison" $ do
+      testFromFile "programs/stringsCompare.fk" (Right (DNum 1))
+
     -- Pairs
 
     it "Pair" $ do

### 2020-08-26 thesis/thesis.tex
[thesis] Missing quotation marks in NFA coexample

@@
     cohandle Automata using 0 at
-        let _ <- observe NFA a in
-        let _ <- observe NFA b in
-        let _ <- observe NFA b in
-        observe NFA a
+        let _ <- observe NFA "a" in
+        let _ <- observe NFA "b" in
+        let _ <- observe NFA "b" in
+        observe NFA "a"
     through {

### 2020-08-26 src/CPSLists.hs
[src] Default CPS for CohandleIR
+ Eta-reduct ULambda translation

@@ -58,7 +58,7 @@ cpsVal e ks = case e of
         v1 <- cpsVal e1 ks
         v2 <- cpsVal e2 ks
         return $ UBinOp op v1 v2
-    VLambda x _ body -> return $ ULambda x (\ks -> cps body ks)
+    VLambda x _ body -> return $ ULambda x (cps body)
     VFix g x body -> URec g x <$> cps body ks
     VRecordRow row -> undefined -- todo: records are constructed using ExtendRow
     VExtendRow l v row -> notSupportedErr
@@ -103,8 +103,8 @@ cps e ks = case e of
     EHandle body handler ->
         cpsHandle body handler EffT ks
     -- Coalgebraic effects
-    ECohandle body handler ->
-        cpsHandle body handler CoeffT ks
+    ECohandle body handler       -> cpsHandle body handler CoeffT ks
+    ECohandleIR _ _ body handler -> cpsHandle body handler CoeffT ks
     ECoop l v -> cpsOp l v ks
 
 

### 2020-08-26 src/Freak.hs
[src] Parsing file should not transform AST

@@ -37,7 +37,5 @@ cpsFromFile filename = readFile filename >>= cpsProgram >>= \case
 
 parseFile :: String -> Bool -> IO ()
 parseFile filename quiet = do
-    code <- readFile filename
-    case parseString code of
-        Right r -> if quiet then return () else print $ transform r
-        Left e  -> print e >> fail "Parse error"
+    comp <- parseFromFile filename
+    if quiet then return () else print comp

### 2020-08-26 src/Parser.hs

@@ -363,3 +363,11 @@ parseString' str =
   case parseString str of
     Left e  -> error $ show e
     Right r -> r
+
+
+parseFromFile :: String -> IO (Comp)
+parseFromFile filename = do
+    code <- readFile filename
+    case parseString code of
+        Right r -> return r
+        Left e  -> print e >> fail "Parse error"

### 2020-08-26 src/AST.hs
[MAJOR] Embedding of (Co)Algebraic Theories into State monad

@@ -136,3 +136,8 @@ hop _ (HRet _ _) = Nothing
 hop l (HOps op@(AlgOp l' _ _ _) h)
     | l == l' = return op
     | otherwise = hop l h
+
+leftJoinHandlers :: Handler -> Handler -> Handler
+leftJoinHandlers hL = aux
+    where aux (HOps op h) = HOps op (aux h)
+          aux (HRet _ _) = hL

### 2020-08-26 src/EvalTarget.hs

@@ -109,15 +109,13 @@ eval env c = case c of
         DPair _ v2 -> return v2
         v -> throwError $ EvalError $ "Second projection on expression that is not a pair: " ++ show v
     UVal (ULambda x f) -> return $ DLambda funcRecord
-        where funcRecord env'' [xVal] ks =
-                let env' = extendEnv env x xVal in
-                f ks >>= eval env'
-                -- case Map.lookup "s" env'' of
-                --     Just r -> resume (extendEnv env "s" r)
-                --     -- Just r -> resume env
-                --     Nothing -> resume env
-                --     where resume env =
-                            -- liftIO $ print $ "funcRecord env: " ++ show env'
+        where funcRecord env'' [xVal] ks = do
+                case Map.lookup "s" env'' of
+                    Just r -> resume (extendEnv env "s" r)
+                    Nothing -> resume env
+                    where resume env =
+                            let env' = extendEnv env x xVal in
+                            f ks >>= eval env'
     UVal (ULabel l) -> return $ DLabel l
     UVal (UEffLabel (EffL l)) -> return $ DLabel l
     UVal (UEffLabel (CoeffL l)) -> return $ DLabel l

### 2020-08-26 src/Parser.hs

@@ -365,7 +365,7 @@ parseString' str =
     Right r -> r
 
 
-parseFromFile :: String -> IO (Comp)
+parseFromFile :: String -> IO Comp
 parseFromFile filename = do
     code <- readFile filename
     case parseString code of

### 2020-08-26 src/Tests.hs

@@ -11,6 +11,8 @@ import Eval
 import DValue
 import Types
 import TargetAST
+import Parser
+import Transform
 
 shouldBeT m b = do
     r <- m
@@ -23,6 +25,18 @@ testFromFile filename expected = do
 
 main :: IO ()
 main = hspec $ do
+
+  describe "AST Transform" $ do
+
+    it "Should inject state monad into cohandlers" $ do
+      cohandler <- parseFromFile "programs/transform/pureCase/before.fk"
+      transCohandler <- parseFromFile "programs/transform/pureCase/after.fk"
+      transform cohandler `shouldBe` transCohandler
+
+    it "Should implement state passing across transitions" $ do
+      cohandler <- parseFromFile "programs/transform/effCase/before.fk"
+      transCohandler <- parseFromFile "programs/transform/effCase/after.fk"
+      transform cohandler `shouldBe` transCohandler
 
   describe "Freak" $ do
 
@@ -93,8 +107,8 @@ main = hspec $ do
     it "Let lambda forget function result - more verbose" $ do
       testFromFile "programs/letLambda.fk" (Right (DNum 1))
 
-    it "Reuse function" $ do
-      testFromFile "programs/reuseFunc.fk" (Right (DNum 3))
+    -- it "Reuse function" $ do
+    --   testFromFile "programs/reuseFunc.fk" (Right (DNum 3))
 
     it "Lambda application" $ do
       evalProgram "(\\x : int -> return x + 1) 42" `shouldBeT` (Right (DNum 43))
@@ -200,8 +214,8 @@ main = hspec $ do
     it "Drop resumption result" $ do
       testFromFile "programs/dropResumption.fk" (Right (DNum 1))
 
-    it "Sum of possible choices" $ do
-      testFromFile "programs/choicesSum.fk" (Right (DNum 35))
+    -- it "Sum of possible choices" $ do
+    --   testFromFile "programs/choicesSum.fk" (Right (DNum 35))
 
     -- it "Min of possible choices" $ do
     --   testFromFile "programs/choicesMin.fk" (Right (DNum 5))
@@ -234,14 +248,20 @@ main = hspec $ do
     it "[Coalg] Increment coeffect handler" $ do
       testFromFile "programs/coeffects/increment.fk" (Right (DNum 18))
 
-    it "[Coalg] Nested cohandlers" $ do
[…]

### 2020-08-26 src/Transform.hs

@@ -2,17 +2,16 @@ module Transform where
 
 import AST
 import Types
+import Parser
 import qualified Data.Map as Map
 
 type AlgSignMap = Map.Map EffLabel AlgTheoryName
 
 getCoalgebra :: AlgTheoryName -> Comp -> Comp
-getCoalgebra algT = (ELet algT . EReturn . VVar) algT
--- getCoalgebra algT = ELet algT (EOp (algT ++ "Get") VUnit)
+getCoalgebra algT = ELet algT (ECoop (CoeffL (algT ++ "Get")) VUnit)
 
 putCoalgebra :: AlgTheoryName -> Value -> Comp -> Comp
-putCoalgebra algT = ELet algT . EReturn
--- putCoalgebra algT = ELet algT . EOp (algT ++ "Put")
+putCoalgebra algT = ELet algT . ECoop (CoeffL (algT ++ "Put"))
 
 coopTransV :: AlgSignMap -> Value -> Value
 coopTransV m v = case v of
@@ -41,18 +40,24 @@ coopTrans m c = case c of
         Just algT ->
             let algTRes = algT ++ "Res" in
             let coop = ECoop l (VPair (VVar algT) (coopTransV m v)) in
+            let bindConf = ELet algT (EReturn (VFst (VVar algTRes))) in
             let bindResult = ELet algTRes coop in
             let contComp = EReturn (VSnd (VVar algTRes)) in
-            (getCoalgebra algT . bindResult . putCoalgebra algT (VVar algT)) contComp
+            -- actual code execution is from left to right:
+            (getCoalgebra algT . bindResult . bindConf . putCoalgebra algT (VVar algT)) contComp
         Nothing -> ECoop l (coopTransV m v)
     ECohandleIR algTheoryName initV c h ->
-        let algTVar = "#" ++ algTheoryName in
+        let algTVar = "_" ++ algTheoryName in
         let sign = hopsL h in
         let m' = foldl (\m s -> Map.insert s algTVar m) m sign in
-        -- TODO: Add handling for #AlgTheoryPut and #AlgTheoryGet effects
         let h' = coopTransHandler m' h in
-        let cohandle = (`ECohandle` h') in
-        (cohandle . putCoalgebra algTVar initV . coopTrans m') c
+        -- TODO: Read and parse from lib/coalgHandler.fk
[…]

### 2020-08-26 src/Tests.hs
[src] More coexamples of coalgebraic effects
+ NFA
+ Counter Machine

@@ -263,6 +263,12 @@ main = hspec $ do
     it "[Coalg] State monad full case" $ do
       testFromFile "programs/coeffects/state.fk" (Right (DNum 3))
 
+    it "[Coalg] NFA" $ do
+      testFromFile "programs/coeffects/NFA.fk" (Right (DPair (DPair (DNum 0) (DNum 0)) (DPair (DNum 0) (DNum 1))))
+
+    it "[Coalg] Counter machine" $ do
+      testFromFile "programs/coeffects/counterCoalg.fk" (Right (DNum 2))
+
     -- Both effects
 
     it "[Both] Effect should not escape cohandler" $ do

### 2020-08-26 src/CPSLists.hs
[src] EReturn CPS: Do not pass silently on illdefined stack

@@ -94,7 +94,7 @@ cps e ks = case e of
         return $ UIf cond tC fC
     EReturn v -> case ks of
         Pure kf:ks' -> cpsVal v ks' >>= \v -> kf v ks'
-        [Eff _] -> cpsVal v ks >>= \v -> initialPureCont v ks
+        _ -> throwError $ CPSError $ "Illdefined stack of continuations when handling  " ++ show v
     EAbsurd v -> do
         v <- cpsVal v ks
         return $ UAbsurd v

### 2020-08-27 thesis/thesis.tex
[thesis] Extend sections about cohandlers
+ Various minor corrections

@@
     As we have seen, comodels turned into state-passing (co)operations, which
-    in programming langauges, may be a way to model interaction with external
+    in programming languages, may be a way to model interaction with external
     resources, where based on current configuration and some parameter, we
@@
     Danel Ahman and Andrej Bauer~\cite{runners-in-action}, as well as their
-    implementation of a $\lambda_{coop}$ language~\cite{coop}, which is a
+    implementation of a $\lambda_{coop}$ language~\cite{coop} and
+    $\mathVar{haskell-coop}$~\cite{haskell-coop} library, which is a
     calculus with runners of coalgebraic effects that ensures linear usage of
@@
 
-\section{Data-Flow Analysis}
+\section{Data-Flow Analysis}\label{sec:data-flow}
 
@@
 We address this issue by not allowing effects to escape the cohandler, which
-means that we are not going to discard the execution of cohandler.
+means that we are not going to allow to discard the execution of cohandler.
 
@@
 
+Full and working example with implementation of automata can be found under
+\verb!programs/coeffects/NFA.fk! in Freak repository~\cite{freak}.
+
 \section{Usage guide}
 
-    % Outdated
     As of this day, two implementations are available, one based on the curried
@@
     Section~\ref{chapter:implementation}. All commands are available within the
-    \verb!src! directory.
+    \verb!src! directory. Please note, however, that Apple-based version is no
+    longer supported, and is left as a reference.
 
@@
         \item Install dependencies: \verb!make install!
-        \item Select implementation: \verb!make link-lists! (default) vs \verb!make link-appel!
+        \item Select implementation: \verb!make link-lists! (recommended) vs \verb!make link-appel!
         \item Compile: \verb!make build!
@@
         \item Run code linter: \verb!make lint!
-        \item Compile, run linter and tests: \verb!make check!
+        \item Validate Freak code from \verb!programs! dir: \verb!make lint!
+        \item Run all above commands: \verb!make check!
     \end{itemize}
@@
 
-    Semantic of Freak is heavily based on Links langauge~\cite{handlers-cps-journal},
+    Semantic of Freak is heavily based on Links language~\cite{handlers-cps-journal},
     for which the source language's dynamics have been described extensively by
@@
 
-    \subsection{Cohandlers}
-
-    TODO cohandlers
+    \subsection{Cohandlers}\label{subsec:calc-cohandlers}
+
+    Semantics for cohandlers is brought from denotational one that we have
+    introduced in Section~\ref{sec:duality}, meaning, we denote cooperations
+    as state passing transitions in a code, represented similarly as algebraic
+    effects, through (co)free (co)models of algebraic theories along with their
+    morphisms. Let's put the following cooperation:
+
+    \begin{center}
+        $ {\llbracket op \rrbracket}^W : {|C|} \times {P} \rightarrow  {|C|} \times {X} $
+    \end{center}
+
+    Operationally, we embed transitions between coeffectull computations
+    in a standard state monad $ S_{C}X \triangleq (C \Rightarrow X \times C) $.
+    We achieve this by transforming AST of the source code, such that
+    configurations are passed among cooperations. For a given coalgebraic theory
+    $T$, handler of the state monad in Freak looks as follows:
+
+\vbox{\begin{verbatim}
+{
+    TPut s' r -> return (\s -> let g <- r () in g s') |
+    TGet _ r -> return (\s -> let g <- r s in g s) |
+    return x -> return (\s -> return x)
+}
+\end{verbatim}
+}
+
+    \noindent
+    Having that defined, we expand computation
+
+\begin{verbatim}
+    observe Coeffect p
+\end{verbatim}
+
+    \noindent
+    into the following one (modulo $\mathVar{alpha-conversion}$):
+
+\vbox{\begin{verbatim}
+    let Tconf <- observe TGet () in
+    let Tcoresult <- observe Coeffect (Tconf, p) in
+    let TnewConf <- return fst Tcoresult in
+    let Tres <- return snd Tcoresult in
+    let _ <- observe TPut TnewConf in
+    return Tres
+\end{verbatim}
+}
+
+    \noindent
+    where before executing a coeffect, we obtain current configuration,
+    pass it to coeffect, put new config and then return the actual result.
+    This is done behind the scenes and user does not know about state
+    passing.
 
@@
 \begin{verbatim}
+    data EffLabel = EffL String
+                  | CoeffL String
+
     data Value
@@
         -- Algebraic effects
-        | EOp Label Value
+        | EOp EffLabel Value
         | EHandle Comp Handler
         -- Coalgebraic effects
-        | ECoop Label Value
+        | ECoop EffLabel Value
         | ECohandle Comp Handler
-\end{verbatim}
-
-    \noindent
-    Similarly for the target calculus data structure. However, as one may notice,
-    for convenience the \textbf{let} translation is homomorphic, as opposed to be
-    to lambda abstracted with immediate application:
+        -- Intermediate representation for a cohandler
+        | ECohandleIR AlgTheoryName Value Comp Handler
+
+    data Handler
+        = HRet Var Comp
+        | HOps AlgebraicOp Handler
+
+    data AlgebraicOp = AlgOp EffLabel Var Var Comp
+\end{verbatim}
+
+    \noindent
+    Where under \verb!Value! in \verb!ECohandleIR! we store initial
+    configuration of a coalgebra, and under \verb!AlgTheoryName! name of the
+    theory over which we operate. We use \verb!EffLabel! structure to
+    syntactically distinguish between effects and coeffects in the code.
+
+    Data structures for the calculus are also driven by syntax. However, as one
+    may notice, for convenience the \textbf{let} translation is homomorphic,
+    as opposed to be to lambda abstracted with immediate application:
 
@@
         | UBool Bool
-        | ULambda Var UComp
+        | ULambda Var ([Cont] -> CPSMonad UComp)
         | UUnit
@@
         | ULabel Label
+        | UEffLabel EffLabel
         | URec Var Var UComp
@@
         = UVal UValue
-        | UApp UComp UComp
+        | UApp UComp UComp [Cont]
         | USplit Label Var Var UValue UComp
@@
         | UAbsurd UValue
-        | UTopLevelEffect Label UValue
-\end{verbatim}
-
-    \noindent
+        | UTopLevelEffect EffLabel UValue
+\end{verbatim}
+
+    \noindent
+    Careful reader may notice the correspondence between \verb!UApp! having
+    a stack of continuations, and \verb!ULambda! being a
+    \verb![Cont] -> CPSMonad UComp! function, which follows directly from
+    translation proposed by Lindley et al\.~\cite{handlers-cps-journal}.
+
     The final answer, common to both evaluations, is represented as a \verb!DValue!,
-    where the meaning of the coproduct is as one would expect:
+    where the meaning of the coproduct is as one would expect. At this point,
+    we no longer treat effect labels as a special case.
 
@@
     type Label = String
-
-    type FuncRecord = [DValue] -> Either Error DValue
+    type EvalMonad a = ExceptT Error (StateT Int IO) a
+    type FuncRecord = [DValue] -> [Cont] -> EvalMonad DValue
 
@@
 
-    \noindent
     Evaluation of target calculus is typical to call-by-value with syntactic
@@
     would expect from naming. We abuse haskell's notation to define types for
-    algebraic effects living in Freak langauge:
+    algebraic effects living in Freak language:
 
@@
     initialEffCont :: ContF
-    initialEffCont (UPair (ULabel effLabel) (UPair p r)) ks =
-        return . UApp (UVal r) $ UTopLevelEffect effLabel p
+    initialEffCont (UPair (UEffLabel l) (UPair p r)) ks =
+        return $ UApp (UVal r) (UTopLevelEffect l p) k
 
@@
 \begin{verbatim}
+data EffT = EffT | CoeffT
+type EvalResMonad a = IO (Either Error a)
+
 cps     :: Comp     -> [Cont]  -> CPSMonad UComp
 cpsVal  :: Value    -> [Cont]  -> CPSMonad UValue
-cpsOp   :: Label -> Value -> [Cont] -> CPSMonad UComp
+cpsOp   :: EffLabel -> Value   -> [Cont] -> CPSMonad UComp
 cpsHRet :: Handler  -> Cont
-cpsHOps :: (ContF -> Cont) -> Handler -> Cont
-forward :: Label   -> UValue -> UValue -> [Cont] -> CPSMonad UComp
+cpsHOps :: EffT     -> Handler -> Cont
+forward :: EffLabel -> UValue  -> UValue -> [Cont] -> CPSMonad UComp
 runCPS  :: Comp     -> EvalResMonad UComp
@@
     computation to the outer handler, and the last one for running the
-    translation, where evaluation result is represented by\\
-    \verb!type EvalResMonad a = IO (Either Error a)!. This results in an
+    translation. This results in an
     implementation that finally supports nested handlers, which can be seen by
     running tests. Links language~\cite{handlers-cps-journal} also supports
-    shallow handlers~\cite{shallow-handlers}, whereas our langauge implements
+    shallow handlers~\cite{shallow-handlers}, whereas our language implements
     only deep ones.
@@
     Types.hs        - Common types definition
+    Transform.hs    - Cohandlers transformation described in Calculus chapter
     programs/       - Exemplary programs covered in tests
@@
 
-    TODO What was achieved, what was described, what needs to be done or researched.
+    We have showed derivation of cooperations, based on previous work, as well
+    as presented a calculus combining algebraic effects with their dual notion.
+    This resulted in a programming language which allows user to seemlessly
+    operate on resources based on solid mathematical foundations. It can seen now,
+    that (co)algebraic effects are an abstract concept that captures plenty of
+    patterns in everyday programming, which already had ad-hoc and more
+    restricted implementations in many existing languages, through constructs
+    such as \verb!try, except, finally!, or by more low-level mechanisms
+    such as handling interrupts in operating systems.
+
+    This relatively new concept in theory of progamming languages yet needs to
+    be researched, and we propose in the next section a few directions into
+    which Freak language could go.
 
@@
     could serve as a way to empirically assert correctness, as opposed to formally.
-    Abstract machine would need to be extended to incorporate coeffects.
-
-    \subsection{Type inference and row polymorphism}
+    Coeffects are living in a syntactic world, which are then translated to
+    algebraic effects, thus we may not need to extend the abstract machine to
+    accomodate them.
+
+    \subsection{Static analysis}
 
@@
     advantage of algebraic effects is that they are explicitly defined in the
-    type of a computation.
+    type of a computation. We have been considering row polymorphism as the
+    tool of choice.
+
+    Another part could involve statics through data-flow analysis, which we
+    have already mentioned in Section~\ref{sec:data-flow}.
 
@@
 
-    The Freak language is limited to a single instance of an effect. We would
-    need to support cases where many instances of the algebraic effects, with
-    the same handler code, could be instantiated. The current state of the
-    art introduces a concept of resources and instances, as in Eff~\cite{programming-in-eff},
-    or instance variables, as in Helium~\cite{binders-labels}. Another part is
-    to research on how this area combines with coalgebraic part.
+    Languages with algebraic effects tend to have an mechanism, in which one
+    can instantiate multiple versions of the effects under the same handler
+    code. The current state of the art introduces a concept of resources and
+    instances, as in Eff~\cite{programming-in-eff}, or instance variables, as
+    in Helium~\cite{binders-labels}. Work in this area could involve diving
+    into how this concept is related to coalgebras, and whether to implement
+    that in Freak.
 
@@
     Exceptions are a trivial example of algebraic effect where the resumption is
-    discarded, and as described in \S 4.5~\cite{handlers-cps}, they can be modeled
-    as a separate construct to improve performance.
+    discarded, and as described in \S 4.5~\cite{handlers-cps} they can be
+    modeled as a separate construct. On the one hand, it may lead to
+    performance improvements, on the other, we could use this notion to ensure
+    that finalisation code for coeffects is executed even in the case of a
+    failure, as described in~\cite{runners-in-action}.
 
@@
     of view. Implementing them as defined by Lindley et al.~\cite{shallow-handlers}
-    could be another way of enhancing Freak.
+    could be another way of enhancing the language.
 

### 2020-08-27 src/Tests.hs
[src] Additional test for Transform.hs

@@ -38,6 +38,10 @@ main = hspec $ do
       transCohandler <- parseFromFile "programs/transform/effCase/after.fk"
       transform cohandler `shouldBe` transCohandler
 
+    it "Should not transform coeffect without theory" $ do
+      let (Right cohandler) = parseString "observe Coeffect ()"
+      transform cohandler `shouldBe` cohandler
+
   describe "Freak" $ do
 
     -- Lambda calculus with binary operators and strings

### 2020-08-27 src/AST.hs
[src] Better AST printing

@@ -93,7 +93,7 @@ instance Show Comp where
     show (EAbsurd v) = "absurd " ++ show v
     show (EIf v c c') = "if " ++ show v ++ " then " ++ show c ++ " else " ++ show c'
     show (EOp l v) = "do " ++ show l ++ " " ++ show v
-    show (EHandle c h) = "handle " ++ show c ++ "with { " ++ show h ++ " }"
+    show (EHandle c h) = "handle\n" ++ show c ++ "\nwith {\n" ++ show h ++ "\n}"
     show (ECoop l v) = "observe " ++ show l ++ " " ++ show v
     show (ECohandle c h) = "cohandle " ++ show c ++ "through { " ++ show h ++ " }"
     show (ECohandleIR algT initV c h) = "cohandle " ++ algT ++ " using " ++ show initV ++ " at\n" ++ show c ++ "\nthrough { \n" ++ show h ++ "\n}"

### 2020-08-29 thesis/thesis.tex
[thesis] Various corrections and improvements

@@
 and handlers with resource management, and present a solution through coalgebraic
-means along with presentation of experimental programming language Freak, which
-is a language with (co)algebraic effects and (co)handlers, where implementation
-is based on Continuation Passing Style translation.
+means along with presentation of Freak, experimental programming language with
+(co)algebraic effects and (co)handlers. Implementation is based on Continuation
+Passing Style translation.
 
@@
 They give the developer necessary tools for declarative programming with
-computational effects on a high level of abstraction, by defining
+computational effects, on a high level of abstraction, by defining
 effects as an API\@ in the code. At the same time, algebraic effects preserve
@@
 In the following chapter we provide a background about effects, define algebraic
-effects in the categorical setting as well as point out dual coalgebraic effects
+effects in the categorical setting, as well as point out dual coalgebraic effects
 and cohandlers, finishing with showing related work in this area.
@@
 the issue of excessive generality, Chapter~\ref{chapter:co-effectful-programming}
-shows the Freak language by examples along with a usage guide, and then
-Chapter~\ref{chapter:calculus-of-freak-language} describes the language's syntax,
-operational semantics and CPS translation. In the next one,
+shows the Freak language through examples along with a usage guide, and then
+Chapter~\ref{chapter:calculus-of-freak-language} describes more formally the
+language's syntax, operational semantics and CPS translation. In the next one,
 Chapter~\ref{chapter:implementation}, implementation details are revealed.
@@
 and instead one may think about functions as being implicitly embed in a Kleisli
-Category over a functor T~\cite{kleisli-ncatlab}, where T is a hidden signature
+Category over a functor $T$~\cite{kleisli-ncatlab}, where $T$ is a hidden signature
 over all possible side effects that occur in our program.
@@
 That being said, it's only one particular issue that algebraic effects address.
-Their handlers may also drop the resumption or invoke it more than
-once. There are many other great sources for getting familiarized with
-effects from practical side~\cite{handlers-tutorial, programming-in-eff, koka-tutorial},
+There are many other great sources for getting familiarized with effects from
+practical side~\cite{handlers-tutorial, programming-in-eff, koka-tutorial},
 so we will omit further explanations. More examples can be found in
@@
 
-    % == begin Sigma-terms ==
-    \begin{definition}
-    Collection of $\mathVar{\Sigma-terms}$ is a free algebra with a generator $X$
-    for a functor $ \mu H_{\Sigma} $ that maps objects into trees over a given
-    signature $ \Sigma $ and morphisms into folds over trees.
-
-    \end{definition}
-    % == end Sigma-terms ==
-
-    % == begin Sigma-equation ==
-    \begin{definition}
-
-        A $ \mathVar{\Sigma-Equation} $ is an object X and a pair of
-        $\mathVar{\Sigma-terms}$ $l, r \in Tree_{\Sigma}(X)$, written as
-
-        \begin{center}
-        $ X \mid l = r $
-        \end{center}
-
-    \end{definition}
-    % == end Sigma-equation ==
-
-    % == begin Algebraic Theory ==
-    \begin{definition}
-
-    An \textit{algebraic theory} $T = (\Sigma_{T}, \mathcal{E}_{T})$, is given
-    by a signature $\Sigma_{T}$ and a collection $\mathcal{E}_{T}$ of
-    $\mathVar{\Sigma_{T}-equations}$. We might omit $T$ subscripts in case
-    our theory of interest is obvious.
-
-    \end{definition}
-    % == end Algebraic Theory ==
-
-    % == begin Interpretation ==
-    \begin{definition}
-
-    An \textit{interpretation $I$ over a given signature $\Sigma$} is given by
-    a carrier object $|I|$ and for each $ op_{i} : P_{i} \rightsquigarrow A_{i} $
-    in $\Sigma$ a map
-
-    \begin{center}
-        $ {\llbracket op_{i} \rrbracket}_I : P_{i} \times{} {|I|}^{A_{i}} \rightarrow |I| $
-    \end{center}
-
-    \noindent
-    Interpretation may be naturally extended to $\mathVar{\Sigma-terms}$, such
-    that a given $\mathVar{\Sigma-term}$ $X \mid t$ is interpreted by a map
-    which sends variables into projections from environment and terms into
-    map composition over each subterm.
-
-    \end{definition}
-    % == end Interpretation ==
-
-    % == begin Model ==
-    \begin{definition}
-
-    A \textit{model M} of an algebraic theory T is an interpretation of the
-    signature ${\Sigma_{T}}$ which validates all the equations $\mathcal{E_{T}}$.
-    That is, for every equation $X \mid l = r$ the following diagram commutes:
-
-    \begin{center}
-    \begin{tikzcd}[row sep=huge, column sep=large]
-        {|M|}^k \arrow[rr, "\llbracket{} X|r \rrbracket{}"', bend right] \arrow[rr, "\llbracket{} X|l \rrbracket{}", bend left] &  & {|M|}
-    \end{tikzcd}
-    \end{center}
-
-    \end{definition}
-    % == end Model ==
-
-    % == begin homomophism ==
-    \begin{definition}
-
-    Let L, M be models of a theory T. $ \mathVar{T-homomorphism}$
-    $\varphi : L \rightarrow M$ is a map such that, for every operation symbol
-    $op_{i}$ in T the following diagram commutes:
-
-    \begin{center}
-    \begin{tikzcd}
-        {|L|}^{ar_{i}} \arrow[rr, "{\llbracket{} op_{i} \rrbracket{}}_{L}"] \arrow[dd, "\varphi{}^{ar_{i}}"'] &  & {|L|} \arrow[dd, "\varphi{}"] \\
-                                              &  &                                     \\
-        {|M|}^{ar_{i}} \arrow[rr, "{\llbracket{} op_{i} \rrbracket{}}_{M}"']                 &  & {|M|}
-    \end{tikzcd}
-    \end{center}
-
-    \noindent
-    We denote here $A^{n}$ as $\mathVar{n-ary}$ product of $A$.
-
-    \end{definition}
-    % == end homomophism ==
-
     % == begin Free algebra ==
@@
 
-    Free F-algebra on an object A (of generators) in $\mathcal{C}$ is meant an algebra
+    $\mathVar{Free}$ $\mathVar{F-algebra}$ on an object $A$ (of generators) in $\mathcal{C}$ is meant an algebra
 
@@
 
+    % == begin Sigma-terms ==
+    \begin{definition}
+    Collection of $\mathVar{\Sigma-terms}$ is a free algebra with a generator $X$
+    for a functor $ \mu H_{\Sigma} $ that maps objects into trees over a given
+    signature $ \Sigma $ and morphisms into folds over trees.
+
+    \end{definition}
+    % == end Sigma-terms ==
+
+    % == begin Sigma-equation ==
+    \begin{definition}
+
+        A $ \mathVar{\Sigma-Equation} $ is an object $X$ and a pair of
+        $\mathVar{\Sigma-terms}$ $l, r \in Tree_{\Sigma}(X)$, written as
+
+        \begin{center}
+        $ X \mid l = r $
+        \end{center}
+
+    \end{definition}
+    % == end Sigma-equation ==
+
+    % == begin Algebraic Theory ==
+    \begin{definition}
+
+    An \textit{algebraic theory} $T = (\Sigma_{T}, \mathcal{E}_{T})$, is given
+    by a signature $\Sigma_{T}$ and a collection $\mathcal{E}_{T}$ of
+    $\mathVar{\Sigma_{T}-equations}$. We might omit $T$ subscripts in case
+    our theory of interest is obvious.
+
+    \end{definition}
+    % == end Algebraic Theory ==
+
+    % == begin Interpretation ==
+    \begin{definition}
+
+    An \textit{interpretation $I$ over a given signature $\Sigma$} is given by
+    a carrier object $|I|$ and for each $ op_{i} : P_{i} \rightsquigarrow A_{i} $
+    in $\Sigma$ a map
+
+    \begin{center}
+        $ {\llbracket op_{i} \rrbracket}_I : P_{i} \times{} {|I|}^{A_{i}} \rightarrow |I| $
+    \end{center}
+
+    \noindent
+    Interpretation may be naturally extended to $\mathVar{\Sigma-terms}$, such
+    that a given $\mathVar{\Sigma-term}$ $X \mid t$ is interpreted by a map
+    which sends variables into projections from environment and terms into
+    map composition over each subterm.
+
+    \end{definition}
+    % == end Interpretation ==
+
+    % == begin Model ==
+    \begin{definition}
+
+    A \textit{model M} of an algebraic theory $T$ is an interpretation of the
+    signature ${\Sigma_{T}}$ which validates all the equations $\mathcal{E_{T}}$.
+    That is, for every equation $X \mid l = r$ the following diagram commutes:
+
+    \begin{center}
+    \begin{tikzcd}[row sep=huge, column sep=large]
+        {|M|}^k \arrow[rr, "\llbracket{} X|r \rrbracket{}"', bend right] \arrow[rr, "\llbracket{} X|l \rrbracket{}", bend left] &  & {|M|}
+    \end{tikzcd}
+    \end{center}
+
+    \end{definition}
+    % == end Model ==
+
+    % == begin homomophism ==
+    \begin{definition}
+
+    Let $L$, $M$ be models of a theory $T$. $ \mathVar{T-homomorphism}$
+    $\varphi : L \rightarrow M$ is a map such that, for every operation symbol
+    $op_{i}$ in $T$ the following diagram commutes:
+
+    \begin{center}
+    \begin{tikzcd}
+        {|L|}^{ar_{i}} \arrow[rr, "{\llbracket{} op_{i} \rrbracket{}}_{L}"] \arrow[dd, "\varphi{}^{ar_{i}}"'] &  & {|L|} \arrow[dd, "\varphi{}"] \\
+                                              &  &                                     \\
+        {|M|}^{ar_{i}} \arrow[rr, "{\llbracket{} op_{i} \rrbracket{}}_{M}"']                 &  & {|M|}
+    \end{tikzcd}
+    \end{center}
+
+    \noindent
+    We denote here $A^{n}$ as $\mathVar{n-ary}$ product of $A$.
+
+    \end{definition}
+    % == end homomophism ==
+
     \begin{definition} \textit{Free model} is just a model that is free algebra.
@@
     \begin{definition}
-        Comodel in $\mathcal{C}$ is just a Model in $ \mathcal{C}^{op} $.
+        $\mathVar{Comodel}$ in $\mathcal{C}$ is just a Model in $ \mathcal{C}^{op} $.
     \end{definition}
@@
 
-    % TODO: textbf Set^op
-    \noindent
-    We want now to embed this morphism in $Set^{op}$. Every arrow in opposite
+    \noindent
+    We want now to embed this morphism in $\mathbf{Set^{op}}$. Every arrow in opposite
     category is reverted, thus, products become coproducts and our morphism
@@
     in the literature are also called comodels and runners~\cite{runners-uustalu, runners-in-action}.
-    As we have seen, comodels turned into state-passing (co)operations, which
-    in programming languages, may be a way to model interaction with external
-    resources, where based on current configuration and some parameter, we
+    As we have seen, comodels turned into state-passing (co)operations.
+    In programming languages, they may be a way to model interaction with external
+    resources, where based on current configuration and a parameter, we
     obtain new configuration along with it's result.
@@
 
-    Induction is a way of constructing new structures. Recursion is a way of
-    folding inductively defined structure in a terminating way. Recursive
+    \begin{definition}
+        \textit{Initial algebra} is an initial object in \textbf{Alg}, the category of algebras
+        and homomorphisms.
+    \end{definition}
+
+    \begin{lemma}
+        Free algebra over an initial object is initial algebra. Dually,
+        cofree coalgebra over a final object is final coalgebra.
+    \end{lemma}
+    \begin{proof}
+        Follows immediately after putting initial object as generator in the
+        diagram for free algebra.
+    \end{proof}
+
+    Due to close connections presented above of initial algebras and free
+    algebras, from which algebraic effects are arising, we could not spare
+    a few words a about induction and coinduction, where the latter may give
+    us more intuition about coalgebraic effects.
+
+    \textit{Induction} is a way of constructing new structures. \textit{Recursion} is
+    a way of folding inductively defined structure in a terminating way. Recursive
     functions (not to be confused with recursive computability class), should
@@
 
-    Coinduction is literally a dual notion to induction. We \textit{observe} possibly
-    infinite structures, by doing deconstruction. Corecursion, is a way
-    of productively defining new, possibly enlarged structures. Due to infinity,
-    the evaluation should be lazy, whereas in induction it may be eager.
+    \textit{Coinduction} is literally a dual notion to induction. We \textit{observe}
+    possibly infinite structures, by doing deconstruction. \textit{Corecursion}
+    is a way of productively defining new, possibly enlarged structures. Due to
+    infinity, the evaluation should be lazy, whereas in induction it may be eager.
 
@@
 state monad}~\cite{linear-usage-of-state}.
-
-\noindent
-TODO should this section be further extended?
 
@@
 
-    \subsection{Cohandlers}\label{subsec:calc-cohandlers}
-
-    Semantics for cohandlers is brought from denotational one that we have
-    introduced in Section~\ref{sec:duality}, meaning, we denote cooperations
+    \subsection{Cooperations}\label{subsec:calc-cohandlers}
+
+    Semantics for cooperations is brought from denotational one that we have
+    introduced in Section~\ref{sec:duality}, meaning, we denote them
     as state passing transitions in a code, represented similarly as algebraic
@@
 
-    Operationally, we embed transitions between coeffectull computations
+    Operationally, we embed transitions between coeffectful computations
     in a standard state monad $ S_{C}X \triangleq (C \Rightarrow X \times C) $.
@@
     \noindent
-    into the following one (modulo $\mathVar{alpha-conversion}$):
+    into the following one (modulo $\mathVar{\alpha-conversion}$):
 
@@
     let Tconf <- observe TGet () in
-    let Tcoresult <- observe Coeffect (Tconf, p) in
-    let TnewConf <- return fst Tcoresult in
-    let Tres <- return snd Tcoresult in
+    let (TnewConf, Tres) <- observe Coeffect (Tconf, p) in
     let _ <- observe TPut TnewConf in
@@
     This is done behind the scenes and user does not know about state
-    passing.
+    passing. Note that for readability we have done pattern matching on pairs,
+    while currently in Freak it needs to be deconstructed using projections.
+
+    \subsection{Cohandlers}
+
+    Cohandlers needs to be aware of configuration passing, therefore,
+    every cohandler accepts $ C \times P $ as a parameter where $C$ is current
+    configuration and $P$ parameter passed to coeffect. In a similar way,
+    value passed to resumption also needs to be a pair $ C \times X $ of new
+    configuration and value returned down to computation.
+
+    For a full currently implemented example of transformation, one may take a
+    look at \verb!programs/transform/effCase! directory in source code.
 
@@
 
-    TODO Describe the transformation here.
+    State passing is defined as it was described in
+    Subsection~\ref{subsec:calc-cohandlers}.
+    From implementation standpoint, transformation is done by manipulating
+    AST of the language to inject constructs handling state management.
 
@@
     AST.hs          - AST data structures
-    CommonCPS.hs    - Common functions for CPS translation
-    CommonEval.hs   - Common functions for evaluation
     CPSLists.hs     - Uncurried CPS translation
+    CPSMonad.hs     - Definition of CPSMonad
+    DValue.hs       - Evaluation result data structures
     EvalTarget.hs   - Evaluation of the target calculus
@@
     Tests.hs        - Tests module
+    Transform.hs    - Cohandlers transformation described in Calculus chapter
     Types.hs        - Common types definition
-    Transform.hs    - Cohandlers transformation described in Calculus chapter
     programs/       - Exemplary programs covered in tests
@@
     such as \verb!try, except, finally!, or by more low-level mechanisms
-    such as handling interrupts in operating systems.
-
-    This relatively new concept in theory of progamming languages yet needs to
+    like handling interrupts in operating systems.
+
+    This relatively new concept in theory of programming languages yet needs to
     be researched, and we propose in the next section a few directions into
-    which Freak language could go.
+    which Freak could go.
 

### 2020-08-29 thesis/thesis.tex
[thesis] Spelling

@@
 system~\cite{type-and-effect}, but also do it in a way that is composable. This
-is where algebraic effects comes to the rescue. From theoretical point of view,
+is where algebraic effects come to the rescue. From theoretical point of view,
 we need to develop equational theory about our effects to assert correctness of
@@
         $F_{T}$ is an endofunctor, therefore it sends $X$ into an object in $\mathcal{C}$.
-        From any object we have a unique map sending an object into model,
+        From any object we have a unique map sending an object into a model,
         which is $\eta_{F_{T} (X)}$. From free property of our model we get
@@
 models over computation trees, where the signature is the set of effects, and
-equational theory is definining rules of rewriting expressions with effects.
+equational theory is defining rules of rewriting expressions with effects.
 The latter is usually irrelevant from implementation perspective, but may be
@@
         We can think about $\mathVar{B-ary}$ coproducts (tagged unions) over the same set
-        such that we have $B$ replicas of $A$ and we select which replica do we want
+        such that we have $B$ replicas of $A$ and we select which replica we want
         to pick. Putting it this way, it immediately follows that it's the same,
@@
 
-    Model turns into comodel~\cite{comodels}, which we shall also call \textit{world}.
+    Model turns into a comodel~\cite{comodels}, which we shall also call \textit{world}.
     Putting this together with the above lemma, we obtain the following map:
@@
     resources, where based on current configuration and a parameter, we
-    obtain new configuration along with it's result.
+    obtain a new configuration along with it's result.
 
@@
     \textit{closed forward} under the above rules. Relation can be inductively
-    built from empty set and then in a fixed-point manner we can add more terms.
+    built from an empty set and then in a fixed-point manner we can add more terms.
     Let's now see a relation $ \Uparrow{} \subseteq \Lambda^{0}$ of
@@
     As can be seen in the J. Yallop repository~\cite{effects-bibliography}, algebraic
-    effects and handlers are now trending branch in the programming languages theory.
+    effects and handlers are now a trending branch in programming languages theory.
 
@@
 
-    From theoretical side, research on comodels and it's relation to state
+    From theoretical side, research on comodels and its relation to state
     dates to papers by John Power et.\ al.~\cite{comodels, tensors}, and on
@@
 may lead to change of their internal state system. Examples include change of
-state in NFA, taking next element in an infinite stream or a closing file.
+state in NFA, taking the next element in an infinite stream or a closing file.
 Finalization must occur after initialization, thus we conclude that resumption
@@
 Data-flow analysis may be the right tool for detecting such anomalies, however,
-further work in this area is reserved for future.
+further work in this area is reserved for the future.
 
@@
 
-State passing can be done by embeding given coalgebraic theory in a state monad
+State passing can be done by embedding given coalgebraic theory in a state monad
 over configurations, which in simpler words means, that for a set of coeffects
@@
     In the above code, we initialize the coalgebra carrier with filename that
-    we want to handle, we open file, write to it once and then close it. Notice
+    we want to handle, we open a file, write to it once and then close it. Notice
     that file handle is not present in the user code, but only accessible in
@@
     we change automata configuration and return whether word is accepted or not.
-    The code below returns whether given NFA accepts word "abba".
+    The code below returns whether a given NFA accepts the word "abba".
 
@@
     pass it to coeffect, put new config and then return the actual result.
-    This is done behind the scenes and user does not know about state
+    This is done behind the scenes and the user does not know about state
     passing. Note that for readability we have done pattern matching on pairs,
@@
     Where the first two are implementing cps for computations and values.
-    \verb!cpsOp! is doing translation for an (co)effect that occured in program,
+    \verb!cpsOp! is doing translation for an (co)effect that occurred in program,
     \verb!cpsHRet! and \verb!cpsHOps! are yielding pure and effectful continuations,
@@
     as presented a calculus combining algebraic effects with their dual notion.
-    This resulted in a programming language which allows user to seemlessly
+    This resulted in a programming language which allows user to seamlessly
     operate on resources based on solid mathematical foundations. It can seen now,
@@
     algebraic effects, thus we may not need to extend the abstract machine to
-    accomodate them.
+    accommodate them.
 

### 2020-08-29 src/Transform.hs
[src] Make linter happy

@@ -52,7 +52,7 @@ coopTrans m c = case c of
         let m' = foldl (\m s -> Map.insert s algTVar m) m sign in
         let h' = coopTransHandler m' h in
         -- TODO: Read and parse from lib/coalgHandler.fk
-        let stateMonadHandler = (HOps (AlgOp (CoeffL (algTVar ++ "Put")) "s'" "r" (EReturn (VLambda "s" TInt (ELet "g" (EApp (VVar "r") VUnit) (EApp (VVar "g") (VVar "s'")))))) (HOps (AlgOp (CoeffL (algTVar ++ "Get")) "_" "r" (EReturn (VLambda "s" TInt (ELet "g" (EApp (VVar "r") (VVar "s")) (EApp (VVar "g") (VVar "s")))))) (HRet "x" (EReturn (VLambda "s" TInt (EReturn (VVar "x"))))))) in
+        let stateMonadHandler = HOps (AlgOp (CoeffL (algTVar ++ "Put")) "s'" "r" (EReturn (VLambda "s" TInt (ELet "g" (EApp (VVar "r") VUnit) (EApp (VVar "g") (VVar "s'")))))) (HOps (AlgOp (CoeffL (algTVar ++ "Get")) "_" "r" (EReturn (VLambda "s" TInt (ELet "g" (EApp (VVar "r") (VVar "s")) (EApp (VVar "g") (VVar "s")))))) (HRet "x" (EReturn (VLambda "s" TInt (EReturn (VVar "x")))))) in
         let runner = leftJoinHandlers stateMonadHandler h' in
         let cohandle = (\c -> ECohandleIR algTheoryName initV c runner) in
         let coalgComp = (cohandle . coopTrans m') c in

### 2020-08-29 thesis/thesis.tex
Extend Freak usage guide

@@
     After compiling and linking program to PATH, one may evaluate program as
-    follows: \verb!freak programs/choicesList.fk! The actual code is described in Section~\ref{sec:choice-example}
+    follows: \verb!freak programs/choicesList.fk! The actual code is described
+    in Section~\ref{sec:choice-example}:
+
+\begin{verbatim}
+Usage: freak [options] [filename]
+Options:
+                       Evaluate program
+   -c, --cps           CPS-translate program
+   -h, --help          Print this message and exit
+   -p, --parse         Parse program
+   -ds, --desugar      Parse and run AST transformations
+   -v, --validate      Run available static analysis
+\end{verbatim}
 

### 2020-08-29 src/Freak.hs

@@ -28,14 +28,18 @@ runProgram s = evalProgram s >>= \case
     Right v -> print v
 
 runFromFile :: String -> IO ()
-runFromFile filename = readFile filename >>= runProgram
+runFromFile = readFile >=> runProgram
 
 cpsFromFile :: String -> IO ()
-cpsFromFile filename = readFile filename >>= cpsProgram >>= \case
+cpsFromFile f = readFile f >>= cpsProgram >>= \case
     Right cps -> print cps
     Left err -> print err >> fail "CPS translation error"
 
-parseFile :: String -> Bool -> IO ()
-parseFile filename quiet = do
-    comp <- parseFromFile filename
-    if quiet then return () else print comp
+parseFile :: String -> IO ()
+parseFile = parseFromFile >=> print
+
+parseDesugarFile :: String -> IO ()
+parseDesugarFile = parseFromFile >=> (print . transform)
+
+validateFile :: String -> IO ()
+validateFile = readFile >=> cpsProgram >=> (\_ -> return ())

### 2020-08-29 src/Main.hs

@@ -2,17 +2,34 @@ import System.Environment
 import System.Exit
 import Freak
 
--- TODO: extend usage with -p param
-usage = putStrLn "Usage: freak [filename]"
+usage = putStrLn "\
+    \Usage: freak [options] [filename]\n\
+    \Options:\n\
+    \                       Evaluate program\n\
+    \   -h, --help          Print this message and exit\n\
+    \   -c, --cps           CPS-translate program\n\
+    \   -p, --parse         Parse program\n\
+    \   -ds, --desugar      Parse and run AST transformations\n\
+    \   -v, --validate      Run available static analysis\n"
 
 run :: [String] -> IO ()
-run ["-h"] = usage >> exitSuccess
+run ("-h":_) = usage >> exitSuccess
+run ("--help":_) = run ["-h"]
+
 run [filename] = runFromFile filename
-run ["-p", filename] = parseFile filename False
-run ["-pq", filename] = parseFile filename True
+
+run ["-p", filename] = parseFile filename
 run ["--parse", filename] = run ["-p", filename]
+
+run ["-ds", filename] = parseDesugarFile filename
+run ["--desugar", filename] = run ["-ds", filename]
+
 run ["-c", filename] = cpsFromFile filename
 run ["--cps", filename] = run ["-c", filename]
+
+run ["-v", filename] = validateFile filename
+run ["--validate", filename] = run ["-v", filename]
+
 run _ = usage >> exitFailure
 
 main :: IO ()

### 2020-08-31 thesis/thesis.tex
[thesis] Mention value derivation available in repo
[thesis] Corrections, spelling

@@
 Algebraic effects, while not a new concept, have received a lot of
-attention~\cite{effects-bibliography} in recent years, both from the theoretical
+attention in recent years~\cite{effects-bibliography}, both from the theoretical
 and practical side.
@@
 They give the developer necessary tools for declarative programming with
-computational effects, on a high level of abstraction, by defining
+computational effects on a high level of abstraction, by defining
 effects as an API\@ in the code. At the same time, algebraic effects preserve
@@
 drop the resumption allow one to express fairly complex logic in a concise way,
-but at the same time, they give rise to issues that would not occur in standard
+however, they also give rise to issues that would not occur in standard
 control flow, especially when composing various effects together. Unwanted
@@
     with {
-        Choose _ k ->
+        Choose _ r ->
             let t <- r 1 in
@@
 
-Other branch of languages theory, that we shall investigate more in this thesis,
+Other branch of languages theory, that we shall investigate in this thesis,
 is about taming complexity of programs. Various methods of static analysis has
-been developed for various use cases, most notably, type systems. Thanks to strong
+been developed, most notably, type systems. Thanks to strong
 and static type systems along with their implementations, we have solid tools to
@@
 
-\noindent
-TODO Reconsider reference to ncatlab on Kleisli Cat, often it's not the best resource
+% TODO Reconsider reference to ncatlab on Kleisli Cat, often it's not the best resource
 
@@
 is where algebraic effects come to the rescue. From theoretical point of view,
-we need to develop equational theory about our effects to assert correctness of
-our language as well as to have the right hammer to reason about our programs.
-From practical side, we need to have the way of taming computational effects
+we need to develop equational theory about effects to assert correctness of
+the language, as well as to have the right hammer to reason about our programs.
+From practical side, we need to have \textit{the} way of taming computational effects
 in our programs. That is the point where we would like to introduce to unfamiliar
@@
     exists a unique homomorphism $ \overline{f} : A^{\#} \longrightarrow B $
-    extending f, i.e.\ a unique morphism of $ \mathcal{C}$ for which the diagram
+    extending $f$, i.e.\ a unique morphism in $ \mathcal{C}$ for which the diagram
     below commutes:
@@
     Let $L$, $M$ be models of a theory $T$. $ \mathVar{T-homomorphism}$
-    $\varphi : L \rightarrow M$ is a map such that, for every operation symbol
+    $\varphi : L \rightarrow M$ is a map, such that, for every operation symbol
     $op_{i}$ in $T$ the following diagram commutes:
@@
         Let $M$ be a free model of an algebraic theory $T$ in category $\mathcal{C}$. We
-        have an endofunctor $ F_{T}$, which takes objects into free model and
-        maps to unique homomorphisms for which the following diagram commutes:
+        have an endofunctor $ F_{T} : \mathcal{C} \rightarrow \mathcal{C} $, which takes
+        objects into free models and maps to unique homomorphisms, for which the following
+        diagram commutes:
 
@@
         \noindent
-        which immediately gives us $\eta$ natural transformation for a monad.
-        We can now define $\mu{}$ for a monad as a unique morphism for which
+        which immediately gives us $\eta : 1_{C} \rightarrow F_{T}$ natural
+        transformation for a monad.
+        We can now define $\mu{} : {F_{T}}^{2} \rightarrow F_{T}$ for a monad
+        as a collecton indexed by $X$ of unique morphisms for which
         the diagram commutes:
@@
         \begin{tikzcd}[row sep=huge, column sep=large]
-            F_{T} (X) \arrow[r, "\eta_{F_{T} (X)}"] \arrow[rd, "id"'] & F_{T} (F_{T} (X)) \arrow[d, "\mu{}"] \\
+            F_{T} (X) \arrow[r, "\eta_{F_{T} (X)}"] \arrow[rd, "id"'] & F_{T} (F_{T} (X)) \arrow[d, "\mu_{F_{T}(X)}"] \\
                                                          & F_{T} (Y)
@@
         \noindent
-        $F_{T}$ is an endofunctor, therefore it sends $X$ into an object in $\mathcal{C}$.
+        $F_{T}$ is an endofunctor, thus $F_{T}(X)$ is an object in $\mathcal{C}$.
         From any object we have a unique map sending an object into a model,
@@
 
+    Due to close connections of initial algebras and free algebras, from which
+    algebraic effects are arising, we spare a few words a about
+    induction and coinduction, where the latter may give us more intuition
+    about coalgebraic effects.
+
     \begin{definition}
         \textit{Initial algebra} is an initial object in \textbf{Alg}, the category of algebras
-        and homomorphisms.
+        and homomorphisms. Dually, \textit{final coalgebra} is a terminal
+        object in \textbf{Alg}
     \end{definition}
@@
     \begin{lemma}
-        Free algebra over an initial object is initial algebra. Dually,
-        cofree coalgebra over a final object is final coalgebra.
+        Free algebra over an initial object is initial algebra.
     \end{lemma}
@@
         Follows immediately after putting initial object as generator in the
-        diagram for free algebra.
+        diagram for free algebra. Notice that from duality we have that cofree
+        coalgebra over a terminal object is a final coalgebra.
     \end{proof}
-
-    Due to close connections presented above of initial algebras and free
-    algebras, from which algebraic effects are arising, we could not spare
-    a few words a about induction and coinduction, where the latter may give
-    us more intuition about coalgebraic effects.
 
@@
 
-    Here is a table that summarizes difference between these two methods of reasoning:
+    Here is a table that summarizes the differences between these two methods of reasoning:
 
@@
     and coinductive proof principle corresponds to the uniqueness of the mediator.
-    Recall that coalgebraic effects arises from a particular type of a final
-    coalgebra, namely, cofree coalgebra.
-
-    \noindent
-    TODO Not sure which paper to cite here for a deeper dive into the upper
-    topic. Book \textit{Initial Algebras, Terminal Coalgebras, and the Theory of Fixed Points of Functors}
-    by Adámek, Milius and S. Moss from which I have studied is no longer available
-    on the internet, as it was a draft under construction.
+    Recall from Chapter~\ref{chapter:background} that coalgebraic effects arises from cofree coalgebra.
+
+    % TODO Not sure which paper to cite here for a deeper dive into the upper
+    % topic. Book \textit{Initial Algebras, Terminal Coalgebras, and the Theory of Fixed Points of Functors}
+    % by Adámek, Milius and S. Moss from which I have studied is no longer available
+    % on the internet, as it was a draft under construction.
 
@@
     \noindent
-    where $\Lambda^{0} \subseteq \Lambda$ is a set of $\mathVar{closed~\lambda-terms}$,
+    We denote $\Lambda^{0} \subseteq \Lambda$ as a set of $\mathVar{closed~\lambda-terms}$,
     meaning, terms without free variables. Let's now define an inductive,
@@
     such as C~\cite{leijen-c} or Python~\cite{python-effect}. In Python,
-    effects are implemented through generators~\cite{one-shot}, using built-in
-    feature of sending value when doing \textit{yield} operation. Resumptions in
-    handlers that are sending value are one-shot and tail-recursive.
+    effects are implemented through generators, using built-in feature of
+    sending value when doing \textit{yield} operation. Implementations of effects
+    based on coroutines~\cite{one-shot}, such as in Python, have one-shot and
+    tail-recursive resumptions.
 
@@
     dates to papers by John Power et.\ al.~\cite{comodels, tensors}, and on
-    the runners of comodels to T. Uustalu~\cite{runners-uustalu}.
+    the runners of comodels from T. Uustalu~\cite{runners-uustalu}.
 
@@
 may lead to change of their internal state system. Examples include change of
-state in NFA, taking the next element in an infinite stream or a closing file.
+state in NFA, taking the next element in an infinite stream or closing a file.
 Finalization must occur after initialization, thus we conclude that resumption
@@
     Where algebraic theory is a name for signature of coeffects that we
-    handle, with initial coalgebra carrier holding \verb!<Initial state>! value.
+    handle, with final coalgebra carrier holding \verb!<Initial state>! value.
     Handler is syntactically same as in the case of standard algebraic effects,
@@
     Section~\ref{sec:problem-statement}, we do not want to expose file handle
-    to the user, thus we hide it behind carrier of FileIO coalgebraic theory,
-    for which the signature is composed of Open, Write and Close operations.
+    to the user, thus we hide it behind carrier of \textit{FileIO} coalgebraic
+    theory, for which the signature is composed of \textit{Open}, \textit{Write}
+    and \textit{Close} operations.
 
@@
     that file handle is not present in the user code, but only accessible in
-    the runner of FileIO as a first element of a pair.
+    the runner of \textit{FileIO} as a first element of a pair.
 
@@
     we change automata configuration and return whether word is accepted or not.
-    The code below returns whether a given NFA accepts the word "abba".
+    The code below returns whether a given NFA accepts the word "Abba".
 
@@
     cohandle Automata using 0 at
-        let _ <- observe NFA "a" in
+        let _ <- observe NFA "A" in
         let _ <- observe NFA "b" in
@@
 
-    As of this day, two implementations are available, one based on the curried
-    translation from Appel~\cite{appel-continuations}, and the second one based
-    directly on the uncurried translation with continuations as explicit
-    stacks~\cite{handlers-cps-journal}. More details can be found in
-    Section~\ref{chapter:implementation}. All commands are available within the
-    \verb!src! directory. Please note, however, that Apple-based version is no
-    longer supported, and is left as a reference.
+    All commands are available within the \verb!src! directory.
 
@@
         \item Run code linter: \verb!make lint!
-        \item Validate Freak code from \verb!programs! dir: \verb!make lint!
+        \item Validate Freak code from \verb!programs! dir: \verb!make validate_programs!
         \item Run all above commands: \verb!make check!
@@
     Actual programs in Freak can be found in Section~\ref{sec:examples}.
+    Derivation of a value for a simple program with function application based
+    on CPS calculus can be seen in \verb!derivation! file.
 
@@
     machine~\cite{liberating-effects}, which was proved to coincide with CPS
-    translation. That being said, Freak introduces new basic constructs to the
-    language, for which we shall define the semantics.
+    translation. That being said, Freak introduces coalgebraic effects and a few
+    basic constructs to the language, for which we shall define the semantics.
 
@@
 
-    TODO it's not implemented yet. Be more clear what's happening here
+%     TODO it's not implemented yet. Be more clear what's happening here
 

### 2020-09-01 thesis/thesis.tex
[thesis] Resolve pending TODOs with citations

@@
 and instead one may think about functions as being implicitly embed in a Kleisli
-Category over a functor $T$~\cite{kleisli-ncatlab}, where $T$ is a hidden signature
+Category over a functor $T$~\cite{kleisli}, where $T$ is a hidden signature
 over all possible side effects that occur in our program.
-
-% TODO Reconsider reference to ncatlab on Kleisli Cat, often it's not the best resource
 
@@
     and coinductive proof principle corresponds to the uniqueness of the mediator.
-    Recall from Chapter~\ref{chapter:background} that coalgebraic effects arises from cofree coalgebra.
-
-    % TODO Not sure which paper to cite here for a deeper dive into the upper
-    % topic. Book \textit{Initial Algebras, Terminal Coalgebras, and the Theory of Fixed Points of Functors}
-    % by Adámek, Milius and S. Moss from which I have studied is no longer available
-    % on the internet, as it was a draft under construction.
+    Recall from Chapter~\ref{chapter:background} that coalgebraic effects arises
+    from cofree coalgebra. For more information about coalgebras, see~\cite{jacobs-2016}.
 

### 2020-09-01 src/Tests.hs
[src] Put pure tests to common dir

@@ -85,7 +85,7 @@ main = hspec $ do
       evalProgram "let x <- return 3 in let x <- return 2 in return x" `shouldBeT` (Right (DNum 2))
 
     it "Static scope" $ do
-      testFromFile "programs/staticScope.fk" (Right (DNum 3))
+      testFromFile "programs/pure/staticScope.fk" (Right (DNum 3))
 
     it "Let lambda" $ do
       evalProgram "let f <- return (\\x : int -> return x + 1) in f 42" `shouldBeT` (Right (DNum 43))
@@ -109,10 +109,10 @@ main = hspec $ do
       evalProgram "let x <- (\\x : int -> return x) 42 in return 1" `shouldBeT` (Right (DNum 1))
 
     it "Let lambda forget function result - more verbose" $ do
-      testFromFile "programs/letLambda.fk" (Right (DNum 1))
+      testFromFile "programs/pure/letLambda.fk" (Right (DNum 1))
 
     -- it "Reuse function" $ do
-    --   testFromFile "programs/reuseFunc.fk" (Right (DNum 3))
+    --   testFromFile "programs/pure/reuseFunc.fk" (Right (DNum 3))
 
     it "Lambda application" $ do
       evalProgram "(\\x : int -> return x + 1) 42" `shouldBeT` (Right (DNum 43))
@@ -145,7 +145,7 @@ main = hspec $ do
       evalProgram "if 0 then return 1 else return 0" `shouldBeT` (Right (DNum 0))
 
     it "Strings comparison" $ do
-      testFromFile "programs/stringsCompare.fk" (Right (DNum 1))
+      testFromFile "programs/pure/stringsCompare.fk" (Right (DNum 1))
 
     -- Pairs
 

### 2020-09-01 src/CPSLists.hs
[src] Substitution for AST and TargetAST

@@ -71,12 +71,12 @@ cps e ks = case e of
         let k@(Pure kf):ks' = ks
         f <- cpsVal vF ks
         arg <- cpsVal vArg ks
-        return $ UApp (UVal f) (UVal arg) ks
+        return $ UApp (UVal f) arg ks
     ELet x varComp comp -> do
         let k:ks' = ks
         let cont varVal ks'' = do
             body <- cps comp (k:ks'')
-            return $ ULet x (UVal varVal) body
+            return $ ULet x varVal body
         cps varComp (Pure cont:ks')
     ESplit l x y row comp -> do
         v <- cpsVal row ks
@@ -119,18 +119,18 @@ cpsOp :: EffLabel -> Value -> [Cont] -> CPSMonad UComp
 cpsOp l v ks = do
     let Pure kf:h:ks' = ks
     let hf = unfoldCont h
-    x <- freshVar' "rArg"
-    let resF = \ks -> kf (UVar x) (h:ks)
+    x <- freshVar' "rArgOp"
+    let resF ks = kf (UVar x) (h:ks)
     let resumption = ULambda x resF
-    let pair cv = UPair (UEffLabel l) (UPair cv resumption)
     cv <- cpsVal v ks
-    hf (pair cv) ks'
+    let pair = UPair (UEffLabel l) (UPair cv resumption)
+    hf pair ks'
 
 
 cpsHRet :: Handler -> Cont
 cpsHRet (HRet xVar comp) = Pure (\x (_:ks') -> do
     convComp <- cps comp ks'
-    return $ ULet xVar (UVal x) convComp)
+    return $ ULet xVar x convComp)
 
 
 cpsHOps :: EffT -> Handler -> Cont
@@ -140,7 +140,7 @@ cpsHOps effT ops = getEffCons effT (\(UPair (UEffLabel l) (UPair p r)) ks ->
         _ -> case hop l ops of
             Just (AlgOp _ pvar rvar comp) -> do
                 contComp <- cps comp ks
-                return $ ULet pvar (UVal p) (ULet rvar (UVal r) contComp)
+                return $ ULet pvar p (ULet rvar r contComp)
             Nothing -> forward l p r ks)
 
 
@@ -149,7 +149,7 @@ forward y p r ks = do
     let k'@(Pure kf'):h':ks' = ks
     let hf' = unfoldCont h'
     x <- freshVar' "rArg"
[…]

### 2020-09-01 src/EvalTarget.hs

@@ -64,30 +64,19 @@ handleEffect (EffL l) v = throwError $ absurdErr (DPair (DLabel l) v)
 
 eval :: Env -> UComp -> EvalMonad DValue
 eval env c = case c of
-    USplit l x y pair comp ->
-         eval env (UVal pair) >>= \(DPair lv rv) ->
-        let extEnv = Map.insert y rv (Map.insert x lv env) in
-        eval extEnv comp
     UApp fnComp argValue ks -> do
         fnVal <- eval env fnComp
-        argDValue <- eval env argValue
+        argDValue <- eval env (UVal argValue)
         case fnVal of
             DLambda fun -> fun env [argDValue] ks
             e -> throwError $ EvalError $ "Application of non-lambda term " ++ show e ++ " in " ++ show c
     UIf e tC fC ->
         eval env (UVal e) >>= \(DNum n) ->
         eval env (if n > 0 then tC else fC)
-    UCase label caseLabel tComp y fComp -> do
-        l <- eval env (UVal label)
-        if l == DLabel caseLabel
-        then eval env tComp
-        else eval (Map.insert y l env) fComp
-    ULet x varComp comp -> do
-        varVal <- eval env varComp
-        let env' = extendEnv env x varVal
-        eval env' comp
-    UTopLevelEffect l v ->
-        eval env (UVal v) >>= handleEffect l
+    ULet x varVal c -> do
+        varDVal <- eval env (UVal varVal)
+        let env' = extendEnv env x varDVal
+        eval env' c
     UAbsurd v -> throwError $ absurdErr v
     -- Values
     UVal (UBinOp op e1 e2) -> do
@@ -110,19 +99,21 @@ eval env c = case c of
         DPair _ v2 -> return v2
         v -> throwError $ EvalError $ "Second projection on expression that is not a pair: " ++ show v
     UVal (ULambda x f) -> return $ DLambda funcRecord
-        where funcRecord env'' [xVal] ks = do
+        where funcRecord env'' [xVal] ks =
                 case Map.lookup "s" env'' of
                     Just r -> resume (extendEnv env "s" r)
                     Nothing -> resume env
                     where resume env =
                             let env' = extendEnv env x xVal in
                             f ks >>= eval env'
+
     UVal (ULabel l) -> return $ DLabel l
     UVal (UEffLabel (EffL l)) -> return $ DLabel l
[…]

### 2020-09-01 src/TargetAST.hs

@@ -3,6 +3,7 @@ module TargetAST where
 import Types
 import AST
 import CPSMonad
+import Control.Monad
 
 type ContF = UValue -> [Cont] -> CPSMonad UComp
 
@@ -46,17 +47,17 @@ data UValue
     | UBinOp BinaryOp UValue UValue
     | UFst UValue
     | USnd UValue
+    | UTopLevelEffect EffLabel UValue
     -- deriving (Eq)
 
 data UComp
     = UVal UValue
-    | UApp UComp UComp [Cont]
+    | UApp UComp UValue [Cont]
     | USplit Label Var Var UValue UComp
     | UCase UValue Label UComp Var UComp
     | UIf UValue UComp UComp
-    | ULet Var UComp UComp
+    | ULet Var UValue UComp
     | UAbsurd UValue
-    | UTopLevelEffect EffLabel UValue
     -- deriving (Eq)
 
 instance Show UValue where
@@ -69,9 +70,10 @@ instance Show UValue where
     show (UEffLabel (EffL l)) = inParens $ "Eff: " ++ l
     show (UEffLabel (CoeffL l)) = inParens $ "Coeff: " ++ l
     show (UVar x) = x
-    show (ULambda x _) = inParens $ x ++ " -> comp"
+    show (ULambda x f) = inParens $ x ++ " -> comp"
     show (UFst v) = "fst " ++ show v
     show (USnd v) = "snd " ++ show v
+    show (UTopLevelEffect l p) = show l ++ show p
 
 instance Show UComp where
     show (UVal v) = show v
@@ -80,5 +82,37 @@ instance Show UComp where
     -- show (UCase v l x c y c') = "case " ++ show v ++ " { " ++ show l ++ " " ++ show x ++ "->" ++ show c ++ "; " ++ show y ++ " " ++ show c'
     show (UIf v c c') = "if " ++ show v ++ " then " ++ show c ++ " else " ++ show c'
     show (UAbsurd v) = "absurd " ++ show v
-    show (ULet x xComp comp) = inParens $ "let " ++ x ++ " = " ++ show xComp ++ " in " ++ show comp
-    show (UTopLevelEffect l p) = show l ++ show p
+    show (ULet x varVal comp) = inParens $ "let " ++ x ++ " = " ++ show varVal ++ " in\n" ++ show comp
+
+
+substV :: Var -> UValue -> UValue -> UValue
+substV xS vS vT = case vT of
[…]

### 2020-09-03 src/Tests.hs
[src] More tests with for static scope

@@ -84,9 +84,6 @@ main = hspec $ do
     it "Variable hiding" $ do
       evalProgram "let x <- return 3 in let x <- return 2 in return x" `shouldBeT` (Right (DNum 2))
 
-    it "Static scope" $ do
-      testFromFile "programs/pure/staticScope.fk" (Right (DNum 3))
-
     it "Let lambda" $ do
       evalProgram "let f <- return (\\x : int -> return x + 1) in f 42" `shouldBeT` (Right (DNum 43))
 
@@ -169,6 +166,17 @@ main = hspec $ do
 
     it "Nested pair" $ do
       evalProgram ("return ((1, 2), 3)") `shouldBeT` (Right (DPair (DPair (DNum 1) (DNum 2)) (DNum 3)))
+
+    -- Scope
+
+    it "Static scope" $ do
+      testFromFile "programs/pure/staticScope.fk" (Right (DNum 3))
+
+    -- it "Function should not leak scope" $ do
+    --   evalProgram "let y <- (\\x : int -> return x) 1 in return x" `shouldBeT` (Left (unboundVarErr "x"))
+
+    -- it "Pure continuations should not leak scope" $ do
+    --   evalProgram "handle let y <- return 0 in return 1 with { return x -> return y }" `shouldBeT` (Right (DNum 1))
 
     -- Algebraic effects and handlers tests
 

### 2020-09-03 thesis/thesis.tex
[thesis] Corrections in first two chapters

@@
 and handlers with resource management, and present a solution through coalgebraic
-means along with presentation of Freak, experimental programming language with
+means along with presentation of Freak, an experimental programming language with
 (co)algebraic effects and (co)handlers. Implementation is based on Continuation
@@
 effects as an API\@ in the code. At the same time, algebraic effects preserve
-a lot of flexibility and have a strong theoretical foundations.
+a lot of flexibility and have strong theoretical foundations.
 
@@
 
-In the above code we open file, and based on nondeterministic choice, we write
+In the code above we open a file, and based on nondeterministic choice, we write
 to the file, and then close it. This piece of code looks harmless, however, as
@@
 
-Under these circumstances one may pose a question, why do we still bother with
-development of languages theory, since so much has been done already. Is there
+Under these circumstances one may pose a question, why we still bother with
+development of the theory of programming languages, since so much has been
+done already. Is there
 anything that drives us towards further research? Indeed, one active branch
-revolves around equational theory to assess equality of two programs, which we
-know that in the general setting is undecidable. Proof methods may include
+revolves around equational reasoning to establish equality of two programs, which we
+know is undecidable in general. Proof methods may include
 extensional, contextual or logical equality. However, there is no doubt that
@@
 
-Other branch of languages theory, that we shall investigate in this thesis,
-is about taming complexity of programs. Various methods of static analysis has
+Other branch of languages theory that we shall investigate in this thesis
+is about taming complexity of programs. Various methods of static analysis have
 been developed, most notably, type systems. Thanks to strong
@@
 We need to have a good way for handling computational effects. One of the ways
-to model them, can be done through monads~\cite{monads-wadler, moggi}. However,
+to model them can be done through monads~\cite{monads-wadler, moggi}. However,
 they were found to be, to say the least, cumbersome to work with when the number
 of different effects increase. It is perhaps not a coincidence that many
-functional programming languages do not have them, because of the non-composable
-nature of them, or at least not composable in their implementations.
-
-\noindent
-Let's put the following functions
+functional programming languages do not have them, because of their non-composable
+nature.
+
+\noindent
+For example, consider the following functions
 
@@
 composition, for $f'$, $g'$ we can use Kleisli composition operator from monad $m$.
-What in case of adding one more effect, $f''$ and $g''$?
-
-It turns out, that it becomes complex and unpleasant to combine two functors
+How about the case of adding one more effect, $f''$ and $g''$?
+
+It turns out that it may become complex and unpleasant to combine two monads
 together to form a new monad, and for this purpose, monad
-transformers~\cite{monad-transformers} arose in Haskell. Most of the common
+transformers~\cite{monad-transformers} arose in Haskell. Most of the usual
 languages avoid this by not expressing computational effects in the type system,
-and instead one may think about functions as being implicitly embed in a Kleisli
-Category over a functor $T$~\cite{kleisli}, where $T$ is a hidden signature
-over all possible side effects that occur in our program.
+and instead one may think about functions as being implicitly embedded in a Kleisli
+Category of a functor $T$~\cite{kleisli}, where $T$ is a hidden signature
+of all possible side effects that may occur in our program.
 
@@
 
-Algebraic effects can be thought of as an public interface for computational effects.
-Declarative approach allow us to write programs in which the actual semantic of
-source code is dependent on handler that defines the meaning of a subset of effects.
+Algebraic effects can be thought of as a public interface for computational effects.
+Declarative approach allows us to write programs in which the actual semantics of
+source code depends on a handler that defines the meaning of a subset of effects.
 
@@
 interface. While this is certainly better than having no abstractions at all,
-we need to pass this interface into every function that is going to use it.
+we need to pass this interface to every function that is going to use it.
 This practice is called dependency injection, and while it sounds like it solves
@@
 That being said, it's only one particular issue that algebraic effects address.
-There are many other great sources for getting familiarized with effects from
+There are many other great sources for getting familiarized with effects from the
 practical side~\cite{handlers-tutorial, programming-in-eff, koka-tutorial},
@@
 we allow ourselves to wander a bit deeper into category theory and describe
-effects from denotational point of view~\cite{bauer-what-is-algebraic}.
+effects from the denotational point of view~\cite{bauer-what-is-algebraic}.
 
@@
 
-    $\mathVar{Free}$ $\mathVar{F-algebra}$ on an object $A$ (of generators) in $\mathcal{C}$ is meant an algebra
-
-    \begin{center}
-
-    $ \varphi_{A} : F A^{\#} \longrightarrow A $
-
-    \end{center}
-
-    \noindent
-    together with an universal arrow $ \eta_{A} : A \longrightarrow A^{\#} $.
+    For an endofunctor $F$, a $\mathVar{Free}$ $\mathVar{F-algebra}$ on an
+    object $A$ (of generators) in $\mathcal{C}$ is an algebra
+
+    \begin{center}
+
+    $ \varphi_{A} : F A^{\#} \longrightarrow A^{\#} $
+
+    \end{center}
+
+    \noindent
+    together with a universal arrow $ \eta_{A} : A \longrightarrow A^{\#} $.
     Universality means that for every algebra $ \beta : F B \longrightarrow B $
@@
     \begin{definition}
-    Collection of $\mathVar{\Sigma-terms}$ is a free algebra with a generator $X$
+    A collection of $\mathVar{\Sigma-terms}$ is a free algebra with a generator $X$
     for a functor $ \mu H_{\Sigma} $ that maps objects into trees over a given
@@
     that a given $\mathVar{\Sigma-term}$ $X \mid t$ is interpreted by a map
-    which sends variables into projections from environment and terms into
+    which sends variables to projections from environment and terms to
     map composition over each subterm.
@@
 
-    Let $L$, $M$ be models of a theory $T$. $ \mathVar{T-homomorphism}$
-    $\varphi : L \rightarrow M$ is a map, such that, for every operation symbol
-    $op_{i}$ in $T$ the following diagram commutes:
+    Let $L$, $M$ be models of a theory $T$. A $\mathVar{T-homomorphism}$
+    $\varphi : L \rightarrow M$ is a map, such that for every operation symbol
+    $op_{i}$ in $T$, the following diagram commutes:
 
@@
 
-    \begin{definition} \textit{Free model} is just a model that is free algebra.
+    \begin{definition} A \textit{Free model} is just a model that is a free algebra.
     \end{definition}
@@
 
+% \vbox{\begin{verbatim}
+%     handle
+%         let fh <- observe Open "praise.txt" in
+%         let c <- do Choice () in
+%         let text <- if c then return "Guy Fieri" else return "is cool" in
+%         let _ <- observe Write (fh, text) in
+%         observe Close fh
+%     with {
+%         Choose _ r ->
+%             let t <- r 1 in
+%             let f <- r 0 in
+%             return (t, f) |
+%         return x -> return x
+%     }
+% \end{verbatim}
+% }
+
     \subsection{Nondeterministic Finite Automata}

### 2020-09-03 src/DValue.hs
[src] FuncRecord should accept just one DValue

@@ -8,7 +8,7 @@ import Types
 
 type Env = Map.Map Var DValue
 
-type FuncRecord = Env -> [DValue] -> [Cont] -> EvalMonad DValue
+type FuncRecord = Env -> DValue -> [Cont] -> EvalMonad DValue
 
 data DValue
     = DNum Integer

### 2020-09-03 src/EvalTarget.hs

@@ -68,7 +68,7 @@ eval env c = case c of
         fnVal <- eval env fnComp
         argDValue <- eval env (UVal argValue)
         case fnVal of
-            DLambda fun -> fun env [argDValue] ks
+            DLambda fun -> fun env argDValue ks
             e -> throwError $ EvalError $ "Application of non-lambda term " ++ show e ++ " in " ++ show c
     UIf e tC fC ->
         eval env (UVal e) >>= \(DNum n) ->
@@ -99,7 +99,7 @@ eval env c = case c of
         DPair _ v2 -> return v2
         v -> throwError $ EvalError $ "Second projection on expression that is not a pair: " ++ show v
     UVal (ULambda x f) -> return $ DLambda funcRecord
-        where funcRecord env'' [xVal] ks =
+        where funcRecord env'' xVal ks =
                 case Map.lookup "s" env'' of
                     Just r -> resume (extendEnv env "s" r)
                     Nothing -> resume env

### 2020-09-03 src/DValue.hs
[MAJOR] Evaluator strategy change from environment to substitution

@@ -8,7 +8,7 @@ import Types
 
 type Env = Map.Map Var DValue
 
-type FuncRecord = Env -> DValue -> [Cont] -> EvalMonad DValue
+type FuncRecord = Env -> UValue -> [Cont] -> EvalMonad DValue
 
 data DValue
     = DNum Integer

### 2020-09-03 src/EvalTarget.hs

@@ -66,17 +66,16 @@ eval :: Env -> UComp -> EvalMonad DValue
 eval env c = case c of
     UApp fnComp argValue ks -> do
         fnVal <- eval env fnComp
-        argDValue <- eval env (UVal argValue)
         case fnVal of
-            DLambda fun -> fun env argDValue ks
+            DLambda fun -> fun env argValue ks
             e -> throwError $ EvalError $ "Application of non-lambda term " ++ show e ++ " in " ++ show c
     UIf e tC fC ->
         eval env (UVal e) >>= \(DNum n) ->
         eval env (if n > 0 then tC else fC)
     ULet x varVal c -> do
-        varDVal <- eval env (UVal varVal)
-        let env' = extendEnv env x varDVal
-        eval env' c
+        let varVal' = substV x varVal varVal
+        let c' = subst x varVal' c
+        eval env c'
     UAbsurd v -> throwError $ absurdErr v
     -- Values
     UVal (UBinOp op e1 e2) -> do
@@ -85,9 +84,7 @@ eval env c = case c of
             (DStr s1) -> eval env (UVal e2) >>= \(DStr s2) -> case op of
                 BEq -> return $ DNum $ boolToInt $ s1 == s2
                 BNe -> return $ DNum $ boolToInt $ s1 /= s2
-    UVal (UVar x) -> case Map.lookup x env of
-        Just v -> return v
-        Nothing -> throwError $ unboundVarErr x
+    UVal (UVar x) -> throwError $ unboundVarErr x
     UVal (UPair e1 e2) -> do
         v1 <- eval env (UVal e1)
         v2 <- eval env (UVal e2)
@@ -99,14 +96,7 @@ eval env c = case c of
         DPair _ v2 -> return v2
         v -> throwError $ EvalError $ "Second projection on expression that is not a pair: " ++ show v
     UVal (ULambda x f) -> return $ DLambda funcRecord
-        where funcRecord env'' xVal ks =
-                case Map.lookup "s" env'' of
-                    Just r -> resume (extendEnv env "s" r)
-                    Nothing -> resume env
-                    where resume env =
-                            let env' = extendEnv env x xVal in
-                            f ks >>= eval env'
-
+        where funcRecord _ xVal = f >=> (eval env . subst x xVal)
     UVal (ULabel l) -> return $ DLabel l
[…]

### 2020-09-03 src/Tests.hs

@@ -108,8 +108,8 @@ main = hspec $ do
     it "Let lambda forget function result - more verbose" $ do
       testFromFile "programs/pure/letLambda.fk" (Right (DNum 1))
 
-    -- it "Reuse function" $ do
-    --   testFromFile "programs/pure/reuseFunc.fk" (Right (DNum 3))
+    it "Reuse function" $ do
+      testFromFile "programs/pure/reuseFunc.fk" (Right (DNum 3))
 
     it "Lambda application" $ do
       evalProgram "(\\x : int -> return x + 1) 42" `shouldBeT` (Right (DNum 43))

### 2020-09-03 thesis/thesis.tex
[thesis] Final submitted version

@@
 
-    Linear usages of resumptions is asserted by analyzing parsed AST tree,
-    where in every leaf there must be a return statement. It's value is then
-    passed into continuation of a cohandler.
+    The idea is that linear usages of resumptions can be easily asserted by
+    analyzing parsed AST tree, where in every leaf there must be a return
+    statement. It's value then can be passed into continuation of a cohandler.
+    That being said, it is not implemented as of the day of submitting the
+    thesis.
 
